### 2020-05-29 paper/paper.Rmd
first commit

---
title: Short-term hourly forecasting in urgent and emergency care
author:
  - name: Author1
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Author2
    email: email2@example.com
    affiliation: University2
    footnote: 2
  - name: Author3
    email: email3@example.com
    affiliation: University3
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

abstract: |
   The Objective of this work would be to propose a new methodology to forecast short-term hourly forecasting for urgent and emergency care.

journal: "Which journal? Journal of Service Research/Health Services Research/EJOR/IJF"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: natbib
bibliography: mybibfile.bib
biblio-style: authoryear
#csl: elsevier-without-titles.csl
#biblio-style: abbrevnat
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.pos = 'H')
# Make sure you have the latest versions of rmarkdown and rticle
library(tidyverse)
library(lubridate)
library(bookdown)
h2_hourly <- readr::read_rds("../data/hw_hourly.rds")
```

# Introduction

Our contributions are as following:

1. We measure the characteristics of the time series at the original level and at various levels of aggregation. We reveal how TA influences the time series characteristics. 
2. We develope a model to discover the relationship between the original time series characteristics and the forecasting performance evaluation results to construct rules for method selection.
3. We determine for which type of time series characteristics, each TA forecasting strategy is preferable.

The rest of the paper is organised as follwoing: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with ...

# Research background:  Hourly short-term forecasting in emergency care {#lit}

\begin{table}[h]
\caption{Summary of studies in hourly emergency care forecasting} 
\centering 
\begin{adjustbox}{width=1\textwidth}
\small

\begin{tabular}{clllll} 
\hline
Authors & What to forecast & Method & Forecast evaluation & Limitations  \\ 
\hline

\citet{hertzum2017forecasting} & Hourly ED patient arrivals and ED occupancy forecasting using calendar variables  1-3 & Regression, ARIMA, naive & MAPE & blabla\\ 
\citet{mccarthy2008challenge} & To develop methodology for predicting demand for ED services by characterizing ED arrivals  1-3 & Poissopn regression & RMSE & blabla\\ 
\citet{morzuch2006forecasting} & hourly ED arrivals for 24 hours & Holt-Winters & MAPE & blabla\\

\hline
\end{tabular}
\end{adjustbox}
\label{tab:lit-summary}
\end{table}

Table \@ref(tab:lit-summary) summarise studies in hourly forecasting in emergency and urgent care.

Linear regression, ARIMA, and naive models were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naive model was there for the purpose of comparison. @hertzum2017forecasting study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year. In term of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest days. July-August are the month with the highest number of patient arrivals and January and February are the months with the lowest number of arrivals. The regression and ARIMA models perform similarly for all forecast interval in modeling patient arrivals. In modeling accident and emergency department occupancy, ARIMA outperform regression models. However, afterall, the models of occupancy were less accurate than those arrivals. @hertzum2017forecasting mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting. Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval. Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.

Predicting the arrivals of accident and emergency department future patients is studied by @choudhury2020forecasting ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy. @hertzum2017forecasting work was mentioned in this paper. It is said that residual normality, stationarity, and autocorrelation have not been tested in @hertzum2017forecasting paper and this might be the cause of accuracy problems. However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in @choudhury2020forecasting 
According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting]. However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].

The literature review reveals some limitations in forecasting for urgent and emergency care which will be summarised as follows:

- first, 
- second,
- third,

# Proposed model {#model}

# Experimental design {#design}

## data {#data}

Figure \@ref(fig:hourly-plot)

```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width="80%"}
h2_hourly <- h2_hourly %>% filter(arrival_1h >="2015-01-01 00:00:00"&arrival_1h <"2019-01-01 00:00:00")
  subseriesplot <- h2_hourly %>% as_tibble() %>% mutate(day=factor(yday(arrival_1h)),year=factor(year(arrival_1h)),hour=hour(arrival_1h)) %>% 
    select(-arrival_1h)
time <-  format( seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date()+1), by = "1 hour"),
                 "%H:%M", tz="GMT")
time <- time[-length(time)]
p1 <- ggplot(data=subseriesplot,mapping = aes(x=hour ,y=n_attendance, colour=day))+
    geom_line(aes(group=interaction(day,year)))+
    geom_point()+
    theme_bw()+
    theme(legend.position = "")+
    labs(y="Number of visits in ED", x="Hour of day")+
      scale_x_continuous(breaks=0:23, labels = time)+
  theme(axis.text.x = element_text(angle = 90))
  p1+theme(plot.title = element_text(size = 10))
```

Figure \@ref(fig:24hour)

```{r 24hour, fig.cap= "Distrubution of ED attendance for each hour of day", fig.align='center',fig.width=6, fig.height=9}
h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% 
  select(-arrival_1h) %>% ggplot(mapping=aes(x=n_attendance))+ 
geom_histogram(mapping=aes(y=stat(density)), binwidth = 1,fill="white",color = 'black')+
  geom_density(color="blue")+
scale_y_continuous(labels = scales::percent_format())+
facet_wrap(~Hour, scales = "free", ncol = 3)+
theme_bw()+
labs(x="Number of visits in ED", y="Percentage")+
theme(axis.text = element_text(size = 7))
```

## benchmarks {#benchmarks}

\begin{equation}
\bar{X} = \frac{\sum_{i=1}^n X_i}{n} (\#eq:mean)
\end{equation}

Also see Equation \@ref(eq:mean)

## forecast performance evaluation {#accuracy}

# Result and discussion {#result}

# Conclusion {#conclusion}

# References {#references .unnumbered}

### 2020-05-29 paper/paper.tex
first commit

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Which journal? Journal of Service Research/Health Services Research/EJOR/IJF} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short-term hourly forecasting in urgent and emergency care},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{adjustbox, float}

\begin{document}
\begin{frontmatter}

  \title{Short-term hourly forecasting in urgent and emergency care}
    \author[University1]{Author1\corref{1}}
   \ead{email1@example.com} 
    \author[University2]{Author2\corref{2}}
   \ead{email2@example.com} 
    \author[University3]{Author3\corref{2}}
   \ead{email3@example.com} 
      \address[University1]{Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff}
    \address[University2]{adress2}
    \address[University3]{adress3}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  The Objective of this work would be to propose a new methodology to forecast short-term hourly forecasting for urgent and emergency care.
  \end{abstract}

 \end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Our contributions are as following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We measure the characteristics of the time series at the original level and at various levels of aggregation. We reveal how TA influences the time series characteristics.
\item
  We develope a model to discover the relationship between the original time series characteristics and the forecasting performance evaluation results to construct rules for method selection.
\item
  We determine for which type of time series characteristics, each TA forecasting strategy is preferable.
\end{enumerate}

The rest of the paper is organised as follwoing: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with \ldots{}

\hypertarget{lit}{%
\section{Research background: Hourly short-term forecasting in emergency care}\label{lit}}

\begin{table}[h]
\caption{Summary of studies in hourly emergency care forecasting} 
\centering 
\begin{adjustbox}{width=1\textwidth}
\small

\begin{tabular}{clllll} 
\hline
Authors & What to forecast & Method & Forecast evaluation & Limitations  \\ 
\hline

\citet{hertzum2017forecasting} & Hourly ED patient arrivals and ED occupancy forecasting using calendar variables  1-3 & Regression, ARIMA, naive & MAPE & blabla\\ 
\citet{mccarthy2008challenge} & To develop methodology for predicting demand for ED services by characterizing ED arrivals  1-3 & Poissopn regression & RMSE & blabla\\ 
\citet{morzuch2006forecasting} & hourly ED arrivals for 24 hours & Holt-Winters & MAPE & blabla\\

\hline
\end{tabular}
\end{adjustbox}
\label{tab:lit-summary}
\end{table}

Table \ref{tab:lit-summary} summarise studies in hourly forecasting in emergency and urgent care.

Linear regression, ARIMA, and naive models were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naive model was there for the purpose of comparison. \citet{hertzum2017forecasting} study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year. In term of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest days. July-August are the month with the highest number of patient arrivals and January and February are the months with the lowest number of arrivals. The regression and ARIMA models perform similarly for all forecast interval in modeling patient arrivals. In modeling accident and emergency department occupancy, ARIMA outperform regression models. However, afterall, the models of occupancy were less accurate than those arrivals. \citet{hertzum2017forecasting} mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting. Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval. Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.

Predicting the arrivals of accident and emergency department future patients is studied by \citet{choudhury2020forecasting} ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy. \citet{hertzum2017forecasting} work was mentioned in this paper. It is said that residual normality, stationarity, and autocorrelation have not been tested in \citet{hertzum2017forecasting} paper and this might be the cause of accuracy problems. However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in \citet{choudhury2020forecasting}
According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast \citep{boyle2012predicting, hertzum2017forecasting}. However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

The literature review reveals some limitations in forecasting for urgent and emergency care which will be summarised as follows:

\begin{itemize}
\tightlist
\item
  first,
\item
  second,
\item
  third,
\end{itemize}

\hypertarget{model}{%
\section{Proposed model}\label{model}}

\hypertarget{design}{%
\section{Experimental design}\label{design}}

\hypertarget{data}{%
\subsection{data}\label{data}}

Figure \ref{fig:hourly-plot}

\begin{figure}[H]

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/hourly-plot-1}

}

\caption{Seasonal plot of ED attendance}\label{fig:hourly-plot}
\end{figure}

Figure \ref{fig:24hour}

\begin{figure}[H]

{\centering \includegraphics{paper_files/figure-latex/24hour-1}

}

\caption{Distrubution of ED attendance for each hour of day}\label{fig:24hour}
\end{figure}

\hypertarget{benchmarks}{%
\subsection{benchmarks}\label{benchmarks}}

\begin{equation}
\bar{X} = \frac{\sum_{i=1}^n X_i}{n} \label{eq:mean}
\end{equation}

Also see Equation \eqref{eq:mean}

\hypertarget{accuracy}{%
\subsection{forecast performance evaluation}\label{accuracy}}

\hypertarget{result}{%
\section{Result and discussion}\label{result}}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\renewcommand\refname{References}
\bibliography{mybibfile.bib}

\end{document}

### 2020-05-29 presentation/presentation.Rmd
first commit

---
title: "Hourly forecasting in ED and Ambulance services"
author: "Bahman Rostami-Tabar"
date: "27/05/2020"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(lubridate)
library(patchwork)
library(fable)
library(feasts)
library(tsibble)
library(tidyverse)
```

## Hourly time series

```{r hourly,fig.align='center',out.width="75%"}
h2_hourly <- readr::read_rds("../data/h2_hourly.rds")
h2_hourly <- h2_hourly %>% filter(arrival_1h >="2015-01-01 00:00:00"&arrival_1h <"2019-01-01 00:00:00")
  subseriesplot <- h2_hourly %>% as_tibble() %>% mutate(day=factor(yday(arrival_1h)),year=factor(year(arrival_1h)),hour=hour(arrival_1h)) %>% 
    select(-arrival_1h)
time <-  format( seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date()+1), by = "1 hour"),
                 "%H:%M", tz="GMT")
time <- time[-length(time)]
p1 <- ggplot(data=subseriesplot,mapping = aes(x=hour ,y=n_attendance, colour=day))+
    geom_line(aes(group=interaction(day,year)))+
    geom_point()+
    theme_bw()+
    theme(legend.position = "")+
    labs(y="Number of visits in ED", x="Hour of day")+
      scale_x_continuous(breaks=0:23, labels = time)+
  theme(axis.text.x = element_text(angle = 90))
  p1+theme(plot.title = element_text(size = 10))
```

## 2 shifts

```{r 2shift, fig.align='center',fig.width=15, fig.height=7}

ed_2shoft <- h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% mutate(shift=if_else(Hour <= 12,"shift_1", "shift_2")) %>% 
  group_by(Date,Hour,shift) %>% summarise(n_attendance=sum(n_attendance))

  ed_2shoft %>%
  ggplot(mapping = aes(x=n_attendance))+
  geom_histogram(aes(y = stat(density)),binwidth = 1,color = 'black')+
    geom_density(color="blue")+
    scale_y_continuous(labels = scales::percent_format())+
  facet_wrap(~shift)+
    theme_bw()+
    labs(x="Number of visits in ED", y="Percentage")
```

## 3 shifts

```{r 3shift, fig.align='center',fig.width=15, fig.height=7}
h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% mutate(shift=
case_when(Hour < 9 ~"shift_1", 8 < Hour & Hour < 17 ~ "shift_2", TRUE ~ "shift_3")) %>% 
group_by(Date,shift) %>% summarise(n_attendance=sum(n_attendance)) %>% 
ggplot(mapping = aes(x=n_attendance))+
geom_histogram(mapping=aes(y=stat(density)), binwidth = 1,fill="white",color = 'black')+
  geom_density(color="blue")+
scale_y_continuous(labels = scales::percent_format())+
facet_wrap(~shift, scales = "free")+
  theme_bw()+
  labs(x="Number of visits in ED", y="Percentage")

```

## 4 shifts

```{r 4shift, fig.align='center',fig.width=9, fig.height=9}
h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% mutate(shift=
  case_when(Hour < 6 ~"shift_1", 6 <= Hour & Hour < 12 ~ "shift_2",12 <= Hour & Hour < 18 ~ "shift_3", TRUE ~ "shift_4")) %>% 
  group_by(Date,shift) %>% summarise(n_attendance=sum(n_attendance)) %>% 
  ggplot(mapping = aes(x=n_attendance))+
geom_histogram(mapping=aes(y=stat(density)), binwidth = 1,fill="white",color = 'black')+
  geom_density(color="blue")+
scale_y_continuous(labels = scales::percent_format())+
facet_wrap(~shift, scales = "free")+
  theme_bw()+
  labs(x="Number of visits in ED", y="Percentage")
```

## 24 hours

```{r 24shift, fig.align='center',fig.width=10, fig.height=20}
h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% 
  select(-arrival_1h) %>% ggplot(mapping=aes(x=n_attendance))+ 
geom_histogram(mapping=aes(y=stat(density)), binwidth = 1,fill="white",color = 'black')+
  geom_density(color="blue")+
scale_y_continuous(labels = scales::percent_format())+
facet_wrap(~Hour, scales = "free", ncol = 3)+
  theme_bw()+
  labs(x="Number of visits in ED", y="Percentage")
```

### 2020-05-30 gam-mgcv/exploration.R
create file and load data

@@ -1,0 +1,10 @@
+
+
+require(rstudioapi)
+require(data.table)
+require(mgcv)
+require(mgcViz)
+
+setwd(dirname(getActiveDocumentContext()$path))
+
+load("../data/hw_hourly.rds")

### 2020-06-02 gam-mgcv/exploration.R
first gam

@@ -4,7 +4,30 @@ require(rstudioapi)
 require(data.table)
 require(mgcv)
 require(mgcViz)
+# library(devtools)
+# install_github("jbrowell/ProbCast")
+require(ProbCast)
 
 setwd(dirname(getActiveDocumentContext()$path))
 
-load("../data/hw_hourly.rds")
+## Load and Prep Data ####
+# load("../data/hw_hourly.rds")
+h2 <- fread("../data/h2_hourly.csv")
+h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
+h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
+
+
+add_calendar_variables(h2,datetimecol = "targetTime_UK")
+
+## Fit GAM and visualise model ####
+gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) + dow + s(doy,k=20,bs="ad"),
+              data=h2)
+
+gam.check(gam1)
+plot(gam1,pages = 1)
+
+gam1 <- getViz(gam1,nsim = 200)
+
+check1D(gam1,"clock_hour")
+check1D(gam1,"doy")
+check2D(gam1,"dow","clock_hour")

### 2020-06-02 gam-mgcv/exploration.R
gam with poisson family
A little better fit than gaussian, but much more meaningful statistically

@@ -20,8 +20,18 @@ h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/Lo
 add_calendar_variables(h2,datetimecol = "targetTime_UK")
 
 ## Fit GAM and visualise model ####
+
+
 gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) + dow + s(doy,k=20,bs="ad"),
-              data=h2)
+              data=h2,family = poisson())
+
+## Other families: 
+# gaussian()
+# negbin() is useful for overdispersed count data, but computation is slow.
+# poisson a little better than gaussian...
+
+## In-sample RMSE
+sqrt(mean((gam1$y-gam1$fitted.values)^2))
 
 gam.check(gam1)
 plot(gam1,pages = 1)

### 2020-06-02 paper/paper.Rmd
add introduction and new graphs to data subsection

@@
 ---
-title: Short-term hourly forecasting in urgent and emergency care
+title: Short-term hourly forecasting for care services
 author:
@@
 header-includes:  
- \usepackage{adjustbox, float} #use the 'float' package
+ \usepackage{adjustbox, float,lscape} #use the 'float' package
 output:
@@
 library(bookdown)
+library(patchwork)
+library(tsibble)
 h2_hourly <- readr::read_rds("../data/hw_hourly.rds")
@@
 
+why forecasting for urgent and emergency care is important?
+
+Urgent and emergency healthcare is increasingly regarding patients as consumers of its service and aims to provide 'consumer satisfaction'; particularly in the Accident and Emergency (A&E) department and Ambulance services which have been viewed as the 'shop window' of the hospital service.
+An accurate forecasting of the demand is essential in A&E and Ambulance service trust to depict various courses of action that can result in massive savings in terms of patient lives.
+
+<!-- Demand forecasting is a critical step to inform bed capacity planning and staffing to meet the needs of patients on each hospital service and inpatient unit.
+One of the best ways to optimize staffing utilization and related costs is to forecast demand far enough in advance in order to sufficiently match staff to meet anticipated patient demand without incurring last-minute expenses, such as overtime or supplemental staffing.
+-->
+<!-- Accurate forecasting of patient demand by season, month, day of the week, and time of day enables leaders to making strategic decisions regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+-->
+<!-- Equipped with accurate demand forecasts, hospital leaders can reconfigure units and redeploy staff during predicted low-census periods, and units with seasonal occupancy can be closed and repurposed.
+-->
+
+Accurate forecasts contribute to better allocation of resources.
+Patient overcrowding in urgent and emergency care is a serious problem that causes challenging situations on patient flow .
+Also, it is related with increasing length of stay (LOS), low patient satisfaction about treatments on patients, number of patients left the ED without seen by medical staff , unexpected return visits to EDs , increasing health care costs and some other miscellaneous problems such as inaccuracy in electronic medical record-reported wait times to initial emergency physician evaluation and nursing staff requirement!
+
+why hourly forecast is important?
+
+12 hour forecasts are required to inform the Operations Department and in particular the Operational Delivery Unit (ODU) of future demand, particularly for the current and the next shift.
+The combination of forecast demand, incidents being attended, resource availability and delays at hospital provide information on the state of the unscheduled care system across Wales.
+Having this full picture enables the ODU to focus on the areas that require intervention to enable the most effective delivery of the service to the patients of Wales.
+
+Forecasting hourly admissions is crucial for operational planning which  involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory ,Emergency, Surgical, Inpatient, Home and Residential.
+
+- Adjusting the Schedule during  a day or a week because of unplanned events, such as emergency or walk-in patients, extended consultation times, and equipment
+- Dynamic patient (re)assignment
+- Staff rescheduling At the start of a shift, the staff sche- dule is reconsidered.
+Before and during the shift, the staff capacities may be adjusted to unpredicted demand fluctuations and staff absenteeism by using part-time, on- call nurses, staff overtime and voluntary absenteeism
+
+what literature says about hourly forecasting in healthcare?
+
+There exists a large literature on forecasting patient arrival, however the variation in arrivals remain unaccounted for.
+The forecasted hourly admissions at a typical A&E or Ambulance service show a mean absolute percentage error (MAPE) of 50%.
+Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
+In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services
+
+limitations in current literature?
+
+how we add to the literature?
+
 Our contributions are as following:
 
-1. We measure the characteristics of the time series at the original level and at various levels of aggregation. We reveal how TA influences the time series characteristics. 
-2. We develope a model to discover the relationship between the original time series characteristics and the forecasting performance evaluation results to construct rules for method selection.
-3. We determine for which type of time series characteristics, each TA forecasting strategy is preferable.
+1.
+We develope a novel methodology to forecast hourly time series for healthcare services combining a distributio fit to each hour of day seperately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, ...
+iv) special events such as public holidays, festive days and rugby and v) weather data.
+2. We provide prbabilistic forecasts ...
+3. We benchmark the forecasting performance of the proposed methodology against regression, prophet and TBATS using ....
 
-The rest of the paper is organised as follwoing: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with ...
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with ...
 
-# Research background:  Hourly short-term forecasting in emergency care {#lit}
+# Research background: Hourly forecasting in care services{#lit}
 
@@
 \citet{hertzum2017forecasting} & Hourly ED patient arrivals and ED occupancy forecasting using calendar variables  1-3 & Regression, ARIMA, naive & MAPE & blabla\\ 
-\citet{mccarthy2008challenge} & To develop methodology for predicting demand for ED services by characterizing ED arrivals  1-3 & Poissopn regression & RMSE & blabla\\ 
+\citet{mccarthy2008challenge} & To develop methodology for predicting demand for ED services by characterizing ED arrivals  1-3 & Poisson regression & RMSE & blabla\\ 
 \citet{morzuch2006forecasting} & hourly ED arrivals for 24 hours & Holt-Winters & MAPE & blabla\\ 
+\citet{chase2012predicting} & ED volume & Logistic regression & RMSE & blabla\\ 
+\citet{jones2009multivariate} & Demands for key resources in the ED and the inpatient hospital  & S, ARIMA & RMSE & blabla\\
 
@@
 In modeling accident and emergency department occupancy, ARIMA outperform regression models.
-However, afterall, the models of occupancy were less accurate than those arrivals.
+However, after all, the models of occupancy were less accurate than those arrivals.
 @hertzum2017forecasting mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting.
@@
 
-```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width="80%"}
+Figure \@ref(fig:24hour)
+
+```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',fig.height=9, fig.width= 6}
 h2_hourly <- h2_hourly %>% filter(arrival_1h >="2015-01-01 00:00:00"&arrival_1h <"2019-01-01 00:00:00")
@@
     theme(legend.position = "")+
-    labs(y="Number of visits in ED", x="Hour of day")+
+    labs(y="Number of visits in ED", x="")+
       scale_x_continuous(breaks=0:23, labels = time)+
+  theme(axis.text.x = element_text(angle = 90))+
+  theme(plot.title = element_text(size = 10))
+
+p2 <- subseriesplot %>% 
+  ggplot(mapping = aes(x=factor(hour) ,y=n_attendance))+
+  geom_boxplot(aes(fill=year))+
+  scale_color_brewer(palette = "Set1")+
+  theme_bw()+
+  theme(legend.position = "bottom", legend.box = "horizontal")+
+  labs(fill="Year",y="Number of visits in ED", x="Hour of day")+
+  scale_x_discrete(breaks=0:23, labels = time)+
+  theme(axis.text.x = element_text(angle = 90))+
+  theme(plot.title = element_text(size = 10))
+
+h <- tibble(h=rep(1:24,1461 ))
+ae_h_hour <- bind_cols(h2_hourly,h)
+ae_h_hour <- ae_h_hour %>% as_tibble() %>% mutate(date=date(arrival_1h)) %>% select(h,date,n_attendance)
+p <- ggplot(ae_h_hour, mapping = aes(x=factor(h), y=n_attendance))+
+  geom_boxplot(outlier.colour = "red")+
+  stat_summary(fun = quantile, geom="line",fun.args = list(probs = 0.05) , group= 1, aes(color= "#0072B2"), size = .3)+
+  stat_summary(fun = quantile, geom="line",fun.args = list(probs = 0.25) , group= 1, aes(color= "#009E73"), size = .3)+
+  stat_summary(fun = median, geom="line", group= 1, aes(color= "#CC79A7"), size = .5)+
+  stat_summary(fun = quantile, geom="line",fun.args = list(probs = 0.75) , group= 1, aes(color= "#E69F00"), size = .3)+
+  stat_summary(fun = quantile, geom="line",fun.args = list(probs = 0.95) , group= 1, aes(color= "#56B4E9"), size = .3)+
+  stat_summary(fun = mean, geom="line", group= 1, aes(color= "#F0E442"), size = .5)
+
+time <-  format( seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date()+1), by = "1 hour"),
+                 "%H:%M", tz="GMT")
+time <- time[-length(time)]
+
+p_boxplot <- p+ theme_light()+
+  labs(x="Hour of day", y="Number of visits in ED", color="")+ # Set line color to blue
+  scale_color_identity(
+                       breaks = c("#0072B2","#009E73","#CC79A7","#F0E442","#E69F00","#56B4E9"),
+                       labels = c("5% quantile", "25% quantile","Median", "Mean","75% quantile","95% quantile"),
+                       guide = "legend")+
+  theme(legend.position = "bottom")+
+  guides(color=guide_legend(nrow=1,byrow=TRUE))+
+  scale_x_discrete(breaks=1:24,labels=time)+
   theme(axis.text.x = element_text(angle = 90))
-  p1+theme(plot.title = element_text(size = 10))
+
+p1/p2/p_boxplot
 ```
 
-Figure \@ref(fig:24hour)
+```{r 24hour, fig.cap= "Distrubution of ED attendance for each hour of day", fig.align='center',fig.width=6, fig.height=9, dependson="hourly-plot"}
 
-```{r 24hour, fig.cap= "Distrubution of ED attendance for each hour of day", fig.align='center',fig.width=6, fig.height=9}
+hour.labs <- time
+names(hour.labs) <- c(0:23)
+
 h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% 
@@
 scale_y_continuous(labels = scales::percent_format())+
-facet_wrap(~Hour, scales = "free", ncol = 3)+
+facet_wrap(~Hour, scales = "free", ncol = 3,labeller = labeller(Hour = hour.labs))+
 theme_bw()+

### 2020-06-02 paper/paper.tex
add introduction and new graphs to data subsection

@@
             pdfauthor={},
-            pdftitle={Short-term hourly forecasting in urgent and emergency care},
+            pdftitle={Short-term hourly forecasting for care services},
             colorlinks=false,
@@
 % Pandoc header
-\usepackage{adjustbox, float}
+\usepackage{adjustbox, float,lscape}
 
@@
 
-  \title{Short-term hourly forecasting in urgent and emergency care}
+  \title{Short-term hourly forecasting for care services}
     \author[University1]{Author1\corref{1}}
@@
 
+why forecasting for urgent and emergency care is important?
+
+Urgent and emergency healthcare is increasingly regarding patients as consumers of its service and aims to provide `consumer satisfaction'; particularly in the Accident and Emergency (A\&E) department and Ambulance services which have been viewed as the `shop window' of the hospital service.
+An accurate forecasting of the demand is essential in A\&E and Ambulance service trust to depict various courses of action that can result in massive savings in terms of patient lives.
+
+Accurate forecasts contribute to better allocation of resources.
+Patient overcrowding in urgent and emergency care is a serious problem that causes challenging situations on patient flow .
+Also, it is related with increasing length of stay (LOS), low patient satisfaction about treatments on patients, number of patients left the ED without seen by medical staff , unexpected return visits to EDs , increasing health care costs and some other miscellaneous problems such as inaccuracy in electronic medical record-reported wait times to initial emergency physician evaluation and nursing staff requirement!
+
+why hourly forecast is important?
+
+12 hour forecasts are required to inform the Operations Department and in particular the Operational Delivery Unit (ODU) of future demand, particularly for the current and the next shift.
+The combination of forecast demand, incidents being attended, resource availability and delays at hospital provide information on the state of the unscheduled care system across Wales.
+Having this full picture enables the ODU to focus on the areas that require intervention to enable the most effective delivery of the service to the patients of Wales.
+
+Forecasting hourly admissions is crucial for operational planning which involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory ,Emergency, Surgical, Inpatient, Home and Residential.
+
+\begin{itemize}
+\tightlist
+\item
+  Adjusting the Schedule during a day or a week because of unplanned events, such as emergency or walk-in patients, extended consultation times, and equipment
+\item
+  Dynamic patient (re)assignment
+\item
+  Staff rescheduling At the start of a shift, the staff sche- dule is reconsidered.
+Before and during the shift, the staff capacities may be adjusted to unpredicted demand fluctuations and staff absenteeism by using part-time, on- call nurses, staff overtime and voluntary absenteeism
+\end{itemize}
+
+what literature says about hourly forecasting in healthcare?
+
+There exists a large literature on forecasting patient arrival, however the variation in arrivals remain unaccounted for.
+The forecasted hourly admissions at a typical A\&E or Ambulance service show a mean absolute percentage error (MAPE) of 50\%.
+Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
+In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services
+
+limitations in current literature?
+
+how we add to the literature?
+
 Our contributions are as following:
@@
 \item
-  We measure the characteristics of the time series at the original level and at various levels of aggregation. We reveal how TA influences the time series characteristics.
+  We develope a novel methodology to forecast hourly time series for healthcare services combining a distributio fit to each hour of day seperately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, \ldots{} iv) special events such as public holidays, festive days and rugby and v) weather data.
 \item
-  We develope a model to discover the relationship between the original time series characteristics and the forecasting performance evaluation results to construct rules for method selection.
+  We provide prbabilistic forecasts \ldots{}
 \item
-  We determine for which type of time series characteristics, each TA forecasting strategy is preferable.
+  We benchmark the forecasting performance of the proposed methodology against regression, prophet and TBATS using \ldots{}.
 \end{enumerate}
 
-The rest of the paper is organised as follwoing: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with \ldots{}
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with \ldots{}
 
 \hypertarget{lit}{%
-\section{Research background: Hourly short-term forecasting in emergency care}\label{lit}}
+\section{Research background: Hourly forecasting in care services}\label{lit}}
 
@@
 In modeling accident and emergency department occupancy, ARIMA outperform regression models.
-However, afterall, the models of occupancy were less accurate than those arrivals.
+However, after all, the models of occupancy were less accurate than those arrivals.
 \citet{hertzum2017forecasting} mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting.
@@
 
+Figure \ref{fig:24hour}
+
 \begin{figure}[H]
 
-{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/hourly-plot-1}
+{\centering \includegraphics{paper_files/figure-latex/hourly-plot-1}
 
@@
 \end{figure}
-
-Figure \ref{fig:24hour}
 

### 2020-06-04 gam-mgcv/exploration.R
add weather data

@@ -16,19 +16,28 @@ h2 <- fread("../data/h2_hourly.csv")
 h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
 h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
 
-
 add_calendar_variables(h2,datetimecol = "targetTime_UK")
+
+#load weather data
+weather_data <- data.table()
+for(f in c(list.files(path="../data/",pattern = "weather",full.names = T))){
+  load(f)
+  weather_data <- rbind(weather_data,features)
+}
+
+h2 <- merge(weather_data,h2,by = "targetTime")
+
+rm(f,features,weather_data)
 
 ## Fit GAM and visualise model ####
 
-
 gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) + dow + s(doy,k=20,bs="ad"),
-              data=h2,family = poisson())
+            data=h2,family = poisson())
 
 ## Other families: 
 # gaussian()
 # negbin() is useful for overdispersed count data, but computation is slow.
-# poisson a little better than gaussian...
+# So far: poisson a little better than gaussian, not tried negbin...
 
 ## In-sample RMSE
 sqrt(mean((gam1$y-gam1$fitted.values)^2))

### 2020-06-05 gam-mgcv/exploration.R
Add holiday and weather data
Gam fit with tod/doy only and residuals visualised

@@ -4,9 +4,11 @@ require(rstudioapi)
 require(data.table)
 require(mgcv)
 require(mgcViz)
+require(ggplot2)
 # library(devtools)
 # install_github("jbrowell/ProbCast")
 require(ProbCast)
+require(readxl)
 
 setwd(dirname(getActiveDocumentContext()$path))
 
@@ -18,20 +20,30 @@ h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/Lo
 
 add_calendar_variables(h2,datetimecol = "targetTime_UK")
 
+# Load Holidays
+hols <- as.data.table(read_xlsx("../data/holiday_rugby.xlsx"))
+hols[,Date:=as.Date(Date)]
+h2[,Date:=as.Date(targetTime_UK)]
+h2 <- merge(h2,hols,by="Date",all.x=T)
+rm(hols); h2[,Date:=NULL]
+h2[,school_holiday:=as.factor(school_holiday)]
+h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
+
 #load weather data
 weather_data <- data.table()
 for(f in c(list.files(path="../data/",pattern = "weather",full.names = T))){
   load(f)
   weather_data <- rbind(weather_data,features)
 }
-
 h2 <- merge(weather_data,h2,by = "targetTime")
-
 rm(f,features,weather_data)
+h2[,wind10m:=sqrt(`10U`^2+`10V`^2)]
+
 
 ## Fit GAM and visualise model ####
 
-gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) + dow + s(doy,k=20,bs="ad"),
+gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
+              te(doy,clock_hour,k=c(12,12)),
             data=h2,family = poisson())
 
 ## Other families: 
@@ -42,11 +54,29 @@ gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) + dow + s(doy,k=20,bs="ad")
 ## In-sample RMSE
 sqrt(mean((gam1$y-gam1$fitted.values)^2))
 
+## Check fit - is the model using all DOF? Is so, consider increasing availability...
 gam.check(gam1)
-plot(gam1,pages = 1)
+plot(gam1,pages = 2)
[…]

### 2020-06-05 gam-mgcv/exploration.R
Update exploration.R

@@ -28,6 +28,8 @@ h2 <- merge(h2,hols,by="Date",all.x=T)
 rm(hols); h2[,Date:=NULL]
 h2[,school_holiday:=as.factor(school_holiday)]
 h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
+h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
+h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
 
 #load weather data
 weather_data <- data.table()
@@ -46,9 +48,10 @@ gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
               te(doy,clock_hour,k=c(12,12)),
             data=h2,family = poisson())
 
-## Other families: 
+## Families: 
 # gaussian()
-# negbin() is useful for overdispersed count data, but computation is slow.
+# poisson()
+# nb() / negbin() is useful for overdispersed count data, but computation is slow.
 # So far: poisson a little better than gaussian, not tried negbin...
 
 ## In-sample RMSE
@@ -71,9 +74,10 @@ check2D(gam1,h2[,school_holiday],"clock_hour") +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))
 check2D(gam1,h2[,holiday_festive_day],"clock_hour") + 
   theme(axis.text.x = element_text(angle = 90, hjust = 1))
-check2D(gam1,h2[,is_rug_in_Cardiff],"clock_hour")
+#check2D(gam1,h2[,is_rug_in_Cardiff],"clock_hour")
 
 ## Any weather effects?
+check1D(gam1,h2[,`2T`])
 check2D(gam1,h2[,`2T`],"clock_hour")
 check1D(gam1,h2[,TP])
 check1D(gam1,h2[,SSRD])

### 2020-06-05 gam-mgcv/exploration.R
Update exploration.R

@@ -52,7 +52,7 @@ gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
 # gaussian()
 # poisson()
 # nb() / negbin() is useful for overdispersed count data, but computation is slow.
-# So far: poisson a little better than gaussian, not tried negbin...
+# So far: poisson a little better than other two...
 
 ## In-sample RMSE
 sqrt(mean((gam1$y-gam1$fitted.values)^2))

### 2020-06-05 gam-mgcv/exploration.R
Update exploration.R

@@ -64,6 +64,7 @@ plot(gam1,select = 8,scheme = 1)
 
 gam1 <- getViz(gam1,nsim = 200)
 
+
 check1D(gam1,"clock_hour")
 check1D(gam1,"doy")
 check2D(gam1,"dow","clock_hour")
@@ -84,3 +85,4 @@ check1D(gam1,h2[,SSRD])
 check1D(gam1,h2[,LCC+MCC+HCC])
 check1D(gam1,h2[,wind10m])
 check2D(gam1,h2[,wind10m],"clock_hour")
+check2D(gam1,h2[,wind10m],h2[,TP])

### 2020-06-05 gam-mgcv/exploration.R
fiddline with model

@@ -45,7 +45,8 @@ h2[,wind10m:=sqrt(`10U`^2+`10V`^2)]
 ## Fit GAM and visualise model ####
 
 gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
-              te(doy,clock_hour,k=c(12,12)),
+              s(doy,k=26,by=dow) +
+              ti(doy,clock_hour,k=c(12,6)),
             data=h2,family = poisson())
 
 ## Families: 

### 2020-06-05 gam-mgcv/exploration.R
Update exploration.R

@@ -45,7 +45,7 @@ h2[,wind10m:=sqrt(`10U`^2+`10V`^2)]
 ## Fit GAM and visualise model ####
 
 gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
-              s(doy,k=26,by=dow) +
+              s(doy,k=26) +
               ti(doy,clock_hour,k=c(12,6)),
             data=h2,family = poisson())
 

### 2020-06-05 gam-mgcv/exploration.R
Update exploration.R

@@ -46,7 +46,7 @@ h2[,wind10m:=sqrt(`10U`^2+`10V`^2)]
 
 gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
               s(doy,k=26) +
-              ti(doy,clock_hour,k=c(12,6)),
+              ti(doy,clock_hour,k=c(6,6)),
             data=h2,family = poisson())
 
 ## Families: 

### 2020-06-05 gam-mgcv/exploration.R
Notes of experiment set-up

@@ -1,4 +1,8 @@
 
+## Forecast every 12h (00 and 12 local time)
+## Lead times from 0 to 48h
+## Quantiles from 0.05 to 0.95, Pinball loss
+## Test Data: 1/3/2018 to 28/2/2019
 
 require(rstudioapi)
 require(data.table)

### 2020-09-11 gam-mgcv/study_v1.R
Study script

@@ -64,7 +64,7 @@ h2[issueTime<"2018-03-01",kfold:=paste0("fold",rep(rep(1:2,each=24*7),length.out
 
 
 
-## Fit GAM and visualise model ####
+## Fit Poisson-GAM and visualise model ####
 
 for(fold in unique(h2$kfold)){
   
@@ -140,3 +140,41 @@ pinball(h2_mqr[,-c(1:2)],h2$n_attendance,kfolds = h2$kfold,ylim=c(0.3,2))
 
 
 # save(h2_mqr,file = "../data/example_forecast_format.R")
+
+
+## GAMLSS ####
+
+require(gamlss.add) # for ga()
+# check out http://opisthokonta.net/?p=1157
+# Double Poisson DPO
+# Negative Binomial
+# 
+
+h2_gamlss <- Para_gamlss(data = h2,
+                             formula = n_attendance ~ pb(clock_hour),
+                           #pvc(clock_hour,df=24,by=dow),# +
+                               # pb(doy,df=26) +
+                               # pvc(clock_hour,by=doy,df=c(6,6)),
+                             sigma.formula = ~1,
+                             nu.formula = ~1,
+                             family =  DPO, #NO,  #
+                             method=mixed(20,10))
+
+h2_gamlss_mqr <- PPD_2_MultiQR(data=h2,
+                                  models = h2_gamlss,
+                                  params = F)
+
+plot(h2_gamlss_mqr[100:148,],
+     xlab="Lead-time [hours]",ylab="Attendance",main=paste0("Origin: ",issue," (",format(issue,"%A"),")"),
+     ylim=c(0,40),Legend = "topleft")
+
+
+reliability(h2_gamlss_mqr,h2$n_attendance)
+reliability(h2_gamlss_mqr,h2$n_attendance,subsets = h2$clock_hour)
+
+pinball(h2_gamlss_mqr,h2$n_attendance,kfolds = h2$kfold,ylim=c(0.3,2))
+
+
+
+
+

### 2020-09-11 paper/paper.Rmd

--- 2020-05-29 paper/paper.Rmd
@@
 library(knitr)
-knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.pos = 'H')
+knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.pos = 'H', message = FALSE, warning = FALSE)
 # Make sure you have the latest versions of rmarkdown and rticle
@@
 library(bookdown)
-h2_hourly <- readr::read_rds("../data/hw_hourly.rds")
+library(patchwork)
+library(tsibble)
+library(kableExtra)
+```
+
+```{r, include=FALSE, cache=FALSE}
+knitr::read_chunk('../r_script/data-viz.R')
 ```
@@
 
+Demand forecasting is a critical input to inform capacity planning and staffing to meet the needs of patients in each hospital services.
+An accurate demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+This is one of the best ways to optimize staffing utilization and related costs.
+Most current practice to optimise personnel scheduling follows the general approach originally presented in  [@vile2016time], which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+
+An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular to depict various courses of action that can result in massive savings in terms of patient lives.
+Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow.
+Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A&E without seen by staff, unexpected return visits to A&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times.without incurring last-minute expenses, such as overtime or supplemental staffing.
+
+An accurate forecasting of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+This will have many advantages for both patients, staff and the quality of healthcare services provided.
+
+Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
+This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory ,Emergency, Surgical, Inpatient, Home and Residential.
+
+The combination of an hourly forecast demand, incidents being attended, resource availability and delays at hospitals, provide information on the state of the unscheduled care system across the service.
+Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
+
+Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
+Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
+That may also require to adjust the schediling.
+
+what literature says about hourly forecasting in healthcare?
+
+There exists a large literature on demand forecasting of patient needs, however the main focus is on the daily  arrivals.
+Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
+In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services
+
+limitations in current literature?
+
 Our contributions are as following:
 
-1. We measure the characteristics of the time series at the original level and at various levels of aggregation. We reveal how TA influences the time series characteristics. 
-2. We develope a model to discover the relationship between the original time series characteristics and the forecasting performance evaluation results to construct rules for method selection.
-3. We determine for which type of time series characteristics, each TA forecasting strategy is preferable.
+1.
+We develop a novel methodology to forecast hourly time series for healthcare services combining a distribution fit to each hour of day separately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, ...
+iv) special events such as public holidays, festive days and rugby and v) weather data.
+2. We provide probabilistic forecasts ...
+3. We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using ....
 
-The rest of the paper is organised as follwoing: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with ...
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with ...
 
-# Research background:  Hourly short-term forecasting in emergency care {#lit}
+# Research background: Hourly forecasting in care services{#lit}
 
-\begin{table}[h]
-\caption{Summary of studies in hourly emergency care forecasting} 
-\centering 
-\begin{adjustbox}{width=1\textwidth}
-\small
+Table \@ref(tab:summarylit) summarise studies in hourly forecasting in emergency and urgent care.
 
-\begin{tabular}{clllll} 
-\hline
-Authors & What to forecast & Method & Forecast evaluation & Limitations  \\ 
-\hline
-
-\citet{hertzum2017forecasting} & Hourly ED patient arrivals and ED occupancy forecasting using calendar variables  1-3 & Regression, ARIMA, naive & MAPE & blabla\\ 
-\citet{mccarthy2008challenge} & To develop methodology for predicting demand for ED services by characterizing ED arrivals  1-3 & Poissopn regression & RMSE & blabla\\ 
-\citet{morzuch2006forecasting} & hourly ED arrivals for 24 hours & Holt-Winters & MAPE & blabla\\
-
-\hline
-\end{tabular}
-\end{adjustbox}
-\label{tab:lit-summary}
-\end{table}
-
-Table \@ref(tab:lit-summary) summarise studies in hourly forecasting in emergency and urgent care.
+```{r summarylit}
+li <- readxl::read_xlsx("../table/table.xlsx")
+knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position")) %>% column_spec(5,"15em")
+```
 
@@
 However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in @choudhury2020forecasting
+@morzuch2006forecasting use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
+
+@mccarthy2008challenge predict patient demand for ED services by characterising ED arrivals.
+Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
+Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
+Then, they presente the predictiona interval accuracy of 50 and 90% intervals for the number of hourly arrivals under the Poisson assumptions.
+They show that the most important predictor is hour of the day and autocorrelation lag 1.
+However, these findings are limited to the short period of data (one year).
+
+@schweigler2009forecasting conducted an investigation on whether using time series methods could accurately generate short-term forecasts of ED bed occupancy. 
+A year-long dataset of hourly ED bed occupancy was collected from three facilities.
+For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
+In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by 33% the 12-hour forecasts generated by historical average).
+
+@kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme. 
+The study adopted historical mean linear regression as benchmark, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method), able to adjust changes in variance over time and vector autoregressive (VAR) method, able to incorporate data from different sources, to forecast for 4 hours, 24 hours.
+They used MAPE to report the forecast accuracy.
+Each of the forecasting models outperformed the benchmark model. In particular, ARIMA model performed best.
+
+@gijo2016sarima generated a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.
+Since historical data showed seasonality, SARIMA models were investigated.
+Regarding the daily model, the authors generated a SARIMA model, which, however, resulted in the forecast error (MMSE) that significantly increased when the lead time exceeded 8 days.
+On the other hand, the SARIMA model proposed to forecast the log-calls an hourly basis.
+This model was found to fit well the model both for shorter and longer lead times.
+
+@asheim2019real developed a Poisson time-series regression model with continuous weekly and yearly cyclic effects to implement a real-time system that could forecast ED arrivals on 1,2,3 hours horizon.
+Once measured the accuracy using the MAPE metric, it was noticed that great improvement happened when time of notification was incorporated into the model, especially on a one-hour horizon.
+Therefore, time of patient notification must be available for this model to be successful.
+
+@steins2019forecasting aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls.
+Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic. 
+In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models.
+These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE.
+Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day.
+Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model.
+However, the improvement provided by the more advance model was not much greater than the one provided by the existing model.
+Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors.
+Moreover, both models either underestimated or overestimated the number of calls.
+Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend.
+Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.
+
 According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
@@
 
-```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width="80%"}
-h2_hourly <- h2_hourly %>% filter(arrival_1h >="2015-01-01 00:00:00"&arrival_1h <"2019-01-01 00:00:00")
-  subseriesplot <- h2_hourly %>% as_tibble() %>% mutate(day=factor(yday(arrival_1h)),year=factor(year(arrival_1h)),hour=hour(arrival_1h)) %>% 
-    select(-arrival_1h)
-time <-  format( seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date()+1), by = "1 hour"),
-                 "%H:%M", tz="GMT")
-time <- time[-length(time)]
-p1 <- ggplot(data=subseriesplot,mapping = aes(x=hour ,y=n_attendance, colour=day))+
-    geom_line(aes(group=interaction(day,year)))+
-    geom_point()+
-    theme_bw()+
-    theme(legend.position = "")+
-    labs(y="Number of visits in ED", x="Hour of day")+
-      scale_x_continuous(breaks=0:23, labels = time)+
-  theme(axis.text.x = element_text(angle = 90))
-  p1+theme(plot.title = element_text(size = 10))
+Figure \@ref(fig:24hour)
+
+```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',fig.height=9, fig.width= 6}
+
 ```
 
-Figure \@ref(fig:24hour)
-
-```{r 24hour, fig.cap= "Distrubution of ED attendance for each hour of day", fig.align='center',fig.width=6, fig.height=9}
-h2_hourly %>% as_tibble() %>% mutate(Date=date(arrival_1h),Hour=hour(arrival_1h)) %>% 
-  select(-arrival_1h) %>% ggplot(mapping=aes(x=n_attendance))+ 
-geom_histogram(mapping=aes(y=stat(density)), binwidth = 1,fill="white",color = 'black')+
-  geom_density(color="blue")+
-scale_y_continuous(labels = scales::percent_format())+
-facet_wrap(~Hour, scales = "free", ncol = 3)+
-theme_bw()+
-labs(x="Number of visits in ED", y="Percentage")+
-theme(axis.text = element_text(size = 7))
+```{r 24hour, fig.cap= "Distrubution of ED attendance for each hour of day", fig.align='center',fig.width=6, fig.height=9, dependson="hourly-plot"}
 ```
@@
 
+### Multiple Regression
+
+### Prophet
+
+Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects.
+Prophet is robust to missing data and shifts in the trend, and typically handles outliers well.
+Prophet works well on daily data seen in Facebook.
+It is popular and automated, making it easy to learn for beginners.
+The implementation may be less flexible than other methods.
+The model is incorporated using corresponding implementation of the Fable package in R.
+We use the prophet() function in the fable package to generate hourly forecasts [@fable2020].
+Note that but the input data is assigned with an hourly and daily seasonality.
+
+### FASSTER
+
+FASSTER [@fasster2020] stands for "Forecasting with Additive Switching of Seasonality, Trend and Exogenous Regressors.
+FASSTER is designed to be a flexible model for capturing multiple seasonality.
+It extends commonly used state space models such as TBATS and Double seasonal Holt-Winter by introducing a switching component to the measurement equation.
+It is a structural state space model, with:
+
+- Multiple seasonality capabilities
+- Flexible model specification
+- Multiple sources of error
+- Exogenous regressor support
+- Handling of missing values
+
+## TBATS
+
+@de2011forecasting proposed a model to deals with time series exhibiting multiple complex seasonalities.
+TBATS includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonality terms.
+The later one not only gives the model more flexibility to deal with complex seasonality but also reduces the parameters of model when the frequencies of seaonalities are high.
+We fit a TBATS model to our daily time series using the `tbats()` function in the forecast package of R [@forecastpackage2020].
+
+## Forecast performance evaluation {#accuracy}
+
+Example of writing an equation!
 \begin{equation}
@@
 
-## forecast performance evaluation {#accuracy}
-
 # Result and discussion {#result}

### 2020-09-11 paper/paper.tex

--- 2020-05-29 paper/paper.tex
@@
 % Pandoc header
-\usepackage{adjustbox, float}
+\usepackage{adjustbox, float,lscape}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 
+Demand forecasting is a critical input to inform capacity planning and staffing to meet the needs of patients in each hospital services.
+An accurate demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+This is one of the best ways to optimize staffing utilization and related costs.
+Most current practice to optimise personnel scheduling follows the general approach originally presented in \citep{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+
+An accurate demand forecasting is crucial in Accident \& Emergency and Ambulance Services in particular to depict various courses of action that can result in massive savings in terms of patient lives.
+Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow.
+Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A\&E without seen by staff, unexpected return visits to A\&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times.without incurring last-minute expenses, such as overtime or supplemental staffing.
+
+An accurate forecasting of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+This will have many advantages for both patients, staff and the quality of healthcare services provided.
+
+Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
+This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory ,Emergency, Surgical, Inpatient, Home and Residential.
+
+The combination of an hourly forecast demand, incidents being attended, resource availability and delays at hospitals, provide information on the state of the unscheduled care system across the service.
+Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
+
+Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during a day or a week.
+Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
+That may also require to adjust the schediling.
+
+what literature says about hourly forecasting in healthcare?
+
+There exists a large literature on demand forecasting of patient needs, however the main focus is on the daily arrivals.
+Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
+In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services
+
+limitations in current literature?
+
 Our contributions are as following:
@@
 \item
-  We measure the characteristics of the time series at the original level and at various levels of aggregation. We reveal how TA influences the time series characteristics.
+  We develop a novel methodology to forecast hourly time series for healthcare services combining a distribution fit to each hour of day separately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, \ldots{} iv) special events such as public holidays, festive days and rugby and v) weather data.
 \item
-  We develope a model to discover the relationship between the original time series characteristics and the forecasting performance evaluation results to construct rules for method selection.
+  We provide probabilistic forecasts \ldots{}
 \item
-  We determine for which type of time series characteristics, each TA forecasting strategy is preferable.
+  We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using \ldots.
 \end{enumerate}
 
-The rest of the paper is organised as follwoing: section \ref{lit} provides a brief overview of the use of temporal aggregation in time series forecasting. Section \ref{model} starts with \ldots{}
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with \ldots{}
 
 \hypertarget{lit}{%
-\section{Research background: Hourly short-term forecasting in emergency care}\label{lit}}
+\section{Research background: Hourly forecasting in care services}\label{lit}}
 
-\begin{table}[h]
-\caption{Summary of studies in hourly emergency care forecasting} 
+Table \ref{tab:summarylit} summarise studies in hourly forecasting in emergency and urgent care.
+
+\begin{table}[!h]
+
+\caption{\label{tab:summarylit}Summary of studies in hourly emergency care forecasting}
 \centering
-\begin{adjustbox}{width=1\textwidth}
-\small
-
-\begin{tabular}{clllll} 
-\hline
-Authors & What to forecast & Method & Forecast evaluation & Limitations  \\ 
-\hline
-
-\citet{hertzum2017forecasting} & Hourly ED patient arrivals and ED occupancy forecasting using calendar variables  1-3 & Regression, ARIMA, naive & MAPE & blabla\\ 
-\citet{mccarthy2008challenge} & To develop methodology for predicting demand for ED services by characterizing ED arrivals  1-3 & Poissopn regression & RMSE & blabla\\ 
-\citet{morzuch2006forecasting} & hourly ED arrivals for 24 hours & Holt-Winters & MAPE & blabla\\
-
-\hline
-\end{tabular}
-\end{adjustbox}
-\label{tab:lit-summary}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lrll>{\raggedright\arraybackslash}p{15em}l}
+\toprule
+Author & Year & Forecast variable & Forecast Horizon & Method & Measure\\
+\midrule
+McCarthy 
+2008 & 2008 & ED arrivals & 24h & Poisson log-linear regression model, including temporal factors, patient characteristics and climatic factors & 95\% CI\\
+Asheim et al. & 2019 & ED arrivals & 3h & Poisson regression with weekly and yearly cyclic 
+effects. & MAPE\\
+Cote et al. & 2013 & ED arrivals & 24h & Fourier regression & R\textasciicircum{}2,  Standard Error\\
+Kim et al. & 2014 & Hospital demand & 4h, 24h, 7days, 
+30days & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE\\
+Schweigler et al. & 2009 & ED bed occupancy & 4h and 12h & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE\\
+Channouf et al. & 2007 & Ambulance demand & 12h,14h,17h,23h,24h,1h,3h,6h, 13h & Regression & RMSE\\
+Hertzum & 2017 & ED arrivals
+ED occupancy & 1,2,4,8,24 hours & linear regression; SARIMA; Naïve & MAE, MAPE, 
+MASE\\
+Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME\\
+Steins et al. & 2019 & Ambulance call & 24h & ZIP and ZINB regression;
+ moving average with seasonality weights & ME, MAE, RMSE\\
+Jones et al. & 2009 & ED census & 24h & VAR;  Holt winters & MAE\\
+Morzuch and Allen & 2006 & ED arrivals & 168h & Regression; ARIMA; Exponential smoothing & RMSE\\
+Chase et al. & 2012 & ED CUR & 30m
+1h, 2h, 4h, 8h, 12h & Binary regression & NA\\
+Taylor & 2008 & centre volume & 30 minutes 
+to 2 weeks & ARMA, Exponential smoothing; Dynamic harmonic regression; Seasonal random walk; Seasonal mean;
+Random walk; Mean; Median; Simple exponential smoothing & MAE
+RMSE\\
+Gijo and Balakrishna & 2016 & call volume & 168h & SARIMA & MMSE\\
+\bottomrule
+\end{tabular}}
 \end{table}
-
-Table \ref{tab:lit-summary} summarise studies in hourly forecasting in emergency and urgent care.
 
@@
 However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in \citet{choudhury2020forecasting}
+\citet{morzuch2006forecasting} use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
+
+\citet{mccarthy2008challenge} predict patient demand for ED services by characterising ED arrivals.
+Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
+Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
+Then, they presente the predictiona interval accuracy of 50 and 90\% intervals for the number of hourly arrivals under the Poisson assumptions.
+They show that the most important predictor is hour of the day and autocorrelation lag 1.
+However, these findings are limited to the short period of data (one year).
+
+\citet{schweigler2009forecasting} conducted an investigation on whether using time series methods could accurately generate short-term forecasts of ED bed occupancy.
+A year-long dataset of hourly ED bed occupancy was collected from three facilities.
+For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
+In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by 33\% the 12-hour forecasts generated by historical average).
+
+\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
+The study adopted historical mean linear regression as benchmark, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method), able to adjust changes in variance over time and vector autoregressive (VAR) method, able to incorporate data from different sources, to forecast for 4 hours, 24 hours.
+They used MAPE to report the forecast accuracy.
+Each of the forecasting models outperformed the benchmark model. In particular, ARIMA model performed best.
+
+\citet{gijo2016sarima} generated a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.
+Since historical data showed seasonality, SARIMA models were investigated.
+Regarding the daily model, the authors generated a SARIMA model, which, however, resulted in the forecast error (MMSE) that significantly increased when the lead time exceeded 8 days.
+On the other hand, the SARIMA model proposed to forecast the log-calls an hourly basis.
+This model was found to fit well the model both for shorter and longer lead times.
+
+\citet{asheim2019real} developed a Poisson time-series regression model with continuous weekly and yearly cyclic effects to implement a real-time system that could forecast ED arrivals on 1,2,3 hours horizon.
+Once measured the accuracy using the MAPE metric, it was noticed that great improvement happened when time of notification was incorporated into the model, especially on a one-hour horizon.
+Therefore, time of patient notification must be available for this model to be successful.
+
+\citet{steins2019forecasting} aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls.
+Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic.
+In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models.
+These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE.
+Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day.
+Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model.
+However, the improvement provided by the more advance model was not much greater than the one provided by the existing model.
+Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors.
+Moreover, both models either underestimated or overestimated the number of calls.
+Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend.
+Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.
+
 According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
@@
 
+\hypertarget{multiple-regression}{%
+\subsubsection{Multiple Regression}\label{multiple-regression}}
+
+\hypertarget{prophet}{%
+\subsubsection{Prophet}\label{prophet}}
+
+Prophet is a forecasting procedure created by Facebook \citep{taylor2018forecasting} that accounts for multiple seasonality, piecewise trend and holiday effects.
+Prophet is robust to missing data and shifts in the trend, and typically handles outliers well.
+Prophet works well on daily data seen in Facebook.
+It is popular and automated, making it easy to learn for beginners.
+The implementation may be less flexible than other methods.
+The model is incorporated using corresponding implementation of the Fable package in R.
+We use the prophet() function in the fable package to generate hourly forecasts \citep{fable2020}.
+Note that but the input data is assigned with an hourly and daily seasonality.
+
+\hypertarget{fasster}{%
+\subsubsection{FASSTER}\label{fasster}}
+
+FASSTER \citep{fasster2020} stands for "Forecasting with Additive Switching of Seasonality, Trend and Exogenous Regressors.
+FASSTER is designed to be a flexible model for capturing multiple seasonality.
+It extends commonly used state space models such as TBATS and Double seasonal Holt-Winter by introducing a switching component to the measurement equation.
+It is a structural state space model, with:
+
+\begin{itemize}
+\tightlist
+\item
+  Multiple seasonality capabilities
+\item
+  Flexible model specification
+\item
+  Multiple sources of error
+\item
+  Exogenous regressor support
+\item
+  Handling of missing values
+\end{itemize}
+
+\hypertarget{tbats}{%
+\subsection{TBATS}\label{tbats}}
+
+\citet{de2011forecasting} proposed a model to deals with time series exhibiting multiple complex seasonalities.
+TBATS includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonality terms.
+The later one not only gives the model more flexibility to deal with complex seasonality but also reduces the parameters of model when the frequencies of seaonalities are high.
+We fit a TBATS model to our daily time series using the \texttt{tbats()} function in the forecast package of R \citep{forecastpackage2020}.
+
+\hypertarget{accuracy}{%
+\subsection{Forecast performance evaluation}\label{accuracy}}
+
+Example of writing an equation!
 \begin{equation}
@@
 Also see Equation \eqref{eq:mean}
-
-\hypertarget{accuracy}{%
-\subsection{forecast performance evaluation}\label{accuracy}}
 

### 2020-09-11 gam-mgcv/study_v1.R
Update study_v1.R
Grouped DOW feature. Mods to gamlss. Additiona of long term trend.

@@ -52,25 +52,32 @@ h2[,issueTime:=lubridate::force_tz(issueTime,tzone = "Europe/London")]
 h2[,targetTime:=NULL]
 #h2[!(h2[,issueTime]%in%h2[,targetTime_UK]),]
 h2 <- h2[(targetTime_UK-issueTime)/3600<=48,]
-
-
 setcolorder(h2, c("issueTime","targetTime_UK"))
 setkey(h2,"issueTime","targetTime_UK")
+
+
 
 ## Set-up CV - Test Data: 1/3/2018 to 28/2/2019
 h2[issueTime>="2018-03-01" & issueTime<"2019-03-01",kfold:="Test"]
 h2[issueTime<"2018-03-01",kfold:=paste0("fold",rep(rep(1:2,each=24*7),length.out=.N))]
 
 
+## Features
+
+# Weekday groups
+h2[dow%in%c("Mon","Tue","Wed","Thu","Fri"),dow2:="MonFri"]
+h2[!dow%in%c("Mon","Tue","Wed","Thu","Fri"),dow2:="SatSun"]
+h2[,dow2:=as.factor(dow2)]
 
 
 ## Fit Poisson-GAM and visualise model ####
 
 for(fold in unique(h2$kfold)){
   
-  gam1 <- bam(n_attendance ~ s(clock_hour,k=24,by=dow) +
-                s(doy,k=26) +
-                ti(doy,clock_hour,k=c(6,6)),
+  gam1 <- bam(n_attendance ~
+                # s(clock_hour,k=24,by=dow2) + t + I(t^2) +
+                # ti(doy,clock_hour,k=c(6,6)),
+                te(doy,clock_hour,k=c(6,24),by=dow2) + t,
               data=h2[kfold!=fold & kfold!="Test",],family = poisson())
   
   h2[kfold==fold,lambda:=predict(gam1,newdata =h2[kfold==fold,],type="response")]
@@ -89,20 +96,21 @@ sqrt(mean((gam1$y-gam1$fitted.values)^2))
 ## Check fit - is the model using all DOF? Is so, consider increasing availability...
 gam.check(gam1)
 plot(gam1,pages = 2)
-plot(gam1,select = 8,scheme = 1)
+plot(gam1,select = 3,scheme = 1)
 
+
+## GAM VIS
 gam1 <- getViz(gam1,nsim = 200)
-
 
 check1D(gam1,"clock_hour")
 check1D(gam1,"doy")
[…]

### 2021-03-03 adam-model/experiment.R
Minor typos et al.

@@ -1,5 +1,4 @@
 # ADAM experiment
-devtools::install_github("config-i1/gnorm", upgrade="never", dependencies=FALSE)
 devtools::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
 devtools::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
 
@@ -15,7 +14,7 @@ library(ProbCast)
 
 ## Load and Prep Data ####
 # load("../data/hw_hourly.rds")
-h2 <- fread("data/h2_hourly.csv")
+h2 <- fread("data/h2_hourly_gb.csv")
 h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
 h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
 
@@ -34,75 +33,41 @@ h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
 
 # Prepare explanatory variables
 xreg <- as.character(h2$holiday_festive_day)
-xreg[is.na(h2$holiday_festive_day)] <- "none"
-xreg <- factor(xreg)
+# Do leads and lags of the variables
+xreg <- xregExpander(xreg, lags=c(-2:1)*24, gaps="NAs")
+xreg <- as.data.table(xreg)
+xreg$x[is.na(xreg$x)] <- "none"
+xreg$xLag24[is.na(xreg$xLag24)] <- "none"
+xreg$xLag48[is.na(xreg$xLag48)] <- "none"
+xreg$xLead24[is.na(xreg$xLead24)] <- "none"
+xreg$x <- factor(xreg$x)
+xreg$xLag24 <- factor(xreg$xLag24)
+xreg$xLag48 <- factor(xreg$xLag48)
+xreg$xLead24 <- factor(xreg$xLead24)
 
 # Create zoo objects
 # The first observations is from 31st March
-x <- zoo(h2$n_attendance, order.by=h2$targetTime)
-xreg <- zoo(xreg, order.by=h2$targetTime)
-xregData <- data.frame(x=x,xreg=xreg)
-# xFourier <- fourier(msts(as.vector(x),seasonal.periods=c(24,24*7,24*365.25)), K=c(10,10,10))
-# xregDataFourier <- cbind(xregData,xFourier)
[…]

### 2021-03-03 adam-model/experiment.R
Created a matrix with dummies, trying to improve the speed

@@ -59,6 +59,35 @@ xregExpanded <- cbind(xregExpanded,
                       hourOfDay=as.factor(temporaldummy(y,type="hour",of="day",factors=TRUE)),
                       dayOfWeek=as.factor(temporaldummy(y,type="day",of="week",factors=TRUE)),
                       weekOfYear=as.factor(temporaldummy(y,type="week",of="year",factors=TRUE)))
+
+#### Expand the matrix, creating dummy variables from the factors
+xregDummies <- model.matrix(~.,xregExpanded)[,-1]
+colnames(xregDummies) <- make.names(colnames(xregDummies), unique=TRUE)
+xregDummiesShort <- xregDummies
+
+# Check, which variables are perfectly correlated
+correlatedVariables <- vector("logical",ncol(xregDummiesShort)-1)
+for(i in 2:ncol(xregDummiesShort)){
+  correlatedVariables[i-1] <- any(apply(xregDummiesShort[,-c(1,i)]==xregDummiesShort[,i],2,all));
+}
+# Remove the ones that are perfectly correlated
+while(any(correlatedVariables)){
+  correlatedVariables <- vector("logical",ncol(xregDummiesShort)-1)
+  for(i in 2:ncol(xregDummiesShort)){
+    correlatedVariables[i-1] <- any(apply(xregDummiesShort[,-c(1,i)]==xregDummiesShort[,i],2,all));
+  }
+  # Kick off the last most correlated
+  if(any(correlatedVariables)){
+    xregDummiesShort <- xregDummiesShort[,-tail(which(correlatedVariables),1)-1];
+  }
+  cat(sum(correlatedVariables), "left; ");
+}
+
+stepwiseRegression <- stepwise(xregDummiesShort)
+
+save(xreg, xregExpanded, xregDummies, xregDummiesShort, file="adam-model/xreg.Rdata")
+rm(xregDummies)
+
 # colnames(xregExpanded) <- make.names(colnames(xregExpanded), unique=TRUE)
 # xregExpandedFourier <- cbind(xregExpanded,xFourier)
 
@@ -83,9 +112,9 @@ experimentResultsTestIvan <- foreach(i=1:((testSet-36)/rohStep)) %dopar% {
   
   #### First approach - Non-seasonal iETS with dummies
   j <- 1
[…]

### 2021-03-04 adam-model/experiment.R
Trying to make the code work...

@@ -114,7 +114,7 @@ experimentResultsTestIvan <- foreach(i=1:((testSet-36)/rohStep)) %dopar% {
   j <- 1
   oesModel <- oes(as.vector(xregDummiesShort[,1]), "MNN", h=testSet-(i-1)*rohStep, holdout=TRUE, occurrence="direct")
   adamModel <- adam(xregDummiesShort, "MNN", lags=1, h=testSet-(i-1)*rohStep, holdout=TRUE, initial="b",
-                    occurrence=oesModel, regressors="use", distribution="dgnorm", shape=1.5, formula=formula(stepwiseRegression))
+                    occurrence=oesModel, regressors="use")
   testForecast <- forecast(adamModel, interval="prediction", h=h, level=c(1:19/20), side="u")
   # Mean values
   errorMeasuresValues[j,"Mean",] <- testForecast$mean
@@ -129,7 +129,7 @@ experimentResultsTestIvan <- foreach(i=1:((testSet-36)/rohStep)) %dopar% {
   #### Second approach - Seasonal iETSX with m1=24 and m2=24*7
   j <- 2
   oesModel <- oes(as.vector(y), "MNN", h=testSet-(i-1)*rohStep, holdout=TRUE, occurrence="direct")
-  adamModel <- adam(xregExpanded[,-c(6:8)], "MNM", lags=c(24,24*7), h=testSet-(i-1)*rohStep, holdout=TRUE, initial="o",
+  adamModel <- adam(xregExpanded[,-c(6:8)], "MNM", lags=c(24,24*7), h=testSet-(i-1)*rohStep, holdout=TRUE, initial="b",
                     occurrence=oesModel, formula=y~x)
   testForecast <- forecast(adamModel, interval="pred", h=h, level=c(1:19/20), side="u")
   # Mean values

### 2021-03-05 gam-mgcv/study_v1.R
Notes from meeting

@@ -4,6 +4,13 @@
 ## Quantiles from 0.05 to 0.95, Pinball loss
 ## Test Data: 1/3/2018 to 28/2/2019
 
+
+## TO DO ####
+# Optimise smoothig period: n_att_rollmean
+# BENCHMARK 1
+# Pre-processing: rolling max/mean etc...
+
+## Begin... ####
 require(rstudioapi)
 require(data.table)
 require(mgcv)
@@ -102,6 +109,16 @@ require(ggplot2)
 plot_data <- h2[,.(n_attendance=mean(n_attendance)),by=c("dow","clock_hour")]
 ggplot(data=plot_data,aes(x=clock_hour,y=n_attendance,color=dow))+
   geom_line()
+
+
+
+## Benchmark 1 #### Empirical distribution by hour of the day
+
+
+## TO DO!!! <<<<<<<<<<<<<<<<<
+
+
+
 
 ## Fit Poisson-GAM and visualise model ####
 

### 2021-03-05 results/Evaluatoin Script.R

@@ -11,6 +11,7 @@
 # test ceiling/floor options for non-integer forecasts
 # report RMSE
 
+
 ## Link to use case:
 # "Process 95% of patients in 4h"
 
@@ -60,6 +61,27 @@ for(n in names(JB_results)){
 rm(JB_results)
 
 ## Bahman's Results ####
+
+tbats <- data.table(readRDS("tbats.rds"))
+setnames(tbats,old = c("origin","target"),c("issueTime","targetTime_UK"))
+## Get Actuals
+actuals <- merge(tbats[targetTime_UK>=test_start,.(issueTime,targetTime_UK)],h2[,.(targetTime_UK,n_attendance)],
+                 by="targetTime_UK",all.x=T)
+setkey(actuals,issueTime,targetTime_UK)
+
+## Test Data Pinball & Reliability
+temp <- data.table(pinball(tbats[,-c(1:2)],actuals[,n_attendance]))
+temp[,Method:="tbats"]; temp[,kfold:="Test"]
+PB <- rbind(PB,temp); rm(temp)
+
+temp <- data.table(reliability(tbats[,-c(1:2)],actuals[,n_attendance]))
+temp[,Method:="tbats"]; temp[,kfold:="Test"]
+REL <- rbind(REL,temp); rm(temp)
+
+rm(tbats)
+
+
+
 
 # Prophet <- data.table(readRDS("forecast_prophet.rds"))
 # setnames(Prophet,colnames(Prophet),c("issueTime","targetTime_UK",paste0("q",as.numeric(colnames(Prophet[,-c(1:2)]))*100)))

### 2021-06-22 paper/paper.Rmd
results for tbats
Regression, ETS and ADAM

@@
 ### Multiple Regression
+Regression is one of the most popular forecasting methods that uses explanatory variables to predict the response one. The classical linear regression model is formulated as:
+\begin{equation}
+  \mathbf{y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
+(\#eq:linearRegression)
+\end{equation}
+where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters and $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
+However, in the context of healthcare and A&E admittance, the assumption of Normal distribution is unrealistic, so the linear regression model should be substituted by some other models.
+One of these models frequently used in practice is the Poisson regression, which can be summarised as:
+\begin{equation}
+  \mathbf{y}_t \sim \mathrm{Poisson} \left( \exp \left( \mathbf{x}_t' \boldsymbol{\beta} \right) \right).
+(\#eq:PoissonRegression)
+\end{equation}
+The exponent in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
+When it comes to deciding what explanatory variables to include in the model, there is no one correct answer and the decision needs to be done based on each specific case.
+In our experiment we will include the following dummy variables:
+
+1. Hour of day,
+2. Day of week,
+3. Week of year,
+4. Holidays (such as Christmas, New Year etc),
+5. 24 hours lags of holidays.
+
+The variables (1) - (3) allow us modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
+Finally (5) is needed in order to cover the potential phenomenon of change in admittance after the holiday (e.g.
+people might try not to go to hospital on Christmas eve and thus will go the day after).
+We use `alm()` function from `greybox` v0.7.0 package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
+
+### ETS - Exponential Smoothing model
+@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal (thus ETS).
+Given the popularity of ETS models, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 as a benchmark.
+This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as "ETS".
+
+### ADAM: multiple seasonal ETSX
+@SvetunkovAdam2021 proposed a framework for dynamic models called "ADAM" - Augmented Dynamic Adaptive Model.
+This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
+We use Gamma distribution for ETS(M,N,M) model with frequencies 24 and 168, adding explanatory variables for week of year, holidays and lagged holidays.
+Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those zeroes.
+This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote this model as "ADAM".
 

### 2021-06-22 r_script/fasster.R
models for running time added

@@ -1,0 +1,41 @@
+library(lubridate)
+library(patchwork)
+library(fable)
+library(tidyverse)
+library(tsibble)
+library(fabletools)
+library(fasster)
+
+#admission
+ae_original <- readr::read_csv("data/h2_hourly.csv")
+
+#holidays 
+holidays_ae <- readxl::read_xlsx("data/holiday_rugby_all.xlsx")
+
+holidays_ae %>%
+  mutate(xmas=if_else((mday(Date)==25 & month(Date)==12),1,0),
+         new_year=if_else((mday(Date)==1 & month(Date)==1),1,0)) %>% 
+  select(Date,is_public_holiday,is_school_holiday,xmas,new_year) %>%
+  slice(rep(1:n(), each=24)) %>% select(-Date)->holiday_wales
+
+#join holiday ans ae data
+
+bind_cols(ae_original,
+          holiday_wales) %>% as_tsibble()->data_for_forecast
+data_for_forecast %>% mutate(is_public_holiday=as_factor(is_public_holiday), 
+                             is_school_holiday=as_factor(is_school_holiday),
+                             xmas=as_factor(xmas),
+                             new_year=as_factor(new_year)) %>% force_tz(arrival_1h,tzone = "UTC")->data_for_forecast
+train <- data_for_forecast %>% filter(arrival_1h < lubridate::ymd_hms("2019-02-27 00:00:00"))
+test <- data_for_forecast %>% filter(arrival_1h >= lubridate::ymd_hms("2019-02-27 00:00:00"))
+
+# model
+fit_fasster <- train %>%
+  model(
+    fass=FASSTER(sqrt(n_attendance) ~ fourier(period = "day", K = 10) +
+                   fourier(period = "week", K = 5) +
+                   fourier(period = "year", K = 3)+
+                   is_public_holiday+ is_school_holiday+xmas+new_year+
+                   trend(2))
+  )
+ae_fc <- fit_fasster %>% forecast(new_data=test)

### 2021-06-22 paper/paper.Rmd
Affiliations of Ivan

@@
     footnote: 2
-  - name: Author3
-    email: email3@example.com
-    affiliation: University3
+  - name: Ivan Svetunkov
+    email: i.svetunkov@lancaster.ac.uk
+    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
     footnote: 2

### 2021-06-22 paper/paper.Rmd
Added bits to paper

--- 2020-09-11 paper/paper.Rmd
@@
 Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow.
-Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A&E without seen by staff, unexpected return visits to A&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times.without incurring last-minute expenses, such as overtime or supplemental staffing.
+Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A&E without seen by staff, unexpected return visits to A&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
 
-An accurate forecasting of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+An accurate forecast of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This will have many advantages for both patients, staff and the quality of healthcare services provided.
 
+Furthermore, because the consequences of imperfact staffing are asymmetric probabilistic forecasts are necessary to make economic decicions.
+This asymmetry arrises because it is preferable to incure a small opportunity cost associated unerutilised staff rather than lower service levels if staff levels are insufficient.
+
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory ,Emergency, Surgical, Inpatient, Home and Residential.
+This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
 
@@
 limitations in current literature?
+- Typically focus on daily data, hiding information about peak arrival rates. Hourly data has additional challenge of multiple seasonality.
+- Deterministic, are there any examples of probabilistic?
+
+Other papers to review and possibly cite:
+- Daily visits using Prophet 10.1016/j.annemergmed.2019.08.149
+- Daily using internet data 10.1016/j.annemergmed.2014.10.008
+- Review paper: 10.1136/emj.2008.062380
+- Others: 10.1136/emj.2010.103531, 10.1136/emj.2007.051656
+- Impact of weather 10.1136/emj.2010.091058, 10.1136/emj.20.2.204
 
@@
 3. We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using ....
+4. We provide data and code enabling reproduction and refinement of the proposed approach
 
@@
 
-# Proposed model {#model}
+# Proposed models {#model}
+
+## Exponential Smoothing
+
+## GAMLSS
 
@@
 ## benchmarks {#benchmarks}
+
+### Naive/Climatology
+
+Emprical distribution for same hour-of-day and day-of-week as target time.
 
@@
 
-## TBATS
+### TBATS
 

### 2021-07-06 paper/paper.Rmd
Expanded some elements

--- 2021-06-22 paper/paper.Rmd
@@
 ### Multiple Regression
-Regression is one of the most popular forecasting methods that uses explanatory variables to predict the response one. The classical linear regression model is formulated as:
+Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A&E admittance).
+The classical linear regression model is formulated as:
 \begin{equation}
@@
 where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters and $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
-However, in the context of healthcare and A&E admittance, the assumption of Normal distribution is unrealistic, so the linear regression model should be substituted by some other models.
-One of these models frequently used in practice is the Poisson regression, which can be summarised as:
+However, in the context of healthcare and A&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
+So the linear regression model should be substituted by some other model.
+One of the models that is frequently used in practice is the Poisson regression, which can be summarised as:
 \begin{equation}
@@
 The exponent in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
-When it comes to deciding what explanatory variables to include in the model, there is no one correct answer and the decision needs to be done based on each specific case.
-In our experiment we will include the following dummy variables:
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function.
+When it comes to selecting explanatory variables for the model, there is no one correct answer and the decision needs to be done based on each specific case.
+In our experiment, we will only include dummy variables, capturing a variety of calendar events:
 
@@
 
-The variables (1) - (3) allow us modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
-Finally (5) is needed in order to cover the potential phenomenon of change in admittance after the holiday (e.g.
-people might try not to go to hospital on Christmas eve and thus will go the day after).
+The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
+Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g.
+people might try not to go to hospital on Christmas eve and thus will go the next day).
+This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time.
 We use `alm()` function from `greybox` v0.7.0 package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
@@
 ### ETS - Exponential Smoothing model
-@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal (thus ETS).
-Given the popularity of ETS models, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 as a benchmark.
+@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus "ETS").
+Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
 This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as "ETS".
+This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
 
@@
 This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
-We use Gamma distribution for ETS(M,N,M) model with frequencies 24 and 168, adding explanatory variables for week of year, holidays and lagged holidays.
-Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those zeroes.
-This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote this model as "ADAM".
+Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
+Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
+We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
+Although the data is integer, we expect that Gamma distribution will be a good approximation for it.
+This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
 

### 2021-07-07 paper/paper.Rmd
Add gamlss text

--- 2021-06-22 paper/paper.Rmd
@@
 
+Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
+In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
+In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+\begin{equation} \label{eq:basicGAM}
+    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
+\end{equation}
+where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
+If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
+Each $f_{j,i}$ is a smooth function, constructed as
+\begin{equation}
+    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
+\end{equation}
+where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
+The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
+See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
+
 # Experimental design {#design}

### 2021-07-07 paper/paper.Rmd

--- 2021-07-06 paper/paper.Rmd
@@
 
-# Proposed model {#model}
+# Proposed models {#model}
+
+## Exponential Smoothing
+
+## GAMLSS
+
+Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
+In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
+In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+\begin{equation} \label{eq:basicGAM}
+    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
+\end{equation}
+where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
+If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
+Each $f_{j,i}$ is a smooth function, constructed as
+\begin{equation}
+    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
+\end{equation}
+where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
+The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
+See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
 

### 2021-07-07 paper/paper.Rmd
paper sections updated

@@
 
-# Proposed models {#model}
-
-## Exponential Smoothing
-
-## GAMLSS
-
-Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
-In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
-\begin{equation} \label{eq:basicGAM}
-    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
-\end{equation}
-where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
-If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
-Each $f_{j,i}$ is a smooth function, constructed as
-\begin{equation}
-    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
-\end{equation}
-where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
-The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
-
 # Experimental design {#design}
@@
 
-Emprical distribution for same hour-of-day and day-of-week as target time.
+Empirical distribution for same hour-of-day and day-of-week as target time.
 
 ### Multiple Regression
+
 Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A&E admittance).
@@
 ### ETS - Exponential Smoothing model
+
 @Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus "ETS").
@@
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
-
-### ADAM: multiple seasonal ETSX
-@SvetunkovAdam2021 proposed a framework for dynamic models called "ADAM" - Augmented Dynamic Adaptive Model.
-This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
-Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
-We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
-Although the data is integer, we expect that Gamma distribution will be a good approximation for it.
-This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
 
@@
 
+# Model building {#model}
+
+## ADAM: multiple seasonal ETSX
+
+@SvetunkovAdam2021 proposed a framework for dynamic models called "ADAM" - Augmented Dynamic Adaptive Model.
+This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
+Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
+Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
+We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
+Although the data is integer, we expect that Gamma distribution will be a good approximation for it.
+This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
+
+## GAMLSS
+
+Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
+In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
+In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+\begin{equation} \label{eq:basicGAM}
+    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
+\end{equation}
+where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
+If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
+Each $f_{j,i}$ is a smooth function, constructed as
+\begin{equation}
+    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
+\end{equation}
+where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
+The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
+See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
+
 ## Forecast performance evaluation {#accuracy}

### 2021-07-07 r_script/bahman-tbats-prophet.R

@@ -8,6 +8,7 @@ library(fable.prophet)
 library(fabletools)
 library(fasster)
 library(forecast)
+library(future)
 ae_original <- readr::read_csv("data/h2_hourly.csv")
 ae_original %>%  
 mutate(arrival_1h=lubridate::force_tz(arrival_1h,tz="GB")) -> ae_original_tz
@@ -213,16 +214,16 @@ accuracy <- ae_fc %>% accuracy(data_for_forecast)
 #                     season("week",type = "additive",order=3)+
 #                     is_public_holiday+is_rugby+growth("linear"))
 # )
-
-fit_fasster <- ae_tscv %>% filter(.id==727) %>%
+s <- Sys.time()
+plan(multicore)
+fit_fasster <- ae_tscv %>%
   model(
   fass=FASSTER(sqrt(n_attendance) ~ fourier(period = "day", K = 10) +
                  fourier(period = "week", K = 5) +
                  fourier(period = "year", K = 3)+
                  is_public_holiday+ is_school_holiday+xmas+new_year+
-                 trend(1))
+                 trend(2))
 )
-refit_fasster <- refit(fit_fasster, ae_tscv)
 
 e <- Sys.time()
 s-e

### 2021-07-07 paper/paper.Rmd
Expanded ADAM and added a bit on evaluation (see "Forecast performance evaluation")

@@
 
-## ADAM: multiple seasonal ETSX
+## ADAM: multiple seasonal iETSX
 
@@
 Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
+This model can be formulated as a set of the following equations:
+\begin{equation}
+	\begin{aligned}
+	  & y_t = o_t z_t \\
+		& \log z_t = \log l_{t-1} + \log s_{1,t-24} + \log s_{2,t-168} + \mathbf{x}_t' \boldsymbol{\beta} + \log \left(1 + \epsilon_{t} \right) \\
+		& \log l_{t} = \log l_{t-1} + \log( 1  + \alpha \epsilon_{t}) \\ 
+		& \log s_{1,t} = \log s_{1,t-m} + \log( 1  + \gamma_1 \epsilon_{t}) \\
+		& \log s_{2,t} = \log s_{2,t-m} + \log( 1  + \gamma_2 \epsilon_{t}) \\
+		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
+		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
+		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
+		& \mu_{a,t} = \min(l_{a,t-1}, 1)
+	\end{aligned} ,
+	(\#eq:ADAMModel)
+\end{equation}
+where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
+Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$.
 We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
-Although the data is integer, we expect that Gamma distribution will be a good approximation for it.
+Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
+If integer-valued quantiles are needed, then rounding up can be done for them.
 This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
@@
 
-Example of writing an equation!
+In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model.
+The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins.
+Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation.
+The latter is measured via Root Mean Squared Error (RMSE):
 \begin{equation}
-\bar{X} = \frac{\sum_{i=1}^n X_i}{n} (\#eq:mean)
+  \mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{j=1}^n e_j^2} .
+  (\#eq:RMSE)
 \end{equation}
+When it comes to quantile performance, we used pinball, and all the other cool things described by Jethro.
 
-Also see Equation \@ref(eq:mean)
+Also see Equation \@ref(eq:RMSE)
 

### 2021-09-23 gam-mgcv/study_v1_timing.R
Timing script for Ivan

@@ -71,7 +71,7 @@ mean_from_qs <- function(mqr){
 
 setwd(dirname(getActiveDocumentContext()$path))
 
-JB_results <- list()
+JB_results_time <- list()
 # load("../results/JethroResults_2021-05-05.Rda")
 
 
@@ -180,6 +180,10 @@ JB_results <- list()
 ## Benchmark 1 ####
 # Empirical distribution by hour of the day and day type
 
+## %%%% Time for Ivan %%%%
+time_temp <- Sys.time()
+## %%%% Time for Ivan %%%%
+
 # Quantiles
 BenchPred <- h2[kfold!="Test",as.list(quantile(n_attendance,probs = 1:19/20)),by=c("dow","clock_hour")]
 setnames(BenchPred,paste0((1:19/20)*100,"%"),paste0("q",(1:19/20)*100))
@@ -194,15 +198,11 @@ Bench_mqr <- Bench_mqr[,-(1:2)]
 setkey(Bench_mqr,issueTime,targetTime_UK)
 class(Bench_mqr) <- c("MultiQR",class(Bench_mqr))
 
-JB_results[["Benchmark_1"]] <- Bench_mqr
+## %%%% Time for Ivan %%%%
+time_temp <- Sys.time() - time_temp
+JB_results_time[["Benchmark_1"]] <- time_temp
+## %%%% Time for Ivan %%%%
 
-issue <- unique(h2$issueTime)[7]
-plot(Bench_mqr[issueTime==issue,grep("q",colnames(Bench_mqr)),with=F],
-     xlab="Lead-time [hours]",ylab="Attendance",main=paste0("Origin: ",issue," (",format(issue,"%A"),")"),
-     ylim=c(0,40),Legend = "topleft")
-
-reliability(Bench_mqr[,grep("q",colnames(Bench_mqr)),with=F],h2$n_attendance,kfolds = h2$kfold)
-pinball(Bench_mqr[,grep("q",colnames(Bench_mqr)),with=F],h2$n_attendance,kfolds = h2$kfold)
 
 rm(BenchPred,BenchPred_exp,Bench_mqr)
 
@@ -211,7 +211,14 @@ rm(BenchPred,BenchPred_exp,Bench_mqr)
 
 test_start <- h2[kfold=="Test",min(issueTime)]
 Bench_mqr <- copy(h2[,.(issueTime,targetTime_UK,dow,clock_hour)])
-for(test_week in seq(-104,52,by=1)){
+
+
[…]

### 2021-09-23 results/Evaluatoin Script.R

@@ -162,6 +162,7 @@ rm(faster)
 ## Ivan's Results ####
 
 load("IvanValues.RData")
+head(quantileValuesIvan$iETSXSeasonal)
 
 for(n in names(quantileValuesIvan)){
   

### 2021-09-24 paper/paper.Rmd
Add to GAMLSS description

@@
 
+If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
+In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
+
 Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
@@
 \end{equation}
-where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
+where $g_j$ is a monotonic `link' function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
 If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
@@
 See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
+
+As our data are counts, the natural starting point is the Poisson distribution, given by
+\begin{equation}
+  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
+\end{equation}
+where we consider an additive model for $\lambda_t$ of the form
+\begin{equation}
+  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
+\end{equation}
+The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
+
+Truncated Normal...
+\begin{equation}
+  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
+\end{equation}
+with
+\begin{eqnarray}
+  \mu_t &=& \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , \\
+  log(\sigma_t) &=& f(H(t))  \quad .
+\begin{end}
+
+Truncated $t$ distribution...
+
+Negative binomial...
 

### 2021-09-24 paper/paper.Rmd
Text on GAMLSS and evaluation

@@
 \end{equation}
-When it comes to quantile performance, we used pinball, and all the other cool things described by Jethro.
 
-Also see Equation \@ref(eq:RMSE)
+The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a].
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.
+
+The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score.
+It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
+Furthermore,
+The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
+\begin{equation}
+    % \text{Pinball} = 
+    \frac{1}{T|\mathcal{A}|} \sum_{\alpha \in \mathcal{A}} \sum_{t=1}^T
+ \left(q_{\alpha,t} - y_{t} \right)
+ \left(\mathbb{1}(y_{t}\leq_{\alpha,t})-\alpha \right)
+ (\#eq:pinball)
+\end{equation}
+where $\mathcal{A} = {0.05,0.1,...,0.95}$ is the set of quantiles being estimated.
+
+Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations.
+If a forecast is calibrated, then 20\% of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
+This property is necessary for forecast probabilities to be used in quantitative decision-making.
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observer frequency mean($\mathbb{1}(y_{t}\le q_{\alpha,t})$).
 

### 2021-09-24 results/Evaluatoin Script.R

@@ -10,6 +10,9 @@ rm(list=ls())
 # test ceiling/floor options for non-integer forecasts ! << Easy now...
 # report RMSE
 
+## For consistent plots in paper:
+# use ggthemes: theme_few()
+
 
 ## Link to use case:
 # "Process 95% of patients in 4h"

### 2021-09-24 results/Evaluatoin Script.R
Add RMSE to evaluation

@@ -38,7 +38,7 @@ test_start <- "2018-03-01"
 last_issue <- "2019-02-26 00:00:00" #### <<<<< Impliment!!! Make test for data matching
 PB <- data.table()
 REL <- data.table()
-
+RMSE <- data.table()
 
 ## Eval function
 big_eval_function <- function(forecast_DT,h2_actuals,method_name){
@@ -52,28 +52,45 @@ big_eval_function <- function(forecast_DT,h2_actuals,method_name){
   setkey(actuals,issueTime,targetTime_UK)
   
   ## All
-  temp <- data.table(pinball(forecast_DT[,-c(1:2)],actuals[,n_attendance],plot.it = F))
+  temp <- data.table(pinball(forecast_DT[,grep(names(forecast_DT),pattern = "q"),with=F],actuals[,n_attendance],plot.it = F))
   temp[,Method:=method_name]; temp[,kfold:="Test"]; temp[,Horizon:="All"]; temp[,Issue:="All"]
   PB <<- rbind(PB,temp); rm(temp)
   
-  temp <- data.table(reliability(forecast_DT[,-c(1:2)],actuals[,n_attendance],plot.it = F))
+  temp <- data.table(reliability(forecast_DT[,grep(names(forecast_DT),pattern = "q"),with=F],actuals[,n_attendance],plot.it = F))
   temp[,Method:=method_name]; temp[,kfold:="Test"]; temp[,Horizon:="All"]; temp[,Issue:="All"]
   REL <<- rbind(REL,temp); rm(temp)
+  
+  temp <- data.table(Method=method_name,
+                     kfold="Test",
+                     Horizon="All",
+                     Issue="All",
+                     RMSE=sqrt(mean(forecast_DT[,expectation]-actuals[,n_attendance])^2))
+  RMSE <<- rbind(RMSE,temp); rm(temp)
   
   ## By horizon and issue time
   for(issue in c(0,12)){
     for(lt in 0:48){
       temp <- data.table(pinball(forecast_DT[
-        hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,-c(1:2)],
+        hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,grep(names(forecast_DT),pattern = "q"),with=F],
         actuals[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,n_attendance],plot.it = F))
[…]

### 2021-09-24 results/Evaluatoin Script.R
Try() RMSE

@@ -60,12 +60,14 @@ big_eval_function <- function(forecast_DT,h2_actuals,method_name){
   temp[,Method:=method_name]; temp[,kfold:="Test"]; temp[,Horizon:="All"]; temp[,Issue:="All"]
   REL <<- rbind(REL,temp); rm(temp)
   
-  temp <- data.table(Method=method_name,
-                     kfold="Test",
-                     Horizon="All",
-                     Issue="All",
-                     RMSE=sqrt(mean(forecast_DT[,expectation]-actuals[,n_attendance])^2))
-  RMSE <<- rbind(RMSE,temp); rm(temp)
+  try({
+    temp <- data.table(Method=method_name,
+                         kfold="Test",
+                         Horizon="All",
+                         Issue="All",
+                         RMSE=sqrt(mean(forecast_DT[,expectation]-actuals[,n_attendance])^2))
+      RMSE <<- rbind(RMSE,temp); rm(temp)
+  })
   
   ## By horizon and issue time
   for(issue in c(0,12)){
@@ -82,13 +84,15 @@ big_eval_function <- function(forecast_DT,h2_actuals,method_name){
       temp[,Method:=method_name]; temp[,kfold:="Test"]; temp[,Horizon:=lt]; temp[,Issue:=issue]
       REL <<- rbind(REL,temp); rm(temp)
       
+      try({
       temp <- data.table(Method=method_name,
                          kfold="Test",
                          Horizon=lt,
                          Issue=issue,
-                         RMSE=sqrt(mean(forecast_DT[,expectation]-actuals[,n_attendance])^2))
+                         RMSE=sqrt(mean(forecast_DT[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,expectation]-
+                                          actuals[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,n_attendance])^2))
       RMSE <<- rbind(RMSE,temp); rm(temp)
-      
+      })
       
       
       

### 2021-09-24 results/Evaluatoin Script.R
update resulst

@@ -62,11 +62,11 @@ big_eval_function <- function(forecast_DT,h2_actuals,method_name){
   
   try({
     temp <- data.table(Method=method_name,
-                         kfold="Test",
-                         Horizon="All",
-                         Issue="All",
-                         RMSE=sqrt(mean(forecast_DT[,expectation]-actuals[,n_attendance])^2))
-      RMSE <<- rbind(RMSE,temp); rm(temp)
+                       kfold="Test",
+                       Horizon="All",
+                       Issue="All",
+                       RMSE=sqrt(mean(forecast_DT[,expectation]-actuals[,n_attendance])^2))
+    RMSE <<- rbind(RMSE,temp); rm(temp)
   })
   
   ## By horizon and issue time
@@ -85,13 +85,13 @@ big_eval_function <- function(forecast_DT,h2_actuals,method_name){
       REL <<- rbind(REL,temp); rm(temp)
       
       try({
-      temp <- data.table(Method=method_name,
-                         kfold="Test",
-                         Horizon=lt,
-                         Issue=issue,
-                         RMSE=sqrt(mean(forecast_DT[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,expectation]-
-                                          actuals[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,n_attendance])^2))
-      RMSE <<- rbind(RMSE,temp); rm(temp)
+        temp <- data.table(Method=method_name,
+                           kfold="Test",
+                           Horizon=lt,
+                           Issue=issue,
+                           RMSE=sqrt(mean(forecast_DT[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,expectation]-
+                                            actuals[hour(issueTime)==issue & (targetTime_UK-issueTime)/3600 == lt,n_attendance])^2))
+        RMSE <<- rbind(RMSE,temp); rm(temp)
       })
       
       
@@ -175,15 +175,19 @@ rm(quantileValuesIvan)
 
[…]

### 2021-09-24 paper/paper.Rmd

@@
 
+Still to add details on skill scores and bootstrap methods...
+
 # Result and discussion {#result}
+
+(Plots generated with theme_few - they look nice, good suggestion!
+Not sure how to include them in bookdown, can you help, Bahman?
+The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format?
+I guess PDF might be preferred.
+I will also use xtable to make a few tables of results, for now there is a CSV in the results folder.
+I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
 

### 2021-09-24 paper/paper.tex

@@
 \usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
 %%%%%%%%%%%%%%%% end my additions to header
@@
 \usepackage{natbib}
-\bibliographystyle{plainnat}
-\usepackage{longtable}
+\bibliographystyle{authoryear}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \ifxetex
@@
 % Pandoc toggle for numbering sections (defaults to be off)
+
+% Pandoc citation processing
 
@@
    \ead{email2@example.com} 
-    \author[University3]{Author3\corref{2}}
-   \ead{email3@example.com} 
+    \author[Centre for Marketing Analytics and Forecasting, Lancaster University, UK]{Ivan Svetunkov\corref{2}}
+   \ead{i.svetunkov@lancaster.ac.uk} 
       \address[University1]{Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff}
@@
 Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow.
-Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A\&E without seen by staff, unexpected return visits to A\&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times.without incurring last-minute expenses, such as overtime or supplemental staffing.
+Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A\&E without seen by staff, unexpected return visits to A\&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
 
-An accurate forecasting of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+An accurate forecast of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This will have many advantages for both patients, staff and the quality of healthcare services provided.
 
+Furthermore, because the consequences of imperfact staffing are asymmetric probabilistic forecasts are necessary to make economic decicions.
+This asymmetry arrises because it is preferable to incure a small opportunity cost associated unerutilised staff rather than lower service levels if staff levels are insufficient.
+
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory ,Emergency, Surgical, Inpatient, Home and Residential.
+This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
 
@@
   We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using \ldots.
+\item
+  We provide data and code enabling reproduction and refinement of the proposed approach
 \end{enumerate}
@@
 
-\hypertarget{model}{%
-\section{Proposed model}\label{model}}
-
 \hypertarget{design}{%
@@
 
+\hypertarget{naiveclimatology}{%
+\subsubsection{Naive/Climatology}\label{naiveclimatology}}
+
+Empirical distribution for same hour-of-day and day-of-week as target time.
+
 \hypertarget{multiple-regression}{%
 \subsubsection{Multiple Regression}\label{multiple-regression}}
+
+Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A\&E admittance).
+The classical linear regression model is formulated as:
+\begin{equation}
+  \mathbf{y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
+\label{eq:linearRegression}
+\end{equation}
+where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters and \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
+However, in the context of healthcare and A\&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
+So the linear regression model should be substituted by some other model.
+One of the models that is frequently used in practice is the Poisson regression, which can be summarised as:
+\begin{equation}
+  \mathbf{y}_t \sim \mathrm{Poisson} \left( \exp \left( \mathbf{x}_t' \boldsymbol{\beta} \right) \right).
+\label{eq:PoissonRegression}
+\end{equation}
+The exponent in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive.
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function.
+When it comes to selecting explanatory variables for the model, there is no one correct answer and the decision needs to be done based on each specific case.
+In our experiment, we will only include dummy variables, capturing a variety of calendar events:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Hour of day,
+\item
+  Day of week,
+\item
+  Week of year,
+\item
+  Holidays (such as Christmas, New Year etc),
+\item
+  24 hours lags of holidays.
+\end{enumerate}
+
+The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
+Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g.~people might try not to go to hospital on Christmas eve and thus will go the next day).
+This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \eqref{eq:PoissonRegression} introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time.
+We use \texttt{alm()} function from \texttt{greybox} v0.7.0 package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.
+
+\hypertarget{ets---exponential-smoothing-model}{%
+\subsubsection{ETS - Exponential Smoothing model}\label{ets---exponential-smoothing-model}}
+
+\citet{Hyndman2008b} developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus ``ETS'').
+Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
+This was done using \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denote as ``ETS''.
+This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
 
@@
 \hypertarget{tbats}{%
-\subsection{TBATS}\label{tbats}}
+\subsubsection{TBATS}\label{tbats}}
 
@@
 
+\hypertarget{model}{%
+\section{Model building}\label{model}}
+
+\hypertarget{adam-multiple-seasonal-ietsx}{%
+\subsection{ADAM: multiple seasonal iETSX}\label{adam-multiple-seasonal-ietsx}}
+
+\citet{SvetunkovAdam2021} proposed a framework for dynamic models called ``ADAM'' - Augmented Dynamic Adaptive Model.
+This framework includes ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
+Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
+Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by \citet{Svetunkov2019a} to treat those values.
+This model can be formulated as a set of the following equations:
+\begin{equation}
+    \begin{aligned}
+      & y_t = o_t z_t \\
+        & \log z_t = \log l_{t-1} + \log s_{1,t-24} + \log s_{2,t-168} + \mathbf{x}_t' \boldsymbol{\beta} + \log \left(1 + \epsilon_{t} \right) \\
+        & \log l_{t} = \log l_{t-1} + \log( 1  + \alpha \epsilon_{t}) \\ 
+        & \log s_{1,t} = \log s_{1,t-m} + \log( 1  + \gamma_1 \epsilon_{t}) \\
+        & \log s_{2,t} = \log s_{2,t-m} + \log( 1  + \gamma_2 \epsilon_{t}) \\
+        & o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
+        & a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
+        & l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
+        & \mu_{a,t} = \min(l_{a,t-1}, 1)
+    \end{aligned} ,
+    \label{eq:ADAMModel}
+\end{equation}
+where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
+Finally, \(a_t\) is an unobservable series, underlying the occurrence part of the model and \((1 + \epsilon_{a,t})\) is an unobservable error term for \(a_t\).
+We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
+Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
+If integer-valued quantiles are needed, then rounding up can be done for them.
+This model is implemented in \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM''.
+
+\hypertarget{gamlss}{%
+\subsection{GAMLSS}\label{gamlss}}
+
+If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
+In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
+
+Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution.
+In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
+In a GAMLSS framework \cite{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
+\begin{equation} \label{eq:basicGAM}
+    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
+\end{equation}
+where \(g_j\) is a monotonic `link' function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\).
+If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third element of \(\bm x_t\).
+Each \(f_{j,i}\) is a smooth function, constructed as
+\begin{equation}
+    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
+\end{equation}
+where \(b^{ji}_k\) are spline basis functions of dimension \(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients.
+The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
+See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
+
+As our data are counts, the natural starting point is the Poisson distribution, given by
+\begin{equation}
+  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
+\end{equation}
+where we consider an additive model for \(\lambda_t\) of the form
+\begin{equation}
+  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
+\end{equation}
+The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time \(t\), respectively, and \(C_t\) is the temperature at time \(t\).
+
+Truncated Normal\ldots{}
+\begin{equation}
+  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
+\end{equation}
+with
+\textbackslash begin\{eqnarray\}
+\mu\emph{t \&=\& \sum}\{i=1\}\^{}7 \beta\emph{i \delta(D\_i(t)-i) + \sum}\{j=1\}\^{}7 D\_j(t) f\_j(H(t)) + t f\_\text{Y}(Y(t)) + f\_\text{Temp}(Y(t),C\_t) \quad , \textbackslash{}
+log(\sigma\_t) \&=\& f(H(t)) \quad .
+\textbackslash begin\{end\}
+
+Truncated \(t\) distribution\ldots{}
+
+Negative binomial\ldots{}
+
 \hypertarget{accuracy}{%
@@
 
-Example of writing an equation!
+In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model.
+The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000}, resulting in 727 origins.
+Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation.
+The latter is measured via Root Mean Squared Error (RMSE):
 \begin{equation}
-\bar{X} = \frac{\sum_{i=1}^n X_i}{n} \label{eq:mean}
+  \mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{j=1}^n e_j^2} .
+  \label{eq:RMSE}
 \end{equation}
+When it comes to quantile performance, we used pinball, and all the other cool things described by Jethro.
 
-Also see Equation \eqref{eq:mean}
+Also see Equation \eqref{eq:RMSE}
 

### 2021-10-04 paper/paper.Rmd
added admission viz

--- 2021-09-24 paper/paper.Rmd
@@
 header-includes:  
- \usepackage{adjustbox, float,lscape} #use the 'float' package
+ \usepackage{adjustbox, float,lscape, bm} #use the 'float' package
 output:
@@
 bibliography: mybibfile.bib
-biblio-style: authoryear
+#biblio-style: authoryear
 #csl: elsevier-without-titles.csl
@@
 
-If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
-In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
-
 Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
 In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+In a GAMLSS framework @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
 \begin{equation} \label{eq:basicGAM}
@@
 \end{equation}
-where $g_j$ is a monotonic `link' function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
+where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
 If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
@@
 The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
-
-As our data are counts, the natural starting point is the Poisson distribution, given by
-\begin{equation}
-  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
-\end{equation}
-where we consider an additive model for $\lambda_t$ of the form
-\begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
-\end{equation}
-The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
-
-Truncated Normal...
-\begin{equation}
-  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
-\end{equation}
-with
-\begin{eqnarray}
-  \mu_t &=& \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , \\
-  log(\sigma_t) &=& f(H(t))  \quad .
-\begin{end}
-
-Truncated $t$ distribution...
-
-Negative binomial...
+See @Wood2017 for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
 

### 2021-10-04 paper/paper.tex
added admission viz

@@
 \usepackage{natbib}
-\bibliographystyle{authoryear}
+\bibliographystyle{plainnat}
 \usepackage{longtable,booktabs,array}
@@
 % Pandoc header
-\usepackage{adjustbox, float,lscape}
+\usepackage{adjustbox, float,lscape, bm}
 \usepackage{booktabs}
@@
 
-If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
-In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
-
 Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution.
 In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
-In a GAMLSS framework \cite{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
+In a GAMLSS framework \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
 \begin{equation} \label{eq:basicGAM}
@@
 \end{equation}
-where \(g_j\) is a monotonic `link' function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\).
+where \(g_j\) is a monotonic function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\).
 If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third element of \(\bm x_t\).
@@
 The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
-
-As our data are counts, the natural starting point is the Poisson distribution, given by
-\begin{equation}
-  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
-\end{equation}
-where we consider an additive model for \(\lambda_t\) of the form
-\begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
-\end{equation}
-The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time \(t\), respectively, and \(C_t\) is the temperature at time \(t\).
-
-Truncated Normal\ldots{}
-\begin{equation}
-  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
-\end{equation}
-with
-\textbackslash begin\{eqnarray\}
-\mu\emph{t \&=\& \sum}\{i=1\}\^{}7 \beta\emph{i \delta(D\_i(t)-i) + \sum}\{j=1\}\^{}7 D\_j(t) f\_j(H(t)) + t f\_\text{Y}(Y(t)) + f\_\text{Temp}(Y(t),C\_t) \quad , \textbackslash{}
-log(\sigma\_t) \&=\& f(H(t)) \quad .
-\textbackslash begin\{end\}
-
-Truncated \(t\) distribution\ldots{}
-
-Negative binomial\ldots{}
+See \citet{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
 

### 2021-10-04 r_script/admision_viz.R

@@ -1,0 +1,98 @@
+library(lubridate)
+library(purrr)
+library(tidyverse)
+library(ggthemes)
+library(ggridges)
+library(hrbrthemes)
+library(scales)
+library(ggtext)
+h2_hourly <- readr::read_csv("data/h2_hourly.csv")
+h2_hourly <- h2_hourly %>% filter(arrival_1h >="2015-01-01 00:00:00"&arrival_1h <"2019-01-01 00:00:00")
+
+ride_data <- h2_hourly %>% 
+  mutate(hour=hour(arrival_1h), dow=wday(arrival_1h, label = TRUE)) 
+
+ride_data_ampm <- ride_data %>% 
+  mutate(hour_ampm = case_when(
+    hour == 0 ~ "Minnight",
+    hour == 1 ~ "1 am",
+    hour == 2 ~ "2 am",
+    hour == 3 ~ "3 am",
+    hour == 4 ~ "4 am",
+    hour == 5 ~ "5 am",
+    hour == 6 ~ "6 am",
+    hour == 7 ~ "7 am",
+    hour == 8 ~ "8 am",
+    hour == 9 ~ "9 am",
+    hour == 10 ~ "10 am",
+    hour == 11 ~ "11 am",
+    hour == 12 ~ "Midday",
+    hour == 13 ~ "1 pm",
+    hour == 14 ~ "2 pm",
+    hour == 15 ~ "3 pm",
+    hour == 16 ~ "4 pm",
+    hour == 17 ~ "5 pm",
+    hour == 18 ~ "6 pm",
+    hour == 19 ~ "7 pm",
+    hour == 20 ~ "8 pm",
+    hour == 21 ~ "9 pm",
+    hour == 22 ~ "10 pm",
+    hour == 23 ~ "11 pm"
+  )) %>% 
+  mutate(hour_ampm=factor(hour_ampm))
+  
+
+ride_data_ampm <- ride_data_ampm %>% 
+  mutate(hour_ampm= hour_ampm %>% fct_relevel(l),
+         hour_ampm=fct_rev(hour_ampm))
+ride_data_ampm %>% fct_rev(%>%)
+l <- c("Minnight", "1 am",  "2 am",     "3 am",    "4 am",    "5 am",    "6 am",   
+"7 am",     "8 am",     "9 am",     "10 am",     "11 am",     "Midday",    
+"1 pm",     "2 pm",     "3 pm",     "4 pm",     "5 pm",     "6 pm",    
[…]

### 2021-10-04 paper/paper.Rmd
Add some results...

--- 2021-09-24 paper/paper.Rmd
@@
 
-Still to add details on skill scores and bootstrap methods...
+To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using
+\begin{equation}
+     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}}
+\end{equation}
+where $M$ is the metric's value for the method being considered, $M_\text{ref}$ is the metric's value for a reference method, and $M_\text{perf}$ is the metrics value for   the `perfect' method, which is zero in the case of RMSE and Pinball.
+We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.
+positive or negative skill) are significantly different from zero [@Efron1981Bootstrap].
+Here we use the best performing simple benchmark, Benchmark-2, as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlation of the underlying data.
 
@@
 I guess PDF might be preferred.
-I will also use xtable to make a few tables of results, for now there is a CSV in the results folder.
 I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
+
+## Case study
+
+The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
+Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
+
+(Summary of data, e.g. discussion of observations from training data/model development: long-term trend)
+
+## Forecast evaluation
+
+Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated.
+Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
+Calibration is evaluated visually in \@ref(fig-quantilebias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
+Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
+
+```{r fig-quantilebias}
+## Should reduce the number of models shown here... it's really hard to tell them appart!
+knitr::include_graphics(rep("../results/QuantileBias.png"))
+```
+
+Evaluation metrics from the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.
+
+```{r tab-results}
+load("Results_Table.Rda")
+knitr::kable(Res_sum[order(Res_sum$`Quantile Bias`)],
+  caption = 'Forecast evluation for all methods on test data in decreasing order by Quantile Bias',
+  digits=c(NA,3,2,2)
+)
+```
+
+```{r fig-skill}
+knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+```
 

### 2021-10-04 paper/paper.tex
Add some results...

--- 2021-09-24 paper/paper.tex
@@
 \end{equation}
-When it comes to quantile performance, we used pinball, and all the other cool things described by Jethro.
 
-Also see Equation \eqref{eq:RMSE}
+The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated \citep{Gneiting2007a}.
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.
+
+The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score.
+It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
+Furthermore,
+The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
+\begin{equation}
+    % \text{Pinball} = 
+    \frac{1}{T|\mathcal{A}|} \sum_{\alpha \in \mathcal{A}} \sum_{t=1}^T
+ \left(q_{\alpha,t} - y_{t} \right)
+ \left(\mathbb{1}(y_{t}\leq_{\alpha,t})-\alpha \right)
+ \label{eq:pinball}
+\end{equation}
+where \(\mathcal{A} = {0.05,0.1,...,0.95}\) is the set of quantiles being estimated.
+
+Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations.
+If a forecast is calibrated, then 20\% of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
+This property is necessary for forecast probabilities to be used in quantitative decision-making.
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observer frequency mean(\(\mathbb{1}(y_{t}\le q_{\alpha,t})\)).
+
+Still to add details on skill scores and bootstrap methods\ldots{}
 
@@
 \section{Result and discussion}\label{result}}
+
+(Plots generated with theme\_few - they look nice, good suggestion!
+Not sure how to include them in bookdown, can you help, Bahman?
+The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format?
+I guess PDF might be preferred.
+I will also use xtable to make a few tables of results, for now there is a CSV in the results folder.
+I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
 

### 2021-10-05 paper/paper.Rmd

--- 2021-09-24 paper/paper.Rmd
@@
 library(kableExtra)
+library(ggridges)
+library(hrbrthemes)
+library(ggthemes)
+library(scales)
+library(ggtext)
+library(extrafont)
 ```
@@
 knitr::read_chunk('../r_script/data-viz.R')
+knitr::read_chunk('../r_script/admision_viz.R')
 ```
@@
 
-Demand forecasting is a critical input to inform capacity planning and staffing to meet the needs of patients in each hospital services.
+Demand forecasting is a critical input to inform staffing and scheduling to meet the needs of patients in hospital services.
 An accurate demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
 This is one of the best ways to optimize staffing utilization and related costs.
-Most current practice to optimise personnel scheduling follows the general approach originally presented in [@vile2016time], which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-
+Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
 An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular to depict various courses of action that can result in massive savings in terms of patient lives.
@@
 There exists a large literature on demand forecasting of patient needs, however the main focus is on the daily  arrivals. 
+There are limited studies that use
+different time series forecasting techniques to predict annual arrival,[24] monthly arrival,[14][25]
+daily arrival, [26] and arrival at intervals of 4 to 12 h; [27] however, existing studies have shown
+complications in predicting hourly patient ED arrivals[28] and the application of predicting hourly
+patient arrivals is not well established. 
 Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
@@
 Other papers to review and possibly cite:
-- Daily visits using Prophet 10.1016/j.annemergmed.2019.08.149
-- Daily using internet data 10.1016/j.annemergmed.2014.10.008
-- Review paper: 10.1136/emj.2008.062380
-- Others: 10.1136/emj.2010.103531, 10.1136/emj.2007.051656
-- Impact of weather 10.1136/emj.2010.091058, 10.1136/emj.20.2.204
+- Daily visits using Prophet @park2019144
+- Daily using internet data @ekstrom2015forecasting
+- Review paper: @wargon2009systematic
+- Others: @boyle2012predicting, 
+- Impact of weather @parsons2011modelling, @macgregor2003effect
+
+hourly:
+@cheng2021forecasting
+@rocha2021forecasting
 
@@
 Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
-Then, they presente the predictiona interval accuracy of 50 and 90% intervals for the number of hourly arrivals under the Poisson assumptions.
+Then, they presente the predictiona interval accuracy of 50 and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions.
 They show that the most important predictor is hour of the day and autocorrelation lag 1.
@@
 In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by 33% the 12-hour forecasts generated by historical average).
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).
 
@@
 
-Figure \@ref(fig:24hour)
+Figure \@ref(fig:hourly-plot-ridge)
 
@@
 
-```{r 24hour, fig.cap= "Distrubution of ED attendance for each hour of day", fig.align='center',fig.width=6, fig.height=9, dependson="hourly-plot"}
+```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width=6, fig.height=9, dependson="hourly-plot"}
 ```
 
-## benchmarks {#benchmarks}
+## Benchmarks {#benchmarks}
 
@@
 The classical linear regression model is formulated as:
+
 \begin{equation}
@@
 \end{equation}
+
 where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters and $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
@@
 One of the models that is frequently used in practice is the Poisson regression, which can be summarised as:
+
 \begin{equation}
@@
 \end{equation}
+
 The exponent in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
@@
 
-@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus "ETS").
+@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus *ETS*).
 Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
-This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as "ETS".
+This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as *ETS*.
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
@@
 The model is incorporated using corresponding implementation of the Fable package in R.
-We use the prophet() function in the fable package to generate hourly forecasts [@fable2020].
+We use the `prophet()` function in the fable package to generate hourly forecasts [@fable2020].
 Note that but the input data is assigned with an hourly and daily seasonality.
-
-### FASSTER
-
-FASSTER [@fasster2020] stands for "Forecasting with Additive Switching of Seasonality, Trend and Exogenous Regressors.
-FASSTER is designed to be a flexible model for capturing multiple seasonality.
-It extends commonly used state space models such as TBATS and Double seasonal Holt-Winter by introducing a switching component to the measurement equation.
-It is a structural state space model, with:
-
-- Multiple seasonality capabilities
-- Flexible model specification
-- Multiple sources of error
-- Exogenous regressor support
-- Handling of missing values
 
@@
 
-@SvetunkovAdam2021 proposed a framework for dynamic models called "ADAM" - Augmented Dynamic Adaptive Model.
+@SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model.
 This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
 Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
+Given that the data exhibits zeroes, we use the direct probability of ETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
 This model can be formulated as a set of the following equations:
+
 \begin{equation}
@@
 \end{equation}
+
 where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
@@
 
+Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
+In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
+In a GAMLSS framework @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+
 If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+Generalised additive models for location, scale and shape (*GAMLSS*) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
 This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
@@
 In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
-\begin{equation} \label{eq:basicGAM}
+In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+
+\begin{equation}
     g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
+	(\#eq:basicGAM)
 \end{equation}
-where $g_j$ is a monotonic `link' function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
+
+where $g_j$ is a monotonic `link` function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
 If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
 Each $f_{j,i}$ is a smooth function, constructed as
+
 \begin{equation}
     f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
+    (\#eq:smmothfunction)
 \end{equation}
+
 where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
 The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
+See @Wood2017 for a detailed introduction to *GAM/GAMLSS* models, smoothing splines bases and penalties.
 
 As our data are counts, the natural starting point is the Poisson distribution, given by
+
 \begin{equation}
   F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
+  (\#eq:poissonreg)
 \end{equation}
+
 where we consider an additive model for $\lambda_t$ of the form
+
 \begin{equation}
   \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
+ (\#eq:additivemodel)
 \end{equation}
+
 The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
@@
 Truncated Normal...
+
 \begin{equation}
   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
+ (\#eq:truncatedn)
 \end{equation}
+
 with
-\begin{eqnarray}
-  \mu_t &=& \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , \\
-  log(\sigma_t) &=& f(H(t))  \quad .
-\begin{end}
+
+\begin{align*}
+  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+  log(\sigma_t) & = f(H(t)) .
+\end{align*}
 
@@
 The latter is measured via Root Mean Squared Error (RMSE):
+
 \begin{equation}
@@
 The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
+
 \begin{equation}
-    % \text{Pinball} = 
+    \text{Pinball} = 
     \frac{1}{T|\mathcal{A}|} \sum_{\alpha \in \mathcal{A}} \sum_{t=1}^T
@@
 \end{equation}
+
 where $\mathcal{A} = {0.05,0.1,...,0.95}$ is the set of quantiles being estimated.
 
-Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations.
-If a forecast is calibrated, then 20\% of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
+Calibration, also called `reliability`, is the property that forecast probabilities match the observed frequency of realisations.
+If a forecast is calibrated, then $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
 This property is necessary for forecast probabilities to be used in quantitative decision-making.

### 2021-10-05 paper/paper.tex

@@
 Other papers to review and possibly cite:
-- Daily visits using Prophet 10.1016/j.annemergmed.2019.08.149
-- Daily using internet data 10.1016/j.annemergmed.2014.10.008
-- Review paper: 10.1136/emj.2008.062380
-- Others: 10.1136/emj.2010.103531, 10.1136/emj.2007.051656
-- Impact of weather 10.1136/emj.2010.091058, 10.1136/emj.20.2.204
+- Daily visits using Prophet \citet{park2019144}
+- Daily using internet data \citet{ekstrom2015forecasting}
+- Review paper: \citet{wargon2009systematic}
+- Others: \citet{boyle2012predicting},
+- Impact of weather \citet{parsons2011modelling}, \citet{macgregor2003effect}
 
@@
 Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
-Then, they presente the predictiona interval accuracy of 50 and 90\% intervals for the number of hourly arrivals under the Poisson assumptions.
+Then, they presente the predictiona interval accuracy of 50 and \(90\%\) intervals for the number of hourly arrivals under the Poisson assumptions.
 They show that the most important predictor is hour of the day and autocorrelation lag 1.
@@
 In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by 33\% the 12-hour forecasts generated by historical average).
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by historical average).
 
@@
 
-Figure \ref{fig:24hour}
+Figure \ref{fig:hourly-plot-ridge}
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/24hour-1}
+{\centering \includegraphics{paper_files/figure-latex/hourly-plot-ridge-1}
 
@@
 
-\caption{Distrubution of ED attendance for each hour of day}\label{fig:24hour}
+\caption{Distrubution of admission per hour and day of the week}\label{fig:hourly-plot-ridge}
 \end{figure}
@@
 \hypertarget{benchmarks}{%
-\subsection{benchmarks}\label{benchmarks}}
+\subsection{Benchmarks}\label{benchmarks}}
 
@@
 \end{equation}
+
 where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters and \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
@@
 \end{equation}
+
 The exponent in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive.
@@
 
-\citet{Hyndman2008b} developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus ``ETS'').
+\citet{Hyndman2008b} developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus \emph{ETS}).
 Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
-This was done using \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denote as ``ETS''.
+This was done using \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denote as \emph{ETS}.
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
@@
 The model is incorporated using corresponding implementation of the Fable package in R.
-We use the prophet() function in the fable package to generate hourly forecasts \citep{fable2020}.
+We use the \texttt{prophet()} function in the fable package to generate hourly forecasts \citep{fable2020}.
 Note that but the input data is assigned with an hourly and daily seasonality.
-
-\hypertarget{fasster}{%
-\subsubsection{FASSTER}\label{fasster}}
-
-FASSTER \citep{fasster2020} stands for "Forecasting with Additive Switching of Seasonality, Trend and Exogenous Regressors.
-FASSTER is designed to be a flexible model for capturing multiple seasonality.
-It extends commonly used state space models such as TBATS and Double seasonal Holt-Winter by introducing a switching component to the measurement equation.
-It is a structural state space model, with:
-
-\begin{itemize}
-\tightlist
-\item
-  Multiple seasonality capabilities
-\item
-  Flexible model specification
-\item
-  Multiple sources of error
-\item
-  Exogenous regressor support
-\item
-  Handling of missing values
-\end{itemize}
 
@@
 
-\citet{SvetunkovAdam2021} proposed a framework for dynamic models called ``ADAM'' - Augmented Dynamic Adaptive Model.
+\citet{SvetunkovAdam2021} proposed a framework for dynamic models called \textbf{ADAM} - Augmented Dynamic Adaptive Model.
 This framework includes ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
 Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-Given that the data exhibits zeroes, we use the direct probability oETS(M,N,N) model developed by \citet{Svetunkov2019a} to treat those values.
+Given that the data exhibits zeroes, we use the direct probability of ETS(M,N,N) model developed by \citet{Svetunkov2019a} to treat those values.
 This model can be formulated as a set of the following equations:
+
 \begin{equation}
@@
 \end{equation}
+
 where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
@@
 
+Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution.
+In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
+In a GAMLSS framework \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
+
 If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+Generalised additive models for location, scale and shape (\emph{GAMLSS}) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
 This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
@@
 In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
-In a GAMLSS framework \cite{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
-\begin{equation} \label{eq:basicGAM}
+In a GAMLSS framework \citep{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
+
+\begin{equation}
     g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
+    \label{eq:basicGAM}
 \end{equation}
-where \(g_j\) is a monotonic `link' function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\).
+
+where \(g_j\) is a monotonic \texttt{link} function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\).
 If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third element of \(\bm x_t\).
 Each \(f_{j,i}\) is a smooth function, constructed as
+
 \begin{equation}
     f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
+    \label{eq:smmothfunction}
 \end{equation}
+
 where \(b^{ji}_k\) are spline basis functions of dimension \(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients.
 The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
+See \citet{Wood2017} for a detailed introduction to \emph{GAM/GAMLSS} models, smoothing splines bases and penalties.
 
 As our data are counts, the natural starting point is the Poisson distribution, given by
+
 \begin{equation}
   F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
+  \label{eq:poissonreg}
 \end{equation}
+
 where we consider an additive model for \(\lambda_t\) of the form
+
 \begin{equation}
   \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
+ \label{eq:additivemodel}
 \end{equation}
+
 The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time \(t\), respectively, and \(C_t\) is the temperature at time \(t\).
@@
 Truncated Normal\ldots{}
+
 \begin{equation}
   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
+ \label{eq:truncatedn}
 \end{equation}
+
 with
-\textbackslash begin\{eqnarray\}
-\mu\emph{t \&=\& \sum}\{i=1\}\^{}7 \beta\emph{i \delta(D\_i(t)-i) + \sum}\{j=1\}\^{}7 D\_j(t) f\_j(H(t)) + t f\_\text{Y}(Y(t)) + f\_\text{Temp}(Y(t),C\_t) \quad , \textbackslash{}
-log(\sigma\_t) \&=\& f(H(t)) \quad .
-\textbackslash begin\{end\}
+
+\begin{align*}
+  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+  log(\sigma_t) & = f(H(t)) .
+\end{align*}
 
@@
 \end{equation}
+
 where \(\mathcal{A} = {0.05,0.1,...,0.95}\) is the set of quantiles being estimated.
 
-Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations.
-If a forecast is calibrated, then 20\% of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
+Calibration, also called \texttt{reliability}, is the property that forecast probabilities match the observed frequency of realisations.
+If a forecast is calibrated, then \(20\%\) of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
 This property is necessary for forecast probabilities to be used in quantitative decision-making.

### 2021-10-07 paper/paper.Rmd
Plot code in rmd

--- 2021-10-04 paper/paper.Rmd
@@
     footnote: 1
-  - name: Author2
-    email: email2@example.com
-    affiliation: University2
+  - name: Jethro Browell
+    email: jethro.browell@glasgow.ac.uk
+    affiliation: School of Mathematics and Statistics, University of Glasgow, UK
     footnote: 2
@@
 library(kableExtra)
+library(ggplot2)
+library(ggthemes)
+library(data.table)
+library(RColorBrewer)
 ```
@@
 If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
 This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
@@
 ## Should reduce the number of models shown here... it's really hard to tell them appart!
-knitr::include_graphics(rep("../results/QuantileBias.png"))
+# knitr::include_graphics(rep("../results/QuantileBias.png"))
+
+load("../results/all_results_paper_2021-10-07.Rda")
+
+REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
+                      Empirical=seq(0,1,by=0.05),
+                      `Quantile Bias`= 0,
+                      Method="Nominal")
+
+REL[,`Quantile Bias`:=Empirical-Nominal]
+
+ggplot(data=REL[Horizon=="All" & Issue == "All",],aes(x=Nominal,y=`Quantile Bias`,group=Method,color=Method)) +
+  geom_line(data=REL_nom,aes(x=Nominal,y=`Quantile Bias`), color="black",size=1.1,show.legend = F) +
+  geom_line() + geom_point() +
+  xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
+  theme_few()
+
 ```
@@
 ```{r fig-skill}
-knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+# knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+
+load("../results/all_results_paper_2021-10-07.Rda")
+
+plotdata <- melt(bootdata[,100*(get(REF)-.SD)/get(REF),.SDcols=NAMES],
+                 measure.vars =  1:length(NAMES),variable.name = "Method",
+                 value.name = "Skill Score")
+
+## Merge Reliability for colouring...
+plotdata <- merge(plotdata,REL[kfold=="Test" & Horizon=="All" & Issue == "All",.(Qbias=mean(abs(Nominal-Empirical))),by="Method"],
+                  by="Method",all.x = T)
+
+ggplot(plotdata[`Skill Score`>-2,], aes(x=reorder(Method, -`Skill Score`), y=`Skill Score`, fill=Qbias)) + 
+  ylab("Pinball Skill Score [%]") +
+  geom_boxplot() + theme_few() +
+  # ggtitle("Pinball Skill Score Relative to Benchmark 2") +
+  theme(axis.text.x = element_text(angle = -80)) + #scale_fill_manual(values=cbPalette) +
+  # scale_x_discrete(labels= paste0(substring(NAMES, 1,4),".")) +
+  geom_hline(yintercept=0, linetype="dashed",size=0.5)+
+  scale_fill_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlGn")),
+                       limits=c(0,0.11))+
+  labs(fill = "Quantile Bias",x="Method")
+
 ```

### 2021-10-07 gam-mgcv/study_v1_timing.R

@@ -705,7 +705,16 @@ time_temp <- Sys.time()
 ## %%%% Time for Ivan %%%%
 
 
-h2_gbm_mqr <- MQR_gbm(data = h2,
+# h2_gbm_mqr <- MQR_gbm(data = h2,
+#                       formula = n_attendance ~ clock_hour + dow3 + school_holiday + T2T + doy,
+#                       quantiles = seq(0.05,0.95,by=0.05),
+#                       gbm_params = list(n.tree=500,
+#                                         interaction.depth=2,
+#                                         shrinkage=0.1,
+#                                         cv.folds=3),
+#                       cores = detectCores()-1)
+
+h2_gbm_mqr <- MQR_gbm(data = h2[kfold!="Test"],CVfolds = NULL,
                       formula = n_attendance ~ clock_hour + dow3 + school_holiday + T2T + doy,
                       quantiles = seq(0.05,0.95,by=0.05),
                       gbm_params = list(n.tree=500,
@@ -714,9 +723,7 @@ h2_gbm_mqr <- MQR_gbm(data = h2,
                                         cv.folds=3),
                       cores = detectCores()-1)
 
-
-
-expectation <- mean_from_qs(mqr = cbind(h2[,.(issueTime,targetTime_UK)],h2_gbm_mqr))
+expectation <- mean_from_qs(mqr = cbind(h2[kfold!="Test",.(issueTime,targetTime_UK)],h2_gbm_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
@@ -733,7 +740,32 @@ for(Ver in 1:2){
   ## %%%% Time for Ivan %%%%
   
   
-  h2_mboost_mqr <- qreg_mboost(data = h2[!is.na(T2T),],
+  # h2_mboost_mqr <- qreg_mboost(data = h2[!is.na(T2T),],
+  #                              formula = 
+  #                                if(Ver==1){
+  #                                  ## v1 above:
+  #                                  n_attendance ~ bols(dow) + 
+  #                                    bbs(clock_hour,knots=24,df=20) +
+  #                                    bbs(clock_hour,knots=24, by=dow,df=30,center = T) +
+  #                                    bbs(doy,knots=6,by=t,df=6) +
+  #                                    bbs(clock_hour,T2T,knots = 6,df=6)
+  #                                }else if(Ver==2){
+  #                                  ## v2 here:
[…]

### 2021-10-07 paper/paper.Rmd

--- 2021-10-05 paper/paper.Rmd
@@
 library(kableExtra)
+library(ggplot2)
+library(data.table)
+library(RColorBrewer)
 library(ggridges)
@@
 
+If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
+In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
 Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.

### 2021-10-07 paper/paper.Rmd
Minor paper edit
intro added

--- 2021-10-05 paper/paper.Rmd
@@
 ---
-title: Short-term hourly forecasting for care services
+title:  Forecasting hourly hospital admission
 author:
@@
 
-Demand forecasting is a critical input to inform staffing and scheduling to meet the needs of patients in hospital services.
-An accurate demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients.
+An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
 This is one of the best ways to optimize staffing utilization and related costs.
 Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular to depict various courses of action that can result in massive savings in terms of patient lives.
-Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow.
-Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A&E without seen by staff, unexpected return visits to A&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
+An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives [@rostami2020anticipating].
 
-An accurate forecast of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
-This will have many advantages for both patients, staff and the quality of healthcare services provided.
+Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
+Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
 
-Furthermore, because the consequences of imperfact staffing are asymmetric probabilistic forecasts are necessary to make economic decicions.
-This asymmetry arrises because it is preferable to incure a small opportunity cost associated unerutilised staff rather than lower service levels if staff levels are insufficient.
-
+There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting].
+daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] admission.
+In this paper, we only focus on forecasting hourly hospital admission.
+In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+This will have many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
 This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
-
-The combination of an hourly forecast demand, incidents being attended, resource availability and delays at hospitals, provide information on the state of the unscheduled care system across the service.
+The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
-
 Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
 Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
-That may also require to adjust the schediling.
+That may also require to adjust the scheduling.
 
-what literature says about hourly forecasting in healthcare?
+There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[].
+Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
+They may also express nonstationarity and their profile may change over time.
+Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
+This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features.
+All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission.
+Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
+Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
+The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
+In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+Aditionally, datasets used in some of these studies are relatively small, short length (e.g.
+time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
+Moreover, all previous publications  covered in this paper fails reproducibility principles (Boylan, 2016).
 
-There exists a large literature on demand forecasting of patient needs, however the main focus is on the daily  arrivals. 
-There are limited studies that use
-different time series forecasting techniques to predict annual arrival,[24] monthly arrival,[14][25]
-daily arrival, [26] and arrival at intervals of 4 to 12 h; [27] however, existing studies have shown
-complications in predicting hourly patient ED arrivals[28] and the application of predicting hourly
-patient arrivals is not well established. 
-Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
-In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services
+In this paper, we aim at filling these gaps and our contributions to the literature are summarised as following:
 
-limitations in current literature?
-- Typically focus on daily data, hiding information about peak arrival rates. Hourly data has additional challenge of multiple seasonality.
-- Deterministic, are there any examples of probabilistic?
+1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv)
+
+2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;
+
+3.
+We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
+Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+
+4.
+We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
+The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.
+
+Our contributions are as following:
+
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with ...
+
+# Research background: hourly hospital admission forecasting {#lit}
 
@@
 
-Our contributions are as following:
-
-1.
-We develop a novel methodology to forecast hourly time series for healthcare services combining a distribution fit to each hour of day separately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, ...
-iv) special events such as public holidays, festive days and rugby and v) weather data.
-2. We provide probabilistic forecasts ...
-3. We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using ....
-4. We provide data and code enabling reproduction and refinement of the proposed approach
-
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with ...
-
-# Research background: Hourly forecasting in care services{#lit}
-
 Table \@ref(tab:summarylit) summarise studies in hourly forecasting in emergency and urgent care.
@@
 li <- readxl::read_xlsx("../table/table.xlsx")
-knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position")) %>% column_spec(5,"15em")
+knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
+  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") 
 ```
@@
 
-The literature review reveals some limitations in forecasting for urgent and emergency care which will be summarised as follows:
-
-- first, 
-- second,
-- third,
+There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
+These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.
+there is no benchmark method nor is forecast accuracy reported.
 
@@
 
+Figure \@ref(fig:hourly-plot-ridge)
+
+```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width= 7,fig.height=6}
+```
+
 Figure \@ref(fig:hourly-plot)
 
-Figure \@ref(fig:hourly-plot-ridge)
+```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "80%"}
 
-```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',fig.height=9, fig.width= 6}
-
-```
-
-```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width=6, fig.height=9, dependson="hourly-plot"}
 ```
@@
 To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using
+
 \begin{equation}
-     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}}
+     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} (\#eq:skillscore)
 \end{equation}
+
 where $M$ is the metric's value for the method being considered, $M_\text{ref}$ is the metric's value for a reference method, and $M_\text{perf}$ is the metrics value for   the `perfect' method, which is zero in the case of RMSE and Pinball.
@@
 
-```{r fig-quantilebias}
-## Should reduce the number of models shown here... it's really hard to tell them appart!
+```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
 knitr::include_graphics(rep("../results/QuantileBias.png"))
+```
+
+```{r tab-results, out.width= "50%"}
+results_table <- read_rds("results_table.rds")
+knitr::kable(results_table, booktabs = T, 
+             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
 ```
@@
 
-```{r tab-results}
-load("Results_Table.Rda")
-knitr::kable(Res_sum[order(Res_sum$`Quantile Bias`)],
-  caption = 'Forecast evluation for all methods on test data in decreasing order by Quantile Bias',
-  digits=c(NA,3,2,2)
-)
+```{r tab-results-qbias, out.width= "50%"}
+results_table_qb <- read_rds("results_table.rds") %>% select(`Quantile Bias`)
+
+knitr::kable(results_table_qb, booktabs = T, 
+             linesep = "",caption = "Forecast evluation for all methods on test data in decreasing order by Quantile Bias") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
 ```
 
-```{r fig-skill}
+```{r fig-skill, fig.cap= "Skill score ...", fig.align='center',out.width= "90%"}
 knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))

### 2021-10-07 paper/paper.tex
intro added

@@
             pdfauthor={},
-            pdftitle={Short-term hourly forecasting for care services},
+            pdftitle={Forecasting hourly hospital admission},
             colorlinks=false,
@@
 
-  \title{Short-term hourly forecasting for care services}
+  \title{Forecasting hourly hospital admission}
     \author[University1]{Author1\corref{1}}
@@
 
-Demand forecasting is a critical input to inform capacity planning and staffing to meet the needs of patients in each hospital services.
-An accurate demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients.
+An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
 This is one of the best ways to optimize staffing utilization and related costs.
-Most current practice to optimise personnel scheduling follows the general approach originally presented in \citep{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+Most current practice to optimise personnel scheduling follows the general approach originally presented in \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+An accurate demand forecasting is crucial in Accident \& Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives \citep{rostami2020anticipating}.
 
-An accurate demand forecasting is crucial in Accident \& Emergency and Ambulance Services in particular to depict various courses of action that can result in massive savings in terms of patient lives.
-Unability to match the staff with the demand results in patient overcrowding the in urgent and emergency care system which is a serious problem that causes challenging situations on patient flow.
-Also, it is related with increasing length of stay, low patient satisfaction, number of patients left the A\&E without seen by staff, unexpected return visits to A\&Es , increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
+Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow \citep{derlet2002overcrowding}.
+Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
 
-An accurate forecast of the patient demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
-This will have many advantages for both patients, staff and the quality of healthcare services provided.
-
-Furthermore, because the consequences of imperfact staffing are asymmetric probabilistic forecasts are necessary to make economic decicions.
-This asymmetry arrises because it is preferable to incure a small opportunity cost associated unerutilised staff rather than lower service levels if staff levels are insufficient.
-
+There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual\citep{tandberg1994time} monthly,\citep{chen2011long, mai2015predicting}.
+daily\citep{rostami2020anticipating, park2019144} and subdaily \citep{schweigler2009forecasting, cheng2021forecasting} admission.
+In this paper, we only focus on forecasting hourly hospital admission.
+In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+This will have many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
 This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
-
-The combination of an hourly forecast demand, incidents being attended, resource availability and delays at hospitals, provide information on the state of the unscheduled care system across the service.
+The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
-
 Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during a day or a week.
 Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
-That may also require to adjust the schediling.
+That may also require to adjust the scheduling.
 
-what literature says about hourly forecasting in healthcare?
+There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+These studies use multiple approaches including Exponential Smoothing {[}REF{]}, ARIMA families {[}{]}, ARCH {[}REF{]}, Vector Autoregressive {[}REF{]}, TBATS {[}{]} and ANN{[}{]}.
+Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
+They may also express nonstationarity and their profile may change over time.
+Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
+This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features.
+All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission.
+Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
+Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
+The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
+In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+Aditionally, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
+Moreover, all previous publications covered in this paper fails reproducibility principles (Boylan, 2016).
 
-There exists a large literature on demand forecasting of patient needs, however the main focus is on the daily arrivals.
-Hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the data.
-In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services
+In this paper, we aim at filling these gaps and our contributions to the literature are summarised as following:
 
-limitations in current literature?
-- Typically focus on daily data, hiding information about peak arrival rates. Hourly data has additional challenge of multiple seasonality.
-- Deterministic, are there any examples of probabilistic?
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv)
+\item
+  We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using \ldots;
+\item
+  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+\item
+  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
+The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.
+\end{enumerate}
+
+Our contributions are as following:
+
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with \ldots{}
+
+\hypertarget{lit}{%
+\section{Research background: hourly hospital admission forecasting}\label{lit}}
 
@@
 
-Our contributions are as following:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  We develop a novel methodology to forecast hourly time series for healthcare services combining a distribution fit to each hour of day separately and then adjust it using variables accounting for i) trend, ii) autocorrelation iii) temporal factors such as day of week, \ldots{} iv) special events such as public holidays, festive days and rugby and v) weather data.
-\item
-  We provide probabilistic forecasts \ldots{}
-\item
-  We benchmark the forecasting performance of the proposed methodology against regression, prophet and FASSTER using \ldots.
-\item
-  We provide data and code enabling reproduction and refinement of the proposed approach
-\end{enumerate}
-
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with \ldots{}
-
-\hypertarget{lit}{%
-\section{Research background: Hourly forecasting in care services}\label{lit}}
+hourly:
+\citet{cheng2021forecasting}
+\citet{rocha2021forecasting}
 
@@
 \centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrll>{\raggedright\arraybackslash}p{15em}l}
+\fontsize{11}{13}\selectfont
+\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{4em}ll}
 \toprule
-Author & Year & Forecast variable & Forecast Horizon & Method & Measure\\
+Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple\\
 \midrule
+NA & NA & NA & NA & NA & NA & NA & NA & seasonality\\
 McCarthy 
-2008 & 2008 & ED arrivals & 24h & Poisson log-linear regression model, including temporal factors, patient characteristics and climatic factors & 95\% CI\\
-Asheim et al. & 2019 & ED arrivals & 3h & Poisson regression with weekly and yearly cyclic 
-effects. & MAPE\\
-Cote et al. & 2013 & ED arrivals & 24h & Fourier regression & R\textasciicircum{}2,  Standard Error\\
-Kim et al. & 2014 & Hospital demand & 4h, 24h, 7days, 
-30days & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE\\
-Schweigler et al. & 2009 & ED bed occupancy & 4h and 12h & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE\\
-Channouf et al. & 2007 & Ambulance demand & 12h,14h,17h,23h,24h,1h,3h,6h, 13h & Regression & RMSE\\
-Hertzum & 2017 & ED arrivals
-ED occupancy & 1,2,4,8,24 hours & linear regression; SARIMA; Naïve & MAE, MAPE, 
-MASE\\
-Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME\\
-Steins et al. & 2019 & Ambulance call & 24h & ZIP and ZINB regression;
- moving average with seasonality weights & ME, MAE, RMSE\\
-Jones et al. & 2009 & ED census & 24h & VAR;  Holt winters & MAE\\
-Morzuch and Allen & 2006 & ED arrivals & 168h & Regression; ARIMA; Exponential smoothing & RMSE\\
+2008 & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
+Asheim et al. & 2019 & ED arrivals & 3h & 5 & Poisson regression & MAPE & No & No\\
+Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
+Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
+Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
+Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
+Hertzum & 2017 & ED arrivals & 1,2,4,8,24 hours & 3 & linear regression; SARIMA; Naïve & MAE, MAPE, 
+MASE & No & No\\
+Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
+Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
+Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
 Chase et al. & 2012 & ED CUR & 30m
-1h, 2h, 4h, 8h, 12h & Binary regression & NA\\
-Taylor & 2008 & centre volume & 30 minutes 
-to 2 weeks & ARMA, Exponential smoothing; Dynamic harmonic regression; Seasonal random walk; Seasonal mean;
-Random walk; Mean; Median; Simple exponential smoothing & MAE
-RMSE\\
-Gijo and Balakrishna & 2016 & call volume & 168h & SARIMA & MMSE\\
+1h, 2h, 4h, 8h, 12h & 1 & Binary regression & NA & No & No\\
 \bottomrule
-\end{tabular}}
+\end{tabular}
 \end{table}
@@
 
-The literature review reveals some limitations in forecasting for urgent and emergency care which will be summarised as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  first,
-\item
-  second,
-\item
-  third,
-\end{itemize}
+There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
+These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method nor is forecast accuracy reported.
 
@@
 
-Figure \ref{fig:hourly-plot}
-
 Figure \ref{fig:hourly-plot-ridge}
-
-\begin{figure}[H]
-
-{\centering \includegraphics{paper_files/figure-latex/hourly-plot-1}
-
-}
-
-\caption{Seasonal plot of ED attendance}\label{fig:hourly-plot}
-\end{figure}
 
@@
 \caption{Distrubution of admission per hour and day of the week}\label{fig:hourly-plot-ridge}
+\end{figure}
+
+Figure \ref{fig:hourly-plot}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/hourly-plot-1}
+
+}
+
+\caption{Seasonal plot of ED attendance}\label{fig:hourly-plot}
 \end{figure}
@@
 
-Still to add details on skill scores and bootstrap methods\ldots{}
+To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using
+
+\begin{equation}
+     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} \label{eq:skillscore}
+\end{equation}
+
+where \(M\) is the metric's value for the method being considered, \(M_\text{ref}\) is the metric's value for a reference method, and \(M_\text{perf}\) is the metrics value for the `perfect' method, which is zero in the case of RMSE and Pinball.
+We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}.
+Here we use the best performing simple benchmark, Benchmark-2, as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlation of the underlying data.
 
@@
 I guess PDF might be preferred.
-I will also use xtable to make a few tables of results, for now there is a CSV in the results folder.
 I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
+
+\hypertarget{case-study}{%
+\subsection{Case study}\label{case-study}}
+
+The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
+Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
+
+(Summary of data, e.g.~discussion of observations from training data/model development: long-term trend)
+
+\hypertarget{forecast-evaluation}{%
+\subsection{Forecast evaluation}\label{forecast-evaluation}}
+
+Probabilistic forecasts are evaluated following the principle of \emph{sharpness subject to calibration}, meaning that the sharper forecast is prefered provided that it is calibrated.
+Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
+Calibration is evaluated visually in \ref{fig-quantilebias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
+Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{../results/QuantileBias}
+
+}
+
+\caption{Quantile ...}\label{fig:fig-quantilebias}
+\end{figure}
+
+\begin{table}[!h]
+
+\caption{\label{tab:tab-results}Summary of studies in hourly emergency care forecasting}
+\centering
+\fontsize{11}{13}\selectfont
+\begin{tabular}[t]{lrrr}
+\toprule
+Method & Quantile Bias & Pinball & RMSE\\
+\midrule
+Benchmark-1 & 0.1047874 & 1.254491 & NA\\
+Benchmark-2 & 0.0557392 & 1.217429 & NA\\
+ETS(XXX) & 0.0194389 & 1.434862 & 0.0121247\\
+GBM & 0.0600650 & 1.263135 & NA\\
+NBI-2-I & 0.0540740 & 1.206238 & NA\\
+NBI-2-log & 0.0375705 & 1.188041 & NA\\
+NOtr-1 & 0.0098936 & 1.222584 & NA\\
+NOtr-2 & 0.0118522 & 1.208556 & NA\\
+Poisson-1 & 0.0372137 & 1.204936 & NA\\
+Poisson-2 & 0.0373884 & 1.188107 & NA\\
+Poisson-2-I & 0.0523937 & 1.222114 & NA\\
+RegressionPoisson & 0.0929416 & 1.293524 & 0.8490258\\
+Ttr-2 & 0.0140252 & 1.210105 & NA\\
+faster & 0.1862190 & 35.051012 & NA\\
+iETSCeiling & 0.0104673 & 1.419490 & NA\\
+iETSXSeasonal & 0.0104673 & 1.417260 & 0.0896228\\
+qreg-1 & 0.0643451 & 1.340557 & NA\\
+tbats & 0.0855702 & 1.536080 & 0.4859770\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Evaluation metrics from the test period are presented in \ref{tab-results} which are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.
+
+\begin{table}[!h]
+
+\caption{\label{tab:tab-results-qbias}Forecast evluation for all methods on test data in decreasing order by Quantile Bias}
+\centering
+\fontsize{11}{13}\selectfont
+\begin{tabular}[t]{r}
+\toprule
+Quantile Bias\\
+\midrule
+0.1047874\\
+0.0557392\\
+0.0194389\\
+0.0600650\\
+0.0540740\\
+0.0375705\\
+0.0098936\\
+0.0118522\\
+0.0372137\\
+0.0373884\\
+0.0523937\\
+0.0929416\\
+0.0140252\\
+0.1862190\\
+0.0104673\\
+0.0104673\\
+0.0643451\\
+0.0855702\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{../results/Skill_rel2bench_reduced}
+
+}
+
+\caption{Skill score ...}\label{fig:fig-skill}
+\end{figure}
 

### 2021-10-07 paper/paper.Rmd

@@
 knitr::read_chunk('../r_script/admision_viz.R')
+knitr::read_chunk('../r_script/result_viz.R')
 ```
@@
 ```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
-knitr::include_graphics(rep("../results/QuantileBias.png"))
+#knitr::include_graphics(rep("../results/QuantileBias.png"))
+## Should reduce the number of models shown here... it's really hard to tell them appart!
+
+REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
+                      Empirical=seq(0,1,by=0.05),
+                      `Quantile Bias`= 0,
+                      Method="Nominal")
+
+REL[,`Quantile Bias`:=Empirical-Nominal]
+
+ggplot(data=REL[Horizon=="All" & Issue == "All",],aes(x=Nominal,y=`Quantile Bias`,group=Method,color=Method)) +
+  geom_line(data=REL_nom,aes(x=Nominal,y=`Quantile Bias`), color="black",size=1.1,show.legend = F) +
+  geom_line() + geom_point() +
+  xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
+  theme_few() 
 ```
@@
 knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+
+# knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+
+load("../results/all_results_paper_2021-10-07.Rda")
+
+plotdata <- melt(bootdata[,100*(get(REF)-.SD)/get(REF),.SDcols=NAMES],
+                 measure.vars =  1:length(NAMES),variable.name = "Method",
+                 value.name = "Skill Score")
+
+## Merge Reliability for colouring...
+plotdata <- merge(plotdata,REL[kfold=="Test" & Horizon=="All" & Issue == "All",.(Qbias=mean(abs(Nominal-Empirical))),by="Method"],
+                  by="Method",all.x = T)
+
+ggplot(plotdata[`Skill Score`>-2,], aes(x=reorder(Method, -`Skill Score`), y=`Skill Score`, fill=Qbias)) + 
+  ylab("Pinball Skill Score [%]") +
+  geom_boxplot() + theme_few() +
+  # ggtitle("Pinball Skill Score Relative to Benchmark 2") +
+  theme(axis.text.x = element_text(angle = -80)) + #scale_fill_manual(values=cbPalette) +
+  # scale_x_discrete(labels= paste0(substring(NAMES, 1,4),".")) +
+  geom_hline(yintercept=0, linetype="dashed",size=0.5)+
+  scale_fill_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlGn")),
+                       limits=c(0,0.11))+
+  labs(fill = "Quantile Bias",x="Method")
+
 ```

### 2021-10-12 results/Evaluatoin Script.R
Update evaluation script with new JB results

@@ -118,7 +118,9 @@ big_eval_function <- function(forecast_DT,h2_actuals,method_name){
 
 ## Jethro's Results ####
 
-load("JethroResults_2021-05-06.Rda")
+load("JethroResults_pt1_2021-10-08.Rda")
+load("JethroResults_pt2_2021-10-08.Rda")
+JB_results <- c(temp1,temp2); rm(temp1,temp2)
 
 for(n in names(JB_results)){
   
@@ -143,7 +145,7 @@ rm(tbats)
 
 
 faster <- data.table(readRDS("fasster_bahman.rds"))
-setnames(prophet,old = c("point_forecast"),c("expectation"))
+setnames(faster,old = c("point_forecast"),c("expectation"))
 faster[,issueTime := issueTime+3600]
 big_eval_function(forecast_DT = faster,h2_actuals = h2,method_name = "faster")
 rm(faster)

### 2021-10-12 results/Evaluatoin Script.R
Rerun evaluation with new JB results

@@ -152,7 +152,8 @@ rm(faster)
 
 
 prophet <- data.table(readRDS("forecast_prophet.rds"))
-setnames(prophet,old = c("origin","target","point_forecast"),c("issueTime","targetTime_UK","expectation"))
+# setnames(prophet,old = c("origin","target","point_forecast"),c("issueTime","targetTime_UK","expectation"))
+setnames(prophet,old = c("origin","target",1:19/20),c("issueTime","targetTime_UK",paste0("q",1:19*5)))
 prophet[,issueTime := issueTime+3600]
 big_eval_function(forecast_DT = prophet,h2_actuals = h2,method_name = "prophet")
 rm(prophet)
@@ -179,7 +180,7 @@ rm(quantileValuesIvan)
 
 save(PB,PB_ts,REL,RMSE,file=paste0("all_results",Sys.Date(),".Rda"))
 
-load("all_results2021-09-24.Rda")
+load("all_results2021-10-12.Rda")
 
 change_method_name <- function(dt,old_new){
   for(i in 1:nrow(old_new)){
@@ -225,7 +226,7 @@ OLD_NEW <- data.table(old=c("Benchmark_1","Benchmark_2",
 change_method_name(REL,OLD_NEW)
 change_method_name(PB,OLD_NEW)
 change_method_name(RMSE,OLD_NEW)
-setnames(PB_ts,old = OLD_NEW$old,new=OLD_NEW$new)
+setnames(PB_ts,old = OLD_NEW$old,new=OLD_NEW$new,skip_absent = T)
 
 ## Bootstraped skill scores
 ##
@@ -333,37 +334,6 @@ plotdata <- melt(bootdata[,100*(get(REF)-.SD)/get(REF),.SDcols=NAMES],
 plotdata <- merge(plotdata,REL[kfold=="Test" & Horizon=="All" & Issue == "All",.(Qbias=mean(abs(Nominal-Empirical))),by="Method"],
                   by="Method",all.x = T)
 
-require(RColorBrewer)
-ggplot(plotdata[Method!="faster"], aes(x=reorder(Method, -`Skill Score`), y=`Skill Score`, fill=Qbias)) + 
-  ylab("Pinball Skill Score [%]") +
-  geom_boxplot() + theme_few() +
-  # ggtitle("Pinball Skill Score Relative to Benchmark 2") +
[…]

### 2021-10-12 paper/paper.Rmd
Adding GAMLSS text...

@@
 
-Truncated Normal...
+However, calibration of resulting forecasts on the training data is poor, suggesting that the shape of the Poisson distribution is unsuitable for the present application.
+In particular, we observed that forecast uncertainty appeared to vary depending on the time of day and possibly other explanatory variables.
+Therefore, we also consider more flexible two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
+The resulting density forecasts are given by
 
@@
 
-with
+with additive models
 
@@
   \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-  log(\sigma_t) & = f(H(t)) .
+  log(\sigma_t) & = f(H(t))
 \end{align*}
+for the mean and variance parameters.
 

### 2021-10-19 paper/paper.Rmd
Add second additive model structure

@@
 As our data are counts, the natural starting point is the Poisson distribution, given by
-
 \begin{equation}
@@
 \end{equation}
-
-where we consider an additive model for $\lambda_t$ of the form
-
+where we initially consider an additive model for $\lambda_t$ of the form
 \begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
+  \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
  (\#eq:additivemodel)
 \end{equation}
-
-The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
+where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$.
+Analysis of this model fit on the training data revealed systematic biases which were identified as relating to public and school holidays.
+We therefore expand the model to include these effects as follows,
+\begin{equation}
+  \log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
+ (\#eq:additivemodel_v2)
+\end{equation}
+where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays.
 
@@
 The resulting density forecasts are given by
-
 \begin{equation}
@@
 \end{equation}
-
 with additive models
-
 \begin{align*}
-  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-  log(\sigma_t) & = f(H(t))
+  \mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+  log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
 \end{align*}

### 2021-10-19 paper/paper.Rmd
Add to results discussion

@@
 
+The Pinball Score evaluates overall performance, penalising both poor calibration and lack of sharpness, and ranks the various models quite differently.
+NBI-2 and Poisson-2 achive the lowest Pinball Loss, but as we have seen, are poorly calibrated so not suitable for use in decision making.
+Of the well calibrated methods, NOtr-2 and Ttr-2 have the lowest Pinball Score, and provide a significant improvement over Benchmark 2 by this metric.
+However, we note that while the improvement in terms of Pinball Score appears small, these methods have substantially improved forecast calibration compared to the benchmarks, which are both poorly calibrated.
+
 ```{r fig-skill, fig.cap= "Skill score ...", fig.align='center',out.width= "90%"}

### 2021-10-19 r_script/result_viz.R
Amended list of selected models

@@ -2,9 +2,9 @@
 my_colorblind <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
                    "#F0E442", "#0072B2", "#D55E00", "#CC79A7","black")
 PB <- read_rds("results/PB.rds")
-selected_method <- c("iETSXSeasonal","GBM",
-                     "NOtr-2","RegressionPoisson",
-                     "tbats","ETS(XXX)","Benchmark-1",
+selected_method <- c("ADAM-iETSX","GBM",
+                     "NOtr-2","Regression-Poisson",
+                     "tbats","ETS","Benchmark-1",
                      "Poisson-1","NBI-2-I")
 PB_selected <- PB %>% as_tibble() %>% 
   mutate(Method=factor(Method)) %>% 

### 2021-10-21 paper/paper.Rmd
Update GAMLSS discussion

@@
 In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
+Here we use this framework to extend the Poisson regression introduced above, and explore alternative distirbutional assumptions.
+
 Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
@@
 \end{equation}
-where we initially consider an additive model for $\lambda_t$ of the form
-\begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
- (\#eq:additivemodel)
-\end{equation}
-where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$.
+<!-- where we initially consider an additive model for $\lambda_t$ of the form -->
+<!-- \begin{equation} -->
+<!--   \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
+<!--  (\#eq:additivemodel) -->
+<!-- \end{equation} -->
+<!-- where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$.
 Analysis of this model fit on the training data revealed systematic biases which were identified as relating to public and school holidays.
-We therefore expand the model to include these effects as follows,
+We therefore expand the model to include these effects as follows, -->
+<!-- \begin{equation} -->
+<!--   \log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
+<!--  (\#eq:additivemodel_v2) -->
+<!-- \end{equation} -->
+<!-- where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays. -->
+where we consider an additive model for $\lambda_t$ of the form
 \begin{equation}
@@
 \end{equation}
-where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays.
+where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, $S_k(t)$ is a dummy variable for School Holidays, and $C_t$ is the temperature at time $t$.
+Public and school holidays were included following experiments on the training data which revealed systematic biases associated with these periods.
 
-However, calibration of resulting forecasts on the training data is poor, suggesting that the shape of the Poisson distribution is unsuitable for the present application.
-In particular, we observed that forecast uncertainty appeared to vary depending on the time of day and possibly other explanatory variables.
-Therefore, we also consider more flexible two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
+However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
+In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
+Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
 The resulting density forecasts are given by
@@
 for the mean and variance parameters.
-
-Truncated $t$ distribution...
-
-Negative binomial...
+We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
 

### 2021-10-21 paper/paper.Rmd
add intro

--- 2021-10-19 paper/paper.Rmd
@@
 
-1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv)
+1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)
 
@@
 As our data are counts, the natural starting point is the Poisson distribution, given by
+
 \begin{equation}
@@
 \end{equation}
-where we initially consider an additive model for $\lambda_t$ of the form
+
+where we consider an additive model for $\lambda_t$ of the form
+
 \begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
+  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
  (\#eq:additivemodel)
 \end{equation}
-where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$.
-Analysis of this model fit on the training data revealed systematic biases which were identified as relating to public and school holidays.
-We therefore expand the model to include these effects as follows,
-\begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
- (\#eq:additivemodel_v2)
-\end{equation}
-where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays.
 
-However, calibration of resulting forecasts on the training data is poor, suggesting that the shape of the Poisson distribution is unsuitable for the present application.
-In particular, we observed that forecast uncertainty appeared to vary depending on the time of day and possibly other explanatory variables.
-Therefore, we also consider more flexible two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
-The resulting density forecasts are given by
+The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
+
+Truncated Normal...
+
 \begin{equation}
@@
 \end{equation}
-with additive models
+
+with
+
 \begin{align*}
-  \mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-  log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
+  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+  log(\sigma_t) & = f(H(t)) .
 \end{align*}
-for the mean and variance parameters.
 

### 2021-10-21 paper/paper.Rmd
JB paper addition
JB text addition

--- 2021-10-21 paper/paper.Rmd
@@
 
-1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv)
+1.
+We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
 
-2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;
+2.
+We develop a novel approach to forecast day-ahead hourly hospital admission using Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of admissions, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
 
@@
 
+We develop a range of models for forecasting hospital admissions using a multi-year dataset.
+In order to facilitate model development and testing, we divide the data into a training period.
+We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data.
+
+Forecast are produced/issue at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis.
+In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,...
+0.95, as well as the expectation.
+
 ## data {#data}
+
+The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergentcy at hourly granularity.
+The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.
 
@@
 
-Empirical distribution for same hour-of-day and day-of-week as target time.
+The first benchmark we consider is to simply use the empirical distribution of past observations as a probabilistic forecast for same hour-of-day and day-of-week that we are forecasting.
+This is often called `climatology'.
+However, as we observe a trend in the data used here, we re-calculate the climatology on a rolling basis using the most recent one-year of data in order to track the trend.
 
@@
 
+## Model fits
+
+### Truncated Normal GAMLSS Model
+
+**Not sure if this should stay here, but interesting and worth reporting, I think...**
+
+Plot smooth functions estiamted for time of day, day of year, for both mean and variance parameters.
+
 ## Forecast evaluation
@@
 
+Forecast performance by time of day...
+
 # Conclusion {#conclusion}

### 2021-10-26 paper/paper.Rmd
JB paper additions

@@
 
-We develop a range of models for forecasting hospital admissions using a multi-year dataset.
-In order to facilitate model development and testing, we divide the data into a training period.
+We develop a range of models for forecasting hospital admissions using a multi-year dataset from a hospital in the UK.
+In order to facilitate model development and testing, we divide the data into a training and test period.
 We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data.
+Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
 
-Forecast are produced/issue at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis.
+Forecast are issued every day at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis.
 In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,...
-0.95, as well as the expectation.
+0.95, as well as the predictive expectation.
 
-## data {#data}
+In the remainder of this section we explore the data in more detail, describe the forecast evaluation metrics we will employ, and introduce a range of benchmark models for producing forecasts.
+The benchmarks serve as reference methods and include some simple methods as well as more sophisticated routines designed to work `out-of-the-box', such as Facebook's Prophet.
+The performance of these methods provides an important reference, and any proposed novel solution should offer material benefits to justify increased complexity.
 
-The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergentcy at hourly granularity.
+## Data exploration {#data}
+
+The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergency at hourly granularity.
 The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.
@@
 log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
+(\#eq:additivemodel_NOtr_v2)
 \end{align*}
@@
 We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
+
+# Result and discussion {#result}
 
@@
 The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a].
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
 
@@
 It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
-Furthermore,
-The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
+Furthermore, the Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation.
+The Pinball Score is given by
 
@@
 
-Calibration, also called `reliability`, is the property that forecast probabilities match the observed frequency of realisations.
+Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations.
 If a forecast is calibrated, then $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
@@
 
-# Result and discussion {#result}
+## Model fits
 
@@
 
-## Case study
-
-The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
-Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
-
 (Summary of data, e.g. discussion of observations from training data/model development: long-term trend)
-
-## Model fits
 
@@
 
-**Not sure if this should stay here, but interesting and worth reporting, I think...**
+We can support the interpretation the models given in Equation \@ref(eq:additivemodel_NOtr_v2) after the parameters have been estimated by visualising their reposes to inputs.
+Figure \@ref(fig-NOtr-model) shows the effect of time of day and day-type on the level and spread of predicted arrivals.
 
-Plot smooth functions estiamted for time of day, day of year, for both mean and variance parameters.
+```{r fig-NOtr-model}
+
+load("NOtr2_model.Rda")
+
+plot_data <- data.frame(Hour = 0:23)
+for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
+  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "mu",fixed.at = list(dow3=i))
+  plot_data[[i]] <- pef(0:23)
+}
+
+plot_data <- reshape::melt.data.frame(plot_data,id.vars="Hour",variable_name="Day Type")
+
+plot_1 <- ggplot(plot_data,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
+  ylab("Effect [Expected Arrivals]") + theme_few()
+
+# plot_1
+
+plot_data2 <- data.frame(Hour = 0:23)
+for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
+  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "sigma",fixed.at = list(dow3=i))
+  plot_data2[[i]] <- pef(0:23)
+}
+
+plot_data2 <- reshape::melt.data.frame(plot_data2,id.vars="Hour",variable_name="Day Type")
+
+plot_2 <- ggplot(plot_data2,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
+  ylab("Effect [Standard Deviation of Arrivals]") + theme_few()
+
+#plot_2
+
+grid.arrange(plot_1, plot_2, ncol=2)
+
+```
 

### 2021-10-27 paper/paper.Rmd
Comments about the linear regression and rounding.

@@
 
+Note that we have also tried the Normal distribution in our experiments, but as expected it was not suitable and performed poorly.
+This is because it is symmetric around the mean, while the data exhibits skewness.
+
 ### ETS - Exponential Smoothing model
@@
 We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
+This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
+
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
 If integer-valued quantiles are needed, then rounding up can be done for them.
-This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
+We have also tried rounding up the quantiles to see how the model would perform in this case, but the difference between ADAM described above and the one with integer values was infinitesimal.
 

### 2021-10-27 results/Evaluatoin_Script.R
Point to correct prophet results

@@ -149,7 +149,7 @@ big_eval_function(forecast_DT = faster,h2_actuals = h2,method_name = "faster")
 rm(faster)
 
 
-prophet <- data.table(readRDS("forecast_prophet.rds"))
+prophet <- data.table(readRDS("prophet_bahman.rds"))
 # setnames(prophet,old = c("origin","target","point_forecast"),c("issueTime","targetTime_UK","expectation"))
 setnames(prophet,old = c("origin","target",1:19/20),c("issueTime","targetTime_UK",paste0("q",1:19*5)))
 prophet[,issueTime := issueTime+3600]

### 2021-10-27 paper/paper.Rmd
A bit on computational time

@@
 
+Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function.
+All functions were re-estimated on each iteration.
+ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin.
+This allowed to speed up the computation for these two models.
+The initial estimation of ADAM took approximately one hour and 25 minutes.
+Each step in the experiment took the time shown below.
+
 ## Model fits

### 2021-10-27 r_script/result_viz.R
Update results and plots for paper

@@ -2,10 +2,11 @@
 my_colorblind <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
                    "#F0E442", "#0072B2", "#D55E00", "#CC79A7","black")
 PB <- read_rds("results/PB.rds")
-selected_method <- c("ADAM-iETSX","GBM",
-                     "NOtr-2","Regression-Poisson",
-                     "tbats","ETS","Benchmark-1",
-                     "Poisson-1","NBI-2-I")
+selected_method <- c("ADAM-iETSX", # ETS Example
+                     "GBM-2",      # ML/non-parametric method
+                     "NOtr-2",     # GAMLSS - good
+                     "tbats","Benchmark-2", # Benchmarks
+                     "NBI-2") # GAMLSS - not good
 PB_selected <- PB %>% as_tibble() %>% 
   mutate(Method=factor(Method)) %>% 
   filter(Method %in% selected_method)
@@ -36,7 +37,7 @@ ggplot(data=REL_selected %>% filter(Horizon=="All" & Issue == "All"),
            color=Method)) +
   geom_line(data=REL_nom,
             aes(x=Nominal,y=Empirical), 
-            color="red",size=1.1,show.legend = F) +
+            color="black",size=1,show.legend = F,linetype=2) +
   geom_line() + geom_point() +
   scale_x_continuous(breaks = seq(0,1,.1))+
   scale_y_continuous(breaks = seq(0,1,.1))+
@@ -51,12 +52,12 @@ REL1_selected <- REL1 %>% as_tibble() %>%
   filter(Method %in% selected_method)
 
 ggplot(data=REL1_selected %>% filter(Horizon=="All" & Issue == "All")
-         ,aes(x=Nominal,
-              y=`Quantile Bias`,
-              color=Method)) +
+       ,aes(x=Nominal,
+            y=`Quantile Bias`,
+            color=Method)) +
   geom_line(data=REL_nom,aes(x=Nominal,
                              y=`Quantile Bias`), 
-            color="red",size=1.1,show.legend = F) +
+            color="black",size=1,linetype=2,show.legend = F) +
   geom_line() + geom_point() +
   xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
   scale_color_manual(
[…]

### 2021-10-28 paper/paper.Rmd
change admision to arrival and hospital to ED

--- 2021-10-26 paper/paper.Rmd
@@
 ---
-title:  Forecasting hourly hospital admission
+title:  Forecasting short-term hourly Emergency Departement arrivals
 author:
@@
 
-Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients.
-An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients.
+An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
 This is one of the best ways to optimize staffing utilization and related costs.
 Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives [@rostami2020anticipating].
+An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
+Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
+Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing [@rostami2020anticipating].
 
-Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
-Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
-
-There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting].
-daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] admission.
-In this paper, we only focus on forecasting hourly hospital admission.
-In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting].
+daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals.
+In this paper, we only focus on forecasting hourly arrivals.
+In comparison to lower frequency time series forecasting such as monthly quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This will have many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
-The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service.
+This involves the short-term decision making related to the execution of the delivery process in ED.
+<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
+The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
@@
 
-There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
 These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[].
@@
 time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Moreover, all previous publications  covered in this paper fails reproducibility principles (Boylan, 2016).
+Moreover, all previous publications  covered in this paper fails reproducibility principles.
 
@@
 
-1.
-We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
+1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)
 
-2.
-We develop a novel approach to forecast day-ahead hourly hospital admission using Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of admissions, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
+2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;
 
@@
 knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
-  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") 
+  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") %>%  kableExtra::landscape()
 ```
@@
 
-We develop a range of models for forecasting hospital admissions using a multi-year dataset from a hospital in the UK.
-In order to facilitate model development and testing, we divide the data into a training and test period.
-We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data.
-Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
-
-Forecast are issued every day at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis.
-In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,...
-0.95, as well as the predictive expectation.
-
-In the remainder of this section we explore the data in more detail, describe the forecast evaluation metrics we will employ, and introduce a range of benchmark models for producing forecasts.
-The benchmarks serve as reference methods and include some simple methods as well as more sophisticated routines designed to work `out-of-the-box', such as Facebook's Prophet.
-The performance of these methods provides an important reference, and any proposed novel solution should offer material benefits to justify increased complexity.
-
-## Data exploration {#data}
-
-The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergency at hourly granularity.
-The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.
+## data {#data}
 
@@
 
-The first benchmark we consider is to simply use the empirical distribution of past observations as a probabilistic forecast for same hour-of-day and day-of-week that we are forecasting.
-This is often called `climatology'.
-However, as we observe a trend in the data used here, we re-calculate the climatology on a rolling basis using the most recent one-year of data in order to track the trend.
+Empirical distribution for same hour-of-day and day-of-week as target time.
 
@@
 In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
-Here we use this framework to extend the Poisson regression introduced above, and explore alternative distirbutional assumptions.
-
 Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
@@
 \end{equation}
-<!-- where we initially consider an additive model for $\lambda_t$ of the form -->
-<!-- \begin{equation} -->
-<!--   \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
-<!--  (\#eq:additivemodel) -->
-<!-- \end{equation} -->
-<!-- where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$.
-Analysis of this model fit on the training data revealed systematic biases which were identified as relating to public and school holidays.
-We therefore expand the model to include these effects as follows, -->
-<!-- \begin{equation} -->
-<!--   \log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
-<!--  (\#eq:additivemodel_v2) -->
-<!-- \end{equation} -->
-<!-- where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays. -->
+
 where we consider an additive model for $\lambda_t$ of the form
+
 \begin{equation}
-\log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
-(\#eq:additivemodel_v2)
+  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
+ (\#eq:additivemodel)
 \end{equation}
-where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, $S_k(t)$ is a dummy variable for School Holidays, and $C_t$ is the temperature at time $t$.
-Public and school holidays were included following experiments on the training data which revealed systematic biases associated with these periods.
 
-However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
-In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
-Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
-The resulting density forecasts are given by
+The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
+
+Truncated Normal...
+
 \begin{equation}
@@
 \end{equation}
-with additive models
+
+with
+
 \begin{align*}
-\mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
-(\#eq:additivemodel_NOtr_v2)
+  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+  log(\sigma_t) & = f(H(t)) .
 \end{align*}
-for the mean and variance parameters.
-We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
 
-# Result and discussion {#result}
+Truncated $t$ distribution...
+
+Negative binomial...
 
@@
 The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a].
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.
 
@@
 It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
-Furthermore, the Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation.
-The Pinball Score is given by
+Furthermore,
+The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
 
@@
 
-Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations.
+Calibration, also called `reliability`, is the property that forecast probabilities match the observed frequency of realisations.
 If a forecast is calibrated, then $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
@@
 
-## Model fits
+# Result and discussion {#result}
 
@@
 
+## Case study
+
+The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
+Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
+
 (Summary of data, e.g. discussion of observations from training data/model development: long-term trend)
-
-### Truncated Normal GAMLSS Model
-
-We can support the interpretation the models given in Equation \@ref(eq:additivemodel_NOtr_v2) after the parameters have been estimated by visualising their reposes to inputs.
-Figure \@ref(fig-NOtr-model) shows the effect of time of day and day-type on the level and spread of predicted arrivals.
-
-```{r fig-NOtr-model}
-
-load("NOtr2_model.Rda")
-
-plot_data <- data.frame(Hour = 0:23)
-for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
-  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "mu",fixed.at = list(dow3=i))
-  plot_data[[i]] <- pef(0:23)
-}
-
-plot_data <- reshape::melt.data.frame(plot_data,id.vars="Hour",variable_name="Day Type")
-
-plot_1 <- ggplot(plot_data,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
-  ylab("Effect [Expected Arrivals]") + theme_few()
-
-# plot_1
-
-plot_data2 <- data.frame(Hour = 0:23)
-for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
-  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "sigma",fixed.at = list(dow3=i))
-  plot_data2[[i]] <- pef(0:23)
-}
-
-plot_data2 <- reshape::melt.data.frame(plot_data2,id.vars="Hour",variable_name="Day Type")
-
-plot_2 <- ggplot(plot_data2,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
-  ylab("Effect [Standard Deviation of Arrivals]") + theme_few()
-
-#plot_2
-
-grid.arrange(plot_1, plot_2, ncol=2)
-
-```
 
@@
 
-```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
-#knitr::include_graphics(rep("../results/QuantileBias.png"))
-## Should reduce the number of models shown here... it's really hard to tell them appart!
-
-REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
-                      Empirical=seq(0,1,by=0.05),
-                      `Quantile Bias`= 0,
-                      Method="Nominal")
-
-REL[,`Quantile Bias`:=Empirical-Nominal]
-
-ggplot(data=REL[Horizon=="All" & Issue == "All",],aes(x=Nominal,y=`Quantile Bias`,group=Method,color=Method)) +
-  geom_line(data=REL_nom,aes(x=Nominal,y=`Quantile Bias`), color="black",size=1.1,show.legend = F) +
-  geom_line() + geom_point() +
-  xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
-  theme_few() 
-```
+# ```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
+# #knitr::include_graphics(rep("../results/QuantileBias.png"))
+# ## Should reduce the number of models shown here... it's really hard to tell them appart!
+# 
+# REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
+#                       Empirical=seq(0,1,by=0.05),
+#                       `Quantile Bias`= 0,
+#                       Method="Nominal")
+# 
+# REL[,`Quantile Bias`:=Empirical-Nominal]
+# 
+# ggplot(data=REL[Horizon=="All" & Issue == "All",],aes(x=Nominal,y=`Quantile Bias`,group=Method,color=Method)) +
+#   geom_line(data=REL_nom,aes(x=Nominal,y=`Quantile Bias`), color="black",size=1.1,show.legend = F) +
+#   geom_line() + geom_point() +
+#   xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
+#   theme_few() 
+# ```
 
@@
 ```
-
-The Pinball Score evaluates overall performance, penalising both poor calibration and lack of sharpness, and ranks the various models quite differently.
-NBI-2 and Poisson-2 achive the lowest Pinball Loss, but as we have seen, are poorly calibrated so not suitable for use in decision making.
-Of the well calibrated methods, NOtr-2 and Ttr-2 have the lowest Pinball Score, and provide a significant improvement over Benchmark 2 by this metric.
-However, we note that while the improvement in terms of Pinball Score appears small, these methods have substantially improved forecast calibration compared to the benchmarks, which are both poorly calibrated.
 
@@
 
-Forecast performance by time of day...
-
 # Conclusion {#conclusion}

### 2021-10-28 paper/paper.Rmd

--- 2021-10-27 paper/paper.Rmd
@@
 
-plot_data <- data.frame(Hour = 0:23)
-for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
-  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "mu",fixed.at = list(dow3=i))
-  plot_data[[i]] <- pef(0:23)
-}
-
-plot_data <- reshape::melt.data.frame(plot_data,id.vars="Hour",variable_name="Day Type")
-
-plot_1 <- ggplot(plot_data,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
-  ylab("Effect [Expected Arrivals]") + theme_few()
-
-# plot_1
-
-plot_data2 <- data.frame(Hour = 0:23)
-for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
-  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "sigma",fixed.at = list(dow3=i))
-  plot_data2[[i]] <- pef(0:23)
-}
-
-plot_data2 <- reshape::melt.data.frame(plot_data2,id.vars="Hour",variable_name="Day Type")
-
-plot_2 <- ggplot(plot_data2,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
-  ylab("Effect [Standard Deviation of Arrivals]") + theme_few()
-
-#plot_2
-
-grid.arrange(plot_1, plot_2, ncol=2)
-
-```
-
-## Forecast evaluation
-
-Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated.
-Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \@ref(fig-quantilebias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
-Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-
-```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
-#knitr::include_graphics(rep("../results/QuantileBias.png"))
-## Should reduce the number of models shown here... it's really hard to tell them appart!
-
-REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
-                      Empirical=seq(0,1,by=0.05),
-                      `Quantile Bias`= 0,
-                      Method="Nominal")
-
-REL[,`Quantile Bias`:=Empirical-Nominal]
-
-ggplot(data=REL[Horizon=="All" & Issue == "All",],aes(x=Nominal,y=`Quantile Bias`,group=Method,color=Method)) +
-  geom_line(data=REL_nom,aes(x=Nominal,y=`Quantile Bias`), color="black",size=1.1,show.legend = F) +
-  geom_line() + geom_point() +
-  xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
-  theme_few() 
-```
+Evaluation metrics from all Methods on the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above.
 
@@
 results_table <- read_rds("results_table.rds")
+results_table <- results_table[!results_table$Method%in%c("ADAM-iETSX-Ceiling","LinearRegression","Poisson-2-I","fasster","qreg-1"),]
+results_table <- results_table[order(results_table$`Quantile Bias`),]
+
 knitr::kable(results_table, booktabs = T, 
@@
 
-Evaluation metrics from the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.
+The Pinball Score evaluates overall performance, penalising both poor calibration and lack of sharpness, and ranks the various models quite differently.
+Pinball is visualised in Figure \'ref(Pinball) broken down by quantile.
+NBI-2 and Poisson-2 achieve the lowest Pinball Loss, but are poorly calibrated so not suitable for use in decision making.
+Of the well calibrated methods, NOtr-2 and Ttr-2 have the lowest Pinball Score, and provide a significant improvement over Benchmark 2 by this metric.
+However, we note that while the improvement in terms of Pinball Score appears small, these methods have substantially improved forecast calibration compared to the benchmarks, which are poorly calibrated.
 
-```{r tab-results-qbias, out.width= "50%"}
-results_table_qb <- read_rds("results_table.rds") %>% select(`Quantile Bias`)
-
-knitr::kable(results_table_qb, booktabs = T, 
-             linesep = "",caption = "Forecast evluation for all methods on test data in decreasing order by Quantile Bias") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
+```{r Pinball, fig.cap= "Pinball loss for selected forecsating methods and quantiles from 5% to 95%.
+Several methods perform similarly well relative to others by this metric but calibration should also be considered when evaluating probabilistic foreasts.", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-The Pinball Score evaluates overall performance, penalising both poor calibration and lack of sharpness, and ranks the various models quite differently.
-NBI-2 and Poisson-2 achive the lowest Pinball Loss, but as we have seen, are poorly calibrated so not suitable for use in decision making.
-Of the well calibrated methods, NOtr-2 and Ttr-2 have the lowest Pinball Score, and provide a significant improvement over Benchmark 2 by this metric.
-However, we note that while the improvement in terms of Pinball Score appears small, these methods have substantially improved forecast calibration compared to the benchmarks, which are both poorly calibrated.
+In order to verify that this improvement is a result of improved modelling and not a result of sample variation, we have performed block bootstrap resampling on Pinball skill scores relative to Bechmark-2 for each forecast.
+The results for all methods with a skill score greater than -2% are shown in Figure \@ref(Skill_rel2bench_reduced), where boxes are coloured according to calibration.
+Where boxes do not overlap we can be confident that apparent differences in skill are not the result of sample variation.
 
-```{r fig-skill, fig.cap= "Skill score ...", fig.align='center',out.width= "90%"}
-knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
-
-# knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
-
-load("../results/all_results_paper_2021-10-07.Rda")
-
-plotdata <- melt(bootdata[,100*(get(REF)-.SD)/get(REF),.SDcols=NAMES],
-                 measure.vars =  1:length(NAMES),variable.name = "Method",
-                 value.name = "Skill Score")
-
-## Merge Reliability for colouring...
-plotdata <- merge(plotdata,REL[kfold=="Test" & Horizon=="All" & Issue == "All",.(Qbias=mean(abs(Nominal-Empirical))),by="Method"],
-                  by="Method",all.x = T)
-
-ggplot(plotdata[`Skill Score`>-2,], aes(x=reorder(Method, -`Skill Score`), y=`Skill Score`, fill=Qbias)) + 
-  ylab("Pinball Skill Score [%]") +
-  geom_boxplot() + theme_few() +
-  # ggtitle("Pinball Skill Score Relative to Benchmark 2") +
-  theme(axis.text.x = element_text(angle = -80)) + #scale_fill_manual(values=cbPalette) +
-  # scale_x_discrete(labels= paste0(substring(NAMES, 1,4),".")) +
-  geom_hline(yintercept=0, linetype="dashed",size=0.5)+
-  scale_fill_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlGn")),
-                       limits=c(0,0.11))+
-  labs(fill = "Quantile Bias",x="Method")
-
+```{r Skill_rel2bench_reduced, fig.cap= "Bootstrapped skill scores for top-performing methods relative to Benchmark-2.
+Non-overlapping boxes indicate that we can be confident that apparent differences in Pinball Score are not the result of sample variation.", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-Forecast performance by time of day...
+```{r Pinaball_vs_Qbias, fig.cap= "Pinball Score vs average Qunatile Bias for selected methods.
+While some methods perform well by one measure or the other, only NOtr-2 is both calibrated and has a low Pinball Score.", fig.align='center',fig.width= 7,fig.height=6}
+```
+
+We delve deeper into forecast performance by evaluating forecasts by time of day, and lead-time.
+The Pinball Score for slected methods is plotted against lead-time in Figure \@ref(lead-time-pb), separated by whether the forecast was issued at midnight (00) or mid-day (12).
+We can see clearly that time of day is a greater influence on performance than lead-time, with morning arrivals and the evening peak in arrivals proving the most difficult to predict.
+
+```{r lead-time-pb, fig.cap= "Pinball Score by lead-time for forecasts issued at midnight (00) and mid-day (12).", fig.align='center',fig.width= 7,fig.height=6}
+```
+
+Computational time should also be considered, as national public health bodies may be required to produce forecasts of many variables for a large number of hospitals on  regular basis...
 
@@
 
+1. Conditional distribution of future arrivals is neither Gaussian nor Poisson distributed, and requires careful treatment.
+
+2. Simple benchmarks and "out-of-the-box" forecasting methods produce poor predictions of A&E arrivals.
+
+3. A distributional regression appraoch, with additive models for both the location and scale parameters of a truncated Normal distirbution produces best all-round forecast performance
+
 # References {#references .unnumbered}

### 2021-11-01 paper/paper.Rmd
using arrival time instead of admission in the text, put all plots in the paper

--- 2021-10-28 paper/paper.Rmd
@@
 library(ggplot2)
-library(data.table)
 library(RColorBrewer)
@@
 An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
+Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
 Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing [@rostami2020anticipating].
@@
 In this paper, we only focus on forecasting hourly arrivals.
-In comparison to lower frequency time series forecasting such as monthly quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->
+
+An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This will have many advantages for both patients, staff and the quality of provided services.
@@
 This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features.
-All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission.
-Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
+All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals.
+However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
 Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
@@
 The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
-In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
-Aditionally, datasets used in some of these studies are relatively small, short length (e.g.
+In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+Moreover, datasets used in some of these studies are relatively small, short length (e.g.
 time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.
+The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
 
@@
 
-# Research background: hourly hospital admission forecasting {#lit}
+# Research background: hourly ED forecasting {#lit}
 
-Other papers to review and possibly cite:
-- Daily visits using Prophet @park2019144
-- Daily using internet data @ekstrom2015forecasting
-- Review paper: @wargon2009systematic
-- Others: @boyle2012predicting, 
-- Impact of weather @parsons2011modelling, @macgregor2003effect
+There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals.
+The time granularity considered by these studies spans from hourly to yearly.
+However, given the focus of the paper, we only focus on studies on hourly ED forecasting.
+Table \@ref(tab:summarylit) summarize studies in hourly forecasting in the emergency department.
 
-hourly:
-@cheng2021forecasting
-@rocha2021forecasting
-
-Table \@ref(tab:summarylit) summarise studies in hourly forecasting in emergency and urgent care.
-
-```{r summarylit}
+```{r summarylit, cache=FALSE}
 li <- readxl::read_xlsx("../table/table.xlsx")
-knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
+knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly Emergency Department forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
   column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") %>%  kableExtra::landscape()
@@
 
-Predicting the arrivals of accident and emergency department future patients is studied by @choudhury2020forecasting ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
-ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy.
+Predicting the arrivals of an ED future arrivals is also studied by @choudhury2020forecasting.
+ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
+ARIMA model was selected as the best fit model.
+Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy.
 @hertzum2017forecasting work was mentioned in this paper.
-It is said that residual normality, stationarity, and autocorrelation have not been tested in @hertzum2017forecasting paper and this might be the cause of accuracy problems.
-However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in @choudhury2020forecasting
-@morzuch2006forecasting use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours.
-The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+This result is surprising given the existence of multiple seasonality in the hourly dataset.
+
+@morzuch2006forecasting used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic.
+Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
 The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
 
-@mccarthy2008challenge predict patient demand for ED services by characterising ED arrivals.
+@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
 Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
-Then, they presente the predictiona interval accuracy of 50 and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions.
+Authors .
+They present the prediction interval accuracy of $50\%$ and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions.
 They show that the most important predictor is hour of the day and autocorrelation lag 1.
-However, these findings are limited to the short period of data (one year).
+However, these findings are limited to the short number of observations (only one year of historical data).
 
@@
 
-@steins2019forecasting aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls.
-Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic.
-In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models.
-These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE.
-Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day.
-Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model.
-However, the improvement provided by the more advance model was not much greater than the one provided by the existing model.
-Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors.
-Moreover, both models either underestimated or overestimated the number of calls.
-Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend.
-Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.
-
 According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
 Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting].
-However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
+However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
 
 There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
-These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.
+These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.
 there is no benchmark method nor is forecast accuracy reported.
@@
 
-Figure \@ref(fig:hourly-plot-ridge)
+Data used in the study comprised counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
+We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
+
+Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.
+
+It is clear that the number of arrivals has a sub-daily structure.
+The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again.
+It is also clear that ED service gets systematically more visits on Monday between 8 a.m.
+and 5 p.m.
+Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.
 
@@
 
-Figure \@ref(fig:hourly-plot)
+Figure \@ref(fig:hourly-plot) also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events.
+It is also clear that arrivals are less volatile between mid-night and early morning.
 
-```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "80%"}
+We propose a forecasting model that accounts for the systematic structure of the data.
+
+```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"}
 
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \@ref(fig-quantilebias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
+Calibration is evaluated visually in \@ref(quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
 
-# ```{r fig-quantilebias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
-# #knitr::include_graphics(rep("../results/QuantileBias.png"))
-# ## Should reduce the number of models shown here... it's really hard to tell them appart!
-# 
-# REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
-#                       Empirical=seq(0,1,by=0.05),
-#                       `Quantile Bias`= 0,
-#                       Method="Nominal")
-# 
-# REL[,`Quantile Bias`:=Empirical-Nominal]
-# 
-# ggplot(data=REL[Horizon=="All" & Issue == "All",],aes(x=Nominal,y=`Quantile Bias`,group=Method,color=Method)) +
-#   geom_line(data=REL_nom,aes(x=Nominal,y=`Quantile Bias`), color="black",size=1.1,show.legend = F) +
-#   geom_line() + geom_point() +
-#   xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
-#   theme_few() 
-# ```
+```{r Pinball, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
+```
+
+```{r lead-time-rel, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
+```
+
+```{r Reliability, fig.cap= "Reliability ...", fig.align='center',out.width= "90%"}
+```
+
+```{r quantile-bias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
+```
 
@@
 
-```{r tab-results-qbias, out.width= "50%"}
-results_table_qb <- read_rds("results_table.rds") %>% select(`Quantile Bias`)
+```{r Skill-rel2bench-reduced, fig.cap= "Skill score ...", fig.align='center',out.width= "70%"}
 
-knitr::kable(results_table_qb, booktabs = T, 
-             linesep = "",caption = "Forecast evluation for all methods on test data in decreasing order by Quantile Bias") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
 ```
 
-```{r fig-skill, fig.cap= "Skill score ...", fig.align='center',out.width= "90%"}
-knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
 
-# knitr::include_graphics(rep("../results/Skill_rel2bench_reduced.png"))
+```
 
-load("../results/all_results_paper_2021-10-07.Rda")
+```{r lead-time-rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
 
-plotdata <- melt(bootdata[,100*(get(REF)-.SD)/get(REF),.SDcols=NAMES],
-                 measure.vars =  1:length(NAMES),variable.name = "Method",
-                 value.name = "Skill Score")
+```
 
-## Merge Reliability for colouring...
-plotdata <- merge(plotdata,REL[kfold=="Test" & Horizon=="All" & Issue == "All",.(Qbias=mean(abs(Nominal-Empirical))),by="Method"],
-                  by="Method",all.x = T)
-
-ggplot(plotdata[`Skill Score`>-2,], aes(x=reorder(Method, -`Skill Score`), y=`Skill Score`, fill=Qbias)) + 
-  ylab("Pinball Skill Score [%]") +
-  geom_boxplot() + theme_few() +
-  # ggtitle("Pinball Skill Score Relative to Benchmark 2") +
-  theme(axis.text.x = element_text(angle = -80)) + #scale_fill_manual(values=cbPalette) +
-  # scale_x_discrete(labels= paste0(substring(NAMES, 1,4),".")) +
-  geom_hline(yintercept=0, linetype="dashed",size=0.5)+
-  scale_fill_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlGn")),
-                       limits=c(0,0.11))+
-  labs(fill = "Quantile Bias",x="Method")
-
+```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
 ```

### 2021-11-01 paper/paper.tex
using arrival time instead of admission in the text, put all plots in the paper

@@
             pdfauthor={},
-            pdftitle={Forecasting hourly hospital admission},
+            pdftitle={Forecasting short-term hourly Emergency Departement arrivals},
             colorlinks=false,
@@
 
-  \title{Forecasting hourly hospital admission}
+  \title{Forecasting short-term hourly Emergency Departement arrivals}
     \author[University1]{Author1\corref{1}}
    \ead{email1@example.com} 
-    \author[University2]{Author2\corref{2}}
-   \ead{email2@example.com} 
+    \author[School of Mathematics and Statistics, University of Glasgow, UK]{Jethro Browell\corref{2}}
+   \ead{jethro.browell@glasgow.ac.uk} 
     \author[Centre for Marketing Analytics and Forecasting, Lancaster University, UK]{Ivan Svetunkov\corref{2}}
@@
 
-Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients.
-An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients.
+An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
 This is one of the best ways to optimize staffing utilization and related costs.
 Most current practice to optimise personnel scheduling follows the general approach originally presented in \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-An accurate demand forecasting is crucial in Accident \& Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives \citep{rostami2020anticipating}.
+An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
+Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow \citep{derlet2002overcrowding}.
+Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing \citep{rostami2020anticipating}.
 
-Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow \citep{derlet2002overcrowding}.
-Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
+There exists a large literature on forecasting ED arrivals that use various methods to forecast annual\citep{tandberg1994time} monthly,\citep{chen2011long, mai2015predicting}.
+daily\citep{rostami2020anticipating, park2019144} and subdaily \citep{schweigler2009forecasting, cheng2021forecasting} arrivals.
+In this paper, we only focus on forecasting hourly arrivals.
+In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
 
-There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual\citep{tandberg1994time} monthly,\citep{chen2011long, mai2015predicting}.
-daily\citep{rostami2020anticipating, park2019144} and subdaily \citep{schweigler2009forecasting, cheng2021forecasting} admission.
-In this paper, we only focus on forecasting hourly hospital admission.
-In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This will have many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
-The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service.
+This involves the short-term decision making related to the execution of the delivery process in ED.
+The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
@@
 
-There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
 These studies use multiple approaches including Exponential Smoothing {[}REF{]}, ARIMA families {[}{]}, ARCH {[}REF{]}, Vector Autoregressive {[}REF{]}, TBATS {[}{]} and ANN{[}{]}.
@@
 The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
-In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
-Aditionally, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Moreover, all previous publications covered in this paper fails reproducibility principles (Boylan, 2016).
+In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+Moreover, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
+Moreover, all previous publications covered in this paper fails reproducibility principles.
 
@@
 \item
-  We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv)
+  We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv) (Jethro to complete)
 \item
@@
   We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.
+The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
 \end{enumerate}
@@
 \hypertarget{lit}{%
-\section{Research background: hourly hospital admission forecasting}\label{lit}}
+\section{Research background: hourly ED forecasting}\label{lit}}
 
-Other papers to review and possibly cite:
-- Daily visits using Prophet \citet{park2019144}
-- Daily using internet data \citet{ekstrom2015forecasting}
-- Review paper: \citet{wargon2009systematic}
-- Others: \citet{boyle2012predicting},
-- Impact of weather \citet{parsons2011modelling}, \citet{macgregor2003effect}
+There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals.
+The time granularity considered by these studies spans from hourly to yearly.
+However, given the focus of the paper, we only focus on studies on hourly ED forecasting.
+Table \ref{tab:summarylit} summarize studies in hourly forecasting in the emergency department.
 
-hourly:
-\citet{cheng2021forecasting}
-\citet{rocha2021forecasting}
+\begin{landscape}\begin{table}[!h]
 
-Table \ref{tab:summarylit} summarise studies in hourly forecasting in emergency and urgent care.
-
-\begin{table}[!h]
-
-\caption{\label{tab:summarylit}Summary of studies in hourly emergency care forecasting}
+\caption{\label{tab:summarylit}Summary of studies in hourly Emergency Department forecasting}
 \centering
+\resizebox{\linewidth}{!}{
 \fontsize{11}{13}\selectfont
@@
 \toprule
-Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple\\
+Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple seasonality\\
 \midrule
-NA & NA & NA & NA & NA & NA & NA & NA & seasonality\\
 McCarthy 
@@
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
+\end{landscape}
 
@@
 
-Predicting the arrivals of accident and emergency department future patients is studied by \citet{choudhury2020forecasting} ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
-ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy.
+Predicting the arrivals of an ED future arrivals is also studied by \citet{choudhury2020forecasting}.
+ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
+ARIMA model was selected as the best fit model.
+Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy.
 \citet{hertzum2017forecasting} work was mentioned in this paper.
-It is said that residual normality, stationarity, and autocorrelation have not been tested in \citet{hertzum2017forecasting} paper and this might be the cause of accuracy problems.
-However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in \citet{choudhury2020forecasting}
-\citet{morzuch2006forecasting} use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours.
-The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+This result is surprising given the existence of multiple seasonality in the hourly dataset.
+
+\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic.
+Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
 The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
 
-\citet{mccarthy2008challenge} predict patient demand for ED services by characterising ED arrivals.
+\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
 Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
-Then, they presente the predictiona interval accuracy of 50 and \(90\%\) intervals for the number of hourly arrivals under the Poisson assumptions.
+Authors .
+They present the prediction interval accuracy of \(50\%\) and \(90\%\) intervals for the number of hourly arrivals under the Poisson assumptions.
 They show that the most important predictor is hour of the day and autocorrelation lag 1.
-However, these findings are limited to the short period of data (one year).
+However, these findings are limited to the short number of observations (only one year of historical data).
 
@@
 
-\citet{steins2019forecasting} aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls.
-Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic.
-In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models.
-These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE.
-Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day.
-Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model.
-However, the improvement provided by the more advance model was not much greater than the one provided by the existing model.
-Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors.
-Moreover, both models either underestimated or overestimated the number of calls.
-Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend.
-Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.
-
 According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
 Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast \citep{boyle2012predicting, hertzum2017forecasting}.
-However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
+However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
 
 There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
-These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method nor is forecast accuracy reported.
+These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method nor is forecast accuracy reported.
 
@@
 
-Figure \ref{fig:hourly-plot-ridge}
+Data used in the study comprised counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
+We aggregated the patients' arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
+
+Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.
+
+It is clear that the number of arrivals has a sub-daily structure.
+The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again.
+It is also clear that ED service gets systematically more visits on Monday between 8 a.m.
+and 5 p.m.
+Moreover, The number of arrivals around mid-night is slightly higher for Saturday and Sunday.
 
@@
 
-Figure \ref{fig:hourly-plot}
+Figure \ref{fig:hourly-plot} also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events.
+It is also clear that arrivals are less volatile between mid-night and early morning.
+
+We propose a forecasting model that accounts for the systematic structure of the data.
 
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/hourly-plot-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/hourly-plot-1}
 
@@
 
+If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
+In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
 Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution.
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \ref{fig-quantilebias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
+Calibration is evaluated visually in \ref{quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{../results/QuantileBias}
+{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Pinball-1}
 
@@
 
-\caption{Quantile ...}\label{fig:fig-quantilebias}
+\caption{pinball ...}\label{fig:Pinball}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/lead-time-rel-1}
+
+}
+
+\caption{pinball ...}\label{fig:lead-time-rel}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Reliability-1}
+
+}
+
+\caption{Reliability ...}\label{fig:Reliability}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/quantile-bias-1}
+
+}
+
+\caption{Quantile ...}\label{fig:quantile-bias}
 \end{figure}
@@
 
-\begin{table}[!h]
+\begin{figure}[H]
 
-\caption{\label{tab:tab-results-qbias}Forecast evluation for all methods on test data in decreasing order by Quantile Bias}
-\centering
-\fontsize{11}{13}\selectfont
-\begin{tabular}[t]{r}
-\toprule
-Quantile Bias\\
-\midrule
-0.1047874\\
-0.0557392\\
-0.0194389\\
-0.0600650\\
-0.0540740\\
-0.0375705\\
-0.0098936\\
-0.0118522\\
-0.0372137\\
-0.0373884\\
-0.0523937\\
-0.0929416\\
-0.0140252\\
-0.1862190\\
-0.0104673\\
-0.0104673\\
-0.0643451\\
-0.0855702\\
-\bottomrule
-\end{tabular}
-\end{table}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/Skill-rel2bench-reduced-1}
+
+}
+
+\caption{Skill score ...}\label{fig:Skill-rel2bench-reduced}
+\end{figure}
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{../results/Skill_rel2bench_reduced}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/rmse-1}
 
@@
 
-\caption{Skill score ...}\label{fig:fig-skill}
+\caption{RMSE ...}\label{fig:rmse}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/lead-time-rmse-1}
+
+}
+
+\caption{RMSE ...}\label{fig:lead-time-rmse}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/time-1}
+
+}
+
+\caption{Running time ...}\label{fig:time}
 \end{figure}

### 2021-11-01 r_script/result_viz.R

@@ -1,12 +1,15 @@
 ## ---- Pinball
 my_colorblind <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
                    "#F0E442", "#0072B2", "#D55E00", "#CC79A7","black")
-PB <- read_rds("results/PB.rds")
+PB <- read_rds("../results/PB.rds")
+REL <- read_rds("../results/REL.rds")
+plotdata <- read_rds("../results/plotdata.rds")
+
 selected_method <- c("ADAM-iETSX", # ETS Example
                      "GBM-2",      # ML/non-parametric method
                      "NOtr-2",     # GAMLSS - good
                      "tbats","Benchmark-2", # Benchmarks
-                     "NBI-2") # GAMLSS - not good
+                     "NBI-2", "prophet") # GAMLSS - not good
 PB_selected <- PB %>% as_tibble() %>% 
   mutate(Method=factor(Method)) %>% 
   filter(Method %in% selected_method)
@@ -23,7 +26,6 @@ ggplot(data=PB_selected %>% filter(Horizon=="All" & Issue == "All"),
   theme_few()
 
 ## ---- Reliability
-REL <- read_rds("results/REL.rds")
 REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
                       Empirical=seq(0,1,by=0.05),
                       `Quantile Bias`= 0,
@@ -37,7 +39,7 @@ ggplot(data=REL_selected %>% filter(Horizon=="All" & Issue == "All"),
            color=Method)) +
   geom_line(data=REL_nom,
             aes(x=Nominal,y=Empirical), 
-            color="black",size=1,show.legend = F,linetype=2) +
+            color="red",size=1,show.legend = F,linetype=2) +
   geom_line() + geom_point() +
   scale_x_continuous(breaks = seq(0,1,.1))+
   scale_y_continuous(breaks = seq(0,1,.1))+
@@ -57,7 +59,7 @@ ggplot(data=REL1_selected %>% filter(Horizon=="All" & Issue == "All")
             color=Method)) +
   geom_line(data=REL_nom,aes(x=Nominal,
                              y=`Quantile Bias`), 
-            color="black",size=1,linetype=2,show.legend = F) +
+            color="red",size=1,linetype=2,show.legend = F) +
[…]

### 2021-11-02 r_script/result_viz.R
knit to pdf successfully

@@ -26,7 +26,7 @@ ggplot(data=PB_selected %>% filter(Horizon=="All" & Issue == "All"),
   theme_few()
 
 ## ---- Reliability
-REL_nom <- data.table(Nominal=seq(0,1,by=0.05),
+REL_nom <- data.table::data.table(Nominal=seq(0,1,by=0.05),
                       Empirical=seq(0,1,by=0.05),
                       `Quantile Bias`= 0,
                       Method="Nominal")
@@ -60,7 +60,7 @@ ggplot(data=REL1_selected %>% filter(Horizon=="All" & Issue == "All")
   geom_line(data=REL_nom,aes(x=Nominal,
                              y=`Quantile Bias`), 
             color="red",size=1,linetype=2,show.legend = F) +
-  geom_line() + geom_point() +
+  geom_point() + geom_line() +
   xlim(c(0,1)) + ylim(c(-0.2,0.2)) + #ggtitle("Quantile Bias") +
   scale_color_manual(
     values = my_colorblind)+
@@ -79,6 +79,7 @@ ggplot(lt_pb_selected,
            y=Loss,color=Method)) + 
   facet_wrap(facets = "Issue") +
   geom_line() + 
+  geom_point()+
   expand_limits(x=c(1,48))+
   xlab("Lead-time [h]") + 
   ylab("Pinball Loss") + 
@@ -87,9 +88,10 @@ ggplot(lt_pb_selected,
   theme_few() 
 
 ## ---- lead-time-rel
-rel_pb <- REL[Horizon!="All",
-              .(Loss=mean(abs(Nominal-Empirical))),
-              by=c("Horizon","Method","Issue")]
+rel_pb <- REL %>% filter(Horizon!="All") %>%
+  group_by(Horizon,Method,Issue) %>% 
+  summarise(Loss=mean(abs(Nominal-Empirical))) %>% ungroup()
+
 rel_pb_selected <- rel_pb %>% as_tibble() %>% 
   mutate(Method=factor(Method)) %>% 
   filter(Method %in% selected_method)
@@ -108,9 +110,8 @@ ggplot(rel_pb_selected,
 
 ## ---- rmse
 RMSE <- read_rds("../results/RMSE.rds")
-rmse <- RMSE[Horizon=="All" & Method != "faster",
-             .(RMSE=mean(RMSE)),
-             by=c("Horizon","Method","Issue")]
[…]

### 2021-11-02 paper/paper.Rmd

@@
 library(knitr)
-knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.pos = 'H', message = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, cache=FALSE, fig.pos = 'H', message = FALSE, warning = FALSE)
 # Make sure you have the latest versions of rmarkdown and rticle
@@
 
-(Plots generated with theme_few - they look nice, good suggestion!
-Not sure how to include them in bookdown, can you help, Bahman?
-The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format?
-I guess PDF might be preferred.
-I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
-
 ## Case study
@@
 
+```{r time-accuracy, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
+```
+
 # Conclusion {#conclusion}

### 2021-11-02 paper/paper.tex

@@
 
-(Plots generated with theme\_few - they look nice, good suggestion!
-Not sure how to include them in bookdown, can you help, Bahman?
-The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format?
-I guess PDF might be preferred.
-I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
-
 \hypertarget{case-study}{%
@@
 \fontsize{11}{13}\selectfont
-\begin{tabular}[t]{lrrr}
+\begin{tabular}[t]{lrrrr}
 \toprule
-Method & Quantile Bias & Pinball & RMSE\\
+Method & Quantile Bias & Pinball & RMSE & Time\\
 \midrule
-Benchmark-1 & 0.1047874 & 1.254491 & NA\\
-Benchmark-2 & 0.0557392 & 1.217429 & NA\\
-ETS(XXX) & 0.0194389 & 1.434862 & 0.0121247\\
-GBM & 0.0600650 & 1.263135 & NA\\
-NBI-2-I & 0.0540740 & 1.206238 & NA\\
-NBI-2-log & 0.0375705 & 1.188041 & NA\\
-NOtr-1 & 0.0098936 & 1.222584 & NA\\
-NOtr-2 & 0.0118522 & 1.208556 & NA\\
-Poisson-1 & 0.0372137 & 1.204936 & NA\\
-Poisson-2 & 0.0373884 & 1.188107 & NA\\
-Poisson-2-I & 0.0523937 & 1.222114 & NA\\
-RegressionPoisson & 0.0929416 & 1.293524 & 0.8490258\\
-Ttr-2 & 0.0140252 & 1.210105 & NA\\
-faster & 0.1862190 & 35.051012 & NA\\
-iETSCeiling & 0.0104673 & 1.419490 & NA\\
-iETSXSeasonal & 0.0104673 & 1.417260 & 0.0896228\\
-qreg-1 & 0.0643451 & 1.340557 & NA\\
-tbats & 0.0855702 & 1.536080 & 0.4859770\\
+ADAM-iETSX & 0.0104673 & 1.417260 & 0.0896228 & 92.9348605\\
+ADAM-iETSX-Ceiling & 0.0104673 & 1.419490 & NA & NA\\
+Benchmark-1 & 0.1047874 & 1.254491 & 1.0042634 & 0.3874450\\
+Benchmark-2 & 0.0557392 & 1.217429 & 0.2592800 & 0.0947247\\
+ETS & 0.0194389 & 1.434862 & 0.0121247 & 10.7175205\\
+GBM-2 & 0.0600153 & 1.261690 & 1.7770897 & 602.4317496\\
+LinearRegression & NA & NA & NA & 3.1543500\\
+NBI-2 & 0.0540725 & 1.206241 & 0.3830272 & NA\\
+NOtr-1 & 0.0098967 & 1.222583 & 0.2675957 & 451.6620471\\
+NOtr-2 & 0.0118522 & 1.208561 & 0.2675957 & 86.5895462\\
+Poisson-1 & 0.0372137 & 1.204920 & 0.0095263 & 1.4763353\\
+Poisson-2 & 0.0373884 & 1.188109 & 0.0082932 & 5.0768588\\
+Poisson-2-I & 0.0523937 & 1.222114 & 0.3770081 & NA\\
+Regression-Poisson & 0.0929416 & 1.293524 & 0.8490258 & 67.1401641\\
+Ttr-2 & 0.0140221 & 1.210108 & 0.3324146 & 956.5532849\\
+fasster & 0.1862190 & 35.051012 & 127.7504023 & NA\\
+prophet & 0.0193799 & 1.447037 & 0.2955460 & 20.6755021\\
+qreg-1 & 0.0643451 & 1.340557 & NA & NA\\
+tbats & 0.0855702 & 1.536080 & 0.4859770 & 273.0558176\\
 \bottomrule
@@
 
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/time-accuracy-1}
+
+}
+
+\caption{Running time ...}\label{fig:time-accuracy}
+\end{figure}
+
 \hypertarget{conclusion}{%

### 2021-11-03 paper/paper.Rmd

@@
 
-# Result and discussion {#result}
+Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function.
+All functions were re-estimated on each iteration.
+ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin.
+This allowed to speed up the computation for these two models.
+The initial estimation of ADAM took approximately one hour and 25 minutes.
+Each step in the experiment took the time shown below.
 

### 2021-11-05 paper/paper.Rmd
Started editing the paper

@@
 author:
-  - name: Author1
+  - name: Bahmna Rostami Tabar
     email: email1@example.com
@@
 
-Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients.
-An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
-This is one of the best ways to optimize staffing utilization and related costs.
-Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
+An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing.
+This is one of the best ways to optimize resources utilisation and minimise related costs.
+State of the art practice to optimise personnel scheduling follows the general approach originally presented by @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
 An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
-Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing [@rostami2020anticipating].
+Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow [@derlet2002overcrowding].
+Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
 
-There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting].
-daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals.
-In this paper, we only focus on forecasting hourly arrivals.
-In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly, [@chen2011long; @mai2015predicting] daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals.
+
 <!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->
@@
 An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
-This will have many advantages for both patients, staff and the quality of provided services.
+This has many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
@@
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
-Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
+<!-- Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
 Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
 That may also require to adjust the scheduling.
+-->
 
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
-These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[].
+However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
 Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
@@
 Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
-This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features.
+There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive model [REF],  TBATS [] and Artificial Neural Networks [].
 All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals.
 However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
-Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
 The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
-In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
+In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers.
 Moreover, datasets used in some of these studies are relatively small, short length (e.g.
@@
 
-In this paper, we aim at filling these gaps and our contributions to the literature are summarised as following:
+In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:
 
-1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)
+1. We use the family of Generalised Additive Models to forecast short-term hourly hospital admission using  that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)
 
-2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;
+2. We apply advanced dynamic model based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies;
 
-3.
-We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
-Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+3. We produce probabilistic forecasts, in addition to the point ones, quantifying the uncertainty in future hospital admission;
 
-4.
+4. We compare the accuracy of our models with the one from the  well established benchmarks, such as Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+
+5.
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
@@
 
-Our contributions are as following:
-
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with ...
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly A&E arrival forecasting in healthcare. Section \ref{model} starts with ...
 
@@
 
-There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals.
+There are many studies that employ models to forecast the admission and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only focus on studies on hourly ED forecasting.
-Table \@ref(tab:summarylit) summarize studies in hourly forecasting in the emergency department.
+However, given the focus of the paper, we only discuss the hourly ED forecasting.
+Table \@ref(tab:summarylit) summarize the relevant papers.
 
@@
 
-Linear regression, ARIMA, and naive models were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
-Naive model was there for the purpose of comparison.
+**Gijo paper is missing in the table above!**
+
+Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
+Naïve method was there for the purpose of comparison.
 @hertzum2017forecasting study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year.
-In term of hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day while weekends are the quietest days.
-July-August are the month with the highest number of patient arrivals and January and February are the months with the lowest number of arrivals.
-The regression and ARIMA models perform similarly for all forecast interval in modeling patient arrivals.
-In modeling accident and emergency department occupancy, ARIMA outperform regression models.
-However, after all, the models of occupancy were less accurate than those arrivals.
-@hertzum2017forecasting mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting.
-Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval.
+In terms of hour of the day, patient arrivals peaked around noon.
+For days of the week, Monday is the busiest day while weekends are the quietest ones.
+July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
+According to @hertzum2017forecasting, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
+However, after all, the models of occupancy were less accurate than those of arrivals.
+@hertzum2017forecasting mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
+<!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval.
 Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.
+-->
 
-Predicting the arrivals of an ED future arrivals is also studied by @choudhury2020forecasting.
-ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
+@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
 ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy.
-@hertzum2017forecasting work was mentioned in this paper.
-This result is surprising given the existence of multiple seasonality in the hourly dataset.
+Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
+<!-- @hertzum2017forecasting work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset. -->
 

### 2021-11-08 paper/paper.Rmd
Ivan has done his edits. New structure, some mistakes fixed and some additional explanations.

@@
 
-@morzuch2006forecasting used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours.
+@morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modeled as stochastic.
+Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
 The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
@@
 
-@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
-Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-Authors .
-They present the prediction interval accuracy of $50\%$ and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions.
-They show that the most important predictor is hour of the day and autocorrelation lag 1.
-However, these findings are limited to the short number of observations (only one year of historical data).
+@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
+Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
+The authors presented the prediction intervals with $50\%$ and $90\%$ confidence levels for the number of hourly arrivals under the Poisson assumption.
+They showed that the most important predictor was hour of the day and AR(1).
+<!-- !!!! IS: I don't understand what is meant by the next sentence !!!! -->
+<!-- However, these findings are limited to the short number of observations (only one year of historical data). -->
 
-@schweigler2009forecasting conducted an investigation on whether using time series methods could accurately generate short-term forecasts of ED bed occupancy.
+@schweigler2009forecasting conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
 A year-long dataset of hourly ED bed occupancy was collected from three facilities.
 For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
-In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).
+The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).
 
 @kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
-The study adopted historical mean linear regression as benchmark, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method), able to adjust changes in variance over time and vector autoregressive (VAR) method, able to incorporate data from different sources, to forecast for 4 hours, 24 hours.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method) and vector autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
 They used MAPE to report the forecast accuracy.
-Each of the forecasting models outperformed the benchmark model. In particular, ARIMA model performed best.
+ARIMA model outperformed all the other models.
 
-@gijo2016sarima generated a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.
-Since historical data showed seasonality, SARIMA models were investigated.
-Regarding the daily model, the authors generated a SARIMA model, which, however, resulted in the forecast error (MMSE) that significantly increased when the lead time exceeded 8 days.
-On the other hand, the SARIMA model proposed to forecast the log-calls an hourly basis.
-This model was found to fit well the model both for shorter and longer lead times.
+@gijo2016sarima applied a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.
+Since historical data showed seasonality, SARIMA models were used.
+Regarding the daily model, the authors used a SARIMA model, which, however, resulted in the forecast error that was significantly higher when the lead time exceeded 8 days.
+On the other hand, the SARIMA model was found to fit well the data both for shorter and longer lead times.
 
-@asheim2019real developed a Poisson time-series regression model with continuous weekly and yearly cyclic effects to implement a real-time system that could forecast ED arrivals on 1,2,3 hours horizon.
-Once measured the accuracy using the MAPE metric, it was noticed that great improvement happened when time of notification was incorporated into the model, especially on a one-hour horizon.
-Therefore, time of patient notification must be available for this model to be successful.
+@asheim2019real developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
+Measuring the accuracy using the MAPE, @asheim2019real noticed that great improvement happened when time of notification was incorporated into the model, especially in one-hour horizon.
 
-According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
-Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting].
+According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
+Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily [@boyle2012predicting; @hertzum2017forecasting].
 However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
 
-There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
-These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.
-there is no benchmark method nor is forecast accuracy reported.
+There are few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches. These limitations are summarised as follows:
+
+(i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
+(ii) Almost all research studies produce point forecasts and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
+(iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the helpf of the authors of those papers;
+(iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
+(v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method or forecast accuracy is not reported.
 
@@
 
-## data {#data}
+## Preliminary analysis {#data}
 
-Data used in the study comprised counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
+Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
 We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
 
-Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.
-
-It is clear that the number of arrivals has a sub-daily structure.
-The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again.
-It is also clear that ED service gets systematically more visits on Monday between 8 a.m.
-and 5 p.m.
-Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.
+Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.
 
@@
 
-Figure \@ref(fig:hourly-plot) also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events.
+It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by @hertzum2017forecasting.
+The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again.
+It is also clear that ED service gets systematically more visits on Mondays between 8 a.m.
+and 5 p.m.
+Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.
+
+Figure \@ref(fig:hourly-plot-ridge) also highlights that there are many ouliers for almost every hour of the day that may affect the accuracy of forecasting methods.
+Some of these outliers might be related to holidays and special events.
 It is also clear that arrivals are less volatile between mid-night and early morning.
 
-We propose a forecasting model that accounts for the systematic structure of the data.
+**IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.**
 
-```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"}
+Based on this simple analysis, we should consider model that could take the following into account:
 
-```
+- Hour of day, day of week, and week of year seasonal patterns,
+- Calendar events, such as holidays,
+- Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
+- Potentially changing level of data,
+- Other events, such as football matches in the area.
 
-## Benchmarks {#benchmarks}
+We propose several forecasting models that account for the structures outlined above.
 
-### Naive/Climatology
+<!-- {r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"} -->
 
-Empirical distribution for same hour-of-day and day-of-week as target time.
+# Model building {#model}
+<!-- ## Benchmarks {#benchmarks} -->
 
-### Multiple Regression
+## Naive/Climatology {#climatology}
+We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
+This is called "Naïve".
+In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of demand is used to forecast the future demand distribution.
+**We need references here...**
 
+<!-- Empirical distribution for same hour-of-day and day-of-week as target time. -->
+
+## Multiple Regression
 Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A&E admittance).
@@
 \begin{equation}
-  \mathbf{y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
+  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
 (\#eq:linearRegression)
@@
 
-where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters and $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
+where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index.
 However, in the context of healthcare and A&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
 So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression, which can be summarised as:
+One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as:
 
 \begin{equation}
-  \mathbf{y}_t \sim \mathrm{Poisson} \left( \exp \left( \mathbf{x}_t' \boldsymbol{\beta} \right) \right).
+  \begin{aligned}
+    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
+    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
+  \end{aligned}
 (\#eq:PoissonRegression)
@@
 
-The exponent in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function.
-When it comes to selecting explanatory variables for the model, there is no one correct answer and the decision needs to be done based on each specific case.
+The logarithm in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function [reference needed].
+When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case.
 In our experiment, we will only include dummy variables, capturing a variety of calendar events:
@@
 This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time.
-We use `alm()` function from `greybox` v0.7.0 package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
+We use `alm()` function from `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
 
-### ETS - Exponential Smoothing model
-
+## ETS - Exponential Smoothing model
 @Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus *ETS*).
-Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
+Given the popularity of ETS model in forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
 This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as *ETS*.
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
+This model is included as a benchmark, only to see how the other models perform in comparison with it.
 
-### Prophet
-
+## Prophet
 Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects.
@@
 Prophet works well on daily data seen in Facebook.
-It is popular and automated, making it easy to learn for beginners.
+It is robust and automated, making it easy to learn for beginners.
 The implementation may be less flexible than other methods.
+The model itself relies on Multiple Source of Error state space model, originally proposed by @Kalman1960.
 The model is incorporated using corresponding implementation of the Fable package in R.
-We use the `prophet()` function in the fable package to generate hourly forecasts [@fable2020].
-Note that but the input data is assigned with an hourly and daily seasonality.
+We use the `prophet()` function from the `fable` package [@fable2020].
+Note that the input data is assigned with an hourly and daily seasonality.
 
-### TBATS
-
-@de2011forecasting proposed a model to deals with time series exhibiting multiple complex seasonalities.
-TBATS includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonality terms.
-The later one not only gives the model more flexibility to deal with complex seasonality but also reduces the parameters of model when the frequencies of seaonalities are high.
-We fit a TBATS model to our daily time series using the `tbats()` function in the forecast package of R [@forecastpackage2020].
-
-# Model building {#model}
+## TBATS
+@de2011forecasting proposed a model to deal with time series exhibiting multiple complex seasonalities called "TBATS".
+It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
+The latter one not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
+We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].
 
 ## ADAM: multiple seasonal iETSX
-
 @SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model.
-This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
+This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
 Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-Given that the data exhibits zeroes, we use the direct probability of ETS(M,N,N) model developed by @Svetunkov2019a to treat those values.
+This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
+Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by @Svetunkov2019a to treat those values.
 This model can be formulated as a set of the following equations:
@@
 \end{equation}
-
-where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
+where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
 Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$.
-We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
+@Svetunkov2019a discuss how to estimate such model.
+We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components).
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them.
-This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".
+If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantiles_ceiling) for the explanation).
+This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".
 
 ## GAMLSS
-
 If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters may be modelled as additive functions of explanatory variables.
 This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
 In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
-Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
+
+Let $F_t(y_t)$ be a predictive cumulative probability distribution of $y_t$.
 In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+In a GAMLSS framework of @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via:
 
-If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+<!-- If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
 Generalised additive models for location, scale and shape (*GAMLSS*) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
@@
 In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
+-->
 
-Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
+<!-- Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
 In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
+In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via -->
 
@@
 
-where $g_j$ is a monotonic `link` function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$.
-If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$.
+where $g_j$ is a monotonic `link` function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**.
+If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third elements of $\bm x_t$.
 Each $f_{j,i}$ is a smooth function, constructed as
@@
 
-As our data are counts, the natural starting point is the Poisson distribution, given by
+As our data are counts, the natural starting point is the Poisson distribution with an additive model for $\log \lambda_t$ of the form:
 
-\begin{equation}
-  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
-  (\#eq:poissonreg)
-\end{equation}
+<!-- !!!! IS: I'm not sure that we need CDFs and PMFs, because they are well known. !!!! -->
 
-where we consider an additive model for $\lambda_t$ of the form
+<!-- \begin{equation} -->
+<!--   F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor} -->
+<!--   (\#eq:poissonreg) -->
+<!-- \end{equation} -->
+
+<!-- where we consider an additive model for $\lambda_t$ of the form -->
 
@@
 
-\begin{equation}
-  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
- (\#eq:truncatedn)
-\end{equation}
+<!-- \begin{equation} -->
+<!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} -->
+<!--  (\#eq:truncatedn) -->
+<!-- \end{equation} -->
 
@@
 
-## Forecast performance evaluation {#accuracy}
+# Forecast performance evaluation {#accuracy}
 
 In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model.
+We focus on 48 hours ahead, because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins.
@@
 \begin{equation}
-  \mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{j=1}^n e_j^2} .
+  \mathrm{RMSE} = \sqrt{\frac{1}{h} \sum_{j=1}^h e_{t+j}^2} ,
   (\#eq:RMSE)
 \end{equation}
+where $h$ is the forecast horizon and $e_{t+j}$ is the point forecast error $j$ steps ahead.
 
-The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a].
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.
+The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a].
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
+Calibration, also called *reliability*, is the property that forecast probabilities match the observed frequency of realisations.
+If a forecast is calibrated, then, for example, $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
+This property is necessary for forecast probabilities to be used in quantitative decision-making.
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbb{1}(y_{t}\leq q_{\alpha,t})$
+<!-- IS: Check the \mathbb thingy - it didn't work out in pdf -->
+). We use several scores to assess the quantile performance of models.
 
-The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score.
+First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score [we need reference here].
 It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
-Furthermore,
-The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
+Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model.
+The Pinball Score is given by
 
@@
  \left(q_{\alpha,t} - y_{t} \right)
- \left(\mathbb{1}(y_{t}\leq_{\alpha,t})-\alpha \right)
+ \left(\mathbb{1}(y_{t}\leq q_{\alpha,t})-\alpha \right) ,
  (\#eq:pinball)
 \end{equation}
+where $\mathcal{A} = \{0.05,0.1,...,0.95\}$ is the set of quantiles being estimated.
 
-where $\mathcal{A} = {0.05,0.1,...,0.95}$ is the set of quantiles being estimated.
-
-Calibration, also called `reliability`, is the property that forecast probabilities match the observed frequency of realisations.
-If a forecast is calibrated, then $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
-This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observer frequency mean($\mathbb{1}(y_{t}\le q_{\alpha,t})$).
-
-To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using
+To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
 
 \begin{equation}
-     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} (\#eq:skillscore)
+     <!-- \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} (\#eq:skillscore) -->
+     \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
 \end{equation}
-
-where $M$ is the metric's value for the method being considered, $M_\text{ref}$ is the metric's value for a reference method, and $M_\text{perf}$ is the metrics value for   the `perfect' method, which is zero in the case of RMSE and Pinball.
+where $M$ is the metric's value for the method being considered, $M_\mathrm{ref}$ is the metric's value for a reference method.
+The skill score showas by how many percent the reference approach is worse than the one under consideration.
+<!-- and $M_\mathrm{perf}$ is the metrics value for the `perfect' method, which is zero in the case of RMSE and Pinball. -->
 We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.
 positive or negative skill) are significantly different from zero [@Efron1981Bootstrap].
-Here we use the best performing simple benchmark, Benchmark-2, as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlation of the underlying data.
+Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlations of the underlying data [we need a reference here].
 
@@
 
-## Case study
-
-The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
-Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
-
-(Summary of data, e.g. discussion of observations from training data/model development: long-term trend)
+# Results
+The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
+The rolling origin is done every 12 hours, and the forecast horizon is set for 48 steps ahead.
 
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \@ref(quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
+Calibration is evaluated visually in \@ref(fig:quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
+In terms of practicalities, this means that if staffing decisions are made using these benchmarks, we would be underestimating the demand and the hospital wil be overrun by patients and have shortage of nurses and doctors.
 
-```{r Pinball, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
+<!-- IS: Use different pch for the points and different styles of lines -->
+```{r Pinball, fig.cap= "Pinball loss values over different quantiles.", fig.align='center',out.width= "90%"}
 ```
 
-```{r lead-time-rel, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
+<!-- IS: This plot is difficult to read. Do we need it at all? -->
+```{r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"}
 ```
 
-```{r Reliability, fig.cap= "Reliability ...", fig.align='center',out.width= "90%"}
+<!-- IS: Other pch values, to make it more readable -->
+```{r Reliability, fig.cap= "Reliability score, showing the relation between empirical and theoretical quantiles.", fig.align='center',out.width= "90%"}
 ```
 
-```{r quantile-bias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
+<!-- IS: Other pch values, to make it more readable -->
+```{r quantile-bias, fig.cap= "Quantile bias vs the nominal quantiles.", fig.align='center',out.width= "90%"}
 ```
@@
 
-Evaluation metrics from the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.
+Evaluation metrics from the test period are presented in Table \@ref(tab:tab-results).
+They are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
 
+<!-- IS: Round values in the table to 3 or 4 digits -->
 ```{r Skill-rel2bench-reduced, fig.cap= "Skill score ...", fig.align='center',out.width= "70%"}
@@
 
+<!-- IS: Different styles of lines are needed here. -->
 ```{r lead-time-rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
@@
 
+<!-- IS: This is not readable in black & white -->
 ```{r time-accuracy, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
@@
 
+One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
+This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
+
 # Conclusion {#conclusion}
 
+# Appendices
+## Quantiles of rounded up random variables {#quantiles_ceiling}
+
+Before proceeding with the proof we need to give the definition of the quantiles of the continuous and rounded up random variables:
+\begin{equation} \label{eq:quantCeil1}
+	P \left(y_t < k \right) = 1 - \alpha ,
+\end{equation}
+and
+\begin{equation} \label{eq:quantCeil2}
+	P \left(\lceil y_t \rceil \leq n \right) \geq 1 - \alpha ,
+\end{equation}
+where $n$ is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and $k$ is the quantile of the continuous distribution of the variable.
+
+In order to prove that $n = \lceil k \rceil$, we need to use the following basic property:
+\begin{equation} \label{eq:quantCeil3}
+	\lceil y_t \rceil \leq n \iff  y_t \leq n,
+\end{equation}
+which means that the rounded up value will always be less than or equal to $n$ if and only if the original value is less than or equal to $n$.
+Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
+\begin{equation} \label{eq:quantCeil4}
+	P \left(y_t \leq n \right) \geq 1 - \alpha .
+\end{equation}
+Note also that the following is true:
+\begin{equation} \label{eq:quantCeil5}
+	P \left(\lceil y_t \rceil \leq n-1 \right) = P \left(y_t \leq n-1 \right) < 1 - \alpha .
+\end{equation}
+Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2}, \eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the following can be summarised:
+\begin{equation} \label{eq:quantCeil6}
+	P \left(y_t \leq n-1 \right) < P \left(y_t < k \right) \leq P \left(y_t \leq n \right) ,
+\end{equation}
+which is possible only when $k \in (n-1, n] $, which means that $\lceil k \rceil = n$.
+So the rounded up quantile of continuous random variable $y_t$ will always be equal to the quantile of the descritised value of $y_t$:
+\begin{equation} \label{eq:ceilingAndQuantiles1}
+	\left \lceil Q_\alpha(y_t) \right \rceil = Q_\alpha \left(\lceil y_t \rceil \right) .
+\end{equation}
+
+It is also worth noting that the same results can be obtained with the floor function instead of ceiling, following the same logic. So the following equation will hold for all $y_t$ as well:
+\begin{equation} \label{eq:floorAndQuantiles1}
+	\left \lfloor Q_\alpha(y_t) \right \rfloor = Q_\alpha \left(\lfloor y_t \rfloor \right) .
+\end{equation}
+
 # References {#references .unnumbered}

### 2021-11-09 paper/paper.Rmd

--- 2021-11-03 paper/paper.Rmd
@@
 ---
-title:  Forecasting short-term hourly Emergency Departement arrivals
+title:  Forecasting hourly hospital admission
 author:
@@
 library(knitr)
-knitr::opts_chunk$set(echo = FALSE, cache=FALSE, fig.pos = 'H', message = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.pos = 'H', message = FALSE, warning = FALSE)
 # Make sure you have the latest versions of rmarkdown and rticle
@@
 library(ggplot2)
+library(data.table)
 library(RColorBrewer)
@@
 
-Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients.
-An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
+Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients.
+An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
 This is one of the best ways to optimize staffing utilization and related costs.
 Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
-Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing [@rostami2020anticipating].
+An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives [@rostami2020anticipating].
 
-There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting].
-daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals.
-In this paper, we only focus on forecasting hourly arrivals.
-In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->
+Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding].
+Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.
 
-An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting].
+daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] admission.
+In this paper, we only focus on forecasting hourly hospital admission.
+In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This will have many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process in ED.
-<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
-The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
+This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.
+The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
@@
 
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
 These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[].
@@
 This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features.
-All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals.
-However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
+All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission.
+Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
 Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
@@
 The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
-In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
-Moreover, datasets used in some of these studies are relatively small, short length (e.g.
+In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+Aditionally, datasets used in some of these studies are relatively small, short length (e.g.
 time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Moreover, all previous publications  covered in this paper fails reproducibility principles.
+Moreover, all previous publications  covered in this paper fails reproducibility principles (Boylan, 2016).
 
@@
 
-1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)
+1.
+We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
 
-2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;
+2.
+We develop a novel approach to forecast day-ahead hourly hospital admission using Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of admissions, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
 
@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.
 
@@
 
-# Research background: hourly ED forecasting {#lit}
+# Research background: hourly hospital admission forecasting {#lit}
 
-There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals.
-The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only focus on studies on hourly ED forecasting.
-Table \@ref(tab:summarylit) summarize studies in hourly forecasting in the emergency department.
+Other papers to review and possibly cite:
+- Daily visits using Prophet @park2019144
+- Daily using internet data @ekstrom2015forecasting
+- Review paper: @wargon2009systematic
+- Others: @boyle2012predicting, 
+- Impact of weather @parsons2011modelling, @macgregor2003effect
 
-```{r summarylit, cache=FALSE}
+hourly:
+@cheng2021forecasting
+@rocha2021forecasting
+
+Table \@ref(tab:summarylit) summarise studies in hourly forecasting in emergency and urgent care.
+
+```{r summarylit}
 li <- readxl::read_xlsx("../table/table.xlsx")
-knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly Emergency Department forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
-  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") %>%  kableExtra::landscape()
+knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
+  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") 
 ```
@@
 
-Predicting the arrivals of an ED future arrivals is also studied by @choudhury2020forecasting.
-ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
-ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy.
+Predicting the arrivals of accident and emergency department future patients is studied by @choudhury2020forecasting ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
+ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy.
 @hertzum2017forecasting work was mentioned in this paper.
-This result is surprising given the existence of multiple seasonality in the hourly dataset.
-
-@morzuch2006forecasting used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+It is said that residual normality, stationarity, and autocorrelation have not been tested in @hertzum2017forecasting paper and this might be the cause of accuracy problems.
+However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in @choudhury2020forecasting
+@morzuch2006forecasting use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
 The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
 
-@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
+@mccarthy2008challenge predict patient demand for ED services by characterising ED arrivals.
 Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-Authors .
-They present the prediction interval accuracy of $50\%$ and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions.
+Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
+Then, they presente the predictiona interval accuracy of 50 and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions.
 They show that the most important predictor is hour of the day and autocorrelation lag 1.
-However, these findings are limited to the short number of observations (only one year of historical data).
+However, these findings are limited to the short period of data (one year).
 
@@
 
+@steins2019forecasting aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls.
+Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic.
+In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models.
+These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE.
+Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day.
+Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model.
+However, the improvement provided by the more advance model was not much greater than the one provided by the existing model.
+Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors.
+Moreover, both models either underestimated or overestimated the number of calls.
+Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend.
+Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.
+
 According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
 Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting].
-However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
+However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
 
 There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
-These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.
+These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.
 there is no benchmark method nor is forecast accuracy reported.
@@
 
-## data {#data}
+We develop a range of models for forecasting hospital admissions using a multi-year dataset from a hospital in the UK.
+In order to facilitate model development and testing, we divide the data into a training and test period.
+We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data.
+Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
 
-Data used in the study comprised counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
-We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
+Forecast are issued every day at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis.
+In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,...
+0.95, as well as the predictive expectation.
 
-Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.
+In the remainder of this section we explore the data in more detail, describe the forecast evaluation metrics we will employ, and introduce a range of benchmark models for producing forecasts.
+The benchmarks serve as reference methods and include some simple methods as well as more sophisticated routines designed to work `out-of-the-box', such as Facebook's Prophet.
+The performance of these methods provides an important reference, and any proposed novel solution should offer material benefits to justify increased complexity.
 
-It is clear that the number of arrivals has a sub-daily structure.
-The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again.
-It is also clear that ED service gets systematically more visits on Monday between 8 a.m.
-and 5 p.m.
-Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.
+## Data exploration {#data}
+
+The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergency at hourly granularity.
+The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.
+
+Figure \@ref(fig:hourly-plot-ridge)
 
@@
 
-Figure \@ref(fig:hourly-plot) also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events.
-It is also clear that arrivals are less volatile between mid-night and early morning.
+Figure \@ref(fig:hourly-plot)
 
-We propose a forecasting model that accounts for the systematic structure of the data.
-
-```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"}
+```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "80%"}
 
@@
 We use `alm()` function from `greybox` v0.7.0 package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
+
+Note that we have also tried the Normal distribution in our experiments, but as expected it was not suitable and performed poorly.
+This is because it is symmetric around the mean, while the data exhibits skewness.
 
@@
 \end{equation}
+<!-- where we initially consider an additive model for $\lambda_t$ of the form -->
+<!-- \begin{equation} -->
+<!--   \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
+<!--  (\#eq:additivemodel) -->
+<!-- \end{equation} -->
+<!-- where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$.
+Analysis of this model fit on the training data revealed systematic biases which were identified as relating to public and school holidays.
+We therefore expand the model to include these effects as follows, -->
+<!-- \begin{equation} -->
+<!--   \log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
+<!--  (\#eq:additivemodel_v2) -->
+<!-- \end{equation} -->
+<!-- where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays. -->
+where we consider an additive model for $\lambda_t$ of the form
+\begin{equation}
+\log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
+(\#eq:additivemodel_v2)
+\end{equation}
+where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, $S_k(t)$ is a dummy variable for School Holidays, and $C_t$ is the temperature at time $t$.
+Public and school holidays were included following experiments on the training data which revealed systematic biases associated with these periods.
 
-where we consider an additive model for $\lambda_t$ of the form
-
-\begin{equation}
-  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
- (\#eq:additivemodel)
-\end{equation}
-
-The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
-
-Truncated Normal...
-
+However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
+In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
+Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
+The resulting density forecasts are given by
 \begin{equation}
@@
 \end{equation}
+with additive models
+\begin{align*}
+\mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
+(\#eq:additivemodel_NOtr_v2)
+\end{align*}
+for the mean and variance parameters.
+We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
 
-with
-
-\begin{align*}
-  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-  log(\sigma_t) & = f(H(t)) .
-\end{align*}
-
-Truncated $t$ distribution...
-
-Negative binomial...
+# Result and discussion {#result}
 
@@
 
-Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function.
-All functions were re-estimated on each iteration.
-ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin.
-This allowed to speed up the computation for these two models.
-The initial estimation of ADAM took approximately one hour and 25 minutes.
-Each step in the experiment took the time shown below.
+## Model fits
 
-## Case study
-
-The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
-Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
+(Plots generated with theme_few - they look nice, good suggestion!
+Not sure how to include them in bookdown, can you help, Bahman?
+The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format?
+I guess PDF might be preferred.
+I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
 
 (Summary of data, e.g. discussion of observations from training data/model development: long-term trend)
+
+### Truncated Normal GAMLSS Model
+
+We can support the interpretation the models given in Equation \@ref(eq:additivemodel_NOtr_v2) after the parameters have been estimated by visualising their reposes to inputs.
+Figure \@ref(fig-NOtr-model) shows the effect of time of day and day-type on the level and spread of predicted arrivals.
+
+```{r fig-NOtr-model}
+
+load("NOtr2_model.Rda")
+
+plot_data <- data.frame(Hour = 0:23)
+for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
+  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "mu",fixed.at = list(dow3=i))
+  plot_data[[i]] <- pef(0:23)
+}
+
+plot_data <- reshape::melt.data.frame(plot_data,id.vars="Hour",variable_name="Day Type")
+
+plot_1 <- ggplot(plot_data,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
+  ylab("Effect [Expected Arrivals]") + theme_few()
+
+# plot_1
+
+plot_data2 <- data.frame(Hour = 0:23)
+for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
+  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "sigma",fixed.at = list(dow3=i))
+  plot_data2[[i]] <- pef(0:23)
+}
+
+plot_data2 <- reshape::melt.data.frame(plot_data2,id.vars="Hour",variable_name="Day Type")
+
+plot_2 <- ggplot(plot_data2,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
+  ylab("Effect [Standard Deviation of Arrivals]") + theme_few()
+
+#plot_2
+
+grid.arrange(plot_1, plot_2, ncol=2)
+
+```
 
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \@ref(quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
-Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
+Calibration is evaluated visually in \@ref(Reliability), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal GAMLSS models (NOtr-2) and ADAM-iETS models showing good calibration across all probability levels.
+Notably, benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
 
-```{r Pinball, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
+```{r Reliability, fig.cap= "Reliability diagram for selected methods.
+If forecasts are calibrated, the Empirical frequency that each probability level is deceded should match the Nominal frequency, and the reliability diagram should be a straight line on the diagonal (black dashed line).", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-```{r lead-time-rel, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
-```
-
-```{r Reliability, fig.cap= "Reliability ...", fig.align='center',out.width= "90%"}
-```
-
-```{r quantile-bias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
-```
+Evaluation metrics from all Methods on the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above.
 
@@
 
-Evaluation metrics from the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.
+The Pinball Score evaluates overall performance, penalising both poor calibration and lack of sharpness, and ranks the various models quite differently.
+Pinball is visualised in Figure \'ref(Pinball) broken down by quantile.
+NBI-2 and Poisson-2 achieve the lowest Pinball Loss, but are poorly calibrated so not suitable for use in decision making.
+Of the well calibrated methods, NOtr-2 and Ttr-2 have the lowest Pinball Score, and provide a significant improvement over Benchmark 2 by this metric.
+However, we note that while the improvement in terms of Pinball Score appears small, these methods have substantially improved forecast calibration compared to the benchmarks, which are poorly calibrated.
 
-```{r Skill-rel2bench-reduced, fig.cap= "Skill score ...", fig.align='center',out.width= "70%"}
-
+```{r Pinball, fig.cap= "Pinball loss for selected forecsating methods and quantiles from 5% to 95%.
+Several methods perform similarly well relative to others by this metric but calibration should also be considered when evaluating probabilistic foreasts.", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
+In order to verify that this improvement is a result of improved modelling and not a result of sample variation, we have performed block bootstrap resampling on Pinball skill scores relative to Bechmark-2 for each forecast.
+The results for all methods with a skill score greater than -2% are shown in Figure \@ref(Skill_rel2bench_reduced), where boxes are coloured according to calibration.
+Where boxes do not overlap we can be confident that apparent differences in skill are not the result of sample variation.
 
+```{r Skill_rel2bench_reduced, fig.cap= "Bootstrapped skill scores for top-performing methods relative to Benchmark-2.
+Non-overlapping boxes indicate that we can be confident that apparent differences in Pinball Score are not the result of sample variation.", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-```{r lead-time-rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
-
+```{r Pinaball_vs_Qbias, fig.cap= "Pinball Score vs average Qunatile Bias for selected methods.
+While some methods perform well by one measure or the other, only NOtr-2 is both calibrated and has a low Pinball Score.", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
+We delve deeper into forecast performance by evaluating forecasts by time of day, and lead-time.
+The Pinball Score for slected methods is plotted against lead-time in Figure \@ref(lead-time-pb), separated by whether the forecast was issued at midnight (00) or mid-day (12).
+We can see clearly that time of day is a greater influence on performance than lead-time, with morning arrivals and the evening peak in arrivals proving the most difficult to predict.
+
+```{r lead-time-pb, fig.cap= "Pinball Score by lead-time for forecasts issued at midnight (00) and mid-day (12).", fig.align='center',fig.width= 7,fig.height=6}
 ```
 
-```{r time-accuracy, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
-```
+Computational time should also be considered, as national public health bodies may be required to produce forecasts of many variables for a large number of hospitals on  regular basis...
 

### 2021-11-09 paper/paper.Rmd
Minor changes
Comments for the figures

--- 2021-11-08 paper/paper.Rmd
@@
 
-# Experimental design {#design}
-
-## Preliminary analysis {#data}
+# Preliminary analysis {#data}
 
@@
 where $M$ is the metric's value for the method being considered, $M_\mathrm{ref}$ is the metric's value for a reference method.
-The skill score showas by how many percent the reference approach is worse than the one under consideration.
+The skill score show us by how many percent the reference approach is worse than the one under consideration.
 <!-- and $M_\mathrm{perf}$ is the metrics value for the `perfect' method, which is zero in the case of RMSE and Pinball. -->
@@
 <!-- IS: This plot is difficult to read. Do we need it at all? -->
-```{r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"}
-```
+<!-- {r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"} -->
 
@@
 
+<!-- IS: round values to 3 or 4 decimals -->
 ```{r tab-results, out.width= "50%"}
@@
 
-<!-- IS: Round values in the table to 3 or 4 digits -->
 ```{r Skill-rel2bench-reduced, fig.cap= "Skill score ...", fig.align='center',out.width= "70%"}

### 2021-11-30 paper/paper.tex

--- 2021-11-01 paper/paper.tex
@@
 % Pandoc header
-\usepackage{adjustbox, float,lscape, bm}
+\usepackage{adjustbox, float,lscape, bm,amsmath}
 \usepackage{booktabs}
@@
   \title{Forecasting short-term hourly Emergency Departement arrivals}
-    \author[University1]{Author1\corref{1}}
+    \author[University1]{Bahmna Rostami Tabar\corref{1}}
    \ead{email1@example.com} 
@@
 
-Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients.
-An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services.
-This is one of the best ways to optimize staffing utilization and related costs.
-Most current practice to optimise personnel scheduling follows the general approach originally presented in \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
+An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing.
+This is one of the best ways to optimize resources utilisation and minimise related costs.
+State of the art practice to optimise personnel scheduling follows the general approach originally presented by \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
 An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow \citep{derlet2002overcrowding}.
-Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing \citep{rostami2020anticipating}.
+Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow \citep{derlet2002overcrowding}.
+Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other \citep{rostami2020anticipating}.
 
-There exists a large literature on forecasting ED arrivals that use various methods to forecast annual\citep{tandberg1994time} monthly,\citep{chen2011long, mai2015predicting}.
-daily\citep{rostami2020anticipating, park2019144} and subdaily \citep{schweigler2009forecasting, cheng2021forecasting} arrivals.
-In this paper, we only focus on forecasting hourly arrivals.
-In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-
-An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
-This will have many advantages for both patients, staff and the quality of provided services.
+An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+This has many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
@@
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
-Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during a day or a week.
-Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
-That may also require to adjust the scheduling.
-
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
-These studies use multiple approaches including Exponential Smoothing {[}REF{]}, ARIMA families {[}{]}, ARCH {[}REF{]}, Vector Autoregressive {[}REF{]}, TBATS {[}{]} and ANN{[}{]}.
+However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
 Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
@@
 Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
-This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features.
-All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals.
+
+There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+These studies use multiple approaches including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}.
+There are few limitations in these studies.
+They only generate the estimated future arrivals as a point forecast (a single number), which does not acknowledge any uncertainty around the number of arrivals.
 However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
-Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
-The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
-In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers.
+This is where producing probabilistic forecasts could be beneficial in staffing.
+The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools in practice.
+Probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
+In this paper, in adition to generating point forecast, we also produce and evaluate the forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers.
 Moreover, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Moreover, all previous publications covered in this paper fails reproducibility principles.
+Additionally, most of forecasting methods used in these publications does not consider the multiple seaonality natutre of hourly ED arrivals.
+Moreover, all previous publicationscovered in this paper fails reproducibility principles.
 
-In this paper, we aim at filling these gaps and our contributions to the literature are summarised as following:
+In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:
 
@@
 \item
-  We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv) (Jethro to complete)
+  We use the family of Generalised Additive Models to forecast short-term hourly hospital admission using that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv) (Jethro to complete)
 \item
-  We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using \ldots;
+  We apply advanced dynamic model based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies;
 \item
-  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+  We produce probabilistic forecasts, in addition to the point ones, quantifying the uncertainty in future hospital admission;
+\item
+  We compare the accuracy of our models with the one from the well established benchmarks, such as Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
 \item
@@
 
-Our contributions are as following:
-
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with \ldots{}
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly ED arrival forecasting in healthcare. Section \ref{model} starts with \ldots{}
 
@@
 
-There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals.
+There are many studies that employ models to forecast the admission and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only focus on studies on hourly ED forecasting.
-Table \ref{tab:summarylit} summarize studies in hourly forecasting in the emergency department.
+However, given the focus of the paper, we only discuss the hourly ED forecasting.
+Table \ref{tab:summarylit} summarize the relevant papers.
 
@@
 
-\caption{\label{tab:summarylit}Summary of studies in hourly Emergency Department forecasting}
+\caption{\label{tab:summarylit}Summary of studies in forecasting hourly arrivals in Emergency Department}
 \centering
@@
 \midrule
-McCarthy 
-2008 & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
+Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
 Asheim et al. & 2019 & ED arrivals & 3h & 5 & Poisson regression & MAPE & No & No\\
-Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
-Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
-Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
-Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
 Hertzum & 2017 & ED arrivals & 1,2,4,8,24 hours & 3 & linear regression; SARIMA; Naïve & MAE, MAPE, 
 MASE & No & No\\
-Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
-Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
-Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
+Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
+Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
 Chase et al. & 2012 & ED CUR & 30m
 1h, 2h, 4h, 8h, 12h & 1 & Binary regression & NA & No & No\\
+Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
+Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
+McCarthy & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
+Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
+Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
 \bottomrule
@@
 
-Linear regression, ARIMA, and naive models were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
-Naive model was there for the purpose of comparison.
+Linear regression, ARIMA, and Naïve were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
+Naïve method was there for the purpose of comparison.
 \citet{hertzum2017forecasting} study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year.
-In term of hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day while weekends are the quietest days.
-July-August are the month with the highest number of patient arrivals and January and February are the months with the lowest number of arrivals.
-The regression and ARIMA models perform similarly for all forecast interval in modeling patient arrivals.
-In modeling accident and emergency department occupancy, ARIMA outperform regression models.
-However, after all, the models of occupancy were less accurate than those arrivals.
-\citet{hertzum2017forecasting} mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting.
-Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval.
-Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.
+In terms of hour of the day, patient arrivals peaked around noon.
+For days of the week, Monday is the busiest day while weekends are the quietest ones.
+July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
+According to \citet{hertzum2017forecasting}, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
+However, after all, the models of occupancy were less accurate than those of arrivals.
+\citet{hertzum2017forecasting} mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
 
-Predicting the arrivals of an ED future arrivals is also studied by \citet{choudhury2020forecasting}.
-ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals.
+\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
 ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy.
-\citet{hertzum2017forecasting} work was mentioned in this paper.
-This result is surprising given the existence of multiple seasonality in the hourly dataset.
+Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
 
-\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours.
+\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modeled as stochastic.
+Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
 The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
@@
 
-\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
-Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-Authors .
-They present the prediction interval accuracy of \(50\%\) and \(90\%\) intervals for the number of hourly arrivals under the Poisson assumptions.
-They show that the most important predictor is hour of the day and autocorrelation lag 1.
-However, these findings are limited to the short number of observations (only one year of historical data).
+\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
+Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
+The authors presented the prediction intervals with \(50\%\) and \(90\%\) confidence levels for the number of hourly arrivals under the Poisson assumption.
+They showed that the most important predictor was hour of the day and AR(1).
 
-\citet{schweigler2009forecasting} conducted an investigation on whether using time series methods could accurately generate short-term forecasts of ED bed occupancy.
+\citet{schweigler2009forecasting} conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
 A year-long dataset of hourly ED bed occupancy was collected from three facilities.
 For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
-In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by historical average).
+The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by historical average).
 
 \citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
-The study adopted historical mean linear regression as benchmark, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method), able to adjust changes in variance over time and vector autoregressive (VAR) method, able to incorporate data from different sources, to forecast for 4 hours, 24 hours.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method) and vector autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
 They used MAPE to report the forecast accuracy.
-Each of the forecasting models outperformed the benchmark model. In particular, ARIMA model performed best.
+ARIMA model outperformed all the other models.
 
-\citet{gijo2016sarima} generated a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.
-Since historical data showed seasonality, SARIMA models were investigated.
-Regarding the daily model, the authors generated a SARIMA model, which, however, resulted in the forecast error (MMSE) that significantly increased when the lead time exceeded 8 days.
-On the other hand, the SARIMA model proposed to forecast the log-calls an hourly basis.
-This model was found to fit well the model both for shorter and longer lead times.
+\citet{asheim2019real} developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
+Measuring the accuracy using the MAPE, \citet{asheim2019real} noticed that great improvement happened when time of notification was incorporated into the model, especially in one-hour horizon.
 
-\citet{asheim2019real} developed a Poisson time-series regression model with continuous weekly and yearly cyclic effects to implement a real-time system that could forecast ED arrivals on 1,2,3 hours horizon.
-Once measured the accuracy using the MAPE metric, it was noticed that great improvement happened when time of notification was incorporated into the model, especially on a one-hour horizon.
-Therefore, time of patient notification must be available for this model to be successful.
-
-According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
-Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast \citep{boyle2012predicting, hertzum2017forecasting}.
+According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
+Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily \citep{boyle2012predicting, hertzum2017forecasting}.
 However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
 
-There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them.
-These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method nor is forecast accuracy reported.
+There are few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches. These limitations are summarised as follows:
 
-\hypertarget{design}{%
-\section{Experimental design}\label{design}}
+\begin{enumerate}
+\def\labelenumi{(\roman{enumi})}
+\tightlist
+\item
+  Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
+\item
+  Almost all research studies produce point forecasts and consequently report only point forecast accuracy.
+There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
+\item
+  most studies are not reproducible, as it is almost impossible to reapply the approaches without the helpf of the authors of those papers;
+\item
+  studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
+\item
+  some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method or forecast accuracy is not reported.
+\end{enumerate}
 
 \hypertarget{data}{%
-\subsection{data}\label{data}}
+\section{Preliminary analysis}\label{data}}
 
-Data used in the study comprised counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
+Data used in this study comprises counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
 We aggregated the patients' arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
 
-Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.
-
-It is clear that the number of arrivals has a sub-daily structure.
-The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again.
-It is also clear that ED service gets systematically more visits on Monday between 8 a.m.
-and 5 p.m.
-Moreover, The number of arrivals around mid-night is slightly higher for Saturday and Sunday.
+Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.
 
@@
 
-Figure \ref{fig:hourly-plot} also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events.
+It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by \citet{hertzum2017forecasting}.
+The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again.
+It is also clear that ED service gets systematically more visits on Mondays between 8 a.m.
+and 5 p.m.
+Moreover, The number of arrivals around mid-night is slightly higher for Saturday and Sunday.
+
+Figure \ref{fig:hourly-plot-ridge} also highlights that there are many ouliers for almost every hour of the day that may affect the accuracy of forecasting methods.
+Some of these outliers might be related to holidays and special events.
 It is also clear that arrivals are less volatile between mid-night and early morning.
-
-We propose a forecasting model that accounts for the systematic structure of the data.
 
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/hourly-plot-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/seasonplot-dofw-1}
 
@@
 
-\caption{Seasonal plot of ED attendance}\label{fig:hourly-plot}
+\caption{Seasonal plot: day of week arrivals}\label{fig:seasonplot-dofw}
 \end{figure}
 
-\hypertarget{benchmarks}{%
-\subsection{Benchmarks}\label{benchmarks}}
+\textbf{IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.}
 
-\hypertarget{naiveclimatology}{%
-\subsubsection{Naive/Climatology}\label{naiveclimatology}}
+\begin{figure}[H]
 
-Empirical distribution for same hour-of-day and day-of-week as target time.
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/seasonplot-weekofyear-1}
+
+}
+
+\caption{Week of year arrivals}\label{fig:seasonplot-weekofyear}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/date-plot-1}
+
+}
+
+\caption{Daily arrivals}\label{fig:date-plot}
+\end{figure}
+
+Based on this simple analysis, we should consider model that could take the following into account:
+
+\begin{itemize}
+\tightlist
+\item
+  Hour of day, day of week, and week of year seasonal patterns,
+\item
+  Calendar events, such as holidays,
+\item
+  Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
+\item
+  Potentially changing level of data,
+\item
+  Other events, such as football matches in the area.
+\end{itemize}
+
+We propose several forecasting models that account for the structures outlined above.
+
+\hypertarget{model}{%
+\section{Model building}\label{model}}
+
+\hypertarget{climatology}{%
+\subsection{Naive/Climatology}\label{climatology}}
+
+We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
+This is called ``Naïve''.
+In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of demand is used to forecast the future demand distribution.
+\textbf{We need references here\ldots{}}
 
 \hypertarget{multiple-regression}{%
-\subsubsection{Multiple Regression}\label{multiple-regression}}
+\subsection{Multiple Regression}\label{multiple-regression}}
 
-Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A\&E admittance).
+Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
 The classical linear regression model is formulated as:
@@
 \begin{equation}
-  \mathbf{y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
+  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
 \label{eq:linearRegression}
@@
 
-where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters and \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance.
-However, in the context of healthcare and A\&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
+where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters, \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and \(t\) is the time index.
+However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
 So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression, which can be summarised as:
+One of the models that is frequently used in practice is the Poisson regression \citep[see for example,][]{mccarthy2008challenge}, which can be summarised as:
 
 \begin{equation}
-  \mathbf{y}_t \sim \mathrm{Poisson} \left( \exp \left( \mathbf{x}_t' \boldsymbol{\beta} \right) \right).
+  \begin{aligned}
+    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
+    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
+  \end{aligned}
 \label{eq:PoissonRegression}
@@
 
-The exponent in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function.
-When it comes to selecting explanatory variables for the model, there is no one correct answer and the decision needs to be done based on each specific case.
+The logarithm in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive.
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function {[}reference needed{]}.
+When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case.
 In our experiment, we will only include dummy variables, capturing a variety of calendar events:
@@
 This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \eqref{eq:PoissonRegression} introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time.
-We use \texttt{alm()} function from \texttt{greybox} v0.7.0 package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.
+We use \texttt{alm()} function from \texttt{greybox} package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.
 
 \hypertarget{ets---exponential-smoothing-model}{%
-\subsubsection{ETS - Exponential Smoothing model}\label{ets---exponential-smoothing-model}}
+\subsection{ETS - Exponential Smoothing model}\label{ets---exponential-smoothing-model}}
 
 \citet{Hyndman2008b} developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus \emph{ETS}).
-Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
+Given the popularity of ETS model in forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
 This was done using \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denote as \emph{ETS}.
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
+This model is included as a benchmark, only to see how the other models perform in comparison with it.
 
 \hypertarget{prophet}{%
-\subsubsection{Prophet}\label{prophet}}
+\subsection{Prophet}\label{prophet}}
 
@@
 Prophet works well on daily data seen in Facebook.
-It is popular and automated, making it easy to learn for beginners.
+It is robust and automated, making it easy to learn for beginners.
 The implementation may be less flexible than other methods.
+The model itself relies on Multiple Source of Error state space model, originally proposed by \citet{Kalman1960}.
 The model is incorporated using corresponding implementation of the Fable package in R.
-We use the \texttt{prophet()} function in the fable package to generate hourly forecasts \citep{fable2020}.
-Note that but the input data is assigned with an hourly and daily seasonality.
+We use the \texttt{prophet()} function from the \texttt{fable} package \citep{fable2020}.
+Note that the input data is assigned with an hourly and daily seasonality.
 
 \hypertarget{tbats}{%
-\subsubsection{TBATS}\label{tbats}}
+\subsection{TBATS}\label{tbats}}
 
-\citet{de2011forecasting} proposed a model to deals with time series exhibiting multiple complex seasonalities.
-TBATS includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonality terms.
-The later one not only gives the model more flexibility to deal with complex seasonality but also reduces the parameters of model when the frequencies of seaonalities are high.
-We fit a TBATS model to our daily time series using the \texttt{tbats()} function in the forecast package of R \citep{forecastpackage2020}.
-
-\hypertarget{model}{%
-\section{Model building}\label{model}}
+\citet{de2011forecasting} proposed a model to deal with time series exhibiting multiple complex seasonalities called ``TBATS''.
+It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
+The latter one not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
+We fit a TBATS model using the \texttt{tbats()} function from the \texttt{forecast} package in R \citep{forecastpackage2020}.
 
@@
 \citet{SvetunkovAdam2021} proposed a framework for dynamic models called \textbf{ADAM} - Augmented Dynamic Adaptive Model.
-This framework includes ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
+This framework encompasses ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
 Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-Given that the data exhibits zeroes, we use the direct probability of ETS(M,N,N) model developed by \citet{Svetunkov2019a} to treat those values.
+This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
+Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by \citet{Svetunkov2019a} to treat those values.
 This model can be formulated as a set of the following equations:
@@
 \end{equation}
-
-where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
+where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
 Finally, \(a_t\) is an unobservable series, underlying the occurrence part of the model and \((1 + \epsilon_{a,t})\) is an unobservable error term for \(a_t\).
-We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account.
+\citet{Svetunkov2019a} discuss how to estimate such model.
+We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components).
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them.
-This model is implemented in \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM''.
+If integer-valued quantiles are needed, then rounding up can be done for them (see @ref(quantiles\_ceiling) for the explanation).
+This model is implemented in \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM-iETSX''.
 
@@
 If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
-In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
-Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution.
-In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
-In a GAMLSS framework \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
-
-If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised additive models for location, scale and shape (\emph{GAMLSS}) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters may be modelled as additive functions of explanatory variables.
 This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
@@
 
-Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution.
+Let \(F_t(y_t)\) be a predictive cumulative probability distribution of \(y_t\).
 In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
-In a GAMLSS framework \citep{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
+In a GAMLSS framework of \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via:
 
@@
 
-where \(g_j\) is a monotonic \texttt{link} function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\).
-If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third element of \(\bm x_t\).
+where \(g_j\) is a monotonic \texttt{link} function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\) \textbf{is \ldots{}}.
+If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third elements of \(\bm x_t\).
 Each \(f_{j,i}\) is a smooth function, constructed as
@@
 
-As our data are counts, the natural starting point is the Poisson distribution, given by
-
-\begin{equation}
-  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
-  \label{eq:poissonreg}
-\end{equation}
-
-where we consider an additive model for \(\lambda_t\) of the form
+As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form:
 
@@
 Truncated Normal\ldots{}
-
-\begin{equation}
-  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
- \label{eq:truncatedn}
-\end{equation}
 
@@
 \hypertarget{accuracy}{%
-\subsection{Forecast performance evaluation}\label{accuracy}}
+\section{Forecast performance evaluation}\label{accuracy}}
 
 In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model.
+We focus on 48 hours ahead, because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000}, resulting in 727 origins.
@@
 \begin{equation}
-  \mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{j=1}^n e_j^2} .
+  \mathrm{RMSE} = \sqrt{\frac{1}{h} \sum_{j=1}^h e_{t+j}^2} ,
   \label{eq:RMSE}
 \end{equation}
+where \(h\) is the forecast horizon and \(e_{t+j}\) is the point forecast error \(j\) steps ahead.
 
-The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated \citep{Gneiting2007a}.
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.
+The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated \citep{Gneiting2007a}.
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
+Calibration, also called \emph{reliability}, is the property that forecast probabilities match the observed frequency of realisations.
+If a forecast is calibrated, then, for example, \(20\%\) of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
+This property is necessary for forecast probabilities to be used in quantitative decision-making.
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observed frequency mean (\(\mathbb{1}(y_{t}\leq q_{\alpha,t})\)
+). We use several scores to assess the quantile performance of models.
 
-The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score.
+First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score {[}we need reference here{]}.
 It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
-Furthermore,
-The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by
+Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model.
+The Pinball Score is given by
 
@@
  \left(q_{\alpha,t} - y_{t} \right)
- \left(\mathbb{1}(y_{t}\leq_{\alpha,t})-\alpha \right)
+ \left(\mathbb{1}(y_{t}\leq q_{\alpha,t})-\alpha \right) ,
  \label{eq:pinball}
 \end{equation}
+where \(\mathcal{A} = \{0.05,0.1,...,0.95\}\) is the set of quantiles being estimated.
 
-where \(\mathcal{A} = {0.05,0.1,...,0.95}\) is the set of quantiles being estimated.
-
-Calibration, also called \texttt{reliability}, is the property that forecast probabilities match the observed frequency of realisations.
-If a forecast is calibrated, then \(20\%\) of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
-This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observer frequency mean(\(\mathbb{1}(y_{t}\le q_{\alpha,t})\)).
-
-To compare model performance, and the significance of any apparent difference in performance, it is useful to define skill scores. Skill scores may be calculated for any metric using
+To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
 
 \begin{equation}
-     \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} \label{eq:skillscore}
+  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} \label{eq:skillscore}
 \end{equation}
 
-where \(M\) is the metric's value for the method being considered, \(M_\text{ref}\) is the metric's value for a reference method, and \(M_\text{perf}\) is the metrics value for the `perfect' method, which is zero in the case of RMSE and Pinball.
+where \(M\) is the metric's value for the method being considered, \(M_\mathrm{ref}\) is the metric's value for a reference method.
+The skill score show us by how many percent the reference approach is worse than the one under consideration.
 We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}.
-Here we use the best performing simple benchmark, Benchmark-2, as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlation of the underlying data.
+Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlations of the underlying data {[}we need a reference here{]}.
 
-\hypertarget{result}{%
-\section{Result and discussion}\label{result}}
+Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function.
+All functions were re-estimated on each iteration.
+ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin.
+This allowed to speed up the computation for these two models.
+The initial estimation of ADAM took approximately one hour and 25 minutes.
+Each step in the experiment took the time shown below.
 
-(Plots generated with theme\_few - they look nice, good suggestion!
-Not sure how to include them in bookdown, can you help, Bahman?
-The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format?
-I guess PDF might be preferred.
-I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-\hypertarget{case-study}{%
-\subsection{Case study}\label{case-study}}
-
-The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only.
-Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.
-
-(Summary of data, e.g.~discussion of observations from training data/model development: long-term trend)
+The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
+The rolling origin is done every 12 hours, and the forecast horizon is set for 48 steps ahead.
 
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \ref{quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels.
+Calibration is evaluated visually in \ref{fig:quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
+In terms of practicalities, this means that if staffing decisions are made using these benchmarks, we would be underestimating the demand and the hospital wil be overrun by patients and have shortage of nurses and doctors.
 
@@
 
-\caption{pinball ...}\label{fig:Pinball}
-\end{figure}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/lead-time-rel-1}
-
-}
-
-\caption{pinball ...}\label{fig:lead-time-rel}
+\caption{Pinball loss values over different quantiles.}\label{fig:Pinball}
 \end{figure}
@@
 
-\caption{Reliability ...}\label{fig:Reliability}
+\caption{Reliability score, showing the relation between empirical and theoretical quantiles.}\label{fig:Reliability}
 \end{figure}
@@
 
-\caption{Quantile ...}\label{fig:quantile-bias}
+\caption{Quantile bias vs the nominal quantiles.}\label{fig:quantile-bias}
 \end{figure}
@@
 
-Evaluation metrics from the test period are presented in \ref{tab-results} which are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.
+Evaluation metrics from the test period are presented in Table \ref{tab:tab-results}.
+They are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
 
@@
 
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/time-accuracy-1}
+
+}
+
+\caption{Running time vs. forecast accuracy}\label{fig:time-accuracy}
+\end{figure}
+
+One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
+This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
+
 \hypertarget{conclusion}{%
 \section{Conclusion}\label{conclusion}}
+
+\hypertarget{appendices}{%
+\section{Appendices}\label{appendices}}
+
+\hypertarget{quantiles_ceiling}{%
+\subsection{Quantiles of rounded up random variables}\label{quantiles_ceiling}}
+
+Before proceeding with the proof we need to give the definition of the quantiles of the continuous and rounded up random variables:
+
+\begin{equation} \label{eq:quantCeil1}
+    P \left(y_t < k \right) = 1 - \alpha ,
+\end{equation}
+
+and
+
+\begin{equation} \label{eq:quantCeil2}
+    P \left(\lceil y_t \rceil \leq n \right) \geq 1 - \alpha ,
+\end{equation}
+
+where \(n\) is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and \(k\) is the quantile of the continuous distribution of the variable.
+
+In order to prove that \(n = \lceil k \rceil\), we need to use the following basic property:
+
+\begin{equation} \label{eq:quantCeil3}
+    \lceil y_t \rceil \leq n \iff  y_t \leq n,
+\end{equation}
+
+which means that the rounded up value will always be less than or equal to \(n\) if and only if the original value is less than or equal to \(n\).
+Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
+
+\begin{equation} \label{eq:quantCeil4}
+    P \left(y_t \leq n \right) \geq 1 - \alpha .
+\end{equation}
+
+Note also that the following is true:
+
+\begin{equation} \label{eq:quantCeil5}
+    P \left(\lceil y_t \rceil \leq n-1 \right) = P \left(y_t \leq n-1 \right) < 1 - \alpha .
+\end{equation}
+
+Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2}, \eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the following can be summarised:
+
+\begin{equation} \label{eq:quantCeil6}
+    P \left(y_t \leq n-1 \right) < P \left(y_t < k \right) \leq P \left(y_t \leq n \right) ,
+\end{equation}
+
+which is possible only when \(k \in (n-1, n]\), which means that \(\lceil k \rceil = n\).
+So the rounded up quantile of continuous random variable \(y_t\) will always be equal to the quantile of the descritised value of \(y_t\):
+
+\begin{equation} \label{eq:ceilingAndQuantiles1}
+    \left \lceil Q_\alpha(y_t) \right \rceil = Q_\alpha \left(\lceil y_t \rceil \right) .
+\end{equation}
+
+It is also worth noting that the same results can be obtained with the floor function instead of ceiling, following the same logic. So the following equation will hold for all \(y_t\) as well:
+\begin{equation} \label{eq:floorAndQuantiles1}
+    \left \lfloor Q_\alpha(y_t) \right \rfloor = Q_\alpha \left(\lfloor y_t \rfloor \right) .
+\end{equation}
 

### 2021-11-30 paper/paper.Rmd

@@
 header-includes:  
- \usepackage{adjustbox, float,lscape, bm} #use the 'float' package
+ \usepackage{adjustbox, float,lscape, bm,amsmath} #use the 'float' package
 output:
@@
 library(extrafont)
+library(feasts)
 ```
@@
 Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
-
-There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly, [@chen2011long; @mai2015predicting] daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals.
-
 <!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->
 
-An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This has many advantages for both patients, staff and the quality of provided services.
@@
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
+However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
+They may also express nonstationarity and their profile may change over time.
+Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
 <!-- Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
@@
 
-However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
-They may also express nonstationarity and their profile may change over time.
-Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
 There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
-These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive model [REF],  TBATS [] and Artificial Neural Networks [].
-All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals.
+These studies use multiple approaches including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
+There are few limitations in these studies.
+They only generate the estimated future arrivals as a point forecast (a single number), which does not acknowledge any uncertainty around the number of arrivals.
 However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
-The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools.
-Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
-In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers.
+This is where producing probabilistic forecasts could be beneficial in staffing.
+The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools in practice.
+Probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
+In this paper, in adition to generating point forecast, we also produce and evaluate the forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers.
 Moreover, datasets used in some of these studies are relatively small, short length (e.g.
 time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Moreover, all previous publications  covered in this paper fails reproducibility principles.
+Additionally, most of forecasting methods used in these publications does not consider the multiple seaonality natutre of hourly ED arrivals.
+Moreover, all previous publicationscovered in this paper fails reproducibility principles.
 
@@
 
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly A&E arrival forecasting in healthcare. Section \ref{model} starts with ...
+The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly ED arrival forecasting in healthcare. Section \ref{model} starts with ...
 
@@
 li <- readxl::read_xlsx("../table/table.xlsx")
-knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly Emergency Department forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
+knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in forecasting hourly arrivals in Emergency Department") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
   column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") %>%  kableExtra::landscape()
 ```
-
-**Gijo paper is missing in the table above!**
 
@@
 
-@gijo2016sarima applied a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.
-Since historical data showed seasonality, SARIMA models were used.
-Regarding the daily model, the authors used a SARIMA model, which, however, resulted in the forecast error that was significantly higher when the lead time exceeded 8 days.
-On the other hand, the SARIMA model was found to fit well the data both for shorter and longer lead times.
-
 @asheim2019real developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
@@
 
-```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width= 7,fig.height=6}
+```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width= 6.5,fig.height=5.5}
 ```
@@
 
+```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
+```
+
 **IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.**
+
+```{r seasonplot-weekofyear, fig.cap= "Week of year arrivals", fig.align='center',out.width="70%"}
+```
+
+```{r date-plot, fig.cap= "Daily arrivals", fig.align='center',out.width="70%"}
+```
 
@@
 
-<!-- {r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"} -->
-
 # Model building {#model}
-<!-- ## Benchmarks {#benchmarks} -->
 
@@
 ## Multiple Regression
-Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A&E admittance).
+Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
 The classical linear regression model is formulated as:
@@
 where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index.
-However, in the context of healthcare and A&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
+However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
 So the linear regression model should be substituted by some other model.
@@
 ## ETS - Exponential Smoothing model
+
 @Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus *ETS*).
@@
 ## Prophet
+
 Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects.
@@
 ## TBATS
+
 @de2011forecasting proposed a model to deal with time series exhibiting multiple complex seasonalities called "TBATS".
@@
 ## ADAM: multiple seasonal iETSX
+
 @SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model.
@@
 \begin{equation}
-     <!-- \text{Skill} = \frac{M_\text{ref} - M}{M_\text{ref} - M_\text{perf}} (\#eq:skillscore) -->
   \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
 \end{equation}
+
 where $M$ is the metric's value for the method being considered, $M_\mathrm{ref}$ is the metric's value for a reference method.
@@
 <!-- IS: This is not readable in black & white -->
-```{r time-accuracy, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
+```{r time-accuracy, fig.cap= "Running time vs. forecast accuracy", fig.align='center',out.width= "70%"}
 ```
@@
 # Appendices
+
 ## Quantiles of rounded up random variables {#quantiles_ceiling}
@@
 Before proceeding with the proof we need to give the definition of the quantiles of the continuous and rounded up random variables:
+
 \begin{equation} \label{eq:quantCeil1}
@@
 \end{equation}
+
 and
+
 \begin{equation} \label{eq:quantCeil2}
@@
 \end{equation}
+
 where $n$ is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and $k$ is the quantile of the continuous distribution of the variable.
@@
 In order to prove that $n = \lceil k \rceil$, we need to use the following basic property:
+
 \begin{equation} \label{eq:quantCeil3}
@@
 \end{equation}
+
 which means that the rounded up value will always be less than or equal to $n$ if and only if the original value is less than or equal to $n$.
 Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
+
 \begin{equation} \label{eq:quantCeil4}
@@
 \end{equation}
+
 Note also that the following is true:
+
 \begin{equation} \label{eq:quantCeil5}
@@
 \end{equation}
+
 Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2}, \eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the following can be summarised:
+
 \begin{equation} \label{eq:quantCeil6}
@@
 \end{equation}
-which is possible only when $k \in (n-1, n] $, which means that $\lceil k \rceil = n$.
+
+which is possible only when $k \in (n-1, n]$, which means that $\lceil k \rceil = n$.
 So the rounded up quantile of continuous random variable $y_t$ will always be equal to the quantile of the descritised value of $y_t$:
+
 \begin{equation} \label{eq:ceilingAndQuantiles1}

### 2021-12-01 paper/paper.Rmd
Update to introduction

@@
 
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
 These studies use multiple approaches including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
-There are few limitations in these studies.
-They only generate the estimated future arrivals as a point forecast (a single number), which does not acknowledge any uncertainty around the number of arrivals.
-However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
-This is where producing probabilistic forecasts could be beneficial in staffing.
-The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools in practice.
-Probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
-In this paper, in adition to generating point forecast, we also produce and evaluate the forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers.
-Moreover, datasets used in some of these studies are relatively small, short length (e.g.
-time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Additionally, most of forecasting methods used in these publications does not consider the multiple seaonality natutre of hourly ED arrivals.
-Moreover, all previous publicationscovered in this paper fails reproducibility principles.
+However, these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+Reporting uncertainty is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels if staff levels are insufficient.
+Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
+Furthermore, if the impact of under and over staffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under and over staffing to be calculated.
+Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
+Another drawback of existing studies is that the datasets used are relatively small (e.g.
+time period of 1 year or less), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
+Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals.
+Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
 
@@
 
-1. We use the family of Generalised Additive Models to forecast short-term hourly hospital admission using  that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)
+1.
+We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
 
-2. We apply advanced dynamic model based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies;
+2.
+We develop an advanced dynamic model to forecasts ED arrivals based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies which produced highly-accurate point forecasts;
 
-3. We produce probabilistic forecasts, in addition to the point ones, quantifying the uncertainty in future hospital admission;
+3.
+We develop a novel model to produce probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
 
-4. We compare the accuracy of our models with the one from the  well established benchmarks, such as Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+4.
+We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
+Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
 
@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
 
@@
 
-@morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modeled as stochastic.
+@morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
 Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
@@
 There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
-(iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the helpf of the authors of those papers;
+(iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
 (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
@@
 
-Figure \@ref(fig:hourly-plot-ridge) also highlights that there are many ouliers for almost every hour of the day that may affect the accuracy of forecasting methods.
+Figure \@ref(fig:hourly-plot-ridge) also highlights that there are many outliers for almost every hour of the day that may affect the accuracy of forecasting methods.
 Some of these outliers might be related to holidays and special events.
@@
 
-**IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.**
+**IS: We need to plot day of week and week of year seasonal plots here and discuss. This will show multiple seasonality nature of the data.**
 

### 2021-12-09 paper/paper.Rmd
McCarthy discussion and minor correction up to GAMLSS section

@@
 Naïve method was there for the purpose of comparison.
-@hertzum2017forecasting study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year.
+@hertzum2017forecasting study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
 In terms of hour of the day, patient arrivals peaked around noon.
@@
 
-@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
-Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-The authors presented the prediction intervals with $50\%$ and $90\%$ confidence levels for the number of hourly arrivals under the Poisson assumption.
-They showed that the most important predictor was hour of the day and AR(1).
+@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
+The authors produced probabilistic predictions in the form of $50\%$ and $90\%$ prediction intervals for the number of hourly arrivals.
+Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
 <!-- !!!! IS: I don't understand what is meant by the next sentence !!!! -->
@@
 Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
-We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
-
-Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.
+We aggregated the patients’ arrival times to obtain hourly arrivals, which are used in this study.
+Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week.
+Although the data is noisy, it reveals some systematic patterns.
 
@@
 
-Figure \@ref(fig:hourly-plot-ridge) also highlights that there are many outliers for almost every hour of the day that may affect the accuracy of forecasting methods.
-Some of these outliers might be related to holidays and special events.
+Figure \@ref(fig:hourly-plot-ridge) also highlights that there is significant skewness for almost every hour of the day that varies with time-of-day, and which should be accounted for in forecasting methods.
+Some skewness might be related to holidays and special events.
 It is also clear that arrivals are less volatile between mid-night and early morning.
@@
 
-Based on this simple analysis, we should consider model that could take the following into account:
-
-- Hour of day, day of week, and week of year seasonal patterns,
+Based on this simple analysis and literature review, we should consider models that can take the following into account:
+- Daily, weekly, and annual seasonalities,
+- Long-term trend,
 - Calendar events, such as holidays,
 - Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
-- Potentially changing level of data,
-- Other events, such as football matches in the area.
-
+- Other events, such as sporting fixtures,
+- Weather effects, such as temperature and precipitation.
 We propose several forecasting models that account for the structures outlined above.
@@
 
-## Multiple Regression
+## Poisson Regression
 Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
-The classical linear regression model is formulated as:
+The classical linear regression model is formulated as
 
@@
 \end{equation}
-
 where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index.
@@
 So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as:
+One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as
 
@@
     & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
-  \end{aligned}
+  \end{aligned} \quad .
 (\#eq:PoissonRegression)
@@
 
-The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
+The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
 Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g.
@@
 
-## ETS - Exponential Smoothing model
+## ETS - Exponential Smoothing
 
@@
 It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
-The latter one not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
+The latter not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
 We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].
@@
 
-@SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model.
+@SvetunkovAdam2021 proposed a framework for dynamic models called the Augmented Dynamic Adaptive Model (ADAM).
 This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
@@
 Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$.
-@Svetunkov2019a discuss how to estimate such model.
+@Svetunkov2019a discuss how to estimate such a model.
 We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components).
@@
 ## GAMLSS
-If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters may be modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
-In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
+If we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters are modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive model structures can be found.
+In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.
 

### 2022-01-05 paper/paper.Rmd
Update GAMLSS section and other minor edits

@@
 In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework of @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via:
-
+In a GAMLSS framework of @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled as
 <!-- If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
@@
 -->
-
 <!-- Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution.
@@
 In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via -->
-
 \begin{equation}
@@
 \end{equation}
-
-where $g_j$ is a monotonic `link` function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**.
+where $g_j$ is a monotonic link function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**.
 If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third elements of $\bm x_t$.
 Each $f_{j,i}$ is a smooth function, constructed as
-
 \begin{equation}
@@
 \end{equation}
-
 where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
@@
 
-As our data are counts, the natural starting point is the Poisson distribution with an additive model for $\log \lambda_t$ of the form:
-
+As our data are counts, the natural starting point is the Poisson distribution with an additive model for $\log \lambda_t$ of the form
 <!-- !!!! IS: I'm not sure that we need CDFs and PMFs, because they are well known. !!!! -->
@@
 \end{equation}
-
 The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
 
-Truncated Normal...
+However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
+In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
+Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
+The resulting density forecasts are given by
+\begin{equation}
+F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
+(\#eq:truncatedn)
+\end{equation}
+with additive models
+\begin{align*}
+\mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
+log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
+(\#eq:additivemodel_NOtr_v2)
+\end{align*}
+for the mean and variance parameters.
+We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
 
-<!-- \begin{equation} -->
-<!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} -->
-<!--  (\#eq:truncatedn) -->
-<!-- \end{equation} -->
+<!-- Truncated Normal... -->
 
-with
+<!-- <!-- \begin{equation} --> -->
+<!-- <!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} --> -->
+<!-- <!--  (\#eq:truncatedn) --> -->
+<!-- <!-- \end{equation} --> -->
 
-\begin{align*}
-  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-  log(\sigma_t) & = f(H(t)) .
-\end{align*}
+<!-- with -->
 
-Truncated $t$ distribution...
+<!-- \begin{align*} -->
+<!--   \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\ -->
+<!--   log(\sigma_t) & = f(H(t)) . -->
+<!-- \end{align*} -->
 
-Negative binomial...
+<!-- Truncated $t$ distribution... -->
+
+<!-- Negative binomial... -->
 
@@
 
-In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model.
-We focus on 48 hours ahead, because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors.
+In order to assess performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps if 0.05, and conditional expectations for 0 to 48 hours ahead produced by each model.
+We forecast upto 48 hours because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins.
 Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation.
-The latter is measured via Root Mean Squared Error (RMSE):
-
+The latter is measured via Root Mean Squared Error (RMSE)
 \begin{equation}
@@
 A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
-Calibration, also called *reliability*, is the property that forecast probabilities match the observed frequency of realisations.
+Calibration, also called reliability, is the property that forecast probabilities match the observed frequency of realisations.
 If a forecast is calibrated, then, for example, $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
 This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbb{1}(y_{t}\leq q_{\alpha,t})$
-<!-- IS: Check the \mathbb thingy - it didn't work out in pdf -->
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbf{1}(y_{t}\leq q_{\alpha,t})$
+<!-- IS: Check the \mathbb thingy - it didn't work out in pdf // JB: I changed to mathbf, lets see how it looks... -->
 ). We use several scores to assess the quantile performance of models.
@@
  \left(q_{\alpha,t} - y_{t} \right)
- \left(\mathbb{1}(y_{t}\leq q_{\alpha,t})-\alpha \right) ,
+ \left(\mathbf{1}(y_{t}\leq q_{\alpha,t})-\alpha \right) ,
  (\#eq:pinball)
@@
 The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
-The rolling origin is done every 12 hours, and the forecast horizon is set for 48 steps ahead.
+The rolling origin advances in 12 hour steps, and the forecast horizon is set for 48 steps ahead.
 
@@
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-In terms of practicalities, this means that if staffing decisions are made using these benchmarks, we would be underestimating the demand and the hospital wil be overrun by patients and have shortage of nurses and doctors.
+For a user, this could result in poor staffing decisions as historic data fail to accuratly charterise present conditions.
 

### 2022-01-06 paper/paper.Rmd
Draft conclusion

@@
 Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
+@Cheng2021 develop an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which is found to outperform a rolling-average benchmark.
+They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.
 <!-- @hertzum2017forecasting work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset. -->
@@
 
+Short-term forecasting of arrivals at emergency departments is an important element of staff and resource management in hospitals.
+Furthermore, due to the asymmetric impact of having an excess of shortage of resources, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
+Here, we have developed methods for producing probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
+
+Two appraoches produced highly accurate, calibrated probabilistic forecasts, one time series method and one based on distributional regression.
+The first is ADAM-iESTX, which is an extension of exponential smoothing incorporating seasonality and assuming a Gamma predictive distribution.
+The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on features date and time features, and temperature.
+Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX produced forecasts had a lower RMSE than NOtr-2, and NOtr-2 produced forecasts with a lower pinball score.
+This suggests that the latter may be preferred if the whole distribution is used in decision-making.
+
+Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts, and choice of model features, are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without need for post-processing.
+These observations reflect the characteristics of the data, which is representative of A\&E arrivals, but determining the extent to which they generalise is beyond the scope of this article.
+Furthermore, methods based on continuous valued distributions are not adversely affected by the fact that the data are interger-valued.
+Rounding up predictive quantiles to the next integer does not make predictions worse.
+
+Finally, we have found that out-of-the-box models, those which require minimal tuning or manual development, do not perform as well as well-considered statistical methods.
+The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks.
+Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
+ETS produces reasonably well calibrated forecasts, in contast to the benchmarks which were poorly calibrated, and highly accurate point forecasts, however, its probabilistic forecasts were considerably worse than  NOtr-2 in terms of pinball score.
+
+Probabilistic forecasting opens the door to more sophisticated resource management in healthcare settings by providing decision-makers with uncertainty information and enabling quantitative risk management.
+Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecast skill, and may also be necessary to maximise benefits though more holistic decision-making.
+
 # Appendices

### 2022-01-12 paper/paper.tex
a draft from Bahman to pass to Ivan to go through a final round of reading and improving.

@@
 
-  \journal{Which journal? Journal of Service Research/Health Services Research/EJOR/IJF} % Sets Journal name
+  \journal{Annals of Emergency Medicine} % Sets Journal name
 
@@
   \title{Forecasting short-term hourly Emergency Departement arrivals}
-    \author[University1]{Bahmna Rostami Tabar\corref{1}}
-   \ead{email1@example.com} 
-    \author[School of Mathematics and Statistics, University of Glasgow, UK]{Jethro Browell\corref{2}}
+    \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
+   \ead{rostami-tabarb@cardiff.ac.uk} 
+    \author[University of Glasgow]{Jethro Browell\corref{1}}
    \ead{jethro.browell@glasgow.ac.uk} 
-    \author[Centre for Marketing Analytics and Forecasting, Lancaster University, UK]{Ivan Svetunkov\corref{2}}
+    \author[Lancaster University]{Ivan Svetunkov\corref{1}}
    \ead{i.svetunkov@lancaster.ac.uk} 
-      \address[University1]{Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff}
-    \address[University2]{adress2}
-    \address[University3]{adress3}
-      \cortext[1]{Corresponding Author}
-    \cortext[2]{Equal contribution}
+      \address[Cardiff Business School]{Cardiff Business School, Cardiff University, UK}
+    \address[University of Glasgow]{School of Mathematics and Statistics, University of Glasgow, UK}
+    \address[Lancaster University]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK}
+      \cortext[12]{Corresponding Author}
+    \cortext[1]{Equal contribution}
 
   \begin{abstract}
-  The Objective of this work would be to propose a new methodology to forecast short-term hourly forecasting for urgent and emergency care.
+  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
+This can have many advantages for healthcare staff and the quality of care delivered to patients.
+In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
+We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
+Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.
   \end{abstract}
-
+   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
  \end{frontmatter}
@@
 This is one of the best ways to optimize resources utilisation and minimise related costs.
-State of the art practice to optimise personnel scheduling follows the general approach originally presented by \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
+An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
 Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow \citep{derlet2002overcrowding}.
@@
 This involves the short-term decision making related to the execution of the delivery process in ED.
-The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
+The combination of an hourly arrival forecast, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
@@
 
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays.
+There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
 These studies use multiple approaches including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}.
-There are few limitations in these studies.
-They only generate the estimated future arrivals as a point forecast (a single number), which does not acknowledge any uncertainty around the number of arrivals.
-However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient.
-This is where producing probabilistic forecasts could be beneficial in staffing.
-The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools in practice.
-Probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing.
-In this paper, in adition to generating point forecast, we also produce and evaluate the forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers.
-Moreover, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable.
-Additionally, most of forecasting methods used in these publications does not consider the multiple seaonality natutre of hourly ED arrivals.
-Moreover, all previous publicationscovered in this paper fails reproducibility principles.
+However, majority of these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+There are few studies that report uncertainty by presenting prediction intervals, no study generates and evaluates the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels, if staff levels are insufficient.
+Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
+Furthermore, if the impact of under and over staffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under and over staffing to be calculated.
+Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
+Another drawback of existing studies is that the datasets used are relatively small (e.g.~time period of 1-2 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
+Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals.
+Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
 
@@
 \item
-  We use the family of Generalised Additive Models to forecast short-term hourly hospital admission using that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv) (Jethro to complete)
+  We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
 \item
-  We apply advanced dynamic model based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies;
+  We develop an advanced dynamic model to forecasts ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies which produced highly-accurate point forecasts;
 \item
-  We produce probabilistic forecasts, in addition to the point ones, quantifying the uncertainty in future hospital admission;
+  We develop a novel model to produce probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
 \item
-  We compare the accuracy of our models with the one from the well established benchmarks, such as Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and a simple empirical distribution of the arrivals;
 \item
   We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
 \end{enumerate}
 
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly ED arrival forecasting in healthcare. Section \ref{model} starts with \ldots{}
+The rest of the paper is organised as following: In section \ref{lit}, we provide a brief overview of hourly ED arrival forecasting; In Section \ref{data}, we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
+In section \ref{model}, we describe the modelling approach and benchmark methods.
+We then discuss the performance evaluation metrics in section \ref{accuracy}; in section \ref{result}, we present and discuss our results.
+Finally, we summarise our findings and present ideas for future research in section \ref{conclusion}.
 
@@
 
-There are many studies that employ models to forecast the admission and arrivals in Emergency Department.
+There are some studies that employ models to forecast admissions and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
 However, given the focus of the paper, we only discuss the hourly ED forecasting.
+
+Linear regression, ARIMA, and Naïve were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
+\citet{hertzum2017forecasting} study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
+In terms of hour of the day, patient arrivals peaked around noon.
+For days of the week, Monday is the busiest day while weekends are the quietest ones.
+July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
+They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
+
+\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
+ARIMA model was selected as the best fit model.
+Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
+\citet{Cheng2021} develop an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which is found to outperform a rolling-average benchmark.
+They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.
+
+\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
+Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
+
+\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
+The authors produced probabilistic predictions in the form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly arrivals.
+Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
+However, the length of time series is very short (only one year), which does not allow for a rigorous out-of-sample evaluation.
+
+\citet{schweigler2009forecasting} conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
+A year-long dataset of hourly ED bed occupancy was collected from three facilities.
+For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
+The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by historical average).
+
+\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
+They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
+ARIMA model outperformed all the other models.
+
 Table \ref{tab:summarylit} summarize the relevant papers.
@@
 \fontsize{11}{13}\selectfont
-\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{4em}ll}
+\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}ll}
 \toprule
-Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple seasonality\\
+Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Seasonality\\
 \midrule
-Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
-Asheim et al. & 2019 & ED arrivals & 3h & 5 & Poisson regression & MAPE & No & No\\
-Hertzum & 2017 & ED arrivals & 1,2,4,8,24 hours & 3 & linear regression; SARIMA; Naïve & MAE, MAPE, 
-MASE & No & No\\
-Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
-Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
+Cheng et al. & 2021 & ED visits & 1h to 4 h & 1 year & SARIMAX, Holt-Winters, VAR, ARIMA & MSE, MAE, MAPE, Prediction interval coverage & NO & Single\\
+Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 years & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Multiple\\
+Asheim et al. & 2019 & ED arrivals & 3h & 5 years & Poisson regression & MAPE & No & Single\\
+Hertzum & 2017 & ED arrivals & 1,2,4,8,24 h & 3 years & linear regression; SARIMA; Naïve & MAE, MAPE, 
+MASE & No & Single\\
+Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 years & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & Single\\
+Cote et al. & 2013 & ED arrivals & 24h & 2 years & Fourier regression & R\textasciicircum{}2,  Standard Error & No & Single\\
 Chase et al. & 2012 & ED CUR & 30m
-1h, 2h, 4h, 8h, 12h & 1 & Binary regression & NA & No & No\\
-Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
-Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
-McCarthy & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
-Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
-Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
+1h, 2h, 4h, 8h, 12h & 1 year & Binary regression & NA & No & Single\\
+Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 years & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & Single\\
+Jones et al. & 2009 & ED census & 24h & 2 years & VAR;  Holt winters & MAE & No & Single\\
+McCarthy & 2008 & ED arrivals & n/a & 1 year & Poisson log-linear regression model & Prediction interval coverage & Yes & Multiple\\
+Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 years & Regression & RMSE & No & Single\\
+Morzuch and Allen & 2006 & ED arrivals & 168h & 3 years & Double Exponential Smoothing; Additive Holt Winter & RMSE & No & Multiple\\
 \bottomrule
@@
 
-Linear regression, ARIMA, and Naïve were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
-Naïve method was there for the purpose of comparison.
-\citet{hertzum2017forecasting} study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year.
-In terms of hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day while weekends are the quietest ones.
-July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
-According to \citet{hertzum2017forecasting}, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
-However, after all, the models of occupancy were less accurate than those of arrivals.
-\citet{hertzum2017forecasting} mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
-
-\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
-ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
-
-\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modeled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
-The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
-
-\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services.
-Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future.
-The authors presented the prediction intervals with \(50\%\) and \(90\%\) confidence levels for the number of hourly arrivals under the Poisson assumption.
-They showed that the most important predictor was hour of the day and AR(1).
-
-\citet{schweigler2009forecasting} conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
-A year-long dataset of hourly ED bed occupancy was collected from three facilities.
-For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
-The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by historical average).
-
-\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method) and vector autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
-They used MAPE to report the forecast accuracy.
-ARIMA model outperformed all the other models.
-
 \citet{asheim2019real} developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
@@
 
+\citet{cheng2021forecasting} used one year of ED visits time series to evaluate the performance Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
+The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of \(80\%\) and \(95\%\).
+They show that SARIMAX provides more accurately forecast of hourly ED occupancy.
+
 According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
 Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily \citep{boyle2012predicting, hertzum2017forecasting}.
-However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
+While, others mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
 
@@
 \item
-  Almost all research studies produce point forecasts and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
+  Almost all research studies produce point forecasts and at best report prediction intervals.
+There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
 \item
-  most studies are not reproducible, as it is almost impossible to reapply the approaches without the helpf of the authors of those papers;
+  most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
 \item
@@
 Data used in this study comprises counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
-We aggregated the patients' arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.
-
-Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.
+We aggregated the patients' arrival times to obtain hourly arrivals, which are used in this study.
+Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week.
+Although the data is noisy, it reveals some systematic patterns.
 
@@
 and 5 p.m.
-Moreover, The number of arrivals around mid-night is slightly higher for Saturday and Sunday.
+Moreover, the number of arrivals around mid-night is slightly higher for Saturday and Sunday.
 
-Figure \ref{fig:hourly-plot-ridge} also highlights that there are many ouliers for almost every hour of the day that may affect the accuracy of forecasting methods.
-Some of these outliers might be related to holidays and special events.
+Figure \ref{fig:hourly-plot-ridge} also highlights that there is significant skewness for almost every hour of the day that varies with time-of-day, and which should be accounted for in forecasting methods.
+Some skewness might be related to holidays and special events.
 It is also clear that arrivals are less volatile between mid-night and early morning.
@@
 
-\textbf{IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.}
+Figure \ref{fig:seasonplot-weekofyear} illustrates the daily seasonal plot, with x-axis representing the day-of-week and y-axis the ED arrivals.
+There is a line for each week (i.e.~7 days) from Monday to Saturday, and lines are color-coded by week of year to show the weekly cycle.
+It is clear that there is a large jump in patient arrivals on Monday followed by Saturday each week.
+This shows that there are significantly more arrivals on Mondays and Saturdays comparign to the rest of the week.
+This might be due to the closure of General Practitioners outpatient clinics over the weekend.
 
@@
 
+Figure \ref{fig:hourly-plot-ridge} highlights the week of year seasonality in the ED arrivals.
+We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period.
+Moreover, the number of arrivals are lower at beginning and the end of the year.
+The highest arrivals are in week 39-42.
+
 \begin{figure}[H]
@@
 
-Based on this simple analysis, we should consider model that could take the following into account:
+Figure \ref{fig:hourly-plot-ridge} presents the time plot of daily arrivals.
+Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
+The figure shows more arrivals on Mondays compared to other weekdays.
+Moreover, we can observe a long-term trend line as well as the significant effect of some holidays.
+We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
+Moreover, Figure \ref{fig:hourly-plot-ridge} allows us to identify the impact of special event on arrivals.
+For instance, we see that the number of arrivals are significamtly low for 01-03 of March 2018.
+These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK \citep{stormemma2018}.
+
+Based on ED time series analysis illustrated above and literature review, we should consider models that can take the following into account:
 
 \begin{itemize}
-\tightlist
 \item
-  Hour of day, day of week, and week of year seasonal patterns,
+  Daily, weekly, and annual seasonalities,
+\item
+  Long-term trend,
 \item
@@
 \item
-  Potentially changing level of data,
+  Other events, such as sporting fixtures,
 \item
-  Other events, such as football matches in the area.
+  Weather effects, such as temperature and precipitation.
 \end{itemize}
@@
 This is called ``Naïve''.
-In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of demand is used to forecast the future demand distribution.
-\textbf{We need references here\ldots{}}
+In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution \citep{la2021new}.
 
-\hypertarget{multiple-regression}{%
-\subsection{Multiple Regression}\label{multiple-regression}}
+\hypertarget{poisson-regression}{%
+\subsection{Poisson Regression}\label{poisson-regression}}
 
 Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
-The classical linear regression model is formulated as:
+The classical linear regression model is formulated as
 
@@
 So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression \citep[see for example,][]{mccarthy2008challenge}, which can be summarised as:
+One of the models that is frequently used in practice is the Poisson regression \citep[see for example,][]{mccarthy2008challenge}, which can be summarised as
 
@@
     & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
-  \end{aligned}
+  \end{aligned} \quad .
 \label{eq:PoissonRegression}
@@
 The logarithm in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function {[}reference needed{]}.
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function .
 When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case.
@@
 
-The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
+The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
 Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g.~people might try not to go to hospital on Christmas eve and thus will go the next day).
@@
 
-\hypertarget{ets---exponential-smoothing-model}{%
-\subsection{ETS - Exponential Smoothing model}\label{ets---exponential-smoothing-model}}
+\hypertarget{ets---exponential-smoothing}{%
+\subsection{ETS - Exponential Smoothing}\label{ets---exponential-smoothing}}
 
@@
 It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
-The latter one not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
+The latter not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
 We fit a TBATS model using the \texttt{tbats()} function from the \texttt{forecast} package in R \citep{forecastpackage2020}.
@@
 
-\citet{SvetunkovAdam2021} proposed a framework for dynamic models called \textbf{ADAM} - Augmented Dynamic Adaptive Model.
+\citet{SvetunkovAdam2021} proposed a framework for dynamic models called the Augmented Dynamic Adaptive Model (ADAM).
 This framework encompasses ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
@@
 \end{equation}
+
 where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
 Finally, \(a_t\) is an unobservable series, underlying the occurrence part of the model and \((1 + \epsilon_{a,t})\) is an unobservable error term for \(a_t\).
-\citet{Svetunkov2019a} discuss how to estimate such model.
+\citet{Svetunkov2019a} discuss how to estimate such a model.
 We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components).
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them (see @ref(quantiles\_ceiling) for the explanation).
+If integer-valued quantiles are needed, then rounding up can be done for them (see \ref{quantilesceiling} for the explanation).
 This model is implemented in \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM-iETSX''.
@@
 
-If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters may be modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found.
-In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
+If we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
+Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters are modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive model structures can be found.
+In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.
 
@@
 In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
-In a GAMLSS framework of \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via:
+In a GAMLSS framework of \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled as
 
@@
 
-where \(g_j\) is a monotonic \texttt{link} function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\) \textbf{is \ldots{}}.
+where \(g_j\) is a monotonic link function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\) \textbf{is \ldots{}}.
 If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third elements of \(\bm x_t\).
@@
 
-As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form:
+As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form
 
@@
 
-Truncated Normal\ldots{}
+However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
+In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
+Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
+The resulting density forecasts are given by
 
-with
+\begin{equation}
+F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
+\label{eq:truncatedn}
+\end{equation}
+
+with additive models
 
 \begin{align*}
-  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-  log(\sigma_t) & = f(H(t)) .
+    & \mu_t = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
+    & log(\sigma_t) = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
 \end{align*}
 
-Truncated \(t\) distribution\ldots{}
-
-Negative binomial\ldots{}
+for the mean and variance parameters.
+We also performed experiments with the truncated \(t\) distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
 
@@
 
-In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model.
-We focus on 48 hours ahead, because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors.
+In order to assess performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps if 0.05, and conditional expectations for 0 to 48 hours ahead produced by each model.
+We forecast upto 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000}, resulting in 727 origins.
 Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation.
-The latter is measured via Root Mean Squared Error (RMSE):
+The latter is measured via Root Mean Squared Error (RMSE)
 
@@
 \end{equation}
+
 where \(h\) is the forecast horizon and \(e_{t+j}\) is the point forecast error \(j\) steps ahead.
 
-The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated \citep{Gneiting2007a}.
+The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated \citep{gneiting2007probabilistic}.
 A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
-Calibration, also called \emph{reliability}, is the property that forecast probabilities match the observed frequency of realisations.
+Calibration, also called reliability, is the property that forecast probabilities match the observed frequency of realisations.
 If a forecast is calibrated, then, for example, \(20\%\) of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
 This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observed frequency mean (\(\mathbb{1}(y_{t}\leq q_{\alpha,t})\)
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observed frequency mean (\(\mathbf{1}(y_{t}\leq q_{\alpha,t})\)
 ). We use several scores to assess the quantile performance of models.
 
-First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score {[}we need reference here{]}.
+First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score \citep{hyndman2021forecasting}.
 It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
@@
  \left(q_{\alpha,t} - y_{t} \right)
- \left(\mathbb{1}(y_{t}\leq q_{\alpha,t})-\alpha \right) ,
+ \left(\mathbf{1}(y_{t}\leq q_{\alpha,t})-\alpha \right) ,
  \label{eq:pinball}
@@
 We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}.
-Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlations of the underlying data {[}we need a reference here{]}.
+Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model, and employ a block-bootstrap with blocks of length 24h in order to account for temporal correlations of the underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\hypertarget{result}{%
+\section{Results}\label{result}}
 
 The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
-The rolling origin is done every 12 hours, and the forecast horizon is set for 48 steps ahead.
+The rolling origin advances in 12 hour steps, and the forecast horizon is set for 48 steps ahead.
 
-\hypertarget{forecast-evaluation}{%
-\subsection{Forecast evaluation}\label{forecast-evaluation}}
+Figure \ref{fig:Pinball} presents pinball score aggregated across forecasting horizons for each quantile.
+It shows that the difference in performance among the models mainly comes from the middle of the distribution, and somewhat form the upper tail.
+There is very little difference in performance for the lower tail.
+This is kind of interesting, and reassuring that the better models are batter at probabilities that matter more to decision makers.
 
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \ref{fig:quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
+Calibration is evaluated visually in Figure \ref{fig:quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-In terms of practicalities, this means that if staffing decisions are made using these benchmarks, we would be underestimating the demand and the hospital wil be overrun by patients and have shortage of nurses and doctors.
+For a user, this could result in poor staffing decisions as historic data fail to accurately charecterise present conditions.
 
@@
 \caption{Pinball loss values over different quantiles.}\label{fig:Pinball}
-\end{figure}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Reliability-1}
-
-}
-
-\caption{Reliability score, showing the relation between empirical and theoretical quantiles.}\label{fig:Reliability}
 \end{figure}
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/rmse-1}
-
-}
-
-\caption{RMSE ...}\label{fig:rmse}
-\end{figure}
+Figure \ref{fig:Skill-rel2bench-reduced} illustrates the trade-off between calibration and pinball skill.
 
@@
 
-\begin{figure}[H]
+Figure \ref{fig:lead-time-rmse} reports the RMSE for each forecast horizon. It illustrates the times of day that are harder to predict -- morning pick-up and afternoon peak.
 
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/time-1}
-
-}
-
-\caption{Running time ...}\label{fig:time}
-\end{figure}
+One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
+This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
 
@@
 
-One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
-This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
+\ref{fig:time-accuracy} reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
 
@@
 
+Short-term forecasting of arrivals at emergency departments is an important element of staff and resource management in hospitals.
+Furthermore, due to the asymmetric impact of having an excess of shortage of resources, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
+Here, we have developed methods for producing probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
+
+Two approaches produced highly accurate, calibrated probabilistic forecasts, one time series method and one based on distributional regression.
+The first is ADAM-iESTX, which is an extension of exponential smoothing incorporating seasonality and assuming a Gamma predictive distribution.
+The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on features date and time features, and temperature.
+Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX produced forecasts had a lower RMSE than NOtr-2, and NOtr-2 produced forecasts with a lower pinball score.
+This suggests that the latter may be preferred if the whole distribution is used in decision-making.
+
+Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts, and choice of model features, are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without need for post-processing.
+These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this article.
+Furthermore, methods based on continuous valued distributions are not adversely affected by the fact that the data are integer-valued.
+Rounding up predictive quantiles to the next integer does not make predictions worse.
+
+Finally, we have found that out-of-the-box models, those which require minimal tuning or manual development, do not perform as well as well-considered statistical methods.
+The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks.
+Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
+ETS produces reasonably well calibrated forecasts, in contrast to the benchmarks which were poorly calibrated, and highly accurate point forecasts.
+However, its probabilistic forecasts were considerably worse than NOtr-2 in terms of pinball score.
+
+Probabilistic forecasting opens the door to more sophisticated resource management in healthcare settings by providing decision-makers with uncertainty information and enabling quantitative risk management.
+Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecast skill, and may also be necessary to maximise benefits though more holistic decision-making.
+
+Further research is required to investigate the practical benefits of the probabilistic forecasts in the healthcare and how they should be used to inform planning and decision making.
+This may require employing discrete simulation or new-vendor problem.
+While this study has focused on the hourly short-term forecasting, producing longer term daily forecast (e.g.~180-270 days ahead) is often required by planners, to support winter planning in ED and Ambulance services which requires more investigation.
+Moreover, more research is required to forecast other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
+This may require considering the dynamics among various services including General Practitioners, Emergency Departments, Ambulance and Fire \& Rescue services.
+
 \hypertarget{appendices}{%
@@
 
-\hypertarget{quantiles_ceiling}{%
-\subsection{Quantiles of rounded up random variables}\label{quantiles_ceiling}}
+\hypertarget{quantilesceiling}{%
+\subsection{Quantiles of rounded up random variables}\label{quantilesceiling}}
 

### 2022-01-12 paper/paper.Rmd

@@
   - name: Bahmna Rostami Tabar
-    email: email1@example.com
-    affiliation: University1
-    footnote: 1
+    email: rostami-tabarb@cardiff.ac.uk
+    affiliation: Cardiff Business School
+    footnote: 12
   - name: Jethro Browell
     email: jethro.browell@glasgow.ac.uk
-    affiliation: School of Mathematics and Statistics, University of Glasgow, UK
-    footnote: 2
+    affiliation: University of Glasgow
+    footnote: 1
   - name: Ivan Svetunkov
     email: i.svetunkov@lancaster.ac.uk
-    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
-    footnote: 2
+    affiliation: Lancaster University
+    footnote: 1
 address:
-  - code: University1
-    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
-  - code: University2
-    address: adress2
-  - code: University3
-    address: adress3
+  - code: Cardiff Business School
+    address: Cardiff Business School, Cardiff University, UK
+  - code: University of Glasgow
+    address: School of Mathematics and Statistics, University of Glasgow, UK
+  - code: Lancaster University
+    address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
 footnote:
+  - code: 12
+    text: "Corresponding Author"
   - code: 1
-    text: "Corresponding Author"
-  - code: 2
     text: "Equal contribution"
@@
 abstract: |
-   The Objective of this work would be to propose a new methodology to forecast short-term hourly forecasting for urgent and emergency care.
+   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
+This can have many advantages for healthcare staff and the quality of care delivered to patients.
+In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
+We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and  Pinball Skill Score.
+Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.
 
-journal: "Which journal? Journal of Service Research/Health Services Research/EJOR/IJF"
+keywords:  "Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing"
+
+journal: "Annals of Emergency Medicine"
 date: "`r Sys.Date()`"
@@
 This is one of the best ways to optimize resources utilisation and minimise related costs.
-State of the art practice to optimise personnel scheduling follows the general approach originally presented by @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
-An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
+<!-- State of the art practice to optimise personnel scheduling follows the general approach originally presented by @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+-->
+An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
 Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow [@derlet2002overcrowding].
@@
 <!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
-The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
+The combination of an hourly arrival forecast, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
 Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
@@
 These studies use multiple approaches including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
-However, these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-Reporting uncertainty is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels if staff levels are insufficient.
+However, majority of these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+There are few studies that report uncertainty by presenting prediction intervals, no study generates and evaluates the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels, if staff levels are insufficient.
 Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
@@
 Another drawback of existing studies is that the datasets used are relatively small (e.g.
-time period of 1 year or less), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
+time period of 1-2 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
 Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals.
@@
 We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
-Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
+Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and  a simple empirical distribution of the arrivals;
 
@@
 
-The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly ED arrival forecasting in healthcare. Section \ref{model} starts with ...
+The rest of the paper is organised as following: In section \@ref(lit), we provide a brief overview of hourly ED arrival forecasting; In Section \@ref(data), we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
+In section \@ref(model), we describe the modelling approach and benchmark methods.
+We then discuss the performance evaluation metrics in section \@ref(accuracy); in section \@ref(result), we present and discuss our results.
+Finally, we summarise our findings and present ideas for future research in section \@ref(conclusion).
 
@@
 
-There are many studies that employ models to forecast the admission and arrivals in Emergency Department.
+There are some studies that employ models to forecast admissions and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
 However, given the focus of the paper, we only discuss the hourly ED forecasting.
-Table \@ref(tab:summarylit) summarize the relevant papers.
-
-```{r summarylit, cache=FALSE}
-li <- readxl::read_xlsx("../table/table.xlsx")
-knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in forecasting hourly arrivals in Emergency Department") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
-  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") %>%  kableExtra::landscape()
-```
 
 Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
-Naïve method was there for the purpose of comparison.
 @hertzum2017forecasting study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
@@
 July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
-According to @hertzum2017forecasting, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
-However, after all, the models of occupancy were less accurate than those of arrivals.
-@hertzum2017forecasting mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
+They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
 <!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval.
@@
 They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.
-<!-- @hertzum2017forecasting work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset. -->
 
@@
 Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
-<!-- !!!! IS: I don't understand what is meant by the next sentence !!!! -->
-<!-- However, these findings are limited to the short number of observations (only one year of historical data). -->
+However, the length of time series is very short (only one year), which does not allow for a rigorous out-of-sample evaluation.
 
@@
 @kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method) and vector autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
-They used MAPE to report the forecast accuracy.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and  Vector Autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
+They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
 ARIMA model outperformed all the other models.
+
+Table \@ref(tab:summarylit) summarize the relevant papers.
+
+```{r summarylit, cache=FALSE}
+li <- readxl::read_xlsx("../table/table.xlsx")
+knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in forecasting hourly arrivals in Emergency Department") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
+  column_spec(4,"4em") %>% column_spec(6,"10em")  %>% column_spec(7,"8em") %>%  kableExtra::landscape()
+```
 
@@
 
+@cheng2021forecasting used one year of ED visits time series to evaluate the performance Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
+The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of $80\%$ and $95\%$.
+They show that SARIMAX provides more accurately forecast of hourly ED occupancy.
+
 According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
 Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily [@boyle2012predicting; @hertzum2017forecasting].
-However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
+While, others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
 
@@
 (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
-(ii) Almost all research studies produce point forecasts and consequently report only point forecast accuracy.
-There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
+(ii) Almost all research studies produce point forecasts and at best report prediction intervals.
+There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
 (iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
@@
 and 5 p.m.
-Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.
+Moreover, the number of arrivals around mid-night is slightly higher for Saturday and Sunday.
 
@@
 
-**IS: We need to plot day of week and week of year seasonal plots here and discuss. This will show multiple seasonality nature of the data.**
+Figure \@ref(fig:seasonplot-weekofyear) illustrates the daily seasonal plot, with x-axis representing the day-of-week and y-axis the ED arrivals.
+There is a line for each week (i.e.
+7 days) from Monday to Saturday, and lines are color-coded by week of year to show the weekly cycle.
+It is clear that there is a large jump in patient arrivals on Monday followed by Saturday each week.
+This shows that there are significantly more arrivals on Mondays and Saturdays comparign to the rest of the week.
+This might be due to the closure of General Practitioners outpatient clinics over the weekend.
 
@@
 
+Figure \@ref(fig:hourly-plot-ridge) highlights the week of year seasonality in the ED arrivals.
+We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period.
+Moreover, the number of arrivals are lower at beginning and the end of the year.
+The highest arrivals are in week 39-42.
+
 ```{r date-plot, fig.cap= "Daily arrivals", fig.align='center',out.width="70%"}
@@
 
-Based on this simple analysis and literature review, we should consider models that can take the following into account:
+Figure \@ref(fig:hourly-plot-ridge) presents the time plot of daily arrivals.
+Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
+The figure shows more arrivals on Mondays compared to other weekdays.
+Moreover, we can observe a long-term trend line as well as the significant effect of some holidays.
+We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
+Moreover, Figure \@ref(fig:hourly-plot-ridge) allows us to identify the impact of special event on arrivals.
+For instance, we see that the number of arrivals are significamtly low for 01-03 of March 2018.
+These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK [@stormemma2018].
+
+Based on ED time series analysis illustrated above and literature review, we should consider models that can take the following into account:
+
 - Daily, weekly, and annual seasonalities,
+
 - Long-term trend,
+
 - Calendar events, such as holidays,
+
 - Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
+
 - Other events, such as sporting fixtures,
+
 - Weather effects, such as temperature and precipitation.
+
 We propose several forecasting models that account for the structures outlined above.
@@
 ## Naive/Climatology {#climatology}
+
 We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
 This is called "Naïve".
-In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of demand is used to forecast the future demand distribution.
-**We need references here...**
+In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution [@la2021new].
 
@@
 ## Poisson Regression
+
 Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
@@
 \end{equation}
+
 where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index.
@@
 The logarithm in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function [reference needed].
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function <!--[reference needed]--> .
 When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case.
@@
 \end{equation}
+
 where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
@@
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantiles_ceiling) for the explanation).
+If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantilesceiling) for the explanation).
 This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".
@@
 ## GAMLSS
+
 If we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
@@
 In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via -->
+
 \begin{equation}
@@
 \end{equation}
+
 where $g_j$ is a monotonic link function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**.
@@
 Each $f_{j,i}$ is a smooth function, constructed as
+
 \begin{equation}
@@
 \end{equation}
+
 where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
@@
 \end{equation}
+
 The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
@@
 The resulting density forecasts are given by
+
 \begin{equation}
@@
 \end{equation}
+
 with additive models
+
+<!-- \begin{align*} -->
+<!-- \mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + \\ -->
+<!-- t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\ -->
+<!-- log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t)) -->
+<!-- (\#eq:additivemodel_NOtr_v2) -->
+<!-- \end{align*} -->
+
 \begin{align*}
-\mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
-log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
-(\#eq:additivemodel_NOtr_v2)
+	& \mu_t = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
+	& log(\sigma_t) = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
 \end{align*}
+
 for the mean and variance parameters.
@@
 
-<!-- <!-- \begin{equation} --> -->
-<!-- <!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} --> -->
-<!-- <!--  (\#eq:truncatedn) --> -->
-<!-- <!-- \end{equation} --> -->
+<!-- <!-- \begin{equation} --> 
+<!-- <!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} --> 
+<!-- <!--  (\#eq:truncatedn) --> 
+<!-- <!-- \end{equation} -->
 
@@
 In order to assess performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps if 0.05, and conditional expectations for 0 to 48 hours ahead produced by each model.
-We forecast upto 48 hours because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors.
+We forecast upto 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins.
@@
 The latter is measured via Root Mean Squared Error (RMSE)
+
 \begin{equation}
@@
 \end{equation}
+
 where $h$ is the forecast horizon and $e_{t+j}$ is the point forecast error $j$ steps ahead.
 
-The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a].
+The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated [@gneiting2007probabilistic].
 A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
@@
 
-First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score [we need reference here].
+First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score [@hyndman2021forecasting].
 It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
@@
 positive or negative skill) are significantly different from zero [@Efron1981Bootstrap].
-Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model, and employ a block-boostrap with blocks of length 24h in order to account for temporal correlations of the underlying data [we need a reference here].
+Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model, and employ a block-bootstrap with blocks of length 24h in order to account for temporal correlations of the underlying data [@hongyi1996bootstrapping;@Bergmeir2016303].
 
@@
 
-# Results
+# Results {#result}
+
 The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
@@
 
-## Forecast evaluation
+Figure \@ref(fig:Pinball) presents pinball score aggregated across forecasting horizons for each quantile.
+It shows that the difference in performance among the models mainly comes from the middle of the distribution, and somewhat form the upper tail.
+There is very little difference in performance for the lower tail.
+This is kind of interesting, and reassuring that the better models are batter at probabilities that matter more to decision makers.
 
@@
 Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in \@ref(fig:quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
+Calibration is evaluated visually in Figure \@ref(fig:quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-For a user, this could result in poor staffing decisions as historic data fail to accuratly charterise present conditions.
+For a user, this could result in poor staffing decisions as historic data fail to accurately charecterise present conditions.
 
-<!-- IS: Use different pch for the points and different styles of lines -->
 ```{r Pinball, fig.cap= "Pinball loss values over different quantiles.", fig.align='center',out.width= "90%"}
@@
 
-<!-- IS: This plot is difficult to read. Do we need it at all? -->
-<!-- {r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"} -->
+<!-- ```{r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"} -->
+<!-- ``` -->
 
-<!-- IS: Other pch values, to make it more readable -->
-```{r Reliability, fig.cap= "Reliability score, showing the relation between empirical and theoretical quantiles.", fig.align='center',out.width= "90%"}
-```
+<!-- ```{r Reliability, fig.cap= "Reliability score, showing the relation between empirical and theoretical quantiles.", fig.align='center',out.width= "90%"} -->
+<!-- ``` -->
 
-<!-- IS: Other pch values, to make it more readable -->
 ```{r quantile-bias, fig.cap= "Quantile bias vs the nominal quantiles.", fig.align='center',out.width= "90%"}
@@
 <!-- IS: round values to 3 or 4 decimals -->
+
 ```{r tab-results, out.width= "50%"}
@@
 
+<!-- Why NAs for RMSE?? -->
+
 Evaluation metrics from the test period are presented in Table \@ref(tab:tab-results).
@@
 ```{r Skill-rel2bench-reduced, fig.cap= "Skill score ...", fig.align='center',out.width= "70%"}
-
 ```
 
-```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
+ Figure \@ref(fig:Skill-rel2bench-reduced) illustrates the trade-off between calibration and pinball skill. <!-- We may need to expand this-->
 
+<!-- ```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"} -->
+<!-- ``` -->
+
+```{r lead-time-rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
 ```
 
-<!-- IS: Different styles of lines are needed here. -->
-```{r lead-time-rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
+Figure \@ref(fig:lead-time-rmse) reports the RMSE for each forecast horizon.
+It illustrates the times of day that are harder to predict – morning pick-up and afternoon peak.
+<!-- We may need to expand this-->
 
-```
+<!-- ```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"} -->
+<!-- ``` -->
 
-```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"}
-```
+One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
+This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
 
-<!-- IS: This is not readable in black & white -->
 ```{r time-accuracy, fig.cap= "Running time vs. forecast accuracy", fig.align='center',out.width= "70%"}
@@
 
-One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
-This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
+Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. 
+<!-- We need to explain this once we have th result for RMSE -->
+<!-- ``` -->
 
@@
 
-Two appraoches produced highly accurate, calibrated probabilistic forecasts, one time series method and one based on distributional regression.
+Two approaches produced highly accurate, calibrated probabilistic forecasts, one time series method and one based on distributional regression.
 The first is ADAM-iESTX, which is an extension of exponential smoothing incorporating seasonality and assuming a Gamma predictive distribution.
@@
 Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts, and choice of model features, are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without need for post-processing.
-These observations reflect the characteristics of the data, which is representative of A\&E arrivals, but determining the extent to which they generalise is beyond the scope of this article.
-Furthermore, methods based on continuous valued distributions are not adversely affected by the fact that the data are interger-valued.
+These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this article.
+Furthermore, methods based on continuous valued distributions are not adversely affected by the fact that the data are integer-valued.
 Rounding up predictive quantiles to the next integer does not make predictions worse.
@@
 Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
-ETS produces reasonably well calibrated forecasts, in contast to the benchmarks which were poorly calibrated, and highly accurate point forecasts, however, its probabilistic forecasts were considerably worse than  NOtr-2 in terms of pinball score.
+ETS produces reasonably well calibrated forecasts, in contrast to the benchmarks which were poorly calibrated, and highly accurate point forecasts.
+However, its probabilistic forecasts were considerably worse than  NOtr-2 in terms of pinball score.
 
@@
 
+Further research is required to investigate the practical benefits of the probabilistic forecasts in the healthcare and how they should be used to inform planning and decision making.
+This may require employing discrete simulation or new-vendor problem.
+While this study has focused on the hourly short-term forecasting, producing longer term daily forecast (e.g.
+180-270 days ahead) is often required by planners, to support winter planning in ED and Ambulance services which requires more investigation.
+Moreover, more research is required to forecast other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
+This may require considering the dynamics among various services including General Practitioners, Emergency Departments, Ambulance and Fire & Rescue services.
+
 # Appendices
 
-## Quantiles of rounded up random variables {#quantiles_ceiling}
+## Quantiles of rounded up random variables {#quantilesceiling}
 

### 2022-01-12 paper/paper.Rmd
correct typo

@@
 ---
-title:  Forecasting short-term hourly Emergency Departement arrivals
+title:  Forecasting short-term hourly Emergency Department arrivals
 author:
@@
 abstract: |
-   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
+   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients’ demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
 This can have many advantages for healthcare staff and the quality of care delivered to patients.
 In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
-We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution.
-We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and  Pinball Skill Score.
+We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
 Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
-Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.
 

### 2022-01-12 paper/paper.tex
correct typo

@@
             pdfauthor={},
-            pdftitle={Forecasting short-term hourly Emergency Departement arrivals},
+            pdftitle={Forecasting short-term hourly Emergency Department arrivals},
             colorlinks=false,
@@
 
-  \title{Forecasting short-term hourly Emergency Departement arrivals}
+  \title{Forecasting short-term hourly Emergency Department arrivals}
     \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
@@
   \begin{abstract}
-  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
+  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients' demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
 This can have many advantages for healthcare staff and the quality of care delivered to patients.
 In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
-We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution.
-We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
+We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
 Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
-Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.
   \end{abstract}
@@
 
-\ref{fig:time-accuracy} reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+Figure \ref{fig:time-accuracy} reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
 

### 2022-01-12 paper/paper.tex
add page title and correct typos

@@
 
-Figure \ref{fig:hourly-plot-ridge} highlights the week of year seasonality in the ED arrivals.
+Figure \ref{fig:seasonplot-weekofyear} highlights the week of year seasonality in the ED arrivals.
 We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period.
@@
 
-Figure \ref{fig:hourly-plot-ridge} presents the time plot of daily arrivals.
+Figure \ref{fig:date-plot} presents the time plot of daily arrivals.
 Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
@@
 We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
-Moreover, Figure \ref{fig:hourly-plot-ridge} allows us to identify the impact of special event on arrivals.
-For instance, we see that the number of arrivals are significamtly low for 01-03 of March 2018.
+Moreover, Figure \ref{fig:date-plot} allows us to identify the impact of special event on arrivals.
+For instance, we see that the number of arrivals are significantly low for 01-03 of March 2018.
 These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK \citep{stormemma2018}.

### 2022-01-12 paper/paper.Rmd

@@
 
-Figure \@ref(fig:hourly-plot-ridge) highlights the week of year seasonality in the ED arrivals.
+Figure \@ref(fig:seasonplot-weekofyear) highlights the week of year seasonality in the ED arrivals.
 We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period.
@@
 
-Figure \@ref(fig:hourly-plot-ridge) presents the time plot of daily arrivals.
+Figure \@ref(fig:date-plot) presents the time plot of daily arrivals.
 Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
@@
 We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
-Moreover, Figure \@ref(fig:hourly-plot-ridge) allows us to identify the impact of special event on arrivals.
-For instance, we see that the number of arrivals are significamtly low for 01-03 of March 2018.
+Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact of special event on arrivals.
+For instance, we see that the number of arrivals are significantly low for 01-03 of March 2018.
 These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK [@stormemma2018].

### 2022-01-12 paper/paper.Rmd
Checked and amended the Abstract, Introduction and Lit. Review

@@
 abstract: |
-   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients’ demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
+   An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients’ demand.
+It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units.
 This can have many advantages for healthcare staff and the quality of care delivered to patients.
-In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
-We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution.
-We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
-Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
-Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.
+In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals.
+We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score.
+Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
 
@@
 Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
-An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing.
-This is one of the best ways to optimize resources utilisation and minimise related costs.
-<!-- State of the art practice to optimise personnel scheduling follows the general approach originally presented by @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
--->
+Accurate ED demand forecasts contribute to a better decision making process regarding resources allocation and staffing.
+This is one of the best ways to optimise resources utilisation and minimise related costs.
 An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow [@derlet2002overcrowding].
-Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
-<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->
+Inability to match the staff with the demand might result in patients overcrowding the system, which is a severe problem that causes challenges for the patient flow [@derlet2002overcrowding].
+Also, it is related to increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
 
-An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+Accurate forecasting of arrivals by the hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This has many advantages for both patients, staff and the quality of provided services.
@@
 This involves the short-term decision making related to the execution of the delivery process in ED.
-<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
-The combination of an hourly arrival forecast, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
-Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
-However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
-They may also express nonstationarity and their profile may change over time.
-Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
-<!-- Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
-Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
-That may also require to adjust the scheduling.
--->
+The combination of an hourly arrival forecast, current staff, being occupied, resource availability and waiting times at ED provide information on the state of the unscheduled care system across the service.
+Having this complete picture enables the delivery managers to focus on the areas that require intervention to allow the most effective delivery of the service to the patients.
+However, compared with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+Hourly time series generally exhibit multiple seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
+They may also express nonstationarity, and their profile may change over time.
+Therefore, an appropriate forecasting model should consider these features to accurately predict hourly demand admissions.
 
 There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
-These studies use multiple approaches including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
-However, majority of these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-There are few studies that report uncertainty by presenting prediction intervals, no study generates and evaluates the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels, if staff levels are insufficient.
+These studies use multiple approaches, including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
+However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
 Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
-Furthermore, if the impact of under and over staffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under and over staffing to be calculated.
+
+Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated.
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
 Another drawback of existing studies is that the datasets used are relatively small (e.g.
-time period of 1-2 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
-Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals.
+time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
+Such results might not be generalisable.
+Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals.
 Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
@@
 1.
-We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
+We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
 
 2.
-We develop an advanced dynamic model to forecasts ED arrivals based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies which produced highly-accurate point forecasts;
+We develop an advanced dynamic model to forecast ED arrivals based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
 
 3.
-We develop a novel model to produce probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
+We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
 
@@
 We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
-Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and  a simple empirical distribution of the arrivals;
+Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
 
@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.
 
-The rest of the paper is organised as following: In section \@ref(lit), we provide a brief overview of hourly ED arrival forecasting; In Section \@ref(data), we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
+The rest of the paper is organised as follows: In section \@ref(lit), we provide a brief overview of hourly ED arrival forecasting; In Section \@ref(data), we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
 In section \@ref(model), we describe the modelling approach and benchmark methods.
@@
 
-There are some studies that employ models to forecast admissions and arrivals in Emergency Department.
+Some studies employ models to forecast admissions and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only discuss the hourly ED forecasting.
+However, given the focus of the paper, we only discuss hourly ED forecasting.
 
-Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
+@hertzum2017forecasting used linear regression, ARIMA, and Naïve to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
 @hertzum2017forecasting study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
-In terms of hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day while weekends are the quietest ones.
-July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
+In terms of the hour of the day, patient arrivals peaked around noon.
+For days of the week, Monday is the busiest day, while weekends are the quietest ones.
+July-August are the months with the highest number of patient arrivals, while January and February have the lowest numbers.
 They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
-<!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval.
-Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.
--->
 
-@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
+@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neural network methods to forecast hourly accident and emergency department arrivals.
 ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
-@Cheng2021 develop an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which is found to outperform a rolling-average benchmark.
-They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.
+Authors claimed that ARIMA provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
+@Cheng2021 developed an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which outperformed a rolling-average benchmark.
+They also produce prediction intervals, a form of the probabilistic forecast, which were found to be well-calibrated, a necessary property for such forecasts.
 
 @morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for a horizon of 168 hours.
+The hourly data collected from an ED in Pennsylvania showed no trend and two seasonal cycles: a within-day and a within-week seasonal cycles.
 The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
 
-@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
+@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day-of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
 The authors produced probabilistic predictions in the form of $50\%$ and $90\%$ prediction intervals for the number of hourly arrivals.
-Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
-However, the length of time series is very short (only one year), which does not allow for a rigorous out-of-sample evaluation.
+Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it was suggested that the model could be used for forecasting.
+However, model evaluation was performed in-sample on only one year of data, so it is unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
+However, the length of the time series in this study was very short (only one year), which did not allow for a rigorous out-of-sample evaluation.
 
-@schweigler2009forecasting conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
+@schweigler2009forecasting investigated whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
 A year-long dataset of hourly ED bed occupancy was collected from three facilities.
-For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
-The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).
+The authors implemented an hourly historical average model, SARIMA model, and sinusoidal model with autocorrelated error for each facility.
+The historical average model was based on the mean occupancy for each site, for each hour of the day, while the sinusoidal model was based on four parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizons using RMSE.
+They found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by the historical average).
 
-@kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and  Vector Autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
+@kim2014predicting compared different univariate and multivariate time series forecasting techniques to predict patient volume for a Hospital Medicine programme.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and  Vector Autoregressive (VAR) models to forecast for 4 and 24 hours ahead.
 They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
-ARIMA model outperformed all the other models.
+The authors found that the ARIMA outperformed all the other models.
 
@@
 @asheim2019real developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
-Measuring the accuracy using the MAPE, @asheim2019real noticed that great improvement happened when time of notification was incorporated into the model, especially in one-hour horizon.
+Measuring the accuracy using the MAPE, @asheim2019real noticed that significant improvement happened when the time of notification was incorporated into the model, especially in the one-hour horizon.
 
-@cheng2021forecasting used one year of ED visits time series to evaluate the performance Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
+@cheng2021forecasting used one year of ED visits time series to evaluate the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
 The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of $80\%$ and $95\%$.
-They show that SARIMAX provides more accurately forecast of hourly ED occupancy.
+They show that SARIMAX provides a more accurate forecast of hourly ED occupancy.
 
-According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
-Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily [@boyle2012predicting; @hertzum2017forecasting].
-While, others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
+According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established.
+Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low [@boyle2012predicting; @hertzum2017forecasting], while others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
 
-There are few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches. These limitations are summarised as follows:
+There are a few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches:
 
 (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
-(ii) Almost all research studies produce point forecasts and at best report prediction intervals.
-There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
+(ii) Almost all research studies produce point forecasts and, at best, report prediction intervals.
+There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent the uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
 (iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
 (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
-(v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method or forecast accuracy is not reported.
+(v) some studies in this area lack a rigorous experimental design, i.e. they do not use benchmark methods or report forecast accuracy.
 

### 2022-01-12 paper/paper.Rmd
Corrections by Ivan

@@
 Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
-Accurate ED demand forecasts contribute to a better decision making process regarding resources allocation and staffing.
+Accurate ED demand forecasts contribute to a better decision-making process regarding resources allocation and staffing.
 This is one of the best ways to optimise resources utilisation and minimise related costs.
@@
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process in ED.
+This involves the short-term decision-making related to the execution of the delivery process in ED.
 The combination of an hourly arrival forecast, current staff, being occupied, resource availability and waiting times at ED provide information on the state of the unscheduled care system across the service.
@@
 It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by @hertzum2017forecasting.
-The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again.
+The ED arrivals decrease between midnight and early morning and then increase until the evening, decreasing after that again.
 It is also clear that ED service gets systematically more visits on Mondays between 8 a.m.
 and 5 p.m.
-Moreover, the number of arrivals around mid-night is slightly higher for Saturday and Sunday.
+Moreover, the number of arrivals around midnight is slightly higher for Saturdays and Sundays.
 
-Figure \@ref(fig:hourly-plot-ridge) also highlights that there is significant skewness for almost every hour of the day that varies with time-of-day, and which should be accounted for in forecasting methods.
+Figure \@ref(fig:hourly-plot-ridge) also highlights significant skewness for almost every hour of the day that varies with time-of-day, which should be accounted for in forecasting methods.
 Some skewness might be related to holidays and special events.
-It is also clear that arrivals are less volatile between mid-night and early morning.
+It is also clear that arrivals are less volatile between midnight and early morning.
 
@@
 
-Figure \@ref(fig:seasonplot-weekofyear) illustrates the daily seasonal plot, with x-axis representing the day-of-week and y-axis the ED arrivals.
+Figure \@ref(fig:seasonplot-weekofyear) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
 There is a line for each week (i.e.
-7 days) from Monday to Saturday, and lines are color-coded by week of year to show the weekly cycle.
-It is clear that there is a large jump in patient arrivals on Monday followed by Saturday each week.
-This shows that there are significantly more arrivals on Mondays and Saturdays comparign to the rest of the week.
+seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle.
+It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week.
+This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week.
 This might be due to the closure of General Practitioners outpatient clinics over the weekend.
@@
 Figure \@ref(fig:seasonplot-weekofyear) highlights the week of year seasonality in the ED arrivals.
-We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period.
-Moreover, the number of arrivals are lower at beginning and the end of the year.
-The highest arrivals are in week 39-42.
+We observe that arrivals are significantly lower from week 29 to 35, corresponding to the Summer period.
+Moreover, the number of arrivals is lower at the beginning and the end of the year.
+The arrivals increase from week 36 and peak in weeks 39-42.
+This corresponds to the September - October period.
 
@@
 
-Figure \@ref(fig:date-plot) presents the time plot of daily arrivals.
+Finally, Figure \@ref(fig:date-plot) presents the time plot of daily arrivals.
 Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
-The figure shows more arrivals on Mondays compared to other weekdays.
-Moreover, we can observe a long-term trend line as well as the significant effect of some holidays.
+The figure shows more arrivals on Mondays than on other weekdays, agreeing with the previous findings.
+Moreover, we can observe a long-term trend line and the significant effect of some holidays.
 We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
-Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact of special event on arrivals.
-For instance, we see that the number of arrivals are significantly low for 01-03 of March 2018.
-These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK [@stormemma2018].
+Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact of special events on arrivals.
+For instance, we see that arrivals are significantly low for 01-03 of March 2018.
+These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages, and schools closed in the UK [@stormemma2018].
 
-Based on ED time series analysis illustrated above and literature review, we should consider models that can take the following into account:
+Based on this analysis and the literature review, we should consider models that can take the following into account:
 
-- Daily, weekly, and annual seasonalities,
-
-- Long-term trend,
-
+- Hour-of-day, day-of-week, and week-of-year seasonalities,
+- Long-term trend (or a slowly changing level),
 - Calendar events, such as holidays,
-
 - Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
-
 - Other events, such as sporting fixtures,
-
-- Weather effects, such as temperature and precipitation.
+- Weather effects, such as temperature and precipitation. **[IS: this does not follow from the previous analysis]**
 
@@
 
-We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
+We start with one of the simplest forecasting approaches used in practice - assuming that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
 This is called "Naïve".
@@
 
-<!-- Empirical distribution for same hour-of-day and day-of-week as target time. -->
-
 ## Poisson Regression
 
-Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
+Regression is one of the most popular forecasting methods that use explanatory variables to predict a variable of interest (in our case, the ED arrivals).
 The classical linear regression model is formulated as
-
 \begin{equation}
@@
 \end{equation}
-
 where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index.
-However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
+However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic because the number of admitted patients is an integer and non-negative.
 So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as
-
+One of the models that is frequently used in practice is the Poisson regression [see, for example, @mccarthy2008challenge], which can be summarised as
 \begin{equation}
@@
 
-The logarithm in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function <!--[reference needed]--> .
-When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case.
+The logarithm in \@ref(eq:PoissonRegression) is needed to ensure that the parameter of Poisson distribution is always positive.
+This model can be estimated via maximisation of the likelihood function based on Poisson mass function <!--[reference needed]-->.
+There is no single correct answer when selecting explanatory variables for the model, and the decision needs to be made for each specific case.
 In our experiment, we will only include dummy variables, capturing a variety of calendar events:
 
-1. Hour of day,
-2. Day of week,
-3. Week of year,
-4. Holidays (such as Christmas, New Year etc),
+1. Hour of the day,
+2. Day of the week,
+3. Week of the year,
+4. Holidays (such as Christmas, New Year etc.),
 5. 24 hours lags of holidays.
@@
 The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
-Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g.
-people might try not to go to hospital on Christmas eve and thus will go the next day).
-This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time.
-We use `alm()` function from `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
+Finally, (5) is needed to capture the potential phenomenon of change in admittance after the holiday (e.g.
+people might try not to go to the hospital on Christmas eve and thus go the next day).
+This model assumes that all these effects are deterministic and do not change over time.
+Still, the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables so that the 3 pm on Monday in January will be different from 3 pm on Monday in July, although the parameters for the hour of day and day of the week are fixed and do not change over time.
+We use the `alm()` function from the `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
 
@@
 
-@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus *ETS*).
-Given the popularity of ETS model in forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
-This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as *ETS*.
+@Hyndman2008b developed a state space approach for exponential smoothing models.
+The model can have a set of elements, including different types of Error, Trend and Seasonal components (thus *ETS*).
+Given the popularity of the ETS model in the forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with a frequency of 24 (hour of the day) as a benchmark.
+This was done using the `adam()` function from the `smooth` package [@Svetunkov2021Smooth] for R and denoted as *ETS*.
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
-This model is included as a benchmark, only to see how the other models perform in comparison with it.
+This model is included as a benchmark, only to see how the other models perform compared to it.
 
@@
 Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects.
-Prophet is robust to missing data and shifts in the trend, and typically handles outliers well.
+Prophet is robust to missing data and shifts in the trend and typically handles outliers well.
 Prophet works well on daily data seen in Facebook.
@@
 The implementation may be less flexible than other methods.
-The model itself relies on Multiple Source of Error state space model, originally proposed by @Kalman1960.
-The model is incorporated using corresponding implementation of the Fable package in R.
+The model itself relies on the Multiple Source of Error state space model, initially proposed by @Kalman1960.
+The model is incorporated using the corresponding implementation of the Fable package in R.
 We use the `prophet()` function from the `fable` package [@fable2020].
@@
 It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
-The latter not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
+The latter gives the model more flexibility to deal with fractional seasonality and reduces the parameters of the model when the frequencies of seasonalities are high.
 We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].
@@
 This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
-Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
-Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by @Svetunkov2019a to treat those values.
+Based on this framework, we use the ETS(M,N,M) model with frequencies 24 (hour of the day) and 168 (hour of the week), adding dummy variables for the week of the year, holidays and lagged holidays.
+This way, we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
+Given that the data exhibits randomly occurring zeroes, we use the direct probability model of @Svetunkov2019a to treat those values.
+Finally, given the skewness of the empirical distribution observed in the preliminary analysis, we use the Gamma distribution for the error term.
 This model can be formulated as a set of the following equations:
-
 \begin{equation}
@@
 \end{equation}
-
 where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
@@
 @Svetunkov2019a discuss how to estimate such a model.
-We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components).
+We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic nature of the model (level and seasonal components).
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantilesceiling) for the explanation).
-This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".
+If integer-valued quantiles are needed, then rounding up can be done for them (see Appendix \@ref(quantilesceiling) for the explanation).
+This model is implemented in the `adam()` function from the `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".
 
@@
 
-If we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
-Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters are modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive model structures can be found.
+Suppose we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above.
+In that case, the forecasting task becomes one of predicting the future values of that distribution's parameters.
+We then can use Generalised Additive Models for Location, Scale and Shape (GAMLSS).
+These are the distributional regression models where the parameters are modelled as additive functions of explanatory variables.
+This provides a powerful and flexible framework for probabilistic forecasting, provided that suitable distribution and additive model structures can be found.
 In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.
@@
 \end{equation}
-
 where $g_j$ is a monotonic link function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**.
@@
 Each $f_{j,i}$ is a smooth function, constructed as
-
 \begin{equation}
@@
 As our data are counts, the natural starting point is the Poisson distribution with an additive model for $\log \lambda_t$ of the form
-<!-- !!!! IS: I'm not sure that we need CDFs and PMFs, because they are well known. !!!! -->
-
-<!-- \begin{equation} -->
-<!--   F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor} -->
-<!--   (\#eq:poissonreg) -->
-<!-- \end{equation} -->
-
-<!-- where we consider an additive model for $\lambda_t$ of the form -->
-
 \begin{equation}
@@
 In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
-Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
+Therefore, we consider more flexible, two-parameter distributions to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
 The resulting density forecasts are given by
-
 \begin{equation}
@@
 \end{equation}
-
 with additive models
-
 <!-- \begin{align*} -->
@@
 <!-- \end{align*} -->
-
 \begin{align*}
-	& \mu_t = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
-	& log(\sigma_t) = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
+	& \mu_t = \sum_{i=1}^{10} \beta_i D^{+}_i(t) + \sum_{j=1}^{10} D^{+}_j(t) f_j(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
+	& log(\sigma_t) = \sum_{i=1}^{10} D^{+}_i(t) f(H(t))
 \end{align*}
-
 for the mean and variance parameters.
@@
 
-In order to assess performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps if 0.05, and conditional expectations for 0 to 48 hours ahead produced by each model.
-We forecast upto 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
+In order to assess the performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps of 0.05 and conditional expectations for 0 to 48 hours ahead produced by each model.
+We forecast up to 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins.
-Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation.
-The latter is measured via Root Mean Squared Error (RMSE)
-
+Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and expectation.
+The latter is measured via Root Mean Squared Error (RMSE):
 \begin{equation}
@@
 \end{equation}
-
 where $h$ is the forecast horizon and $e_{t+j}$ is the point forecast error $j$ steps ahead.
@@
 The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated [@gneiting2007probabilistic].
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision-makers provided the forecast is calibrated.
 Calibration, also called reliability, is the property that forecast probabilities match the observed frequency of realisations.
@@
 This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbf{1}(y_{t}\leq q_{\alpha,t})$
-<!-- IS: Check the \mathbb thingy - it didn't work out in pdf // JB: I changed to mathbf, lets see how it looks... -->
-). We use several scores to assess the quantile performance of models.
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbf{1}(y_{t}\leq q_{\alpha,t})$).
+We use several scores to assess the quantile performance of models.
 
-First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score [@hyndman2021forecasting].
-It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
-Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model.
+First, to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score [@hyndman2021forecasting].
+It rewards sharpness and penalises miscalibration, so it measures overall performance.
+However, calibration should still be verified separately.
+Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in a quantile regression model.
 The Pinball Score is given by
-
 \begin{equation}
@@
 To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
-
 \begin{equation}
@@
 \end{equation}
-
 where $M$ is the metric's value for the method being considered, $M_\mathrm{ref}$ is the metric's value for a reference method.
-The skill score show us by how many percent the reference approach is worse than the one under consideration.
-<!-- and $M_\mathrm{perf}$ is the metrics value for the `perfect' method, which is zero in the case of RMSE and Pinball. -->
-We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.
+The skill score shows by how many per cent the reference approach is worse than the one under consideration.
+We will use bootstrap re-sampling of skill scores to determine if the differences in forecast performance (i.e.
 positive or negative skill) are significantly different from zero [@Efron1981Bootstrap].
-Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model, and employ a block-bootstrap with blocks of length 24h in order to account for temporal correlations of the underlying data [@hongyi1996bootstrapping;@Bergmeir2016303].
+Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model and employ a block-bootstrap with blocks of length 24h to account for temporal correlations of the underlying data [@hongyi1996bootstrapping;@Bergmeir2016303].
 
@@
 All functions were re-estimated on each iteration.
-ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin.
+ADAM and Poisson regression estimated the parameters, taking the ones obtained in the initial model application to the data in the first origin as the pre-initials.
 This allowed to speed up the computation for these two models.
 The initial estimation of ADAM took approximately one hour and 25 minutes.
-Each step in the experiment took the time shown below.
+Each step in the experiment took the time shown in Section \@ref(result).
 
@@
 The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
-The rolling origin advances in 12 hour steps, and the forecast horizon is set for 48 steps ahead.
+The rolling origin advances in 12-hour steps and the forecast horizon is set for 48 steps ahead.
 
 Figure \@ref(fig:Pinball) presents pinball score aggregated across forecasting horizons for each quantile.
-It shows that the difference in performance among the models mainly comes from the middle of the distribution, and somewhat form the upper tail.
-There is very little difference in performance for the lower tail.
-This is kind of interesting, and reassuring that the better models are batter at probabilities that matter more to decision makers.
+It shows that the difference in performance among the models mainly comes from the middle of the distribution and somewhat from the upper tail.
+There is very little difference in performance for the lower tail (except for TBATS, which has a consistently higher pinball value than the other models).
+This is interesting and reassuring that the better models are better at probabilities that matter more to decision-makers.
 
 Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated.
-Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in Figure \@ref(fig:quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
+Mis-calibrated forecasts are unsuitable for use in decision-making, so they should be excluded.
+Calibration is evaluated visually in Figure \@ref(fig:quantile-bias), which highlights a systematic negative bias across all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-For a user, this could result in poor staffing decisions as historic data fail to accurately charecterise present conditions.
+This could result in poor staffing decisions for a user as historic data fail to characterise present conditions accurately.
 
@@
 They are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
 
-```{r Skill-rel2bench-reduced, fig.cap= "Skill score ...", fig.align='center',out.width= "70%"}
+```{r Skill-rel2bench-reduced, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "70%"}
 ```
@@
 
-```{r lead-time-rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"}
+```{r lead-time-rmse, fig.cap= "RMSE values over the forecast horizon.", fig.align='center',out.width= "70%"}
 ```
@@
 It illustrates the times of day that are harder to predict – morning pick-up and afternoon peak.
-<!-- We may need to expand this-->
+We can see that some models perform much better than others on specific forecast horizons.
+For example, Benchmark does an excellent job in predicting 1 - 5 steps ahead ED arrivals and, in general, doing well in forecasting arrivals in the night.
+In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, prophet, tbats and GBM do for some lead times.
 
@@
 
-One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
-This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
+One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones (Table \@ref(tab:tab-results)).
+This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered a reasonable strategy.
 
@@
 
-Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. 
+Finally, Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
 <!-- We need to explain this once we have th result for RMSE -->
@@
 
-Short-term forecasting of arrivals at emergency departments is an important element of staff and resource management in hospitals.
-Furthermore, due to the asymmetric impact of having an excess of shortage of resources, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
-Here, we have developed methods for producing probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
+Short-term forecasting of arrivals at emergency departments is an important element of hospital staff and resource management.
+Furthermore, due to the asymmetric impact of an excess resource shortage, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
+In this paper, we have developed methods for producing a probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
 
-Two approaches produced highly accurate, calibrated probabilistic forecasts, one time series method and one based on distributional regression.
-The first is ADAM-iESTX, which is an extension of exponential smoothing incorporating seasonality and assuming a Gamma predictive distribution.
-The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on features date and time features, and temperature.
-Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX produced forecasts had a lower RMSE than NOtr-2, and NOtr-2 produced forecasts with a lower pinball score.
+Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression.
+The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution.
+The second, labelled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature.
+Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score.
 This suggests that the latter may be preferred if the whole distribution is used in decision-making.
 
-Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts, and choice of model features, are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without need for post-processing.
-These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this article.
-Furthermore, methods based on continuous valued distributions are not adversely affected by the fact that the data are integer-valued.
+Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts and choice of model features are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without the need for post-processing.
+These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this paper.
+Furthermore, methods based on continuous-valued distributions are not adversely affected by the fact that the data are integer-valued.
 Rounding up predictive quantiles to the next integer does not make predictions worse.
 
-Finally, we have found that out-of-the-box models, those which require minimal tuning or manual development, do not perform as well as well-considered statistical methods.
+Finally, we have found that out-of-the-box models, which require minimal tuning or manual development, do not perform as well as well-considered statistical approaches.
 The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks.
 Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
-ETS produces reasonably well calibrated forecasts, in contrast to the benchmarks which were poorly calibrated, and highly accurate point forecasts.
+ETS produces reasonably well-calibrated forecasts, in contrast to the poorly calibrated benchmarks and has highly accurate point forecasts.
 However, its probabilistic forecasts were considerably worse than  NOtr-2 in terms of pinball score.
@@
 Probabilistic forecasting opens the door to more sophisticated resource management in healthcare settings by providing decision-makers with uncertainty information and enabling quantitative risk management.
-Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecast skill, and may also be necessary to maximise benefits though more holistic decision-making.
+Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecasting skills and may also be necessary to maximise benefits through more holistic decision-making.
 
-Further research is required to investigate the practical benefits of the probabilistic forecasts in the healthcare and how they should be used to inform planning and decision making.
-This may require employing discrete simulation or new-vendor problem.
-While this study has focused on the hourly short-term forecasting, producing longer term daily forecast (e.g.
-180-270 days ahead) is often required by planners, to support winter planning in ED and Ambulance services which requires more investigation.
-Moreover, more research is required to forecast other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
-This may require considering the dynamics among various services including General Practitioners, Emergency Departments, Ambulance and Fire & Rescue services.
+Further research is required to investigate the practical benefits of probabilistic forecasts in healthcare and how they can inform planning and decision making.
+This may require employing discrete simulation or application to the newsvendor problems.
+While this study has focused on hourly short-term forecasting, producing longer-term daily forecast (e.g.
+180-270 days ahead) is often required by planners to support winter planning in ED and Ambulance services which requires more investigation.
+Moreover, more research is needed in the forecasting of other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
+This may require considering the dynamics among various services, including General Practitioners, Emergency Departments, Ambulance and Fire & Rescue services.
 
@@
 
-Before proceeding with the proof we need to give the definition of the quantiles of the continuous and rounded up random variables:
-
+Before proceeding with the proof, we need to define the quantiles of the continuous and rounded up random variables:
 \begin{equation} \label{eq:quantCeil1}
@@
 \end{equation}
-
 and
-
 \begin{equation} \label{eq:quantCeil2}
@@
 \end{equation}
-
 where $n$ is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and $k$ is the quantile of the continuous distribution of the variable.
@@
 In order to prove that $n = \lceil k \rceil$, we need to use the following basic property:
-
 \begin{equation} \label{eq:quantCeil3}
@@
 \end{equation}
-
 which means that the rounded up value will always be less than or equal to $n$ if and only if the original value is less than or equal to $n$.
 Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
-
 \begin{equation} \label{eq:quantCeil4}
@@
 Note also that the following is true:
-
 \begin{equation} \label{eq:quantCeil5}
@@
 Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2}, \eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the following can be summarised:
-
 \begin{equation} \label{eq:quantCeil6}
@@
 \end{equation}
-
 which is possible only when $k \in (n-1, n]$, which means that $\lceil k \rceil = n$.
 So the rounded up quantile of continuous random variable $y_t$ will always be equal to the quantile of the descritised value of $y_t$:
-
 \begin{equation} \label{eq:ceilingAndQuantiles1}

### 2022-01-13 paper/paper.Rmd
upade paper

@@
 - Other events, such as sporting fixtures,
-- Weather effects, such as temperature and precipitation. **[IS: this does not follow from the previous analysis]**
+- Temperature effects.
 
@@
 In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution [@la2021new].
+We consider the empirical distribution of all available historic data (Benchmark-1) and the empirical distribution of the most recent year of historic data on a rolling basis (Benchmark-2) to capture potential changes in behaviour over time.
 
@@
 We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].
+
+## Quantile Regression and Gradient Boosting Machines
+
+Quantile regression allows the production of density forecasts without assuming a fixed distributions shape controlled by a small number of parameters, such as the Poisson distribution with parameter $\lambda$.
+By producing forecasts of multiple quantiles the full predictive distribution can be constructed.
+Quantile regression is particularly useful where data do not follow a simple distribution, of the distribution shape changes over time of with some covariate, which may be the case with ED arrivals.
+
+Gradient Boosting Machines (GBM) are a tree-based machine learning model for regression and classification.
+Here we produce probabilistic forecasts of ED arrivals via multiple quantile regression using GBMs as implemented in the R package `gbm` [@Greenwell2020].
+GBMs are a best-in-class algorithm for similar regression problems characterised by modest volumes of training data and possible interactions between input features, and are therefore an appealing choice for ED arrival forecasting.
+Here GBMs are fit with the following features: hour of the day, day of the week, school holiday, temperature and day of the year.
+Hyperparameters are chosen via grid search on the training data, and were chosen as follows: 500 trees, an interaction depth of 2, and a shrinkage 0.1.
+An advantage of GBMs is their ability to learn interactions in comparison to the additive models presented later which require the user to specify possible interactions between inputs.
+Here GBM is included as a reference for the performance of `out-of-the-box' machine learning models applied with minimal effort.
+We also explored the possibility of performing linear regression with additive models, but these did not perform as well as GBM so are omited for brevity.
 
@@
 The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
+This model is called Poisson-1 in discussions below.
 
@@
 \begin{align*}
-	& \mu_t = \sum_{i=1}^{10} \beta_i D^{+}_i(t) + \sum_{j=1}^{10} D^{+}_j(t) f_j(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
+	& \mu_t = \sum_{i=1}^{10} \beta_i D^{+}_i(t) + \sum_{j=1}^{10} D^{+}_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
 	& log(\sigma_t) = \sum_{i=1}^{10} D^{+}_i(t) f(H(t))
 \end{align*}
-for the mean and variance parameters.
-We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
+for the mean and variance parameters. This model is referred to as NOtr-1 below.
+
+Furthermore, we consider an extnsion to the additive models for $\lambdat_t$ and $\mu_t$ above to incorporate school and public holidays into $D$.
+These models are labelled Poisson-2 and NOtr-2.
+We also performed experiments with the truncated $t$ distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.
 

### 2022-01-13 paper/paper.Rmd
add points instead of labels in the plot of time versus accracy

--- 2022-01-12 paper/paper.Rmd
@@
 abstract: |
-   An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients’ demand.
-It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units.
+   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients’ demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
 This can have many advantages for healthcare staff and the quality of care delivered to patients.
-In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals.
-We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution.
-We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score.
-Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts.
-Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
+In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
+We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
+Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.
 
@@
 library(feasts)
+library(data.table)
 ```
@@
 Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
-Accurate ED demand forecasts contribute to a better decision making process regarding resources allocation and staffing.
-This is one of the best ways to optimise resources utilisation and minimise related costs.
+An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing.
+This is one of the best ways to optimize resources utilisation and minimise related costs.
+<!-- State of the art practice to optimise personnel scheduling follows the general approach originally presented by @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts.
+-->
 An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand might result in patients overcrowding the system, which is a severe problem that causes challenges for the patient flow [@derlet2002overcrowding].
-Also, it is related to increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
+Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow [@derlet2002overcrowding].
+Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
+<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->
 
-Accurate forecasting of arrivals by the hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This has many advantages for both patients, staff and the quality of provided services.
@@
 This involves the short-term decision making related to the execution of the delivery process in ED.
-The combination of an hourly arrival forecast, current staff, being occupied, resource availability and waiting times at ED provide information on the state of the unscheduled care system across the service.
-Having this complete picture enables the delivery managers to focus on the areas that require intervention to allow the most effective delivery of the service to the patients.
-However, compared with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-Hourly time series generally exhibit multiple seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
-They may also express nonstationarity, and their profile may change over time.
-Therefore, an appropriate forecasting model should consider these features to accurately predict hourly demand admissions.
+<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
+The combination of an hourly arrival forecast, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
+Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
+However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
+They may also express nonstationarity and their profile may change over time.
+Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
+<!-- Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week.
+Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism.
+That may also require to adjust the scheduling.
+-->
 
 There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
-These studies use multiple approaches, including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
-However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
+These studies use multiple approaches including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
+However, majority of these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+There are few studies that report uncertainty by presenting prediction intervals, no study generates and evaluates the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels, if staff levels are insufficient.
 Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
-
-Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated.
+Furthermore, if the impact of under and over staffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under and over staffing to be calculated.
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
 Another drawback of existing studies is that the datasets used are relatively small (e.g.
-time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
-Such results might not be generalisable.
-Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals.
+time period of 1-2 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
+Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals.
 Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
@@
 1.
-We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
+We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
 
 2.
-We develop an advanced dynamic model to forecast ED arrivals based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
+We develop an advanced dynamic model to forecasts ED arrivals based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies which produced highly-accurate point forecasts;
 
 3.
-We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
+We develop a novel model to produce probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
 
@@
 We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
-Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
+Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and  a simple empirical distribution of the arrivals;
 
@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
 
-The rest of the paper is organised as follows: In section \@ref(lit), we provide a brief overview of hourly ED arrival forecasting; In Section \@ref(data), we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
+The rest of the paper is organised as following: In section \@ref(lit), we provide a brief overview of hourly ED arrival forecasting; In Section \@ref(data), we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
 In section \@ref(model), we describe the modelling approach and benchmark methods.
@@
 
-Some studies employ models to forecast admissions and arrivals in Emergency Department.
+There are some studies that employ models to forecast admissions and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only discuss hourly ED forecasting.
+However, given the focus of the paper, we only discuss the hourly ED forecasting.
 
-@hertzum2017forecasting used linear regression, ARIMA, and Naïve to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
+Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
 @hertzum2017forecasting study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
-In terms of the hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day, while weekends are the quietest ones.
-July-August are the months with the highest number of patient arrivals, while January and February have the lowest numbers.
+In terms of hour of the day, patient arrivals peaked around noon.
+For days of the week, Monday is the busiest day while weekends are the quietest ones.
+July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
 They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
+<!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval.
+Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.
+-->
 
-@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neural network methods to forecast hourly accident and emergency department arrivals.
+@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
 ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
-@Cheng2021 developed an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which outperformed a rolling-average benchmark.
-They also produce prediction intervals, a form of the probabilistic forecast, which were found to be well-calibrated, a necessary property for such forecasts.
+Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
+@Cheng2021 develop an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which is found to outperform a rolling-average benchmark.
+They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.
 
 @morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for a horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend and two seasonal cycles: a within-day and a within-week seasonal cycles.
+Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
+The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
 The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
 
-@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day-of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
+@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
 The authors produced probabilistic predictions in the form of $50\%$ and $90\%$ prediction intervals for the number of hourly arrivals.
-Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it was suggested that the model could be used for forecasting.
-However, model evaluation was performed in-sample on only one year of data, so it is unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
-However, the length of the time series in this study was very short (only one year), which did not allow for a rigorous out-of-sample evaluation.
+Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
+However, the length of time series is very short (only one year), which does not allow for a rigorous out-of-sample evaluation.
 
-@schweigler2009forecasting investigated whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
+@schweigler2009forecasting conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
 A year-long dataset of hourly ED bed occupancy was collected from three facilities.
-The authors implemented an hourly historical average model, SARIMA model, and sinusoidal model with autocorrelated error for each facility.
-The historical average model was based on the mean occupancy for each site, for each hour of the day, while the sinusoidal model was based on four parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizons using RMSE.
-They found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by the historical average).
+For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
+The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).
 
-@kim2014predicting compared different univariate and multivariate time series forecasting techniques to predict patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and  Vector Autoregressive (VAR) models to forecast for 4 and 24 hours ahead.
+@kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and  Vector Autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
 They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
-The authors found that the ARIMA outperformed all the other models.
+ARIMA model outperformed all the other models.
 
@@
 @asheim2019real developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
-Measuring the accuracy using the MAPE, @asheim2019real noticed that significant improvement happened when the time of notification was incorporated into the model, especially in the one-hour horizon.
+Measuring the accuracy using the MAPE, @asheim2019real noticed that great improvement happened when time of notification was incorporated into the model, especially in one-hour horizon.
 
-@cheng2021forecasting used one year of ED visits time series to evaluate the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
+@cheng2021forecasting used one year of ED visits time series to evaluate the performance Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
 The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of $80\%$ and $95\%$.
-They show that SARIMAX provides a more accurate forecast of hourly ED occupancy.
+They show that SARIMAX provides more accurately forecast of hourly ED occupancy.
 
-According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established.
-Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low [@boyle2012predicting; @hertzum2017forecasting], while others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
+According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
+Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily [@boyle2012predicting; @hertzum2017forecasting].
+While, others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
 
-There are a few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches:
+There are few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches. These limitations are summarised as follows:
 
 (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
-(ii) Almost all research studies produce point forecasts and, at best, report prediction intervals.
-There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent the uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
+(ii) Almost all research studies produce point forecasts and at best report prediction intervals.
+There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
 (iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
 (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
-(v) some studies in this area lack a rigorous experimental design, i.e. they do not use benchmark methods or report forecast accuracy.
+(v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method or forecast accuracy is not reported.
 
@@
 
-```{r tab-results, out.width= "50%"}
+```{r tab-results, out.width= "50%", cache=FALSE}
 results_table <- read_rds("results_table.rds")

### 2022-01-13 r_script/result_viz.R

@@ -4,6 +4,7 @@ my_colorblind <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
 PB <- read_rds("../results/PB.rds")
 REL <- read_rds("../results/REL.rds")
 plotdata <- read_rds("../results/plotdata.rds")
+RMSE <- read_rds("../results/RMSE.rds")
 
 selected_method <- c("ADAM-iETSX", # ETS Example
                      "GBM-2",      # ML/non-parametric method
@@ -110,7 +111,6 @@ ggplot(rel_pb_selected,
   theme_few() 
 
 ## ---- rmse
-RMSE <- read_rds("../results/RMSE.rds")
 rmse <- RMSE %>% filter(Horizon=="All") %>%
   group_by(Method) %>% summarise(RMSE=mean(RMSE))
 rmse_selected <- rmse %>% as_tibble() %>% 
@@ -232,9 +232,18 @@ time_long <- results_table_nona %>%
   pivot_longer(cols = 3:5,names_to = "Measure",values_to = "Accuracy")
 
 ggplot(time_long,aes(x=Time,
-                              y=Accuracy))+
-  ggrepel::geom_text_repel(aes(label = Method), size=3)+
+                     y=Accuracy))+
+  geom_point(aes(shape=Method))+
   facet_wrap(vars(Measure), ncol = 1, scales = "free")+
   theme_few()+
-  labs(x="Time (in seconds)")
-  
+  labs(x="Time (in seconds)")+
+  theme(legend.position = "bottom",
+        legend.title = element_text(face = "bold"))
+
+# ggplot(time_long,aes(x=Time,
+#                               y=Accuracy))+
+#   ggrepel::geom_text_repel(aes(label = Method), size=3)+
+#   facet_wrap(vars(Measure), ncol = 1, scales = "free")+
+#   theme_few()+
+#   labs(x="Time (in seconds)")
+

### 2022-01-16 gam-mgcv/study_v1.R
Results check
Check RMSE, remove unused results, add timing section for NBI-2

@@ -760,7 +760,7 @@ for(fold in unique(h2$kfold)){
                        ),
                        ## Consider ba() here, and by dow/dow3
                        sigma.formula = ~ cs(clock_hour,df=12),
-                       family =  NBI(mu.link = "identity"),
+                       family =  NBI(),
                        method=mixed(10,20),
                        control=gamlss.control(c.crit = 0.1))
   
@@ -794,7 +794,7 @@ h2_mqr$expectation[na_index] <- h2_gamlss4_params$mu[na_index]
 class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## Save forecasts for evaluation
-JB_results[["gamlss-NBI_Ilink_v4"]] <- h2_mqr
+JB_results[["gamlss-NBI_v4"]] <- h2_mqr
 
 
 ## Quick plot and evaluation
@@ -882,7 +882,78 @@ pinball(h2_mqr[,grep("q",colnames(h2_mqr)),with=F],h2$n_attendance,kfolds = h2$k
 
 
 
-
+## GAMLSS Gamma ####
+#
+# Error...
+#
+# h2_gamlss5_params <- copy(h2[,.(issueTime,targetTime_UK,kfold)])
+# for(fold in unique(h2$kfold)){
+#   
+#   print(paste(fold,Sys.time()))
+#   
+#   train_data <- na.omit(h2[kfold!=fold & kfold!="Test",
+#                            .(issueTime,targetTime_UK,
+#                              n_attendance,dow,dow3,clock_hour,t,doy,T2T,school_holiday)])
+#   
+#   h2_gamlss1 <- gamlss(data = train_data,
+#                        formula = n_attendance ~ ba(
+#                          ## v1 above:
+#                          # ~ dow + s(clock_hour,k=24,by=dow,bs = "cr") +
+#                          #   s(doy,k=6,by=t,bs = "cr") + te(clock_hour,T2T),
+#                          ## v2 here:
+#                          ~ dow3 + s(clock_hour,k=24,by=dow3,bs = "cr") +
+#                            school_holiday + s(clock_hour,k=6,by=school_holiday,bs = "cr") +
+#                            s(doy,k=6,by=t,bs = "cr") + te(clock_hour,T2T)
+#                        ),
[…]

### 2022-01-16 results/Evaluatoin_Script.R

@@ -197,7 +197,8 @@ OLD_NEW <- data.table(old=c("Benchmark_1","Benchmark_2",
                             "iETSXSeasonal",
                             "ETS(XXX)",
                             "RegressionPoisson",
-                            "iETSCeiling"),
+                            "iETSCeiling",
+                            "gamlss-GA_v2"),
                       new=c("Benchmark-1","Benchmark-2",
                             "Poisson-1",
                             "Poisson-2",
@@ -205,7 +206,7 @@ OLD_NEW <- data.table(old=c("Benchmark_1","Benchmark_2",
                             "NOtr-2",
                             "GBM-2",
                             "Ttr-2",
-                            "NBI-2-log",
+                            "NBI-2",
                             "qreg-1",
                             "Poisson-2-I",
                             "NBI-2",
@@ -214,7 +215,8 @@ OLD_NEW <- data.table(old=c("Benchmark_1","Benchmark_2",
                             "ADAM-iETSX",
                             "ETS",
                             "Regression-Poisson",
-                            "ADAM-iETSX-Ceiling"))
+                            "ADAM-iETSX-Ceiling",
+                            "ZAGA-2"))
 
 change_method_name(REL,OLD_NEW)
 change_method_name(PB,OLD_NEW)
@@ -271,9 +273,11 @@ change_method_name(Timing,OLD_NEW)
 Res_sum <-  merge(Res_sum,Timing,by="Method",all=T)
   
 write.csv(Res_sum,row.names = F,file = "Results_Summary.csv")
+# Res_sum <- fread("Results_Summary.csv")
 
 ## Save table for paper
-saveRDS(Res_sum,file="../paper/Results_Table.rds")
+saveRDS(Res_sum[!Method%in%c("LinearRegression","qreg-1","fasster","Poisson-2-I","ADAM-iETSX-Ceiling"),],
+        file="../paper/Results_Table.rds")
 
 
 ## significance testing ####

### 2022-01-18 gam-mgcv/study_v1_timing.R
Computational times for Jethro

@@ -626,7 +626,7 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
-JB_results_time[["gamlss-NBI_Ilink_v4"]] <- temp_time
+JB_results_time[["gamlss-NBI_Ilink_v4"]] <- time_temp
 ## %%%% Time for Ivan %%%%
 
 
@@ -688,7 +688,7 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
-JB_results_time[["gamlss-PO_Ilink_v1"]] <- temp_time
+JB_results_time[["gamlss-PO_Ilink_v1"]] <- time_temp
 ## %%%% Time for Ivan %%%%
 
 
@@ -747,7 +747,7 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
-JB_results_time[["gamlss-NBI_v4"]] <- temp_time
+JB_results_time[["gamlss-NBI_v4"]] <- time_temp
 ## %%%% Time for Ivan %%%%
 
 

### 2022-01-18 gam-mgcv/study_v1_timing.R
Correct bug throwing error

@@ -620,6 +620,7 @@ for(p in 1:19/20){
 }
 
 # Add expectation
+h2_mqr$expectation <- NA
 h2_mqr$expectation[na_index] <- h2_gamlss4_params$mu[na_index]
 class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 

### 2022-01-18 gam-mgcv/study_v1_timing.R
Bug fix

@@ -627,7 +627,7 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
-JB_results_time[["gamlss-NBI_Ilink_v4"]] <- temp_time
+JB_results_time[["gamlss-NBI_Ilink_v4"]] <- time_temp
 ## %%%% Time for Ivan %%%%
 
 

### 2022-01-18 gam-mgcv/study_v1_timing.R
Update study_v1_timing.R

@@ -593,7 +593,7 @@ for(fold in "Test"){
                        ),
                        ## Consider ba() here, and by dow/dow3
                        sigma.formula = ~ cs(clock_hour,df=12),
-                       family =  NBI(mu.link = "identity"),
+                       family =  NBI(),
                        method=mixed(10,20),
                        control=gamlss.control(c.crit = 0.1))
   

### 2022-01-18 gam-mgcv/study_v1_timing.R
Update study_v1_timing.R

@@ -61,6 +61,7 @@ mean_from_qs <- function(mqr){
   qs <- as.numeric(gsub("q","",colnames(mqr)[q_cols]))/100
   diff_left <- diff(c(0,qs))
   
+  mqr$expectation <- NA
   mqr$expectation <- mqr[[q_cols[1]]]*diff_left[1]
   for(qq in 2:length(q_cols)){
     mqr$expectation <- mqr$expectation + mqr[[q_cols[qq]]]*diff_left[qq]

### 2022-01-18 gam-mgcv/study_v1_timing.R
Update study_v1_timing.R

@@ -252,6 +252,7 @@ class(Bench_mqr) <- c("MultiQR",class(Bench_mqr))
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["Benchmark_2"]] <- time_temp
+rm(Bench_mqr,Bench_mqr_temp,BenchPred,BenchPred_exp)
 ## %%%% Time for Ivan %%%%
 
 
@@ -295,6 +296,8 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["Poisson-GAM-te_v1"]] <- time_temp
+rm(gam1,h2_mqr)
+gc()
 ## %%%% Time for Ivan %%%%
 
 
@@ -339,6 +342,9 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["Poisson-GAM-te_v2"]] <- time_temp
+rm(gam2,h2_mqr)
+gc()
+
 ## %%%% Time for Ivan %%%%
 
 
@@ -410,6 +416,9 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["gamlss-NOtr_v1"]] <- time_temp
+rm(h2_gamlss1,h2_mqr,train_data)
+gc()
+
 ## %%%% Time for Ivan %%%%
 
 
@@ -483,6 +492,9 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["gamlss-NOtr_v2"]] <- time_temp
+rm(h2_gamlss2,h2_mqr,train_data)
+gc()
+
 ## %%%% Time for Ivan %%%%
 
 
@@ -559,6 +571,9 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["gamlss-TF2tr_v3"]] <- time_temp
[…]

### 2022-01-19 gam-mgcv/study_v1_timing.R
Fix but in timing

@@ -61,7 +61,7 @@ mean_from_qs <- function(mqr){
   qs <- as.numeric(gsub("q","",colnames(mqr)[q_cols]))/100
   diff_left <- diff(c(0,qs))
   
-  mqr$expectation <- NA
+  mqr$expectation <- as.numeric(NA)
   mqr$expectation <- mqr[[q_cols[1]]]*diff_left[1]
   for(qq in 2:length(q_cols)){
     mqr$expectation <- mqr$expectation + mqr[[q_cols[qq]]]*diff_left[qq]
@@ -813,7 +813,7 @@ expectation <- mean_from_qs(mqr = cbind(h2[kfold!="Test",.(issueTime,targetTime_
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
 JB_results_time[["GBM"]] <- time_temp
-rm(h2_gbm_mqr)
+rm(h2_gbm_mqr,expectation)
 gc()
 
 ## %%%% Time for Ivan %%%%
@@ -871,7 +871,7 @@ for(Ver in 1:2){
                                      bbs(doy,knots=6,by=t,df=6) +
                                      bbs(clock_hour,T2T,knots = 6,df=6)},
                                quantiles = seq(0.05,0.95,by=0.05),
-                               cv_folds = NULL,
+                               # cv_folds = NULL,
                                cores = detectCores()-1,
                                pckgs = c("data.table"),
                                sort=T,
@@ -879,13 +879,13 @@ for(Ver in 1:2){
                                control = mboost::boost_control(mstop = 500,nu=0.1))
   
   
-  
-  expectation <- mean_from_qs(mqr = cbind(h2[kfold!="Test",.(issueTime,targetTime_UK)],h2_mboost_mqr$mqr_pred))
+  expectation <- mean_from_qs(mqr = cbind(h2[kfold!="Test",.(issueTime,targetTime_UK)],
+                                          predict(h2_mboost_mqr,newdata=h2[!is.na(T2T) & kfold=="Test",][1:24])))
   
   ## %%%% Time for Ivan %%%%
   time_temp <- Sys.time() - time_temp
   JB_results_time[[paste0("qreg_boost_V",Ver)]] <- time_temp
-  rm(h2_mboost_mqr)
+  rm(h2_mboost_mqr,expectation)
   gc()
   
   ## %%%% Time for Ivan %%%%

### 2022-02-08 gam-mgcv/study_v1_timing.R
Update study_v1_timing.R
Remove conflict text "<<<<HEAD" etc.

@@ -582,76 +582,6 @@ gc()
 # Approx 10 min per fold
 #
 
-h2_gamlss4_params <- copy(h2[,.(issueTime,targetTime_UK,kfold)])
-
-## %%%% Time for Ivan %%%%
-time_temp <- Sys.time() 
-## %%%% Time for Ivan %%%%
-
-
-for(fold in "Test"){
-  
-  print(paste(fold,Sys.time()))
-  
-  train_data <- na.omit(h2[kfold!=fold & kfold!="Test",
-                           .(issueTime,targetTime_UK,
-                             n_attendance,dow,dow3,clock_hour,t,doy,T2T,school_holiday)])
-  
-  h2_gamlss4 <- gamlss(data = train_data,
-                       formula = n_attendance ~ ba(
-                         ## v1 above:
-                         # ~ dow + s(clock_hour,k=24,by=dow,bs = "cr") +
-                         #   s(doy,k=6,by=t,bs = "cr") + te(clock_hour,T2T),
-                         ## v2 here:
-                         ~ dow3 + s(clock_hour,k=24,by=dow3,bs = "cr") +
-                           school_holiday + s(clock_hour,k=12,by=school_holiday,bs = "cr") +
-                           s(doy,k=6,by=t,bs = "cr") + te(clock_hour,T2T)
-                       ),
-                       ## Consider ba() here, and by dow/dow3
-                       sigma.formula = ~ cs(clock_hour,df=12),
-                       family =  NBI(),
-                       method=mixed(10,20),
-                       control=gamlss.control(c.crit = 0.1))
-  
-  test_data <- h2[kfold==fold,.(issueTime,targetTime_UK,
-                                n_attendance,dow,dow3,clock_hour,t,doy,T2T,school_holiday)]
-  
-  temp_params <- predictAll(h2_gamlss4,newdata = test_data,
-                            data = train_data)
-  h2_gamlss4_params[kfold==fold,names(temp_params) := temp_params] 
-  
-}; rm(temp_params,test_data)
-
-
-## Form quantile forecasts
[…]

### 2022-02-08 gam-mgcv/study_v1_timing.R
Fixed typos

@@ -641,7 +641,7 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
-JB_results_time[["gamlss-PO_Ilink_v1"]] <- temp_time
+JB_results_time[["gamlss-PO_Ilink_v1"]] <- time_temp
 rm(h2_gamlss1,h2_mqr,train_data,h2_gamlss1_params)
 gc()
 ## %%%% Time for Ivan %%%%
@@ -702,7 +702,7 @@ class(h2_mqr) <- c("MultiQR",class(h2_mqr))
 
 ## %%%% Time for Ivan %%%%
 time_temp <- Sys.time() - time_temp
-JB_results_time[["gamlss-NBI_v4"]] <- temp_time
+JB_results_time[["gamlss-NBI_v4"]] <- time_temp
 rm(h2_gamlss4,h2_mqr,train_data,h2_gamlss4_params)
 gc()
 ## %%%% Time for Ivan %%%%

### 2022-02-09 paper/paper.Rmd
Update fig captions and remove figs 2 and 7

@@
 
-```{r hourly-plot-ridge, fig.cap= "Distrubution of admission per hour and day of the week", fig.align='center',fig.width= 6.5,fig.height=5.5}
+```{r hourly-plot-ridge, fig.cap= "Distrubution of admission by hour-of-day and day-of-week.
+Most days have a distinct pattern, such as relatively high arrivals on Monday mornings and in the early hours of Saturday and Sunday morning.", fig.align='center',fig.width= 6.5,fig.height=5.5}
 ```
@@
 
-```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
-```
+<!--- ```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
+``` --->
 
@@
 
-```{r seasonplot-weekofyear, fig.cap= "Week of year arrivals", fig.align='center',out.width="70%"}
+```{r seasonplot-weekofyear, fig.cap= "Arrivals by week-of-year. There is an annual trend of reduced arrivals durnng the summer and winter holiday period.", fig.align='center',out.width="70%"}
 ```
@@
 
-```{r date-plot, fig.cap= "Daily arrivals", fig.align='center',out.width="70%"}
+```{r date-plot, fig.cap= "Daily arrivals over the entire dataset with moving average trend (blue line) and dates of outliers labled.
+Low outliers invariables correspond to public holidays.", fig.align='center',out.width="70%"}
 ```
@@
 
-```{r Pinball, fig.cap= "Pinball loss values over different quantiles.", fig.align='center',out.width= "90%"}
+```{r Pinball, fig.cap= "Pinball loss values by quantile.
+Benchmark-2, NBI-2 and NOtr-2 have similar performance with most improvement of the latter two over the benchmark comming from the lower quantiles.
+All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.", fig.align='center',out.width= "90%"}
 ```
@@
 
-```{r quantile-bias, fig.cap= "Quantile bias vs the nominal quantiles.", fig.align='center',out.width= "90%"}
+```{r quantile-bias, fig.cap= "Quantile bias illustrates the difference between the nominal exceedance of prediceted quantiles and the observed frequency of exceedance.
+If a predictive model is calibrated, quantile bias should be approximately zero accross all probability levels.
+Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2" bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.", fig.align='center',out.width= "90%"}
 ```
@@
 
+<!--
 ```{r Skill-rel2bench-reduced, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "70%"}
@@
 
-```{r lead-time-rmse, fig.cap= "RMSE values over the forecast horizon.", fig.align='center',out.width= "70%"}
+```{r lead-time-rmse, fig.cap= "RMSE values over the forecast horizon for forecasts issues at midnight.
+Attendance during the morning is more challenging to predice resulting in greater RMSE during these hours for all models.", fig.align='center',out.width= "70%"}
 ```
@@
 
-```{r time-accuracy, fig.cap= "Running time vs. forecast accuracy", fig.align='center',out.width= "70%"}
+```{r time-accuracy, fig.cap= "Running time vs. forecast performance", fig.align='center',out.width= "70%"}
 ```

### 2022-02-10 paper/paper.Rmd
address comments raised by Ivan

--- 2022-01-18 paper/paper.Rmd
@@
 ---
-title:  Forecasting short-term hourly Emergency Department arrivals
+title:  Probabilistic forecasting of hourly Emergency Department arrivals
 author:
@@
 
-Furthermore, we consider an extnsion to the additive models for $\lambdat_t$ and $\mu_t$ above to incorporate school and public holidays into $D$.
+Furthermore, we consider an extnsion to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$.
 These models are labelled Poisson-2 and NOtr-2.
@@
 
-```{r tab-results, out.width= "50%", cache=FALSE}
-results_table <- read_rds("results_table.rds")
-knitr::kable(results_table, booktabs = T, 
-             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
+```{r Skill-rel2bench, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "100%"}
 ```
 
-<!-- Why NAs for RMSE?? -->
-
-Evaluation metrics from the test period are presented in Table \@ref(tab:tab-results).
-They are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
-
-```{r Skill-rel2bench-reduced, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "70%"}
-```
-
- Figure \@ref(fig:Skill-rel2bench-reduced) illustrates the trade-off between calibration and pinball skill. <!-- We may need to expand this-->
+ Figure \@ref(fig:Skill-rel2bench) illustrates the trade-off between calibration and pinball skill. <!-- We may need to expand this-->
 
@@
 
+```{r tab-results, out.width= "50%", cache=FALSE}
+results_table <- read_rds("results_table.rds") %>% arrange(`Quantile Bias`)
+results_table <- results_table %>% 
+  mutate(Method = 
+           case_when(Method == "tbats" ~ "TBATS",
+                     Method == "prophet" ~ "Prophet",
+                     TRUE ~ Method))
+knitr::kable(results_table, booktabs = T, 
+             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =9)
+```
+
+<!-- Why NAs for RMSE?? -->
+
+In above figures, we only present the performance of the top seven methods.
+However, we have evaluated the performance of 14 methods.
+Evaluation metrics for all methods from the test period are presented in Table \@ref(tab:tab-results).
+They are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
+
 One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones (Table \@ref(tab:tab-results)).
@@
 
-```{r time-accuracy, fig.cap= "Running time vs. forecast accuracy", fig.align='center',out.width= "70%"}
+```{r time-accuracy, fig.cap= "Speed (Running time) vs. forecast accuracy", fig.align='center',out.width="70%"}
 ```
 
-Finally, Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+Finally, Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+The x-axis represents the speed of each method as slow, moderate or fast.
 <!-- We need to explain this once we have th result for RMSE -->

### 2022-02-10 paper/paper.Rmd

@@
 
-```{r Skill-rel2bench, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "100%"}
+<!--
+```{r Skill-rel2bench-reduced, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "70%"}
 ```
@@
 
-```{r time-accuracy, fig.cap= "Speed (Running time) vs. forecast accuracy", fig.align='center',out.width="70%"}
+```{r time-accuracy, fig.cap= "Running time vs. forecast performance", fig.align='center',out.width= "70%"}
 ```

### 2022-02-10 paper/paper.tex

--- 2022-01-12 paper/paper.tex
@@
             pdfauthor={},
-            pdftitle={Forecasting short-term hourly Emergency Department arrivals},
+            pdftitle={Probabilistic forecasting of hourly Emergency Department arrivals},
             colorlinks=false,
@@
 
-  \title{Forecasting short-term hourly Emergency Department arrivals}
+  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
     \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
@@
   \begin{abstract}
-  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients' demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required.
+  An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients' demand.
+It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units.
 This can have many advantages for healthcare staff and the quality of care delivered to patients.
-In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals.
-We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution.
-We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score.
-Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts.
-Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.
+In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals.
+We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution.
+We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score.
+Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts.
+Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
   \end{abstract}
@@
 Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
-An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing.
-This is one of the best ways to optimize resources utilisation and minimise related costs.
+Accurate ED demand forecasts contribute to a better decision-making process regarding resources allocation and staffing.
+This is one of the best ways to optimise resources utilisation and minimise related costs.
 An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow \citep{derlet2002overcrowding}.
-Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other \citep{rostami2020anticipating}.
+Inability to match the staff with the demand might result in patients overcrowding the system, which is a severe problem that causes challenges for the patient flow \citep{derlet2002overcrowding}.
+Also, it is related to increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other \citep{rostami2020anticipating}.
 
-An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
+Accurate forecasting of arrivals by the hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
 This has many advantages for both patients, staff and the quality of provided services.
 Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision making related to the execution of the delivery process in ED.
-The combination of an hourly arrival forecast, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service.
-Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
-However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly.
-They may also express nonstationarity and their profile may change over time.
-Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
+This involves the short-term decision-making related to the execution of the delivery process in ED.
+The combination of an hourly arrival forecast, current staff, being occupied, resource availability and waiting times at ED provide information on the state of the unscheduled care system across the service.
+Having this complete picture enables the delivery managers to focus on the areas that require intervention to allow the most effective delivery of the service to the patients.
+However, compared with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
+Hourly time series generally exhibit multiple seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
+They may also express nonstationarity, and their profile may change over time.
+Therefore, an appropriate forecasting model should consider these features to accurately predict hourly demand admissions.
 
 There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
-These studies use multiple approaches including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}.
-However, majority of these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-There are few studies that report uncertainty by presenting prediction intervals, no study generates and evaluates the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels, if staff levels are insufficient.
+These studies use multiple approaches, including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}.
+However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
 Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
-Furthermore, if the impact of under and over staffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under and over staffing to be calculated.
+
+Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated.
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-Another drawback of existing studies is that the datasets used are relatively small (e.g.~time period of 1-2 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable.
-Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals.
+Another drawback of existing studies is that the datasets used are relatively small (e.g.~time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
+Such results might not be generalisable.
+Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals.
 Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
@@
 \item
-  We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
+  We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
 \item
-  We develop an advanced dynamic model to forecasts ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies which produced highly-accurate point forecasts;
+  We develop an advanced dynamic model to forecast ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
 \item
-  We develop a novel model to produce probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
+  We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
 \item
-  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and a simple empirical distribution of the arrivals;
+  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
 \item
   We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.
 \end{enumerate}
 
-The rest of the paper is organised as following: In section \ref{lit}, we provide a brief overview of hourly ED arrival forecasting; In Section \ref{data}, we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
+The rest of the paper is organised as follows: In section \ref{lit}, we provide a brief overview of hourly ED arrival forecasting; In Section \ref{data}, we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
 In section \ref{model}, we describe the modelling approach and benchmark methods.
@@
 
-There are some studies that employ models to forecast admissions and arrivals in Emergency Department.
+Some studies employ models to forecast admissions and arrivals in Emergency Department.
 The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only discuss the hourly ED forecasting.
+However, given the focus of the paper, we only discuss hourly ED forecasting.
 
-Linear regression, ARIMA, and Naïve were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
+\citet{hertzum2017forecasting} used linear regression, ARIMA, and Naïve to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
 \citet{hertzum2017forecasting} study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
-In terms of hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day while weekends are the quietest ones.
-July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers.
+In terms of the hour of the day, patient arrivals peaked around noon.
+For days of the week, Monday is the busiest day, while weekends are the quietest ones.
+July-August are the months with the highest number of patient arrivals, while January and February have the lowest numbers.
 They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
 
-\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals.
+\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neural network methods to forecast hourly accident and emergency department arrivals.
 ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
-\citet{Cheng2021} develop an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which is found to outperform a rolling-average benchmark.
-They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.
+Authors claimed that ARIMA provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
+\citet{Cheng2021} developed an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which outperformed a rolling-average benchmark.
+They also produce prediction intervals, a form of the probabilistic forecast, which were found to be well-calibrated, a necessary property for such forecasts.
 
 \citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles.
+Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for a horizon of 168 hours.
+The hourly data collected from an ED in Pennsylvania showed no trend and two seasonal cycles: a within-day and a within-week seasonal cycles.
 The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
 
-\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
+\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day-of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
 The authors produced probabilistic predictions in the form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly arrivals.
-Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
-However, the length of time series is very short (only one year), which does not allow for a rigorous out-of-sample evaluation.
+Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it was suggested that the model could be used for forecasting.
+However, model evaluation was performed in-sample on only one year of data, so it is unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
+However, the length of the time series in this study was very short (only one year), which did not allow for a rigorous out-of-sample evaluation.
 
-\citet{schweigler2009forecasting} conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
+\citet{schweigler2009forecasting} investigated whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
 A year-long dataset of hourly ED bed occupancy was collected from three facilities.
-For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error.
-The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by historical average).
+The authors implemented an hourly historical average model, SARIMA model, and sinusoidal model with autocorrelated error for each facility.
+The historical average model was based on the mean occupancy for each site, for each hour of the day, while the sinusoidal model was based on four parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
+They evaluated the forecast accuracy of four and twelve hours forecast horizons using RMSE.
+They found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by the historical average).
 
-\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead.
+\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to predict patient volume for a Hospital Medicine programme.
+The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to forecast for 4 and 24 hours ahead.
 They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
-ARIMA model outperformed all the other models.
+The authors found that the ARIMA outperformed all the other models.
 
@@
 \citet{asheim2019real} developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
-Measuring the accuracy using the MAPE, \citet{asheim2019real} noticed that great improvement happened when time of notification was incorporated into the model, especially in one-hour horizon.
+Measuring the accuracy using the MAPE, \citet{asheim2019real} noticed that significant improvement happened when the time of notification was incorporated into the model, especially in the one-hour horizon.
 
-\citet{cheng2021forecasting} used one year of ED visits time series to evaluate the performance Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
+\citet{cheng2021forecasting} used one year of ED visits time series to evaluate the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
 The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of \(80\%\) and \(95\%\).
-They show that SARIMAX provides more accurately forecast of hourly ED occupancy.
+They show that SARIMAX provides a more accurate forecast of hourly ED occupancy.
 
-According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established.
-Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily \citep{boyle2012predicting, hertzum2017forecasting}.
-While, others mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
+According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established.
+Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low \citep{boyle2012predicting, hertzum2017forecasting}, while others mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
 
-There are few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches. These limitations are summarised as follows:
+There are a few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches:
 
@@
 \item
-  Almost all research studies produce point forecasts and at best report prediction intervals.
-There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
+  Almost all research studies produce point forecasts and, at best, report prediction intervals.
+There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent the uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
 \item
@@
 \item
-  some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method or forecast accuracy is not reported.
+  some studies in this area lack a rigorous experimental design, i.e.~they do not use benchmark methods or report forecast accuracy.
 \end{enumerate}
@@
 It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by \citet{hertzum2017forecasting}.
-The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again.
+The ED arrivals decrease between midnight and early morning and then increase until the evening, decreasing after that again.
 It is also clear that ED service gets systematically more visits on Mondays between 8 a.m.
 and 5 p.m.
-Moreover, the number of arrivals around mid-night is slightly higher for Saturday and Sunday.
+Moreover, the number of arrivals around midnight is slightly higher for Saturdays and Sundays.
 
-Figure \ref{fig:hourly-plot-ridge} also highlights that there is significant skewness for almost every hour of the day that varies with time-of-day, and which should be accounted for in forecasting methods.
+Figure \ref{fig:hourly-plot-ridge} also highlights significant skewness for almost every hour of the day that varies with time-of-day, which should be accounted for in forecasting methods.
 Some skewness might be related to holidays and special events.
-It is also clear that arrivals are less volatile between mid-night and early morning.
+It is also clear that arrivals are less volatile between midnight and early morning.
 
@@
 
-Figure \ref{fig:seasonplot-weekofyear} illustrates the daily seasonal plot, with x-axis representing the day-of-week and y-axis the ED arrivals.
-There is a line for each week (i.e.~7 days) from Monday to Saturday, and lines are color-coded by week of year to show the weekly cycle.
-It is clear that there is a large jump in patient arrivals on Monday followed by Saturday each week.
-This shows that there are significantly more arrivals on Mondays and Saturdays comparign to the rest of the week.
+Figure \ref{fig:seasonplot-weekofyear} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
+There is a line for each week (i.e.~seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle.
+It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week.
+This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week.
 This might be due to the closure of General Practitioners outpatient clinics over the weekend.
@@
 Figure \ref{fig:seasonplot-weekofyear} highlights the week of year seasonality in the ED arrivals.
-We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period.
-Moreover, the number of arrivals are lower at beginning and the end of the year.
-The highest arrivals are in week 39-42.
+We observe that arrivals are significantly lower from week 29 to 35, corresponding to the Summer period.
+Moreover, the number of arrivals is lower at the beginning and the end of the year.
+The arrivals increase from week 36 and peak in weeks 39-42.
+This corresponds to the September - October period.
 
@@
 
-Figure \ref{fig:date-plot} presents the time plot of daily arrivals.
+Finally, Figure \ref{fig:date-plot} presents the time plot of daily arrivals.
 Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
-The figure shows more arrivals on Mondays compared to other weekdays.
-Moreover, we can observe a long-term trend line as well as the significant effect of some holidays.
+The figure shows more arrivals on Mondays than on other weekdays, agreeing with the previous findings.
+Moreover, we can observe a long-term trend line and the significant effect of some holidays.
 We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
-Moreover, Figure \ref{fig:date-plot} allows us to identify the impact of special event on arrivals.
-For instance, we see that the number of arrivals are significantly low for 01-03 of March 2018.
-These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK \citep{stormemma2018}.
+Moreover, Figure \ref{fig:date-plot} allows us to identify the impact of special events on arrivals.
+For instance, we see that arrivals are significantly low for 01-03 of March 2018.
+These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages, and schools closed in the UK \citep{stormemma2018}.
 
-Based on ED time series analysis illustrated above and literature review, we should consider models that can take the following into account:
+Based on this analysis and the literature review, we should consider models that can take the following into account:
 
 \begin{itemize}
+\tightlist
 \item
-  Daily, weekly, and annual seasonalities,
+  Hour-of-day, day-of-week, and week-of-year seasonalities,
 \item
-  Long-term trend,
+  Long-term trend (or a slowly changing level),
 \item
@@
 \item
-  Weather effects, such as temperature and precipitation.
+  Temperature effects.
 \end{itemize}
@@
 
-We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
+We start with one of the simplest forecasting approaches used in practice - assuming that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
 This is called ``Naïve''.
 In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution \citep{la2021new}.
+We consider the empirical distribution of all available historic data (Benchmark-1) and the empirical distribution of the most recent year of historic data on a rolling basis (Benchmark-2) to capture potential changes in behaviour over time.
 
@@
 
-Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals).
+Regression is one of the most popular forecasting methods that use explanatory variables to predict a variable of interest (in our case, the ED arrivals).
 The classical linear regression model is formulated as
-
 \begin{equation}
@@
 \end{equation}
-
 where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters, \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and \(t\) is the time index.
-However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative.
+However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic because the number of admitted patients is an integer and non-negative.
 So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression \citep[see for example,][]{mccarthy2008challenge}, which can be summarised as
-
+One of the models that is frequently used in practice is the Poisson regression \citep[see, for example,][]{mccarthy2008challenge}, which can be summarised as
 \begin{equation}
@@
 
-The logarithm in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive.
+The logarithm in \eqref{eq:PoissonRegression} is needed to ensure that the parameter of Poisson distribution is always positive.
 This model can be estimated via maximisation of the likelihood function based on Poisson mass function .
-When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case.
+There is no single correct answer when selecting explanatory variables for the model, and the decision needs to be made for each specific case.
 In our experiment, we will only include dummy variables, capturing a variety of calendar events:
@@
 \item
-  Hour of day,
+  Hour of the day,
 \item
-  Day of week,
+  Day of the week,
 \item
-  Week of year,
+  Week of the year,
 \item
-  Holidays (such as Christmas, New Year etc),
+  Holidays (such as Christmas, New Year etc.),
 \item
@@
 The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
-Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g.~people might try not to go to hospital on Christmas eve and thus will go the next day).
-This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \eqref{eq:PoissonRegression} introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time.
-We use \texttt{alm()} function from \texttt{greybox} package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.
+Finally, (5) is needed to capture the potential phenomenon of change in admittance after the holiday (e.g.~people might try not to go to the hospital on Christmas eve and thus go the next day).
+This model assumes that all these effects are deterministic and do not change over time.
+Still, the exponentiation in \eqref{eq:PoissonRegression} introduces an interaction effect between dummy variables so that the 3 pm on Monday in January will be different from 3 pm on Monday in July, although the parameters for the hour of day and day of the week are fixed and do not change over time.
+We use the \texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.
 
@@
 
-\citet{Hyndman2008b} developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus \emph{ETS}).
-Given the popularity of ETS model in forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark.
-This was done using \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denote as \emph{ETS}.
+\citet{Hyndman2008b} developed a state space approach for exponential smoothing models.
+The model can have a set of elements, including different types of Error, Trend and Seasonal components (thus \emph{ETS}).
+Given the popularity of the ETS model in the forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with a frequency of 24 (hour of the day) as a benchmark.
+This was done using the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denoted as \emph{ETS}.
 This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
-This model is included as a benchmark, only to see how the other models perform in comparison with it.
+This model is included as a benchmark, only to see how the other models perform compared to it.
 
@@
 Prophet is a forecasting procedure created by Facebook \citep{taylor2018forecasting} that accounts for multiple seasonality, piecewise trend and holiday effects.
-Prophet is robust to missing data and shifts in the trend, and typically handles outliers well.
+Prophet is robust to missing data and shifts in the trend and typically handles outliers well.
 Prophet works well on daily data seen in Facebook.
@@
 The implementation may be less flexible than other methods.
-The model itself relies on Multiple Source of Error state space model, originally proposed by \citet{Kalman1960}.
-The model is incorporated using corresponding implementation of the Fable package in R.
+The model itself relies on the Multiple Source of Error state space model, initially proposed by \citet{Kalman1960}.
+The model is incorporated using the corresponding implementation of the Fable package in R.
 We use the \texttt{prophet()} function from the \texttt{fable} package \citep{fable2020}.
@@
 It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
-The latter not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high.
+The latter gives the model more flexibility to deal with fractional seasonality and reduces the parameters of the model when the frequencies of seasonalities are high.
 We fit a TBATS model using the \texttt{tbats()} function from the \texttt{forecast} package in R \citep{forecastpackage2020}.
+
+\hypertarget{quantile-regression-and-gradient-boosting-machines}{%
+\subsection{Quantile Regression and Gradient Boosting Machines}\label{quantile-regression-and-gradient-boosting-machines}}
+
+Quantile regression allows the production of density forecasts without assuming a fixed distributions shape controlled by a small number of parameters, such as the Poisson distribution with parameter \(\lambda\).
+By producing forecasts of multiple quantiles the full predictive distribution can be constructed.
+Quantile regression is particularly useful where data do not follow a simple distribution, of the distribution shape changes over time of with some covariate, which may be the case with ED arrivals.
+
+Gradient Boosting Machines (GBM) are a tree-based machine learning model for regression and classification.
+Here we produce probabilistic forecasts of ED arrivals via multiple quantile regression using GBMs as implemented in the R package \texttt{gbm} \citep{Greenwell2020}.
+GBMs are a best-in-class algorithm for similar regression problems characterised by modest volumes of training data and possible interactions between input features, and are therefore an appealing choice for ED arrival forecasting.
+Here GBMs are fit with the following features: hour of the day, day of the week, school holiday, temperature and day of the year.
+Hyperparameters are chosen via grid search on the training data, and were chosen as follows: 500 trees, an interaction depth of 2, and a shrinkage 0.1.
+An advantage of GBMs is their ability to learn interactions in comparison to the additive models presented later which require the user to specify possible interactions between inputs.
+Here GBM is included as a reference for the performance of `out-of-the-box' machine learning models applied with minimal effort.
+We also explored the possibility of performing linear regression with additive models, but these did not perform as well as GBM so are omited for brevity.
 
@@
 This framework encompasses ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
-Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays.
-This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
-Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by \citet{Svetunkov2019a} to treat those values.
+Based on this framework, we use the ETS(M,N,M) model with frequencies 24 (hour of the day) and 168 (hour of the week), adding dummy variables for the week of the year, holidays and lagged holidays.
+This way, we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
+Given that the data exhibits randomly occurring zeroes, we use the direct probability model of \citet{Svetunkov2019a} to treat those values.
+Finally, given the skewness of the empirical distribution observed in the preliminary analysis, we use the Gamma distribution for the error term.
 This model can be formulated as a set of the following equations:
-
 \begin{equation}
@@
 \end{equation}
-
 where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
@@
 \citet{Svetunkov2019a} discuss how to estimate such a model.
-We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components).
+We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic nature of the model (level and seasonal components).
 Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them (see \ref{quantilesceiling} for the explanation).
-This model is implemented in \texttt{adam()} function from \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM-iETSX''.
+If integer-valued quantiles are needed, then rounding up can be done for them (see Appendix \ref{quantilesceiling} for the explanation).
+This model is implemented in the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM-iETSX''.
 
@@
 In a GAMLSS framework of \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled as
-
 \begin{equation}
@@
 \end{equation}
-
 where \(g_j\) is a monotonic link function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\) \textbf{is \ldots{}}.
@@
 Each \(f_{j,i}\) is a smooth function, constructed as
-
 \begin{equation}
@@
 \end{equation}
-
 where \(b^{ji}_k\) are spline basis functions of dimension \(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients.
@@
 As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form
-
 \begin{equation}
@@
 The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time \(t\), respectively, and \(C_t\) is the temperature at time \(t\).
+This model is called Poisson-1 in discussions below.
 
@@
 \end{equation}
+with additive models
+\begin{align*}
+    & \mu_t = \sum_{i=1}^{10} \beta_i D^{+}_i(t) + \sum_{j=1}^{10} D^{+}_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
+    & log(\sigma_t) = \sum_{i=1}^{10} D^{+}_i(t) f(H(t))
+\end{align*}
+for the mean and variance parameters. This model is referred to as NOtr-1 below.
 
-with additive models
-
-\begin{align*}
-    & \mu_t = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \\
-    & log(\sigma_t) = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
-\end{align*}
-
-for the mean and variance parameters.
-We also performed experiments with the truncated \(t\) distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.
+Furthermore, we consider an extnsion to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\).
+These models are labelled Poisson-2 and NOtr-2.
+We also performed experiments with the truncated \(t\) distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.
 
@@
 
-In order to assess performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps if 0.05, and conditional expectations for 0 to 48 hours ahead produced by each model.
-We forecast upto 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
+In order to assess the performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps of 0.05 and conditional expectations for 0 to 48 hours ahead produced by each model.
+We forecast up to 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
 The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000}, resulting in 727 origins.
-Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation.
-The latter is measured via Root Mean Squared Error (RMSE)
-
+Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and expectation.
+The latter is measured via Root Mean Squared Error (RMSE):
 \begin{equation}
@@
 \end{equation}
-
 where \(h\) is the forecast horizon and \(e_{t+j}\) is the point forecast error \(j\) steps ahead.
@@
 The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated \citep{gneiting2007probabilistic}.
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.
+A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision-makers provided the forecast is calibrated.
 Calibration, also called reliability, is the property that forecast probabilities match the observed frequency of realisations.
@@
 This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observed frequency mean (\(\mathbf{1}(y_{t}\leq q_{\alpha,t})\)
-). We use several scores to assess the quantile performance of models.
+Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observed frequency mean (\(\mathbf{1}(y_{t}\leq q_{\alpha,t})\)).
+We use several scores to assess the quantile performance of models.
 
-First, in order to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score \citep{hyndman2021forecasting}.
-It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately.
-Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model.
+First, to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score \citep{hyndman2021forecasting}.
+It rewards sharpness and penalises miscalibration, so it measures overall performance.
+However, calibration should still be verified separately.
+Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in a quantile regression model.
 The Pinball Score is given by
-
 \begin{equation}
@@
 \end{equation}
-
 where \(M\) is the metric's value for the method being considered, \(M_\mathrm{ref}\) is the metric's value for a reference method.
-The skill score show us by how many percent the reference approach is worse than the one under consideration.
-We will use bootstrap re-sampling of skill scores to determine if apparent differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}.
-Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model, and employ a block-bootstrap with blocks of length 24h in order to account for temporal correlations of the underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
+The skill score shows by how many per cent the reference approach is worse than the one under consideration.
+We will use bootstrap re-sampling of skill scores to determine if the differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}.
+Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model and employ a block-bootstrap with blocks of length 24h to account for temporal correlations of the underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
 
@@
 All functions were re-estimated on each iteration.
-ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin.
+ADAM and Poisson regression estimated the parameters, taking the ones obtained in the initial model application to the data in the first origin as the pre-initials.
 This allowed to speed up the computation for these two models.
 The initial estimation of ADAM took approximately one hour and 25 minutes.
-Each step in the experiment took the time shown below.
+Each step in the experiment took the time shown in Section \ref{result}.
 
@@
 The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
-The rolling origin advances in 12 hour steps, and the forecast horizon is set for 48 steps ahead.
+The rolling origin advances in 12-hour steps and the forecast horizon is set for 48 steps ahead.
 
 Figure \ref{fig:Pinball} presents pinball score aggregated across forecasting horizons for each quantile.
-It shows that the difference in performance among the models mainly comes from the middle of the distribution, and somewhat form the upper tail.
-There is very little difference in performance for the lower tail.
-This is kind of interesting, and reassuring that the better models are batter at probabilities that matter more to decision makers.
+It shows that the difference in performance among the models mainly comes from the middle of the distribution and somewhat from the upper tail.
+There is very little difference in performance for the lower tail (except for TBATS, which has a consistently higher pinball value than the other models).
+This is interesting and reassuring that the better models are better at probabilities that matter more to decision-makers.
 
 Probabilistic forecasts are evaluated following the principle of \emph{sharpness subject to calibration}, meaning that the sharper forecast is prefered provided that it is calibrated.
-Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded.
-Calibration is evaluated visually in Figure \ref{fig:quantile-bias}, which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
+Mis-calibrated forecasts are unsuitable for use in decision-making, so they should be excluded.
+Calibration is evaluated visually in Figure \ref{fig:quantile-bias}, which highlights a systematic negative bias across all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
 Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-For a user, this could result in poor staffing decisions as historic data fail to accurately charecterise present conditions.
+This could result in poor staffing decisions for a user as historic data fail to characterise present conditions accurately.
 
@@
 
-\begin{table}[!h]
-
-\caption{\label{tab:tab-results}Summary of studies in hourly emergency care forecasting}
-\centering
-\fontsize{11}{13}\selectfont
-\begin{tabular}[t]{lrrr}
-\toprule
-Method & Quantile Bias & Pinball & RMSE\\
-\midrule
-Benchmark-1 & 0.1047874 & 1.254491 & NA\\
-Benchmark-2 & 0.0557392 & 1.217429 & NA\\
-ETS(XXX) & 0.0194389 & 1.434862 & 0.0121247\\
-GBM & 0.0600650 & 1.263135 & NA\\
-NBI-2-I & 0.0540740 & 1.206238 & NA\\
-NBI-2-log & 0.0375705 & 1.188041 & NA\\
-NOtr-1 & 0.0098936 & 1.222584 & NA\\
-NOtr-2 & 0.0118522 & 1.208556 & NA\\
-Poisson-1 & 0.0372137 & 1.204936 & NA\\
-Poisson-2 & 0.0373884 & 1.188107 & NA\\
-Poisson-2-I & 0.0523937 & 1.222114 & NA\\
-RegressionPoisson & 0.0929416 & 1.293524 & 0.8490258\\
-Ttr-2 & 0.0140252 & 1.210105 & NA\\
-faster & 0.1862190 & 35.051012 & NA\\
-iETSCeiling & 0.0104673 & 1.419490 & NA\\
-iETSXSeasonal & 0.0104673 & 1.417260 & 0.0896228\\
-qreg-1 & 0.0643451 & 1.340557 & NA\\
-tbats & 0.0855702 & 1.536080 & 0.4859770\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Evaluation metrics from the test period are presented in Table \ref{tab:tab-results}.
-They are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
-
 \begin{figure}[H]
 
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/Skill-rel2bench-reduced-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/Skill-rel2bench-1}
 
@@
 
-\caption{Skill score ...}\label{fig:Skill-rel2bench-reduced}
+\caption{Pinball skill score for the seven best performing approaches.}\label{fig:Skill-rel2bench}
 \end{figure}
 
-Figure \ref{fig:Skill-rel2bench-reduced} illustrates the trade-off between calibration and pinball skill.
+Figure \ref{fig:Skill-rel2bench} illustrates the trade-off between calibration and pinball skill.
 
@@
 
-\caption{RMSE ...}\label{fig:lead-time-rmse}
+\caption{RMSE values over the forecast horizon.}\label{fig:lead-time-rmse}
 \end{figure}
 
-Figure \ref{fig:lead-time-rmse} reports the RMSE for each forecast horizon. It illustrates the times of day that are harder to predict -- morning pick-up and afternoon peak.
+Figure \ref{fig:lead-time-rmse} reports the RMSE for each forecast horizon.
+It illustrates the times of day that are harder to predict -- morning pick-up and afternoon peak.
+We can see that some models perform much better than others on specific forecast horizons.
+For example, Benchmark does an excellent job in predicting 1 - 5 steps ahead ED arrivals and, in general, doing well in forecasting arrivals in the night.
+In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, prophet, tbats and GBM do for some lead times.
 
-One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones.
-This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered as a reasonable strategy.
+\begin{table}[!h]
+
+\caption{\label{tab:tab-results}Summary of studies in hourly emergency care forecasting}
+\centering
+\fontsize{9}{11}\selectfont
+\begin{tabular}[t]{lrrrr}
+\toprule
+Method & Quantile Bias & Pinball & RMSE & Time\\
+\midrule
+NOtr-1 & 0.0098967 & 1.222583 & 0.2675957 & 451.6620471\\
+ADAM-iETSX & 0.0104673 & 1.417260 & 0.0896228 & 92.9348605\\
+NOtr-2 & 0.0118522 & 1.208561 & 0.2675957 & 86.5895462\\
+Ttr-2 & 0.0140221 & 1.210108 & 0.3324146 & 956.5532849\\
+Prophet & 0.0193799 & 1.447037 & 0.2955460 & 20.6755021\\
+ETS & 0.0194389 & 1.434862 & 0.0121247 & 10.7175205\\
+Poisson-1 & 0.0372137 & 1.204920 & 0.0095263 & 1.4763353\\
+Poisson-2 & 0.0373884 & 1.188109 & 0.0082932 & 5.0768588\\
+NBI-2 & 0.0540725 & 1.206241 & 0.3830272 & NA\\
+Benchmark-2 & 0.0557392 & 1.217429 & 0.2592800 & 0.0947247\\
+GBM-2 & 0.0600153 & 1.261690 & 1.7770897 & 602.4317496\\
+TBATS & 0.0855702 & 1.536080 & 0.4859770 & 273.0558176\\
+Regression-Poisson & 0.0929416 & 1.293524 & 0.8490258 & 67.1401641\\
+Benchmark-1 & 0.1047874 & 1.254491 & 1.0042634 & 0.3874450\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+In above figures, we only present the performance of the top seven methods.
+However, we have evaluated the performance of 14 methods.
+Evaluation metrics for all methods from the test period are presented in Table \ref{tab:tab-results}.
+They are ordered by Quantile Bias.
+The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
+
+One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones (Table \ref{tab:tab-results}).
+This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered a reasonable strategy.
 
@@
 
-\caption{Running time vs. forecast accuracy}\label{fig:time-accuracy}
+\caption{Speed (Running time) vs. forecast accuracy}\label{fig:time-accuracy}
 \end{figure}
 
-Figure \ref{fig:time-accuracy} reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+Finally, Figure \ref{fig:time-accuracy} reports the forecast accuracy of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+The x-axis represents the speed of each method as slow, moderate or fast.
 
@@
 
-Short-term forecasting of arrivals at emergency departments is an important element of staff and resource management in hospitals.
-Furthermore, due to the asymmetric impact of having an excess of shortage of resources, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
-Here, we have developed methods for producing probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
+Short-term forecasting of arrivals at emergency departments is an important element of hospital staff and resource management.
+Furthermore, due to the asymmetric impact of an excess resource shortage, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
+In this paper, we have developed methods for producing a probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
 
-Two approaches produced highly accurate, calibrated probabilistic forecasts, one time series method and one based on distributional regression.
-The first is ADAM-iESTX, which is an extension of exponential smoothing incorporating seasonality and assuming a Gamma predictive distribution.
-The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on features date and time features, and temperature.
-Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX produced forecasts had a lower RMSE than NOtr-2, and NOtr-2 produced forecasts with a lower pinball score.
+Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression.
+The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution.
+The second, labelled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature.
+Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score.
 This suggests that the latter may be preferred if the whole distribution is used in decision-making.
 
-Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts, and choice of model features, are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without need for post-processing.
-These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this article.
-Furthermore, methods based on continuous valued distributions are not adversely affected by the fact that the data are integer-valued.
+Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts and choice of model features are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without the need for post-processing.
+These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this paper.
+Furthermore, methods based on continuous-valued distributions are not adversely affected by the fact that the data are integer-valued.
 Rounding up predictive quantiles to the next integer does not make predictions worse.
 
-Finally, we have found that out-of-the-box models, those which require minimal tuning or manual development, do not perform as well as well-considered statistical methods.
+Finally, we have found that out-of-the-box models, which require minimal tuning or manual development, do not perform as well as well-considered statistical approaches.
 The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks.
 Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
-ETS produces reasonably well calibrated forecasts, in contrast to the benchmarks which were poorly calibrated, and highly accurate point forecasts.
+ETS produces reasonably well-calibrated forecasts, in contrast to the poorly calibrated benchmarks and has highly accurate point forecasts.
 However, its probabilistic forecasts were considerably worse than NOtr-2 in terms of pinball score.
@@
 Probabilistic forecasting opens the door to more sophisticated resource management in healthcare settings by providing decision-makers with uncertainty information and enabling quantitative risk management.
-Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecast skill, and may also be necessary to maximise benefits though more holistic decision-making.
+Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecasting skills and may also be necessary to maximise benefits through more holistic decision-making.
 
-Further research is required to investigate the practical benefits of the probabilistic forecasts in the healthcare and how they should be used to inform planning and decision making.
-This may require employing discrete simulation or new-vendor problem.
-While this study has focused on the hourly short-term forecasting, producing longer term daily forecast (e.g.~180-270 days ahead) is often required by planners, to support winter planning in ED and Ambulance services which requires more investigation.
-Moreover, more research is required to forecast other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
-This may require considering the dynamics among various services including General Practitioners, Emergency Departments, Ambulance and Fire \& Rescue services.
+Further research is required to investigate the practical benefits of probabilistic forecasts in healthcare and how they can inform planning and decision making.
+This may require employing discrete simulation or application to the newsvendor problems.
+While this study has focused on hourly short-term forecasting, producing longer-term daily forecast (e.g.~180-270 days ahead) is often required by planners to support winter planning in ED and Ambulance services which requires more investigation.
+Moreover, more research is needed in the forecasting of other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
+This may require considering the dynamics among various services, including General Practitioners, Emergency Departments, Ambulance and Fire \& Rescue services.
 
@@
 \end{equation}
-
 where \(n\) is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and \(k\) is the quantile of the continuous distribution of the variable.
@@
 In order to prove that \(n = \lceil k \rceil\), we need to use the following basic property:
-
 \begin{equation} \label{eq:quantCeil3}
@@
 \end{equation}
-
 which means that the rounded up value will always be less than or equal to \(n\) if and only if the original value is less than or equal to \(n\).
 Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
-
 \begin{equation} \label{eq:quantCeil4}
@@
 \end{equation}
-
 which is possible only when \(k \in (n-1, n]\), which means that \(\lceil k \rceil = n\).
 So the rounded up quantile of continuous random variable \(y_t\) will always be equal to the quantile of the descritised value of \(y_t\):
-
 \begin{equation} \label{eq:ceilingAndQuantiles1}

### 2022-02-10 paper/paper.Rmd
add a description for figure 8. Add title page and cover letter

@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
 
@@
 
-<!--- ```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
-``` --->
+ ```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
+ ```
 
-Figure \@ref(fig:seasonplot-weekofyear) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
+Figure \@ref(fig:seasonplot-dofw) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
 There is a line for each week (i.e.
@@
 
-Furthermore, we consider an extnsion to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$.
+Furthermore, we consider an extension to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$.
 These models are labelled Poisson-2 and NOtr-2.
@@
 Benchmark-2, NBI-2 and NOtr-2 have similar performance with most improvement of the latter two over the benchmark comming from the lower quantiles.
-All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.", fig.align='center',out.width= "90%"}
+All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.", fig.align='center',out.width= "70%"}
 ```
@@
 If a predictive model is calibrated, quantile bias should be approximately zero accross all probability levels.
-Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2" bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.", fig.align='center',out.width= "90%"}
+Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2 bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.", fig.align='center',out.width= "70%"}
 ```
@@
 
-Finally, Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
-The x-axis represents the speed of each method as slow, moderate or fast.
-<!-- We need to explain this once we have th result for RMSE -->
-<!-- ``` -->
+Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+The X-axis shows the speed of each method presented as slow, moderate or fast.
+We observe that there is no clear association between speed and accuracy improvement of the models used in this study.
+We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure.
+Figure \@ref(fig:time-accuracy) also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures.
+The figure also show that the $t$ family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures.
+Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
 
@@
 The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution.
-The second, labelled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature.
+The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature.
 Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score.

### 2022-02-10 paper/paper.tex
add a description for figure 8. Add title page and cover letter

@@
   We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
 \end{enumerate}
@@
 
-\caption{Distrubution of admission per hour and day of the week}\label{fig:hourly-plot-ridge}
+\caption{Distrubution of admission by hour-of-day and day-of-week.
+Most days have a distinct pattern, such as relatively high arrivals on Monday mornings and in the early hours of Saturday and Sunday morning.}\label{fig:hourly-plot-ridge}
 \end{figure}
@@
 
-Figure \ref{fig:seasonplot-weekofyear} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
+Figure \ref{fig:seasonplot-dofw} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
 There is a line for each week (i.e.~seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle.
@@
 
-\caption{Week of year arrivals}\label{fig:seasonplot-weekofyear}
+\caption{Arrivals by week-of-year. There is an annual trend of reduced arrivals durnng the summer and winter holiday period.}\label{fig:seasonplot-weekofyear}
 \end{figure}
@@
 
-\caption{Daily arrivals}\label{fig:date-plot}
+\caption{Daily arrivals over the entire dataset with moving average trend (blue line) and dates of outliers labled. Low outliers invariables correspond to public holidays.}\label{fig:date-plot}
 \end{figure}
@@
 
-Furthermore, we consider an extnsion to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\).
+Furthermore, we consider an extension to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\).
 These models are labelled Poisson-2 and NOtr-2.
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Pinball-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/Pinball-1}
 
@@
 
-\caption{Pinball loss values over different quantiles.}\label{fig:Pinball}
+\caption{Pinball loss values by quantile.
+Benchmark-2, NBI-2 and NOtr-2 have similar performance with most improvement of the latter two over the benchmark comming from the lower quantiles.
+All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.}\label{fig:Pinball}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/quantile-bias-1}
+{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/quantile-bias-1}
 
@@
 
-\caption{Quantile bias vs the nominal quantiles.}\label{fig:quantile-bias}
+\caption{Quantile bias illustrates the difference between the nominal exceedance of prediceted quantiles and the observed frequency of exceedance.
+If a predictive model is calibrated, quantile bias should be approximately zero accross all probability levels.
+Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2 bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.}\label{fig:quantile-bias}
 \end{figure}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/Skill-rel2bench-1}
-
-}
-
-\caption{Pinball skill score for the seven best performing approaches.}\label{fig:Skill-rel2bench}
-\end{figure}
-
-Figure \ref{fig:Skill-rel2bench} illustrates the trade-off between calibration and pinball skill.
 
@@
 
-\caption{RMSE values over the forecast horizon.}\label{fig:lead-time-rmse}
+\caption{RMSE values over the forecast horizon for forecasts issues at midnight.
+Attendance during the morning is more challenging to predice resulting in greater RMSE during these hours for all models.}\label{fig:lead-time-rmse}
 \end{figure}
@@
 
-\caption{Speed (Running time) vs. forecast accuracy}\label{fig:time-accuracy}
+\caption{Running time vs. forecast performance}\label{fig:time-accuracy}
 \end{figure}
 
-Finally, Figure \ref{fig:time-accuracy} reports the forecast accuracy of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
-The x-axis represents the speed of each method as slow, moderate or fast.
+Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+The X-axis shows the speed of each method presented as slow, moderate or fast.
+We observe that there is no clear association between speed and accuracy improvement of the models used in this study.
+We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure.
+Figure \ref{fig:time-accuracy} also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures.
+The figure also show that the \(t\) family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures.
+Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
 

### 2022-02-11 paper/paper.Rmd
correct description of figure 8

@@
 For example, Benchmark does an excellent job in predicting 1 - 5 steps ahead ED arrivals and, in general, doing well in forecasting arrivals in the night.
-In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, prophet, tbats and GBM do for some lead times.
+In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, Prophet, TBATS and GBM do for some lead times.
 
@@
 We observe that there is no clear association between speed and accuracy improvement of the models used in this study.
-We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure.
-Figure \@ref(fig:time-accuracy) also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures.
-The figure also show that the $t$ family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures.
-Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
+
+We found that most of the methods considered in this paper fall into the fast category with a different range of performance, depending on the type of error measure.
+Among the fastest methods, ETS is very competitive when assessed using Quantile Bias and RMSE.
+Moreover, Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by RMSE and Pinball.
+
+Figure \@ref(fig:time-accuracy) shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures.
+This figure also indicates that while the $t$ family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures.
+Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
 

### 2022-11-17 paper/paper.Rmd

@@
 This can have many advantages for healthcare staff and the quality of care delivered to patients.
-In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals.
+In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours.
 We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution.
@@
 
-In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:
+In this paper, we aim at filling these gaps and generate forecasts for a  prediction window of 48 hours. Our contributions to the literature are summarised as follows:
 
@@
 We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements for longer horizons and in other services [@al2021empirical], such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
 
@@
 
-Some studies employ models to forecast admissions and arrivals in Emergency Department.
+Some studies employ models to forecast admissions and arrivals in Emergency Department [@gul2020exhaustive].
 The time granularity considered by these studies spans from hourly to yearly.
@@
 results_table <- read_rds("results_table.rds") %>% arrange(`Quantile Bias`)
+results_table <- results_table %>% mutate(Time=if_else(is.na(Time),1.279187,Time)) %>% rename(`Time (minutes)` = Time)
 results_table <- results_table %>% 
@@
 
-<!-- Why NAs for RMSE?? -->
-
-In above figures, we only present the performance of the top seven methods.
-However, we have evaluated the performance of 14 methods.
-Evaluation metrics for all methods from the test period are presented in Table \@ref(tab:tab-results).
+In above figures (Figures \@ref(fig:Pinball)- \@ref(fig:lead-time-rmse)), we only present the performance of the top seven methods.
+However, we have evaluated the performance of 14 methods from the test period.
+Evaluation metrics and computational time (in minutes) for all methods are presented in Table \@ref(tab:tab-results).
 They are ordered by Quantile Bias.
@@
 
-Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours, presented in Table \@ref(tab:tab-results).
 The X-axis shows the speed of each method presented as slow, moderate or fast.
@@
 Figure \@ref(fig:time-accuracy) shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures.
-This figure also indicates that while the $t$ family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures.
+The figure also indicates that while the $t$ family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures.
 Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.

### 2022-11-17 paper/paper.tex

@@
 
-In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:
+In this paper, we aim at filling these gaps and generate forecasts for a prediction window of 48 hours. Our contributions to the literature are summarised as follows:
 
@@
   We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
+The proposed approach could also be generalised to forecast hourly requirements for longer horizons and in other services \citep{al2021empirical}, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
 \end{enumerate}
@@
 
-Some studies employ models to forecast admissions and arrivals in Emergency Department.
+Some studies employ models to forecast admissions and arrivals in Emergency Department \citep{gul2020exhaustive}.
 The time granularity considered by these studies spans from hourly to yearly.
@@
 \toprule
-Method & Quantile Bias & Pinball & RMSE & Time\\
+Method & Quantile Bias & Pinball & RMSE & Time (minutes)\\
 \midrule
@@
 Poisson-2 & 0.0373884 & 1.188109 & 0.0082932 & 5.0768588\\
-NBI-2 & 0.0540725 & 1.206241 & 0.3830272 & NA\\
+NBI-2 & 0.0540725 & 1.206241 & 0.3830272 & 1.2791870\\
 Benchmark-2 & 0.0557392 & 1.217429 & 0.2592800 & 0.0947247\\
@@
 
-In above figures, we only present the performance of the top seven methods.
-However, we have evaluated the performance of 14 methods.
-Evaluation metrics for all methods from the test period are presented in Table \ref{tab:tab-results}.
+In above figures (Figures \ref{fig:Pinball}- \ref{fig:lead-time-rmse}), we only present the performance of the top seven methods.
+However, we have evaluated the performance of 14 methods from the test period.
+Evaluation metrics and computational time (in minutes) for all methods are presented in Table \ref{tab:tab-results}.
 They are ordered by Quantile Bias.
@@
 
-Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
+Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours, presented in Table \ref{tab:tab-results}.
 The X-axis shows the speed of each method presented as slow, moderate or fast.
 We observe that there is no clear association between speed and accuracy improvement of the models used in this study.
-We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure.
-Figure \ref{fig:time-accuracy} also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures.
-The figure also show that the \(t\) family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures.
-Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
+
+We found that most of the methods considered in this paper fall into the fast category with a different range of performance, depending on the type of error measure.
+Among the fastest methods, ETS is very competitive when assessed using Quantile Bias and RMSE.
+Moreover, Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by RMSE and Pinball.
+
+Figure \ref{fig:time-accuracy} shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures.
+The figure also indicates that while the \(t\) family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures.
+Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
 

### 2022-12-28 paper/paper.Rmd
comments adressed by bahman

@@
 title:  Probabilistic forecasting of hourly Emergency Department arrivals
-author:
-  - name: Bahmna Rostami Tabar
-    email: rostami-tabarb@cardiff.ac.uk
-    affiliation: Cardiff Business School
-    footnote: 12
-  - name: Jethro Browell
-    email: jethro.browell@glasgow.ac.uk
-    affiliation: University of Glasgow
-    footnote: 1
-  - name: Ivan Svetunkov
-    email: i.svetunkov@lancaster.ac.uk
-    affiliation: Lancaster University
-    footnote: 1
-address:
-  - code: Cardiff Business School
-    address: Cardiff Business School, Cardiff University, UK
-  - code: University of Glasgow
-    address: School of Mathematics and Statistics, University of Glasgow, UK
-  - code: Lancaster University
-    address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
-footnote:
-  - code: 12
-    text: "Corresponding Author"
-  - code: 1
-    text: "Equal contribution"
-
+# author:
+#   - name: Bahmna Rostami Tabar
+#     email: rostami-tabarb@cardiff.ac.uk
+#     affiliation: Cardiff Business School
+#     footnote: 12
+#   - name: Jethro Browell
+#     email: jethro.browell@glasgow.ac.uk
+#     affiliation: University of Glasgow
+#     footnote: 1
+#   - name: Ivan Svetunkov
+#     email: i.svetunkov@lancaster.ac.uk
+#     affiliation: Lancaster University
+#     footnote: 1
+# address:
+#   - code: Cardiff Business School
+#     address: Cardiff Business School, Cardiff University, UK
+#   - code: University of Glasgow
+#     address: School of Mathematics and Statistics, University of Glasgow, UK
+#   - code: Lancaster University
+#     address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
+# footnote:
+#   - code: 12
+#     text: "Corresponding Author"
+#   - code: 1
+#     text: "Equal contribution"
 abstract: |
@@
 Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
-
-keywords:  "Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing"
+keywords:  "Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing"
 
@@
 date: "`r Sys.Date()`"
-geometry: "top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm"
+geometry: "top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm"
 #linenumbers: true
@@
 header-includes:  
- \usepackage{adjustbox, float,lscape, bm,amsmath} #use the 'float' package
+ \usepackage{adjustbox, float,lscape,amsmath,bm} #use the 'float' package
 output:
@@
 #biblio-style: abbrevnat
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
-Accurate ED demand forecasts contribute to a better decision-making process regarding resources allocation and staffing.
-This is one of the best ways to optimise resources utilisation and minimise related costs.
-An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand might result in patients overcrowding the system, which is a severe problem that causes challenges for the patient flow [@derlet2002overcrowding].
-Also, it is related to increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].
+Forecasting Emergency Department (ED) arrivals is critical for informing
+staffing and scheduling decisions to meet the needs of patients.
+Accurate ED demand forecasts contribute to a better decision-making
+process regarding resources allocation and staffing. This is one of the
+best ways to optimise resources utilisation and minimise related costs.
+An accurate forecast of patient arrivals is crucial in ED services to
+depict various courses of action that can result in massive savings in
+terms of patient lives. Inability to match the staff with the demand
+might result in patients overcrowding the system, which is a severe
+problem that causes challenges for the patient flow
+[@derlet2002overcrowding]. Also, it is related to increasing length of
+stay [@muhammet2015forecasting], low patient satisfaction, unexpected
+return visits to services, increased health care costs, inaccuracy in
+electronic medical records and other [@rostami2020anticipating].
 
-Accurate forecasting of arrivals by the hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
-This has many advantages for both patients, staff and the quality of provided services.
-Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision-making related to the execution of the delivery process in ED.
-The combination of an hourly arrival forecast, current staff, being occupied, resource availability and waiting times at ED provide information on the state of the unscheduled care system across the service.
-Having this complete picture enables the delivery managers to focus on the areas that require intervention to allow the most effective delivery of the service to the patients.
-However, compared with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-Hourly time series generally exhibit multiple seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
-They may also express nonstationarity, and their profile may change over time.
-Therefore, an appropriate forecasting model should consider these features to accurately predict hourly demand admissions.
+Accurate forecasting of arrivals by the hour of the day enables planners
+to match staff to meet anticipated patients, reconfigure units and
+redeploy staff. This has many advantages for both patients, staff and
+the quality of provided services. Hourly forecasts are required to
+inform the short-term operational planning for the current and the
+upcoming shifts of the day. This involves the short-term decision-making
+related to the execution of the delivery process in ED. The combination
+of an hourly arrival forecast, current staff, being occupied, resource
+availability and waiting times at ED provide information on the state of
+the unscheduled care system across the service. Having this complete
+picture enables the delivery managers to focus on the areas that require
+intervention to allow the most effective delivery of the service to the
+patients. However, compared with lower frequency time series forecasting
+such as monthly, quarterly and yearly, hourly forecasts are challenging
+because the noise caused by random variation may overshadow any pattern
+in the time series. Hourly time series generally exhibit multiple
+seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
+They may also express nonstationarity, and their profile may change over
+time. Therefore, an appropriate forecasting model should consider these
+features to accurately predict hourly demand admissions.
 
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
-These studies use multiple approaches, including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting].
-However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric.
+There are few studies that look at forecasting hourly arrivals in ED and
+other hospital services using historical time series data and/or
+predictors such as patient characteristics, weather, holidays and public
+events. These studies use multiple approaches, including Exponential
+Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average
+(ARIMA) [@hyndman2021forecasting], Autoregressive Conditional
+Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive
+model [@lutkepohl2013vector], TBATS [@de2011forecasting] and Artificial
+Neural Networks [@hyndman2021forecasting].
+\textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study.
+These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
+forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
 Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
-
-Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated.
+Furthermore, if the impact of under- and overstaffing can be quantified,
+probabilistic forecasts allow _optimal_ decisions that balance the cost associated with under- and overstaffing to be calculated.
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-Another drawback of existing studies is that the datasets used are relatively small (e.g.
+\textcolor{blue}{Another drawback of existing studies is that the
+datasets used are relatively small (e.g.
 time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
@@
 Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals.
-Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
+Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
+Finally, all previous publications
+referenced in this paper are not fully reproducible as underlying data and code are not available.}
 
@@
 
-1.
-We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
+1.  We produce probabilistic forecasts, in addition to the point
+    estimation, quantifying uncertainties in future hospital admission,
+    and comparing different forecasting methods using a suite of
+    well-established evaluation metrics;
 
-2.
-We develop an advanced dynamic model to forecast ED arrivals based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
+2.  We develop an advanced dynamic model to forecast ED arrivals based
+    on iETS [@Svetunkov2019a] and ETSX models with a modification for
+    multiple frequencies, which produced highly-accurate point
+    forecasts;
 
-3.
-We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;
+3.  We develop a novel model to produce a probabilistic forecast of ED
+    arrivals based on Generalised Additive Models for Location Scale and
+    Shape, which accounts for i) the bounded and non-Gaussian
+    distribution of arrivals, ii) multiple seasonalities, weather and
+    holiday effects, and iii) variation in forecast uncertainty;
 
-4.
-We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.
-Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
+4.  We benchmark the accuracy of our model against appropriate models
+    used when multiple seasonality is present, i.e. Prophet, TBATS,
+    Poisson Regression, Exponential Smoothing State Space model (ETS)
+    and the simple empirical distribution of the arrivals;
 
-5.
-We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements for longer horizons and in other services [@al2021empirical], such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
+5.  We provide data and code enabling reproduction and refinement of the
+    proposed approach and benchmarks. The proposed approach could also
+    be generalised to forecast hourly requirements for longer horizons
+    and in other services [@al2021empirical], such as inpatient and
+    outpatient care services, the number of attended incidents in
+    ambulance services, or call volumes in clinical desk services.
 
-The rest of the paper is organised as follows: In section \@ref(lit), we provide a brief overview of hourly ED arrival forecasting; In Section \@ref(data), we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
-In section \@ref(model), we describe the modelling approach and benchmark methods.
-We then discuss the performance evaluation metrics in section \@ref(accuracy); in section \@ref(result), we present and discuss our results.
-Finally, we summarise our findings and present ideas for future research in section \@ref(conclusion).
+The rest of the paper is organised as follows: In section \@ref(lit), we
+provide a brief overview of hourly ED arrival forecasting; In Section
+\@ref(data), we present the hourly time series of an ED arrival and use
+various plots to highlight important patterns. In section \@ref(model),
+we describe the modelling approach and benchmark methods. We then
+discuss the performance evaluation metrics in section \@ref(accuracy);
+in section \@ref(result), we present and discuss our results. Finally,
+we summarise our findings and present ideas for future research in
+section \@ref(conclusion).
 
 # Research background: hourly ED forecasting {#lit}
+\textcolor{blue}{
+There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
+Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention.
+We refer interested readers to  some extensive reviews of the relevant literature} [@mingliterature2022;@gul2020review;@ibrahim2016modeling].
+\textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare.
+However, given
+the focus of this study, we only discuss hourly ED forecasting.}
 
-Some studies employ models to forecast admissions and arrivals in Emergency Department [@gul2020exhaustive].
-The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only discuss hourly ED forecasting.
+@hertzum2017forecasting used linear regression, ARIMA, and Naïve to
+investigate whether accurate hourly accident and emergency department
+patient arrivals and occupancy forecasts can be generated using calendar
+variables. @hertzum2017forecasting study found that patient arrival
+variation is larger across the hours of the day than across the days of
+the week and the months of the year. In terms of the hour of the day,
+patient arrivals peaked around noon. For days of the week, Monday is the
+busiest day, while weekends are the quietest ones. July-August are the
+months with the highest number of patient arrivals, while January and
+February have the lowest numbers. They indicate that regression and
+ARIMA models performed similarly in modelling patient arrivals, while
+ARIMA outperformed regression models in modelling accident and emergency
+department occupancy.
 
-@hertzum2017forecasting used linear regression, ARIMA, and Naïve to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
-@hertzum2017forecasting study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
-In terms of the hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day, while weekends are the quietest ones.
-July-August are the months with the highest number of patient arrivals, while January and February have the lowest numbers.
-They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
+@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neural
+network methods to forecast hourly accident and emergency department
+arrivals. ARIMA model was selected as the best fit model. Authors
+claimed that ARIMA provided high and acceptable hourly ED forecasting
+accuracy, even outperforming TBATS. @Cheng2021 developed an ARIMA model
+for ED occupancy with a seasonal component and exogenous variables,
+which outperformed a rolling-average benchmark. They also produce
+prediction intervals, a form of the probabilistic forecast, which were
+found to be well-calibrated, a necessary property for such forecasts.
 
-@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neural network methods to forecast hourly accident and emergency department arrivals.
-ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
-@Cheng2021 developed an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which outperformed a rolling-average benchmark.
-They also produce prediction intervals, a form of the probabilistic forecast, which were found to be well-calibrated, a necessary property for such forecasts.
+@morzuch2006forecasting used the Unobserved Components Model (UCM), in
+which each component of the time series is separately modelled as
+stochastic. Double-seasonal exponential smoothing and standard
+Holt-Winters were used to forecast ED arrival for a horizon of 168
+hours. The hourly data collected from an ED in Pennsylvania showed no
+trend and two seasonal cycles: a within-day and a within-week seasonal
+cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly
+seasonality of the time series.
 
-@morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for a horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend and two seasonal cycles: a within-day and a within-week seasonal cycles.
-The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
+@mccarthy2008challenge employed a Poisson log-linear regression model,
+including independent variables such as temporal factors (e.g.,
+hour-of-day, day-of-week, type-of-day, season, and calendar year),
+patient characteristics (i.e., age, gender, insurance status, triage
+level, mode of arrival, and ambulance diversion status) and climatic
+factors (i.e., temperature and precipitation) to model patient demand
+for ED services. The authors produced probabilistic predictions in the
+form of $50\%$ and $90\%$ prediction intervals for the number of hourly
+arrivals. Hourly data of ED arrivals in the 1-year study period was
+modelled and analysed, and it was suggested that the model could be used
+for forecasting. However, model evaluation was performed in-sample on
+only one year of data, so it is unclear how this approach would perform
+in a forecasting setting or compare to simpler approaches. However, the
+length of the time series in this study was very short (only one year),
+which did not allow for a rigorous out-of-sample evaluation.
 
-@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day-of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
-The authors produced probabilistic predictions in the form of $50\%$ and $90\%$ prediction intervals for the number of hourly arrivals.
-Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it was suggested that the model could be used for forecasting.
-However, model evaluation was performed in-sample on only one year of data, so it is unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
-However, the length of the time series in this study was very short (only one year), which did not allow for a rigorous out-of-sample evaluation.
+@schweigler2009forecasting investigated whether time series methods
+could accurately generate short-term forecasts of ED bed occupancy. A
+year-long dataset of hourly ED bed occupancy was collected from three
+facilities. The authors implemented an hourly historical average model,
+SARIMA model, and sinusoidal model with autocorrelated error for each
+facility. The historical average model was based on the mean occupancy
+for each site, for each hour of the day, while the sinusoidal model was
+based on four parameters: an AR term, a sine coefficient, a cosine
+coefficient and an intercept. They evaluated the forecast accuracy of
+four and twelve hours forecast horizons using RMSE. They found that both
+SARIMA and the sinusoidal models outperformed the historical average
+(for example, at site 2, the two models improved by $33\%$ the 12-hour
+forecasts generated by the historical average).
 
-@schweigler2009forecasting investigated whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
-A year-long dataset of hourly ED bed occupancy was collected from three facilities.
-The authors implemented an hourly historical average model, SARIMA model, and sinusoidal model with autocorrelated error for each facility.
-The historical average model was based on the mean occupancy for each site, for each hour of the day, while the sinusoidal model was based on four parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizons using RMSE.
-They found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by the historical average).
+@kim2014predicting compared different univariate and multivariate time
+series forecasting techniques to predict patient volume for a Hospital
+Medicine programme. The study evaluated linear regression, exponential
+smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional
+Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to
+forecast for 4 and 24 hours ahead. They used Mean Absolute Percentage
+Error (MAPE) to report the forecast accuracy. The authors found that the
+ARIMA outperformed all the other models.
 
-@kim2014predicting compared different univariate and multivariate time series forecasting techniques to predict patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and  Vector Autoregressive (VAR) models to forecast for 4 and 24 hours ahead.
-They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
-The authors found that the ARIMA outperformed all the other models.
-
-Table \@ref(tab:summarylit) summarize the relevant papers.
+Table \@ref(tab:summarylit) summarises the relevant papers.
 
@@
 
-@asheim2019real developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
-Measuring the accuracy using the MAPE, @asheim2019real noticed that significant improvement happened when the time of notification was incorporated into the model, especially in the one-hour horizon.
+@asheim2019real developed a Poisson time-series regression model with
+continuous day-of-week and week-of-year effects to implement a real-time
+system that could forecast ED arrivals on 1, 2, 3 hours ahead. Measuring
+the accuracy using the MAPE, @asheim2019real noticed that significant
+improvement happened when the time of notification was incorporated into
+the model, especially in the one-hour horizon.
 
-@cheng2021forecasting used one year of ED visits time series to evaluate the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
-The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of $80\%$ and $95\%$.
-They show that SARIMAX provides a more accurate forecast of hourly ED occupancy.
+@cheng2021forecasting used one year of ED visits time series to evaluate
+the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED
+occupancy up to 4-hours ahead. The forecast accuracy is evaluated using
+Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point
+forecast and coverage for prediction intervals of $80\%$ and $95\%$.
+They show that SARIMAX provides a more accurate forecast of hourly ED
+occupancy.
 
-According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established.
-Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low [@boyle2012predicting; @hertzum2017forecasting], while others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].
+According to the studies mentioned above, it can be said that they have
+shown complications in forecasting hourly patient accident and emergency
+department visits, and the application of forecasting hourly patient
+visits is not well established. Some of the studies claimed that the
+accuracy of forecasting models on hourly accident and emergency
+department data is low [@boyle2012predicting; @hertzum2017forecasting],
+while others mentioned that the accuracy of ED hourly forecast is at an
+acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;
+@schweigler2009forecasting].
 
-There are a few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches:
+There are a few limitations in the literature which encourage us to
+undertake this research and examine different forecasting approaches:
 
-(i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
-(ii) Almost all research studies produce point forecasts and, at best, report prediction intervals.
-There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent the uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
-(iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
-(iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
-(v) some studies in this area lack a rigorous experimental design, i.e. they do not use benchmark methods or report forecast accuracy.
+(i) Current approaches to forecast hourly ED arrivals do not fully
+    consider the feature of data such as multiple seasonal cycles and
+    changing profile over time;
+(ii) Almost all research studies produce point forecasts and, at best,
+     report prediction intervals. There is a lack of studies presenting
+     the entire forecast distribution of hourly ED arrivals that better
+     represent the uncertainty of future arrivals, providing a holistic
+     picture of future demand for a planner;
+(iii) most studies are not reproducible, as it is almost impossible to
+      reapply the approaches without the help of the authors of those
+      papers;
+(iv) studies are limited in terms of the length of historical data used
+     for training purposes and forecast performance evaluation and
+(v) some studies in this area lack a rigorous experimental design, i.e.
+    they do not use benchmark methods or report forecast accuracy.
 
@@
 
-Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
-We aggregated the patients’ arrival times to obtain hourly arrivals, which are used in this study.
-Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week.
-Although the data is noisy, it reveals some systematic patterns.
+Data used in this study comprises counts of patients' arrival times at
+one of the largest ED units in the UK between April 2014 and February
+2019, extracted from the ED administrative database of the hospital. We
+aggregated the patients' arrival times to obtain hourly arrivals, which
+are used in this study. Figure \@ref(fig:hourly-plot-ridge) illustrates
+the distribution of arrivals for each hour of the day and the day of the
+week. Although the data is noisy, it reveals some systematic patterns.
 
@@
 
-It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by @hertzum2017forecasting.
-The ED arrivals decrease between midnight and early morning and then increase until the evening, decreasing after that again.
-It is also clear that ED service gets systematically more visits on Mondays between 8 a.m.
-and 5 p.m.
-Moreover, the number of arrivals around midnight is slightly higher for Saturdays and Sundays.
+It is clear that the number of arrivals has a sub-daily structure,
+similar to the one summarised by @hertzum2017forecasting. The ED
+arrivals decrease between midnight and early morning and then increase
+until the evening, decreasing after that again. It is also clear that ED
+service gets systematically more visits on Mondays between 8 a.m. and 5
+p.m. Moreover, the number of arrivals around midnight is slightly higher
+for Saturdays and Sundays.
 
-Figure \@ref(fig:hourly-plot-ridge) also highlights significant skewness for almost every hour of the day that varies with time-of-day, which should be accounted for in forecasting methods.
-Some skewness might be related to holidays and special events.
-It is also clear that arrivals are less volatile between midnight and early morning.
+Figure \@ref(fig:hourly-plot-ridge) also highlights significant skewness
+for almost every hour of the day that varies with time-of-day, which
+should be accounted for in forecasting methods. Some skewness might be
+related to holidays and special events. It is also clear that arrivals
+are less volatile between midnight and early morning.
 
-```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
+```{r seasonplot-dofw, fig.cap= "Subseries plot: day of weekly arrivals", fig.align='center',out.width="70%"}
 ```
 
-Figure \@ref(fig:seasonplot-dofw) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
-There is a line for each week (i.e.
-seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle.
-It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week.
-This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week.
-This might be due to the closure of General Practitioners outpatient clinics over the weekend.
+Figure \@ref(fig:seasonplot-dofw) \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals.
+Each individual plot illustrates how arrivals change over time for a each day of week (i.e.
+seven days) from Monday to Saturday.
+The blue line shows the average arrival for the given day.
+It is clear that ED arrivals on Monday are higher followed by Saturday.
+This indicates that there are
+significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}
 
@@
 
-Figure \@ref(fig:seasonplot-weekofyear) highlights the week of year seasonality in the ED arrivals.
-We observe that arrivals are significantly lower from week 29 to 35, corresponding to the Summer period.
-Moreover, the number of arrivals is lower at the beginning and the end of the year.
-The arrivals increase from week 36 and peak in weeks 39-42.
-This corresponds to the September - October period.
+Figure \@ref(fig:seasonplot-weekofyear) highlights the week of year
+seasonality in the ED arrivals. We observe that arrivals are
+significantly lower from week 29 to 35, corresponding to the Summer
+period. Moreover, the number of arrivals is lower at the beginning and
+the end of the year. The arrivals increase from week 36 and peak in
+weeks 39-42. This corresponds to the September - October period.
 
@@
 
-Finally, Figure \@ref(fig:date-plot) presents the time plot of daily arrivals.
-Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
-The figure shows more arrivals on Mondays than on other weekdays, agreeing with the previous findings.
-Moreover, we can observe a long-term trend line and the significant effect of some holidays.
-We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
-Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact of special events on arrivals.
-For instance, we see that arrivals are significantly low for 01-03 of March 2018.
-These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages, and schools closed in the UK [@stormemma2018].
+Finally, Figure \@ref(fig:date-plot) presents the time plot of daily
+arrivals. Each point represents one day, and points are shape-coded by
+day-of-week to show the weekly cycle. The figure shows more arrivals on
+Mondays than on other weekdays, agreeing with the previous findings.
+Moreover, we can observe a long-term trend line and the significant
+effect of some holidays. We can see that arrivals near Christmas and New
+Year's day are significantly lower than other days of the year.
+Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact
+of special events on arrivals. For instance, we see that arrivals are
+significantly low for 01-03 of March 2018. These days correspond to the
+Storm Emma with heavy snowfall that resulted in travel disruption, mass
+power outages, and schools closed in the UK [@stormemma2018].
 
-Based on this analysis and the literature review, we should consider models that can take the following into account:
+Based on this analysis and the literature review, we should consider
+models that can take the following into account:
 
@@
 -   Calendar events, such as holidays,
-- Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
+-   Lags of calendar events to accommodate the potential changes in
+    demand the next day after a holiday,
 -   Other events, such as sporting fixtures,
@@
 
-We propose several forecasting models that account for the structures outlined above.
+We propose several forecasting models that account for the structures
+outlined above.
 
@@
 
-## Naive/Climatology {#climatology}
+## Naive {#naive}
 
-We start with one of the simplest forecasting approaches used in practice - assuming that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
-This is called "Naïve".
-In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution [@la2021new].
-We consider the empirical distribution of all available historic data (Benchmark-1) and the empirical distribution of the most recent year of historic data on a rolling basis (Benchmark-2) to capture potential changes in behaviour over time.
+We start with one of the simplest forecasting approaches used in
+practice - assuming that in the next few hours, everything will be the
+same as in the similar hours of a similar day in the past. This is
+called "Naïve". In our case, given that we need a distribution of
+values, we will use a modified approach, where the
+empirical distribution of the hourly arrival time series is used to
+forecast the future arrival distribution [@la2021new]. We consider the
+empirical distribution of all available historic data (Benchmark-1) and
+the empirical distribution of the most recent year of historic data on a
+rolling basis (Benchmark-2) to capture potential changes in behaviour
+over time.
 
@@
 
-Regression is one of the most popular forecasting methods that use explanatory variables to predict a variable of interest (in our case, the ED arrivals).
-The classical linear regression model is formulated as
+Regression is one of the most popular forecasting methods that use
+explanatory variables to predict a variable of interest (in our case,
+the ED arrivals). The classical linear regression model is formulated as
 \begin{equation}
@@
 (\#eq:linearRegression)
-\end{equation}
-where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index.
-However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic because the number of admitted patients is an integer and non-negative.
-So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression [see, for example, @mccarthy2008challenge], which can be summarised as
-\begin{equation}
+\end{equation} where $\mathbf{x}_t$ is the vector of explanatory
+variables, $\boldsymbol{\beta}$ is the vector of parameters,
+$\epsilon_t$ is the error term, which is typically assumed to follow
+Normal distribution with zero mean and a fixed variance, and $t$ is the
+time index. However, in the context of healthcare and ED arrivals, the
+assumption of Normality is unrealistic because the number of admitted
+patients is an integer and non-negative. So the linear regression model
+should be substituted by some other model. One of the models that is
+frequently used in practice is the Poisson regression [see, for example,
+@mccarthy2008challenge], which can be summarised as \begin{equation}
   \begin{aligned}
@@
 
-The logarithm in \@ref(eq:PoissonRegression) is needed to ensure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function <!--[reference needed]-->.
-There is no single correct answer when selecting explanatory variables for the model, and the decision needs to be made for each specific case.
-In our experiment, we will only include dummy variables, capturing a variety of calendar events:
+The logarithm in \@ref(eq:PoissonRegression) is needed to ensure that
+the parameter of Poisson distribution is always positive. This model can
+be estimated via maximisation of the likelihood function based on
+Poisson mass function <!--[reference needed]-->. There is no single
+correct answer when selecting explanatory variables for the model, and
+the decision needs to be made for each specific case. In our experiment,
+we will only include dummy variables, capturing a variety of calendar
+events:
 
@@
 
-The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
-Finally, (5) is needed to capture the potential phenomenon of change in admittance after the holiday (e.g.
-people might try not to go to the hospital on Christmas eve and thus go the next day).
-This model assumes that all these effects are deterministic and do not change over time.
-Still, the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables so that the 3 pm on Monday in January will be different from 3 pm on Monday in July, although the parameters for the hour of day and day of the week are fixed and do not change over time.
-We use the `alm()` function from the `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".
+The variables _1--3_ allow modelling the seasonal patterns on the
+appropriate level of detail throughout the year, while _4_ covers the
+changes in admittance due to calendar events. Finally, _5_ is needed to
+capture the potential phenomenon of change in admittance after the
+holiday (e.g. people might try not to go to the hospital on Christmas
+eve and thus go the next day). This model assumes that all these effects
+are deterministic and do not change over time. Still, the exponentiation
+in \@ref(eq:PoissonRegression) introduces an interaction effect between
+dummy variables so that the 3 pm on Monday in January will be different
+from 3 pm on Monday in July, although the parameters for the hour of day
+and day of the week are fixed and do not change over time. We use the
+`alm()` function from the `greybox` package [@Svetunkov2021Greybox] for
+R [@RTeam2021] for the experiments and denote this model as "Poisson
+Regression".
 
@@
 
-@Hyndman2008b developed a state space approach for exponential smoothing models.
-The model can have a set of elements, including different types of Error, Trend and Seasonal components (thus *ETS*).
-Given the popularity of the ETS model in the forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with a frequency of 24 (hour of the day) as a benchmark.
-This was done using the `adam()` function from the `smooth` package [@Svetunkov2021Smooth] for R and denoted as *ETS*.
-This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
-This model is included as a benchmark, only to see how the other models perform compared to it.
+@Hyndman2008b developed a state space approach for exponential smoothing
+models. The model can have a set of elements, including different types
+of Error, Trend and Seasonal components (thus *ETS*). Given the
+popularity of the ETS model in the forecasting community, we decided to
+include the basic ETS(A,N,A) model with the seasonal component with a
+frequency of 24 (hour of the day) as a benchmark. This was done using
+the `adam()` function from the `smooth` package [@Svetunkov2021Smooth]
+for R and denoted as *ETS*. This model does not capture the day of week
+or week of year effects, does not include explanatory variables, but its
+seasonal component and level change over time. This model is included as
+a benchmark, only to see how the other models perform compared to it.
 
@@
 
-Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects.
-Prophet is robust to missing data and shifts in the trend and typically handles outliers well.
-Prophet works well on daily data seen in Facebook.
-It is robust and automated, making it easy to learn for beginners.
-The implementation may be less flexible than other methods.
-The model itself relies on the Multiple Source of Error state space model, initially proposed by @Kalman1960.
-The model is incorporated using the corresponding implementation of the Fable package in R.
-We use the `prophet()` function from the `fable` package [@fable2020].
-Note that the input data is assigned with an hourly and daily seasonality.
+Prophet is a forecasting procedure created by Facebook
+[@taylor2018forecasting] that accounts for multiple seasonality,
+piecewise trend and holiday effects. Prophet is robust to missing data
+and shifts in the trend and typically handles outliers well. Prophet
+works well on daily data seen in Facebook. It is robust and automated,
+making it easy to learn for beginners. The implementation may be less
+flexible than other methods. The model itself relies on the Multiple
+Source of Error state space model, initially proposed by @Kalman1960.
+The model is incorporated using the corresponding implementation of the
+Fable package in R. We use the `prophet()` function from the `fable`
+package [@fable2020]. Note that the input data is assigned with an
+hourly and daily seasonality.
+\textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our benchmarks.}
 
@@
 
-@de2011forecasting proposed a model to deal with time series exhibiting multiple complex seasonalities called "TBATS".
-It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
-The latter gives the model more flexibility to deal with fractional seasonality and reduces the parameters of the model when the frequencies of seasonalities are high.
-We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].
+@de2011forecasting proposed a model to deal with time series exhibiting
+multiple complex seasonalities called "TBATS". It includes a Box-Cox
+Transformation, ARMA model for residuals and a trigonometric expression
+of seasonal terms. The latter gives the model more flexibility to deal
+with fractional seasonality and reduces the parameters of the model when
+the frequencies of seasonalities are high. We fit a TBATS model using
+the `tbats()` function from the `forecast` package in R
+[@forecastpackage2020].
 
@@
 
-Quantile regression allows the production of density forecasts without assuming a fixed distributions shape controlled by a small number of parameters, such as the Poisson distribution with parameter $\lambda$.
-By producing forecasts of multiple quantiles the full predictive distribution can be constructed.
-Quantile regression is particularly useful where data do not follow a simple distribution, of the distribution shape changes over time of with some covariate, which may be the case with ED arrivals.
+Quantile regression allows the production of density forecasts without
+assuming a fixed distributions shape controlled by a small number of
+parameters, such as the Poisson distribution with parameter $\lambda$.
+By producing forecasts of multiple quantiles the full predictive
+distribution can be constructed. Quantile regression is particularly
+useful where data do not follow a simple distribution, the
+distribution shape changes over time with some covariate, which may
+be the case with ED arrivals.
 
-Gradient Boosting Machines (GBM) are a tree-based machine learning model for regression and classification.
-Here we produce probabilistic forecasts of ED arrivals via multiple quantile regression using GBMs as implemented in the R package `gbm` [@Greenwell2020].
-GBMs are a best-in-class algorithm for similar regression problems characterised by modest volumes of training data and possible interactions between input features, and are therefore an appealing choice for ED arrival forecasting.
-Here GBMs are fit with the following features: hour of the day, day of the week, school holiday, temperature and day of the year.
-Hyperparameters are chosen via grid search on the training data, and were chosen as follows: 500 trees, an interaction depth of 2, and a shrinkage 0.1.
-An advantage of GBMs is their ability to learn interactions in comparison to the additive models presented later which require the user to specify possible interactions between inputs.
-Here GBM is included as a reference for the performance of `out-of-the-box' machine learning models applied with minimal effort.
-We also explored the possibility of performing linear regression with additive models, but these did not perform as well as GBM so are omited for brevity.
+Gradient Boosting Machines (GBM) are a tree-based machine learning model
+for regression and classification. Here we produce probabilistic
+forecasts of ED arrivals via multiple quantile regression using GBMs as
+implemented in the R package `gbm` [@Greenwell2020]. GBMs are a
+best-in-class algorithm for similar regression problems characterised by
+modest volumes of training data and possible interactions between input
+features, and are therefore an appealing choice for ED arrival
+forecasting. Here GBMs are fit with the following features: hour of the
+day, day of the week, school holiday, temperature and day of the year.
+Hyperparameters are chosen via grid search on the training data, and
+were chosen as follows: 500 trees, an interaction depth of 2, and a
+shrinkage 0.1. An advantage of GBMs is their ability to learn
+interactions in comparison to the additive models presented later which
+require the user to specify possible interactions between inputs. Here
+GBM is included as a reference for the performance of \`out-of-the-box'
+machine learning models applied with minimal effort. We also explored
+the possibility of performing linear regression with additive models,
+but these did not perform as well as GBM so are omited for brevity.
 
@@
 
-@SvetunkovAdam2021 proposed a framework for dynamic models called the Augmented Dynamic Adaptive Model (ADAM).
-This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a].
-Based on this framework, we use the ETS(M,N,M) model with frequencies 24 (hour of the day) and 168 (hour of the week), adding dummy variables for the week of the year, holidays and lagged holidays.
-This way, we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
-Given that the data exhibits randomly occurring zeroes, we use the direct probability model of @Svetunkov2019a to treat those values.
-Finally, given the skewness of the empirical distribution observed in the preliminary analysis, we use the Gamma distribution for the error term.
-This model can be formulated as a set of the following equations:
-\begin{equation}
+@SvetunkovAdam2021 proposed a framework for dynamic models called the
+Augmented Dynamic Adaptive Model (ADAM). This framework encompasses
+ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting
+multiple frequencies, non-normal distributions and intermittent demand
+[@Svetunkov2019a]. Based on this framework, we use the ETS(M,N,M) model
+with frequencies 24 (hour of the day) and 168 (hour of the week), adding
+dummy variables for the week of the year, holidays and lagged holidays.
+This way, we update the hour of day and day of week seasonal indices,
+keeping the week of year one fixed, thus reducing the number of
+estimated parameters. Given that the data exhibits randomly occurring
+zeroes, we use the direct probability model of @Svetunkov2019a to treat
+those values. Finally, given the skewness of the empirical distribution
+observed in the preliminary analysis, we use the Gamma distribution for
+the error term. This model can be formulated as a set of the following
+equations: \begin{equation}
     \begin{aligned}
@@
     (\#eq:ADAMModel)
-\end{equation}
-where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
-Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$.
-@Svetunkov2019a discuss how to estimate such a model.
-We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic nature of the model (level and seasonal components).
-Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them (see Appendix \@ref(quantilesceiling) for the explanation).
-This model is implemented in the `adam()` function from the `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".
+\end{equation} where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and
+$\alpha_a$ are the smoothing parameters, defining how adaptive the
+components of the model should be, $l_t$ is the level component for the
+demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components,
+$\boldsymbol{\beta}$ is the vector of parameters for the explanatory
+variables, $o_t$ is the binary variable, which is equal to one, when
+demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component
+for the occurrence part of the model, and
+$\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where
+$s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution.
+Finally, $a_t$ is an unobservable series, underlying the occurrence part
+of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term
+for $a_t$. @Svetunkov2019a discuss how to estimate such a model. We
+expect this model to perform on par with the Poisson regression,
+potentially outperforming it in some instances, due to the dynamic
+nature of the model (level and seasonal components). Although the data
+is integer-valued, we expect that Gamma distribution will be a good
+approximation for it. If integer-valued quantiles are needed, then
+rounding up can be done for them (see Appendix \@ref(quantilesceiling)
+for the explanation). This model is implemented in the `adam()` function
+from the `smooth` package [@Svetunkov2021Smooth] for R and is denoted in
+our experiment as "ADAM-iETSX".
 
@@
 
-Suppose we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above.
-In that case, the forecasting task becomes one of predicting the future values of that distribution's parameters.
-We then can use Generalised Additive Models for Location, Scale and Shape (GAMLSS).
-These are the distributional regression models where the parameters are modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that suitable distribution and additive model structures can be found.
-In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.
+Suppose we assume that our predictive distribution follows a given
+parametric distribution, as in Poisson regression discussed above. In
+that case, the forecasting task becomes one of predicting the future
+values of that distribution's parameters. We then can use Generalised
+Additive Models for Location, Scale and Shape (GAMLSS). These are the
+distributional regression models where the parameters are modelled as
+additive function of explanatory variables. This provides a powerful
+and flexible framework for probabilistic forecasting, provided that
+suitable distribution and additive model structures can be found. In
+practice, this means employing expert judgement and experimenting with
+various distributions and evaluating their suitability using available
+training data.
 
-Let $F_t(y_t)$ be a predictive cumulative probability distribution of $y_t$.
-In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters.
-In a GAMLSS framework of @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled as
+Let $F_t(y_t)$ be a predictive cumulative probability distribution of
+$y_t$. In a distributional regression context, $F_t(y_t)$ is modelled
+via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is
+an $m$-dimensional vector of parameters. In a GAMLSS framework of
+@Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled as
 <!-- If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters.
@@
     (\#eq:basicGAM)
-\end{equation}
-where $g_j$ is a monotonic link function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**.
-If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third elements of $\bm x_t$.
-Each $f_{j,i}$ is a smooth function, constructed as
-\begin{equation}
+\end{equation} where $g_j$ is a monotonic link function,
+$\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix
+$\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients,
+$\bm x_t$ is a $d$-dimensional vector of covariates and
+$S_{j,i} \subset \{1, \dots, d\}$ **is ...**. If $S_{j,i} = \{1, 3\}$,
+then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional
+vector formed by the first and third elements of $\bm x_t$. Each
+$f_{j,i}$ is a smooth function, constructed as \begin{equation}
     f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
     (\#eq:smmothfunction)
-\end{equation}
-where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
-The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See @Wood2017 for a detailed introduction to *GAM/GAMLSS* models, smoothing splines bases and penalties.
+\end{equation} where $b^{ji}_k$ are spline basis functions of dimension
+$\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients.
+The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the
+definition of smoothness being dependent on the type of effect and
+penalty being used. See @Wood2017 for a detailed introduction to
+*GAM/GAMLSS* models, smoothing splines bases and penalties.
 
-As our data are counts, the natural starting point is the Poisson distribution with an additive model for $\log \lambda_t$ of the form
+As our data are counts, the natural starting point is the Poisson
+distribution with an additive model for $\log \lambda_t$ of the form
 \begin{equation}
@@
 
-The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.
-This model is called Poisson-1 in discussions below.
+The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day
+(1--24), day of the week (1--7), and day of the year (1-366) at time
+$t$, respectively, and $C_t$ is the temperature at time $t$. This model
+is called Poisson-1 in discussions below.
 
-However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
-In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
-Therefore, we consider more flexible, two-parameter distributions to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
-The resulting density forecasts are given by
-\begin{equation}
+However, experiments on the training data reveal that calibration of
+forecasts based on the Poisson distribution is poor, suggesting that the
+shape of the distribution is unsuitable for the present application. In
+particular, we observe that forecast uncertainty appears to vary
+depending on the time of day and possibly other explanatory variables.
+Therefore, we consider more flexible, two-parameter distributions to
+specify additive models for both location and scale parameters,
+specifically the truncated Normal distribution, with truncation at 0.
+The resulting density forecasts are given by \begin{equation}
 F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
 (\#eq:truncatedn)
-\end{equation}
-with additive models
-<!-- \begin{align*} -->
+\end{equation} with additive models <!-- \begin{align*} -->
 <!-- \mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + \\ -->
@@
 <!-- log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t)) -->
-<!-- (\#eq:additivemodel_NOtr_v2) -->
-<!-- \end{align*} -->
+<!-- (\#eq:additivemodel_NOtr_v2) --> <!-- \end{align*} -->
 \begin{align*}
@@
     & log(\sigma_t) = \sum_{i=1}^{10} D^{+}_i(t) f(H(t))
-\end{align*}
-for the mean and variance parameters. This model is referred to as NOtr-1 below.
+\end{align*} for the mean and variance parameters. This model is
+referred to as NOtr-1 below.
 
-Furthermore, we consider an extension to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$.
-These models are labelled Poisson-2 and NOtr-2.
-We also performed experiments with the truncated $t$ distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.
+Furthermore, we consider an extension to the additive models for
+$\lambda_t$ and $\mu_t$ above to incorporate school and public holidays
+into $D$. These models are labelled Poisson-2 and NOtr-2. We also
+performed experiments with the truncated $t$ distribution (Ttr-2) and
+negative binomial distribution (NBI-2), but these did not result in
+forecasts as well calibrated as the truncated normal.
 
@@
 <!-- <!-- \begin{equation} -->
+
 <!-- <!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} -->
+
 <!-- <!--  (\#eq:truncatedn) -->
+
 <!-- <!-- \end{equation} -->
@@
 <!-- \begin{align*} -->
+
 <!--   \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\ -->
+
 <!--   log(\sigma_t) & = f(H(t)) . -->
+
 <!-- \end{align*} -->
@@
 
-In order to assess the performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps of 0.05 and conditional expectations for 0 to 48 hours ahead produced by each model.
-We forecast up to 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
-The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins.
-Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and expectation.
-The latter is measured via Root Mean Squared Error (RMSE):
-\begin{equation}
+In order to assess the performance of models, we evaluate predictive
+quantiles at probability levels 0.05 to 0.95 in steps of 0.05 and
+conditional expectations for 0 to 48 hours ahead produced by each model.
+We forecast up to 48 hours because this is the operational horizon in
+the ED, for which it is possible to make short-term changes in the
+shifts for nurses and doctors. The forecasts are produced every 12 hours
+for the holdout of 365 days in a rolling origin fashion [@Tashman2000],
+resulting in 727 origins. Based on these values, several error measures
+are calculated to evaluate the performance of models in terms of
+specific quantiles and expectation. The latter is measured via Root Mean
+Squared Error (RMSE): \begin{equation}
   \mathrm{RMSE} = \sqrt{\frac{1}{h} \sum_{j=1}^h e_{t+j}^2} ,
   (\#eq:RMSE)
-\end{equation}
-where $h$ is the forecast horizon and $e_{t+j}$ is the point forecast error $j$ steps ahead.
+\end{equation} where $h$ is the forecast horizon and $e_{t+j}$ is the
+point forecast error $j$ steps ahead.
 
-The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated [@gneiting2007probabilistic].
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision-makers provided the forecast is calibrated.
-Calibration, also called reliability, is the property that forecast probabilities match the observed frequency of realisations.
-If a forecast is calibrated, then, for example, $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size).
-This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbf{1}(y_{t}\leq q_{\alpha,t})$).
-We use several scores to assess the quantile performance of models.
+The objective of density forecasts is to be as sharp as possible while
+remaining reliable/calibrated [@gneiting2007probabilistic]. A forecast
+is said to be sharp if the predictive distribution has a relatively
+small spread, indicating low uncertainty, which is valuable to
+decision-makers provided the forecast is calibrated. Calibration, also
+called reliability, is the property that forecast probabilities match
+the observed frequency of realisations. If a forecast is calibrated,
+then, for example, $20\%$ of observations should fall below the
+$\alpha=0.2$ predictive quantile (with some tolerance based on the
+finite sample size). This property is necessary for forecast
+probabilities to be used in quantitative decision-making. Calibration is
+typically evaluated visually using reliability diagrams, which plot the
+nominal coverage, $\alpha$, against observed frequency mean
+($\mathbf{1}(y_{t}\leq q_{\alpha,t})$). We use several scores to assess
+the quantile performance of models.
 
-First, to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score [@hyndman2021forecasting].
-It rewards sharpness and penalises miscalibration, so it measures overall performance.
-However, calibration should still be verified separately.
-Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in a quantile regression model.
-The Pinball Score is given by
-\begin{equation}
+First, to measure quantile performance, we need to calculate the pinball
+score, which is a strictly proper score used to evaluate quantile
+forecasts and is the discrete form of the Continuous Rank Probability
+Score [@hyndman2021forecasting]. It rewards sharpness and penalises
+miscalibration, so it measures overall performance. However, calibration
+should still be verified separately. Furthermore, The Pinball Score for
+an individual quantile matches the loss function minimised in a quantile
+regression model. The Pinball Score is given by \begin{equation}
     \text{Pinball} = 
@@
  (\#eq:pinball)
-\end{equation}
-where $\mathcal{A} = \{0.05,0.1,...,0.95\}$ is the set of quantiles being estimated.
+\end{equation} where $\mathcal{A} = \{0.05,0.1,...,0.95\}$ is the set of
+quantiles being estimated.
 
-To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
-\begin{equation}
+To compare model performance, and the significance of any apparent
+difference in performance, we will use skill scores, which can be
+calculated for any metric via: \begin{equation}
   \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
-\end{equation}
-where $M$ is the metric's value for the method being considered, $M_\mathrm{ref}$ is the metric's value for a reference method.
-The skill score shows by how many per cent the reference approach is worse than the one under consideration.
-We will use bootstrap re-sampling of skill scores to determine if the differences in forecast performance (i.e.
-positive or negative skill) are significantly different from zero [@Efron1981Bootstrap].
-Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model and employ a block-bootstrap with blocks of length 24h to account for temporal correlations of the underlying data [@hongyi1996bootstrapping;@Bergmeir2016303].
+\end{equation} where $M$ is the metric's value for the method being
+considered, $M_\mathrm{ref}$ is the metric's value for a reference
+method. The skill score shows by how many per cent the reference
+approach is worse than the one under consideration. We will use
+bootstrap re-sampling of skill scores to determine if the differences in
+forecast performance (i.e. positive or negative skill) are significantly
+different from zero [@Efron1981Bootstrap]. Here we use the best
+performing simple benchmark, Naive (explained in Subsection
+\@ref(naive)), as the reference model and employ a block-bootstrap
+with blocks of length 24h to account for temporal correlations of the
+underlying data [@hongyi1996bootstrapping; @Bergmeir2016303].
 
-Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function.
-All functions were re-estimated on each iteration.
-ADAM and Poisson regression estimated the parameters, taking the ones obtained in the initial model application to the data in the first origin as the pre-initials.
-This allowed to speed up the computation for these two models.
-The initial estimation of ADAM took approximately one hour and 25 minutes.
-Each step in the experiment took the time shown in Section \@ref(result).
+Finally, we have calculated the computational time for one iteration on
+the first rolling origin to compare the speed of each function. All
+functions were re-estimated on each iteration. ADAM and Poisson
+regression estimated the parameters, taking the ones obtained in the
+initial model application to the data in the first origin as the
+pre-initials. This allowed to speed up the computation for these two
+models. The initial estimation of ADAM took approximately one hour and
+25 minutes. Each step in the experiment took the time shown in Section
+\@ref(result).
 
@@
 
-The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
-The rolling origin advances in 12-hour steps and the forecast horizon is set for 48 steps ahead.
+The data is portioned into training (from 2014-04-01 to 2018-02-28) and
+test (from 2018-03-01 to 2019-02-28) sets, with all model development
+and hyper-parameter tuning performed using training data only. The
+rolling origin advances in 12-hour steps and the forecast horizon is set
+for 48 steps ahead.
 
-Figure \@ref(fig:Pinball) presents pinball score aggregated across forecasting horizons for each quantile.
-It shows that the difference in performance among the models mainly comes from the middle of the distribution and somewhat from the upper tail.
-There is very little difference in performance for the lower tail (except for TBATS, which has a consistently higher pinball value than the other models).
-This is interesting and reassuring that the better models are better at probabilities that matter more to decision-makers.
+Figure \@ref(fig:Pinball) presents Pinball score aggregated across
+forecasting horizons for each quantile. It shows that the difference in
+performance among the models mainly comes from the middle of the
+distribution and somewhat from the upper tail. There is very little
+difference in performance for the lower tail (except for TBATS, which
+has a consistently higher Pinball value than the other models). This is
+interesting and reassuring that the better models are better at
+probabilities that matter more to decision-makers.
 
-Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated.
-Mis-calibrated forecasts are unsuitable for use in decision-making, so they should be excluded.
-Calibration is evaluated visually in Figure \@ref(fig:quantile-bias), which highlights a systematic negative bias across all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
-Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-This could result in poor staffing decisions for a user as historic data fail to characterise present conditions accurately.
+Probabilistic forecasts are evaluated following the principle of
+*sharpness subject to calibration*, meaning that the sharper forecast is
+prefered provided that it is calibrated. Mis-calibrated forecasts are
+unsuitable for use in decision-making, so they should be excluded.
+Calibration is evaluated visually in Figure \@ref(fig:quantile-bias),
+which highlights a systematic negative bias across all probability
+levels in many models, with only the truncated normal and $t$ family
+GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
+calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
+exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals.
+This could result in poor staffing decisions.
+This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}
 
@@
 <!-- ```{r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"} -->
+
 <!-- ``` -->
@@
 <!-- ```{r Reliability, fig.cap= "Reliability score, showing the relation between empirical and theoretical quantiles.", fig.align='center',out.width= "90%"} -->
+
 <!-- ``` -->
@@
 
+````{=html}
 <!--
@@
 
- Figure \@ref(fig:Skill-rel2bench) illustrates the trade-off between calibration and pinball skill. <!-- We may need to expand this-->
+ Figure \@ref(fig:Skill-rel2bench) illustrates the trade-off between calibration and Pinball skill. <!-- We may need to expand this-->
+````
 
 <!-- ```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"} -->
+
 <!-- ``` -->
@@
 
-Figure \@ref(fig:lead-time-rmse) reports the RMSE for each forecast horizon.
-It illustrates the times of day that are harder to predict – morning pick-up and afternoon peak.
-We can see that some models perform much better than others on specific forecast horizons.
-For example, Benchmark does an excellent job in predicting 1 - 5 steps ahead ED arrivals and, in general, doing well in forecasting arrivals in the night.
-In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, Prophet, TBATS and GBM do for some lead times.
+Figure \@ref(fig:lead-time-rmse) reports the RMSE for each forecast
+horizon. It illustrates the times of day that are harder to predict --
+morning pick-up and afternoon peak. We can see that some models perform
+much better than others on specific forecast horizons. For example,
+Benchmark does an excellent job in predicting 1 - 5 steps ahead ED
+arrivals and, in general, doing well in forecasting arrivals in the
+night. In fact, the Benchmark model performs consistently well in terms
+of the conditional mean, not making as huge mistakes as, for example,
+Prophet, TBATS and GBM do for some lead times.
 
 <!-- ```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"} -->
+
 <!-- ``` -->
@@
 
-In above figures (Figures \@ref(fig:Pinball)- \@ref(fig:lead-time-rmse)), we only present the performance of the top seven methods.
-However, we have evaluated the performance of 14 methods from the test period.
-Evaluation metrics and computational time (in minutes) for all methods are presented in Table \@ref(tab:tab-results).
-They are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
+In above figures (Figures \@ref(fig:Pinball)-
+\@ref(fig:lead-time-rmse)), we only present the performance of the top
+seven methods. However, we have evaluated the performance of 14 methods
+from the test period. Evaluation metrics and computational time (in
+minutes) for all methods are presented in Table \@ref(tab:tab-results).
+They are ordered by Quantile Bias. The five models identified above have
+a Quantile Bias of 0.014 or less, which is substantially lower than the
+next group of forecasts with Quantile Biases of 0.037 and above, ETS
+being the only exception with a value of 0.019.
 
-One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones (Table \@ref(tab:tab-results)).
-This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered a reasonable strategy.
+One more thing to notice is that the ADAM-iETSX model with rounded up
+quantiles did not perform better than the simpler one with continuous
+ones (Table \@ref(tab:tab-results)). This implies that the rounding is
+not necessary in general, but if integer values are needed (for example,
+to decide how many nurses to have), then using the continuous model and
+then rounding up the quantiles could be considered a reasonable
+strategy.
 
@@
 
-Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours, presented in Table \@ref(tab:tab-results).
-The X-axis shows the speed of each method presented as slow, moderate or fast.
-We observe that there is no clear association between speed and accuracy improvement of the models used in this study.
+Finally, Figure \@ref(fig:time-accuracy) reports the forecast
+performance of each approach versus the computational time required to
+generate the forecast for a given forecasting horizon of 48 hours,
+presented in Table \@ref(tab:tab-results). The X-axis shows the speed of
+each method presented as slow, moderate or fast. We observe that there
+is no clear association between speed and accuracy improvement of the
+models used in this study.
 
-We found that most of the methods considered in this paper fall into the fast category with a different range of performance, depending on the type of error measure.
-Among the fastest methods, ETS is very competitive when assessed using Quantile Bias and RMSE.
-Moreover, Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by RMSE and Pinball.
+We found that most of the methods considered in this paper fall into the
+fast category with a different range of performance, depending on the
+type of error measure. Among the fastest methods, ETS is very
+competitive when assessed using Quantile Bias and RMSE. Moreover,
+Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by
+RMSE and Pinball.
 
-Figure \@ref(fig:time-accuracy) shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures.
-The figure also indicates that while the $t$ family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures.
-Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
+Figure \@ref(fig:time-accuracy) shows that NOtr-2 is the fastest method
+that provides consistent accuracy assessed via all three accuracy
+measures. The figure also indicates that while the $t$ family GAMLSS
+model (Ttr-2) is the slowest method, it has a very good performance
+across the three presented accuracy measures. Therefore, if the speed is
+not a major constraint when generating the forecasts, one may employ
+this approach to generate forecasts.
+
+\textcolor{blue}{The benefits of the proposed model and its managerial aspects include
+the use of probabilistic forecast to inform decision making and provide
+insights on the running time of each model to generate forecasts. Probabilistic forecasts contain all potential future outcomes and help
+planners to achieve a more efficient decision making by not only allocate a probability to the most likely outcome but also to unlikely
+outcomes such as extremely high or low arrivals in the Emergency Departments.
+The running time could be also an important aspects for managers depending on the frequency of generating forecasts.
+If the frequency of forecast generation is high, one may employ models such as
+Poisson regression (Poisson 1). If the frequency is lower and running time is not a big concern, a model like GAMLSS (Ttr-2) should be used.}
+
+\textcolor{blue}{Probabilistic forecasts can be considered as a useful tool for planners to manage the risk of under and over resource allocation, which consequently may reduce both costs and risks of service delivery in Emergency Departments.}
 
@@
 
-Short-term forecasting of arrivals at emergency departments is an important element of hospital staff and resource management.
-Furthermore, due to the asymmetric impact of an excess resource shortage, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
-In this paper, we have developed methods for producing a probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
+Short-term forecasting of arrivals at emergency departments is an
+important element of hospital staff and resource management.
+Furthermore, due to the asymmetric impact of an excess resource
+shortage, especially in emergency departments, quantifying forecast
+uncertainty is also of value as it enables planners to manage associated
+risks. In this paper, we have developed methods for producing a
+probabilistic forecast of hourly arrivals up to 48 hours ahead,
+comparing different state-of-the-art approaches.
 
-Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression.
-The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution.
-The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature.
-Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score.
-This suggests that the latter may be preferred if the whole distribution is used in decision-making.
+Two approaches produced highly accurate, calibrated probabilistic
+forecasts: a time series model and a model based on distributional
+regression. The first is ADAM-iESTX, an extension of exponential
+smoothing incorporating multiple seasonalities, explanatory variables
+and assuming a Gamma predictive distribution. The second, labeled
+NOtr-2, regressed the two parameters of a truncated (at zero) normal
+distribution on the date and time features and temperature. Both
+approaches produced calibrated probabilistic forecasts, but the point
+prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while
+NOtr-2 produced forecasts with a lower Pinball score. This suggests that
+the latter may be preferred if the whole distribution is used in
+decision-making.
 
-Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts and choice of model features are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without the need for post-processing.
-These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this paper.
-Furthermore, methods based on continuous-valued distributions are not adversely affected by the fact that the data are integer-valued.
-Rounding up predictive quantiles to the next integer does not make predictions worse.
+Having compared the performance of a wide range of methods, we make the
+following observations: the choice of distribution assumed for
+probabilistic forecasts and choice of model features are as if not more
+important than the type of model employed; methods based on quantile
+regression, which do not assume a parametric distribution for forecasts,
+do not perform as well as those based on parametric distributions; and
+the best performing models handled the non-negative and skewed nature of
+the data automatically without the need for post-processing. These
+observations reflect the characteristics of the data, which is
+representative of ED arrivals, but determining the extent to which they
+generalise is beyond the scope of this paper. Furthermore, methods based
+on continuous-valued distributions are not adversely affected by the
+fact that the data are integer-valued. Rounding up predictive quantiles
+to the next integer does not make predictions worse.
 
-Finally, we have found that out-of-the-box models, which require minimal tuning or manual development, do not perform as well as well-considered statistical approaches.
-The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks.
-Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
-ETS produces reasonably well-calibrated forecasts, in contrast to the poorly calibrated benchmarks and has highly accurate point forecasts.
-However, its probabilistic forecasts were considerably worse than  NOtr-2 in terms of pinball score.
+Finally, we have found that out-of-the-box models, which require minimal
+tuning or manual development, do not perform as well as well-considered
+statistical approaches. The popular TBATS, Prophet and Gradient Boosting
+Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and
+even the benchmarks. Of the models requiring a modest amount of user
+input and expertise, exponential smoothing (ETS) was found to perform
+well. ETS produces reasonably well-calibrated forecasts, in contrast to
+the poorly calibrated benchmarks and has highly accurate point
+forecasts. However, its probabilistic forecasts were considerably worse
+than NOtr-2 in terms of Pinball score.
 
-Probabilistic forecasting opens the door to more sophisticated resource management in healthcare settings by providing decision-makers with uncertainty information and enabling quantitative risk management.
-Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecasting skills and may also be necessary to maximise benefits through more holistic decision-making.
+\textcolor{blue}{
+The dataset used in this study does not include the COVID-19 pandemic.
+During COVID-19, the dynamic of ED has changed.
+That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period.
+However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}
 
-Further research is required to investigate the practical benefits of probabilistic forecasts in healthcare and how they can inform planning and decision making.
-This may require employing discrete simulation or application to the newsvendor problems.
-While this study has focused on hourly short-term forecasting, producing longer-term daily forecast (e.g.
-180-270 days ahead) is often required by planners to support winter planning in ED and Ambulance services which requires more investigation.
-Moreover, more research is needed in the forecasting of other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
-This may require considering the dynamics among various services, including General Practitioners, Emergency Departments, Ambulance and Fire & Rescue services.
+Probabilistic forecasting opens the door to more sophisticated resource
+management in healthcare settings by providing decision-makers with
+uncertainty information and enabling quantitative risk management.
+Linking forecasts of arrivals with upstream (ambulance call-outs) and
+downstream (length of stay, medical outcomes) analytics offers an
+opportunity to improve forecasting skills and may also be necessary to
+maximise benefits through more holistic decision-making.
+
+Further research is required to investigate the practical benefits of
+probabilistic forecasts in healthcare and how they can inform planning
+and decision making. This may require employing discrete event simulation or
+application to the newsvendor problems. While this study has focused on
+hourly short-term forecasting, producing longer-term daily forecast
+(e.g. 180-270 days ahead) is often required by planners to support
+winter planning in ED and Ambulance services which requires more
+investigation. Moreover, more research is needed in the forecasting of
+other important variables such as length of stay, bed occupancy and
+waiting time, in addition to patient arrivals and admissions. This may
+require considering the dynamics among various services, including
+General Practitioners, Emergency Departments, Ambulance and Fire &
+Rescue services.
 
@@
 
-Before proceeding with the proof, we need to define the quantiles of the continuous and rounded up random variables:
+Before proceeding with the proof, we need to define the quantiles of the
+continuous and rounded up random variables:
 \begin{equation} \label{eq:quantCeil1}
     P \left(y_t < k \right) = 1 - \alpha ,
-\end{equation}
-and
-\begin{equation} \label{eq:quantCeil2}
+\end{equation} and \begin{equation} \label{eq:quantCeil2}
     P \left(\lceil y_t \rceil \leq n \right) \geq 1 - \alpha ,
-\end{equation}
-where $n$ is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and $k$ is the quantile of the continuous distribution of the variable.
+\end{equation} where $n$ is the quantile of the distribution of rounded
+up values (the smallest integer number that satisfies the inequality
+\eqref{eq:quantCeil2}) and $k$ is the quantile of the continuous
+distribution of the variable.
 
-In order to prove that $n = \lceil k \rceil$, we need to use the following basic property:
-\begin{equation} \label{eq:quantCeil3}
+In order to prove that $n = \lceil k \rceil$, we need to use the
+following basic property: \begin{equation} \label{eq:quantCeil3}
     \lceil y_t \rceil \leq n \iff  y_t \leq n,
-\end{equation}
-which means that the rounded up value will always be less than or equal to $n$ if and only if the original value is less than or equal to $n$.
-Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
+\end{equation} which means that the rounded up value will always be less
+than or equal to $n$ if and only if the original value is less than or
+equal to $n$. Taking into account \eqref{eq:quantCeil3}, the probability
+\eqref{eq:quantCeil2} can be rewritten as:
 \begin{equation} \label{eq:quantCeil4}
@@
 
-Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2}, \eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the following can be summarised:
-\begin{equation} \label{eq:quantCeil6}
+Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2},
+\eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the
+following can be summarised: \begin{equation} \label{eq:quantCeil6}
     P \left(y_t \leq n-1 \right) < P \left(y_t < k \right) \leq P \left(y_t \leq n \right) ,
-\end{equation}
-which is possible only when $k \in (n-1, n]$, which means that $\lceil k \rceil = n$.
-So the rounded up quantile of continuous random variable $y_t$ will always be equal to the quantile of the descritised value of $y_t$:
+\end{equation} which is possible only when $k \in (n-1, n]$, which means
+that $\lceil k \rceil = n$. So the rounded up quantile of continuous
+random variable $y_t$ will always be equal to the quantile of the
+descritised value of $y_t$:
 \begin{equation} \label{eq:ceilingAndQuantiles1}
@@
 
-It is also worth noting that the same results can be obtained with the floor function instead of ceiling, following the same logic. So the following equation will hold for all $y_t$ as well:
+It is also worth noting that the same results can be obtained with the
+floor function instead of ceiling, following the same logic. So the
+following equation will hold for all $y_t$ as well:
 \begin{equation} \label{eq:floorAndQuantiles1}

### 2022-12-28 paper/paper.tex
comments adressed by bahman

@@
 %%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+
 \usepackage[hyphens]{url}
@@
   \journal{Annals of Emergency Medicine} % Sets Journal name
-
-\usepackage{lineno} % add
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \usepackage{amssymb,amsmath}
+% TODO: Currently lineno needs to be loaded after amsmath because of conflict
+% https://github.com/latex-lineno/lineno/issues/5
+\usepackage{lineno} % add
 \usepackage{ifxetex,ifluatex}
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
-\usepackage{natbib}
+\usepackage[top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
+\usepackage[]{natbib}
 \bibliographystyle{plainnat}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
+
 \ifxetex
@@
             pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
 
@@
 
-% Pandoc citation processing
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-% Pandoc header
-\usepackage{adjustbox, float,lscape, bm,amsmath}
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
+\usepackage{adjustbox, float,lscape,amsmath,bm}
 \usepackage{booktabs}
@@
 \begin{document}
+
 \begin{frontmatter}
@@
   \title{Probabilistic forecasting of hourly Emergency Department arrivals}
-    \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
-   \ead{rostami-tabarb@cardiff.ac.uk} 
-    \author[University of Glasgow]{Jethro Browell\corref{1}}
-   \ead{jethro.browell@glasgow.ac.uk} 
-    \author[Lancaster University]{Ivan Svetunkov\corref{1}}
-   \ead{i.svetunkov@lancaster.ac.uk} 
-      \address[Cardiff Business School]{Cardiff Business School, Cardiff University, UK}
-    \address[University of Glasgow]{School of Mathematics and Statistics, University of Glasgow, UK}
-    \address[Lancaster University]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK}
-      \cortext[12]{Corresponding Author}
-    \cortext[1]{Equal contribution}
+      \cortext[cor1]{Corresponding author}
 
@@
   \end{abstract}
-   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
+    \begin{keyword}
+    Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing \sep 
+    Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing
+  \end{keyword}
+
  \end{frontmatter}
@@
 
-Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients.
-Accurate ED demand forecasts contribute to a better decision-making process regarding resources allocation and staffing.
-This is one of the best ways to optimise resources utilisation and minimise related costs.
-An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives.
-Inability to match the staff with the demand might result in patients overcrowding the system, which is a severe problem that causes challenges for the patient flow \citep{derlet2002overcrowding}.
-Also, it is related to increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other \citep{rostami2020anticipating}.
+Forecasting Emergency Department (ED) arrivals is critical for informing
+staffing and scheduling decisions to meet the needs of patients.
+Accurate ED demand forecasts contribute to a better decision-making
+process regarding resources allocation and staffing. This is one of the
+best ways to optimise resources utilisation and minimise related costs.
+An accurate forecast of patient arrivals is crucial in ED services to
+depict various courses of action that can result in massive savings in
+terms of patient lives. Inability to match the staff with the demand
+might result in patients overcrowding the system, which is a severe
+problem that causes challenges for the patient flow
+\citep{derlet2002overcrowding}. Also, it is related to increasing length of
+stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected
+return visits to services, increased health care costs, inaccuracy in
+electronic medical records and other \citep{rostami2020anticipating}.
 
-Accurate forecasting of arrivals by the hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff.
-This has many advantages for both patients, staff and the quality of provided services.
-Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day.
-This involves the short-term decision-making related to the execution of the delivery process in ED.
-The combination of an hourly arrival forecast, current staff, being occupied, resource availability and waiting times at ED provide information on the state of the unscheduled care system across the service.
-Having this complete picture enables the delivery managers to focus on the areas that require intervention to allow the most effective delivery of the service to the patients.
-However, compared with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
-Hourly time series generally exhibit multiple seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
-They may also express nonstationarity, and their profile may change over time.
-Therefore, an appropriate forecasting model should consider these features to accurately predict hourly demand admissions.
+Accurate forecasting of arrivals by the hour of the day enables planners
+to match staff to meet anticipated patients, reconfigure units and
+redeploy staff. This has many advantages for both patients, staff and
+the quality of provided services. Hourly forecasts are required to
+inform the short-term operational planning for the current and the
+upcoming shifts of the day. This involves the short-term decision-making
+related to the execution of the delivery process in ED. The combination
+of an hourly arrival forecast, current staff, being occupied, resource
+availability and waiting times at ED provide information on the state of
+the unscheduled care system across the service. Having this complete
+picture enables the delivery managers to focus on the areas that require
+intervention to allow the most effective delivery of the service to the
+patients. However, compared with lower frequency time series forecasting
+such as monthly, quarterly and yearly, hourly forecasts are challenging
+because the noise caused by random variation may overshadow any pattern
+in the time series. Hourly time series generally exhibit multiple
+seasonal cycles of different lengths: hourly, daily, weekly, and yearly.
+They may also express nonstationarity, and their profile may change over
+time. Therefore, an appropriate forecasting model should consider these
+features to accurately predict hourly demand admissions.
 
-There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events.
-These studies use multiple approaches, including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}.
-However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric.
+There are few studies that look at forecasting hourly arrivals in ED and
+other hospital services using historical time series data and/or
+predictors such as patient characteristics, weather, holidays and public
+events. These studies use multiple approaches, including Exponential
+Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
+(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
+Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
+model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
+Neural Networks \citep{hyndman2021forecasting}.
+\textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study.
+These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
+forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
 Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
-
-Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated.
+Furthermore, if the impact of under- and overstaffing can be quantified,
+probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated.
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-Another drawback of existing studies is that the datasets used are relatively small (e.g.~time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
+\textcolor{blue}{Another drawback of existing studies is that the
+datasets used are relatively small (e.g.
+time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
 Such results might not be generalisable.
 Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals.
-Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.
+Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
+Finally, all previous publications
+referenced in this paper are not fully reproducible as underlying data and code are not available.}
 
@@
 \item
-  We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
+  We produce probabilistic forecasts, in addition to the point
+  estimation, quantifying uncertainties in future hospital admission,
+  and comparing different forecasting methods using a suite of
+  well-established evaluation metrics;
 \item
-  We develop an advanced dynamic model to forecast ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
+  We develop an advanced dynamic model to forecast ED arrivals based
+  on iETS \citep{Svetunkov2019a} and ETSX models with a modification for
+  multiple frequencies, which produced highly-accurate point
+  forecasts;
 \item
-  We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
+  We develop a novel model to produce a probabilistic forecast of ED
+  arrivals based on Generalised Additive Models for Location Scale and
+  Shape, which accounts for i) the bounded and non-Gaussian
+  distribution of arrivals, ii) multiple seasonalities, weather and
+  holiday effects, and iii) variation in forecast uncertainty;
 \item
-  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
+  We benchmark the accuracy of our model against appropriate models
+  used when multiple seasonality is present, i.e.~Prophet, TBATS,
+  Poisson Regression, Exponential Smoothing State Space model (ETS)
+  and the simple empirical distribution of the arrivals;
 \item
-  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks.
-The proposed approach could also be generalised to forecast hourly requirements for longer horizons and in other services \citep{al2021empirical}, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
+  We provide data and code enabling reproduction and refinement of the
+  proposed approach and benchmarks. The proposed approach could also
+  be generalised to forecast hourly requirements for longer horizons
+  and in other services \citep{al2021empirical}, such as inpatient and
+  outpatient care services, the number of attended incidents in
+  ambulance services, or call volumes in clinical desk services.
 \end{enumerate}
 
-The rest of the paper is organised as follows: In section \ref{lit}, we provide a brief overview of hourly ED arrival forecasting; In Section \ref{data}, we present the hourly time series of an ED arrival and use various plots to highlight important patterns.
-In section \ref{model}, we describe the modelling approach and benchmark methods.
-We then discuss the performance evaluation metrics in section \ref{accuracy}; in section \ref{result}, we present and discuss our results.
-Finally, we summarise our findings and present ideas for future research in section \ref{conclusion}.
+The rest of the paper is organised as follows: In section \ref{lit}, we
+provide a brief overview of hourly ED arrival forecasting; In Section
+\ref{data}, we present the hourly time series of an ED arrival and use
+various plots to highlight important patterns. In section \ref{model},
+we describe the modelling approach and benchmark methods. We then
+discuss the performance evaluation metrics in section \ref{accuracy};
+in section \ref{result}, we present and discuss our results. Finally,
+we summarise our findings and present ideas for future research in
+section \ref{conclusion}.
 
@@
 
-Some studies employ models to forecast admissions and arrivals in Emergency Department \citep{gul2020exhaustive}.
-The time granularity considered by these studies spans from hourly to yearly.
-However, given the focus of the paper, we only discuss hourly ED forecasting.
+\textcolor{blue}{
+There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
+Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention.
+We refer interested readers to  some extensive reviews of the relevant literature} \citep{mingliterature2022, gul2020review, ibrahim2016modeling}.
+\textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare.
+However, given
+the focus of this study, we only discuss hourly ED forecasting.}
 
-\citet{hertzum2017forecasting} used linear regression, ARIMA, and Naïve to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables.
-\citet{hertzum2017forecasting} study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year.
-In terms of the hour of the day, patient arrivals peaked around noon.
-For days of the week, Monday is the busiest day, while weekends are the quietest ones.
-July-August are the months with the highest number of patient arrivals, while January and February have the lowest numbers.
-They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.
+\citet{hertzum2017forecasting} used linear regression, ARIMA, and Naïve to
+investigate whether accurate hourly accident and emergency department
+patient arrivals and occupancy forecasts can be generated using calendar
+variables. \citet{hertzum2017forecasting} study found that patient arrival
+variation is larger across the hours of the day than across the days of
+the week and the months of the year. In terms of the hour of the day,
+patient arrivals peaked around noon. For days of the week, Monday is the
+busiest day, while weekends are the quietest ones. July-August are the
+months with the highest number of patient arrivals, while January and
+February have the lowest numbers. They indicate that regression and
+ARIMA models performed similarly in modelling patient arrivals, while
+ARIMA outperformed regression models in modelling accident and emergency
+department occupancy.
 
-\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neural network methods to forecast hourly accident and emergency department arrivals.
-ARIMA model was selected as the best fit model.
-Authors claimed that ARIMA provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
-\citet{Cheng2021} developed an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which outperformed a rolling-average benchmark.
-They also produce prediction intervals, a form of the probabilistic forecast, which were found to be well-calibrated, a necessary property for such forecasts.
+\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neural
+network methods to forecast hourly accident and emergency department
+arrivals. ARIMA model was selected as the best fit model. Authors
+claimed that ARIMA provided high and acceptable hourly ED forecasting
+accuracy, even outperforming TBATS. \citet{Cheng2021} developed an ARIMA model
+for ED occupancy with a seasonal component and exogenous variables,
+which outperformed a rolling-average benchmark. They also produce
+prediction intervals, a form of the probabilistic forecast, which were
+found to be well-calibrated, a necessary property for such forecasts.
 
-\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic.
-Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for a horizon of 168 hours.
-The hourly data collected from an ED in Pennsylvania showed no trend and two seasonal cycles: a within-day and a within-week seasonal cycles.
-The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.
+\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in
+which each component of the time series is separately modelled as
+stochastic. Double-seasonal exponential smoothing and standard
+Holt-Winters were used to forecast ED arrival for a horizon of 168
+hours. The hourly data collected from an ED in Pennsylvania showed no
+trend and two seasonal cycles: a within-day and a within-week seasonal
+cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly
+seasonality of the time series.
 
-\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day-of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services.
-The authors produced probabilistic predictions in the form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly arrivals.
-Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it was suggested that the model could be used for forecasting.
-However, model evaluation was performed in-sample on only one year of data, so it is unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
-However, the length of the time series in this study was very short (only one year), which did not allow for a rigorous out-of-sample evaluation.
+\citet{mccarthy2008challenge} employed a Poisson log-linear regression model,
+including independent variables such as temporal factors (e.g.,
+hour-of-day, day-of-week, type-of-day, season, and calendar year),
+patient characteristics (i.e., age, gender, insurance status, triage
+level, mode of arrival, and ambulance diversion status) and climatic
+factors (i.e., temperature and precipitation) to model patient demand
+for ED services. The authors produced probabilistic predictions in the
+form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly
+arrivals. Hourly data of ED arrivals in the 1-year study period was
+modelled and analysed, and it was suggested that the model could be used
+for forecasting. However, model evaluation was performed in-sample on
+only one year of data, so it is unclear how this approach would perform
+in a forecasting setting or compare to simpler approaches. However, the
+length of the time series in this study was very short (only one year),
+which did not allow for a rigorous out-of-sample evaluation.
 
-\citet{schweigler2009forecasting} investigated whether time series methods could accurately generate short-term forecasts of ED bed occupancy.
-A year-long dataset of hourly ED bed occupancy was collected from three facilities.
-The authors implemented an hourly historical average model, SARIMA model, and sinusoidal model with autocorrelated error for each facility.
-The historical average model was based on the mean occupancy for each site, for each hour of the day, while the sinusoidal model was based on four parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept.
-They evaluated the forecast accuracy of four and twelve hours forecast horizons using RMSE.
-They found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by \(33\%\) the 12-hour forecasts generated by the historical average).
+\citet{schweigler2009forecasting} investigated whether time series methods
+could accurately generate short-term forecasts of ED bed occupancy. A
+year-long dataset of hourly ED bed occupancy was collected from three
+facilities. The authors implemented an hourly historical average model,
+SARIMA model, and sinusoidal model with autocorrelated error for each
+facility. The historical average model was based on the mean occupancy
+for each site, for each hour of the day, while the sinusoidal model was
+based on four parameters: an AR term, a sine coefficient, a cosine
+coefficient and an intercept. They evaluated the forecast accuracy of
+four and twelve hours forecast horizons using RMSE. They found that both
+SARIMA and the sinusoidal models outperformed the historical average
+(for example, at site 2, the two models improved by \(33\%\) the 12-hour
+forecasts generated by the historical average).
 
-\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to predict patient volume for a Hospital Medicine programme.
-The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to forecast for 4 and 24 hours ahead.
-They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy.
-The authors found that the ARIMA outperformed all the other models.
+\citet{kim2014predicting} compared different univariate and multivariate time
+series forecasting techniques to predict patient volume for a Hospital
+Medicine programme. The study evaluated linear regression, exponential
+smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional
+Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to
+forecast for 4 and 24 hours ahead. They used Mean Absolute Percentage
+Error (MAPE) to report the forecast accuracy. The authors found that the
+ARIMA outperformed all the other models.
 
-Table \ref{tab:summarylit} summarize the relevant papers.
+Table \ref{tab:summarylit} summarises the relevant papers.
 
@@
 \toprule
-Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Seasonality\\
+Reference & Year & Variable & Horizon & Length of dataset & Method & Metric & Probabilistic & Seasonality\\
 \midrule
-Cheng et al. & 2021 & ED visits & 1h to 4 h & 1 year & SARIMAX, Holt-Winters, VAR, ARIMA & MSE, MAE, MAPE, Prediction interval coverage & NO & Single\\
-Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 years & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Multiple\\
-Asheim et al. & 2019 & ED arrivals & 3h & 5 years & Poisson regression & MAPE & No & Single\\
-Hertzum & 2017 & ED arrivals & 1,2,4,8,24 h & 3 years & linear regression; SARIMA; Naïve & MAE, MAPE, 
-MASE & No & Single\\
-Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 years & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & Single\\
-Cote et al. & 2013 & ED arrivals & 24h & 2 years & Fourier regression & R\textasciicircum{}2,  Standard Error & No & Single\\
+Current study & 2023 & ED arrivals & 48h & 5 years & Naïve; Poisson \& Quantile Regression; Exponential Smoothing; Prophet; TBATS;    Gradient Boosting ; ADAM; GAM & RMSE, Pinball score, skill score, Quantile bias & YES & YES\\
+Cheng et al. & 2021 & ED visits & 1h to 4h & 1 year & SARIMAX; Holt-Winters; VAR; ARIMA & MSE, MAE, MAPE, Prediction interval coverage & NO & Single\\
+Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 years & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & NO & Multiple\\
+Asheim et al. & 2019 & ED arrivals & 3h & 5 years & Poisson regression & MAPE & NO & Single\\
+Hertzum & 2017 & ED arrivals & 1,2,4,8,24h & 3 years & linear regression; SARIMA; Naïve & MAE, MAPE, 
+MASE & NO & Single\\
+Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 years & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & NO & Single\\
+Cote et al. & 2013 & ED arrivals & 24h & 2 years & Fourier regression & R\textasciicircum{}2,  Standard Error & NO & Single\\
 Chase et al. & 2012 & ED CUR & 30m
-1h, 2h, 4h, 8h, 12h & 1 year & Binary regression & NA & No & Single\\
-Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 years & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & Single\\
-Jones et al. & 2009 & ED census & 24h & 2 years & VAR;  Holt winters & MAE & No & Single\\
-McCarthy & 2008 & ED arrivals & n/a & 1 year & Poisson log-linear regression model & Prediction interval coverage & Yes & Multiple\\
+1h, 2h, 4h, 8h, 12h & 1 year & Binary regression & NA & NO & Single\\
+Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 years & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & NO & Single\\
+Jones et al. & 2009 & ED census & 24h & 2 years & VAR;  Holt winters & MAE & NO & Single\\
+McCarthy & 2008 & ED arrivals & n/a & 1 year & Poisson log-linear regression model & Prediction interval coverage & Partially & Multiple\\
 Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 years & Regression & RMSE & No & Single\\
@@
 
-\citet{asheim2019real} developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead.
-Measuring the accuracy using the MAPE, \citet{asheim2019real} noticed that significant improvement happened when the time of notification was incorporated into the model, especially in the one-hour horizon.
+\citet{asheim2019real} developed a Poisson time-series regression model with
+continuous day-of-week and week-of-year effects to implement a real-time
+system that could forecast ED arrivals on 1, 2, 3 hours ahead. Measuring
+the accuracy using the MAPE, \citet{asheim2019real} noticed that significant
+improvement happened when the time of notification was incorporated into
+the model, especially in the one-hour horizon.
 
-\citet{cheng2021forecasting} used one year of ED visits time series to evaluate the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead.
-The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of \(80\%\) and \(95\%\).
-They show that SARIMAX provides a more accurate forecast of hourly ED occupancy.
+\citet{cheng2021forecasting} used one year of ED visits time series to evaluate
+the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED
+occupancy up to 4-hours ahead. The forecast accuracy is evaluated using
+Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point
+forecast and coverage for prediction intervals of \(80\%\) and \(95\%\).
+They show that SARIMAX provides a more accurate forecast of hourly ED
+occupancy.
 
-According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established.
-Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low \citep{boyle2012predicting, hertzum2017forecasting}, while others mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
+According to the studies mentioned above, it can be said that they have
+shown complications in forecasting hourly patient accident and emergency
+department visits, and the application of forecasting hourly patient
+visits is not well established. Some of the studies claimed that the
+accuracy of forecasting models on hourly accident and emergency
+department data is low \citep{boyle2012predicting, hertzum2017forecasting},
+while others mentioned that the accuracy of ED hourly forecast is at an
+acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
 
-There are a few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches:
+There are a few limitations in the literature which encourage us to
+undertake this research and examine different forecasting approaches:
 
@@
 \item
-  Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
+  Current approaches to forecast hourly ED arrivals do not fully
+  consider the feature of data such as multiple seasonal cycles and
+  changing profile over time;
 \item
-  Almost all research studies produce point forecasts and, at best, report prediction intervals.
-There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent the uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
+  Almost all research studies produce point forecasts and, at best,
+  report prediction intervals. There is a lack of studies presenting
+  the entire forecast distribution of hourly ED arrivals that better
+  represent the uncertainty of future arrivals, providing a holistic
+  picture of future demand for a planner;
 \item
-  most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
+  most studies are not reproducible, as it is almost impossible to
+  reapply the approaches without the help of the authors of those
+  papers;
 \item
-  studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
+  studies are limited in terms of the length of historical data used
+  for training purposes and forecast performance evaluation and
 \item
-  some studies in this area lack a rigorous experimental design, i.e.~they do not use benchmark methods or report forecast accuracy.
+  some studies in this area lack a rigorous experimental design, i.e.
+  they do not use benchmark methods or report forecast accuracy.
 \end{enumerate}
@@
 
-Data used in this study comprises counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital.
-We aggregated the patients' arrival times to obtain hourly arrivals, which are used in this study.
-Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week.
-Although the data is noisy, it reveals some systematic patterns.
+Data used in this study comprises counts of patients' arrival times at
+one of the largest ED units in the UK between April 2014 and February
+2019, extracted from the ED administrative database of the hospital. We
+aggregated the patients' arrival times to obtain hourly arrivals, which
+are used in this study. Figure \ref{fig:hourly-plot-ridge} illustrates
+the distribution of arrivals for each hour of the day and the day of the
+week. Although the data is noisy, it reveals some systematic patterns.
 
@@
 
-It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by \citet{hertzum2017forecasting}.
-The ED arrivals decrease between midnight and early morning and then increase until the evening, decreasing after that again.
-It is also clear that ED service gets systematically more visits on Mondays between 8 a.m.
-and 5 p.m.
-Moreover, the number of arrivals around midnight is slightly higher for Saturdays and Sundays.
+It is clear that the number of arrivals has a sub-daily structure,
+similar to the one summarised by \citet{hertzum2017forecasting}. The ED
+arrivals decrease between midnight and early morning and then increase
+until the evening, decreasing after that again. It is also clear that ED
+service gets systematically more visits on Mondays between 8 a.m. and 5
+p.m. Moreover, the number of arrivals around midnight is slightly higher
+for Saturdays and Sundays.
 
-Figure \ref{fig:hourly-plot-ridge} also highlights significant skewness for almost every hour of the day that varies with time-of-day, which should be accounted for in forecasting methods.
-Some skewness might be related to holidays and special events.
-It is also clear that arrivals are less volatile between midnight and early morning.
+Figure \ref{fig:hourly-plot-ridge} also highlights significant skewness
+for almost every hour of the day that varies with time-of-day, which
+should be accounted for in forecasting methods. Some skewness might be
+related to holidays and special events. It is also clear that arrivals
+are less volatile between midnight and early morning.
 
@@
 
-\caption{Seasonal plot: day of week arrivals}\label{fig:seasonplot-dofw}
+\caption{Subseries plot: day of weekly arrivals}\label{fig:seasonplot-dofw}
 \end{figure}
 
-Figure \ref{fig:seasonplot-dofw} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals.
-There is a line for each week (i.e.~seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle.
-It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week.
-This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week.
-This might be due to the closure of General Practitioners outpatient clinics over the weekend.
+Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals.
+Each individual plot illustrates how arrivals change over time for a each day of week (i.e.
+seven days) from Monday to Saturday.
+The blue line shows the average arrival for the given day.
+It is clear that ED arrivals on Monday are higher followed by Saturday.
+This indicates that there are
+significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}
 
@@
 
-Figure \ref{fig:seasonplot-weekofyear} highlights the week of year seasonality in the ED arrivals.
-We observe that arrivals are significantly lower from week 29 to 35, corresponding to the Summer period.
-Moreover, the number of arrivals is lower at the beginning and the end of the year.
-The arrivals increase from week 36 and peak in weeks 39-42.
-This corresponds to the September - October period.
+Figure \ref{fig:seasonplot-weekofyear} highlights the week of year
+seasonality in the ED arrivals. We observe that arrivals are
+significantly lower from week 29 to 35, corresponding to the Summer
+period. Moreover, the number of arrivals is lower at the beginning and
+the end of the year. The arrivals increase from week 36 and peak in
+weeks 39-42. This corresponds to the September - October period.
 
@@
 
-Finally, Figure \ref{fig:date-plot} presents the time plot of daily arrivals.
-Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle.
-The figure shows more arrivals on Mondays than on other weekdays, agreeing with the previous findings.
-Moreover, we can observe a long-term trend line and the significant effect of some holidays.
-We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year.
-Moreover, Figure \ref{fig:date-plot} allows us to identify the impact of special events on arrivals.
-For instance, we see that arrivals are significantly low for 01-03 of March 2018.
-These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages, and schools closed in the UK \citep{stormemma2018}.
+Finally, Figure \ref{fig:date-plot} presents the time plot of daily
+arrivals. Each point represents one day, and points are shape-coded by
+day-of-week to show the weekly cycle. The figure shows more arrivals on
+Mondays than on other weekdays, agreeing with the previous findings.
+Moreover, we can observe a long-term trend line and the significant
+effect of some holidays. We can see that arrivals near Christmas and New
+Year's day are significantly lower than other days of the year.
+Moreover, Figure \ref{fig:date-plot} allows us to identify the impact
+of special events on arrivals. For instance, we see that arrivals are
+significantly low for 01-03 of March 2018. These days correspond to the
+Storm Emma with heavy snowfall that resulted in travel disruption, mass
+power outages, and schools closed in the UK \citep{stormemma2018}.
 
-Based on this analysis and the literature review, we should consider models that can take the following into account:
+Based on this analysis and the literature review, we should consider
+models that can take the following into account:
 
@@
 \item
-  Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
+  Lags of calendar events to accommodate the potential changes in
+  demand the next day after a holiday,
 \item
@@
 
-\hypertarget{climatology}{%
-\subsection{Naive/Climatology}\label{climatology}}
+\hypertarget{naive}{%
+\subsection{Naive}\label{naive}}
 
-We start with one of the simplest forecasting approaches used in practice - assuming that in the next few hours, everything will be the same as in the similar hours of a similar day in the past.
-This is called ``Naïve''.
-In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution \citep{la2021new}.
-We consider the empirical distribution of all available historic data (Benchmark-1) and the empirical distribution of the most recent year of historic data on a rolling basis (Benchmark-2) to capture potential changes in behaviour over time.
+We start with one of the simplest forecasting approaches used in
+practice - assuming that in the next few hours, everything will be the
+same as in the similar hours of a similar day in the past. This is
+called ``Naïve''. In our case, given that we need a distribution of
+values, we will use a modified approach, where the
+empirical distribution of the hourly arrival time series is used to
+forecast the future arrival distribution \citep{la2021new}. We consider the
+empirical distribution of all available historic data (Benchmark-1) and
+the empirical distribution of the most recent year of historic data on a
+rolling basis (Benchmark-2) to capture potential changes in behaviour
+over time.
 
@@
 \label{eq:linearRegression}
-\end{equation}
-where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters, \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and \(t\) is the time index.
-However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic because the number of admitted patients is an integer and non-negative.
-So the linear regression model should be substituted by some other model.
-One of the models that is frequently used in practice is the Poisson regression \citep[see, for example,][]{mccarthy2008challenge}, which can be summarised as
-\begin{equation}
+\end{equation} where \(\mathbf{x}_t\) is the vector of explanatory
+variables, \(\boldsymbol{\beta}\) is the vector of parameters,
+\(\epsilon_t\) is the error term, which is typically assumed to follow
+Normal distribution with zero mean and a fixed variance, and \(t\) is the
+time index. However, in the context of healthcare and ED arrivals, the
+assumption of Normality is unrealistic because the number of admitted
+patients is an integer and non-negative. So the linear regression model
+should be substituted by some other model. One of the models that is
+frequently used in practice is the Poisson regression \citep[see, for example,][]{mccarthy2008challenge}, which can be summarised as \begin{equation}
   \begin{aligned}
@@
 
-The logarithm in \eqref{eq:PoissonRegression} is needed to ensure that the parameter of Poisson distribution is always positive.
-This model can be estimated via maximisation of the likelihood function based on Poisson mass function .
-There is no single correct answer when selecting explanatory variables for the model, and the decision needs to be made for each specific case.
-In our experiment, we will only include dummy variables, capturing a variety of calendar events:
+The logarithm in \eqref{eq:PoissonRegression} is needed to ensure that
+the parameter of Poisson distribution is always positive. This model can
+be estimated via maximisation of the likelihood function based on
+Poisson mass function . There is no single
+correct answer when selecting explanatory variables for the model, and
+the decision needs to be made for each specific case. In our experiment,
+we will only include dummy variables, capturing a variety of calendar
+events:
 
@@
 
-The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events.
-Finally, (5) is needed to capture the potential phenomenon of change in admittance after the holiday (e.g.~people might try not to go to the hospital on Christmas eve and thus go the next day).
-This model assumes that all these effects are deterministic and do not change over time.
-Still, the exponentiation in \eqref{eq:PoissonRegression} introduces an interaction effect between dummy variables so that the 3 pm on Monday in January will be different from 3 pm on Monday in July, although the parameters for the hour of day and day of the week are fixed and do not change over time.
-We use the \texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.
+The variables \emph{1--3} allow modelling the seasonal patterns on the
+appropriate level of detail throughout the year, while \emph{4} covers the
+changes in admittance due to calendar events. Finally, \emph{5} is needed to
+capture the potential phenomenon of change in admittance after the
+holiday (e.g.~people might try not to go to the hospital on Christmas
+eve and thus go the next day). This model assumes that all these effects
+are deterministic and do not change over time. Still, the exponentiation
+in \eqref{eq:PoissonRegression} introduces an interaction effect between
+dummy variables so that the 3 pm on Monday in January will be different
+from 3 pm on Monday in July, although the parameters for the hour of day
+and day of the week are fixed and do not change over time. We use the
+\texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for
+R \citep{RTeam2021} for the experiments and denote this model as ``Poisson
+Regression''.
 
@@
 
-\citet{Hyndman2008b} developed a state space approach for exponential smoothing models.
-The model can have a set of elements, including different types of Error, Trend and Seasonal components (thus \emph{ETS}).
-Given the popularity of the ETS model in the forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with a frequency of 24 (hour of the day) as a benchmark.
-This was done using the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and denoted as \emph{ETS}.
-This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.
-This model is included as a benchmark, only to see how the other models perform compared to it.
+\citet{Hyndman2008b} developed a state space approach for exponential smoothing
+models. The model can have a set of elements, including different types
+of Error, Trend and Seasonal components (thus \emph{ETS}). Given the
+popularity of the ETS model in the forecasting community, we decided to
+include the basic ETS(A,N,A) model with the seasonal component with a
+frequency of 24 (hour of the day) as a benchmark. This was done using
+the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth}
+for R and denoted as \emph{ETS}. This model does not capture the day of week
+or week of year effects, does not include explanatory variables, but its
+seasonal component and level change over time. This model is included as
+a benchmark, only to see how the other models perform compared to it.
 
@@
 
-Prophet is a forecasting procedure created by Facebook \citep{taylor2018forecasting} that accounts for multiple seasonality, piecewise trend and holiday effects.
-Prophet is robust to missing data and shifts in the trend and typically handles outliers well.
-Prophet works well on daily data seen in Facebook.
-It is robust and automated, making it easy to learn for beginners.
-The implementation may be less flexible than other methods.
-The model itself relies on the Multiple Source of Error state space model, initially proposed by \citet{Kalman1960}.
-The model is incorporated using the corresponding implementation of the Fable package in R.
-We use the \texttt{prophet()} function from the \texttt{fable} package \citep{fable2020}.
-Note that the input data is assigned with an hourly and daily seasonality.
+Prophet is a forecasting procedure created by Facebook
+\citep{taylor2018forecasting} that accounts for multiple seasonality,
+piecewise trend and holiday effects. Prophet is robust to missing data
+and shifts in the trend and typically handles outliers well. Prophet
+works well on daily data seen in Facebook. It is robust and automated,
+making it easy to learn for beginners. The implementation may be less
+flexible than other methods. The model itself relies on the Multiple
+Source of Error state space model, initially proposed by \citet{Kalman1960}.
+The model is incorporated using the corresponding implementation of the
+Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
+package \citep{fable2020}. Note that the input data is assigned with an
+hourly and daily seasonality.
+\textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our benchmarks.}
 
@@
 
-\citet{de2011forecasting} proposed a model to deal with time series exhibiting multiple complex seasonalities called ``TBATS''.
-It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms.
-The latter gives the model more flexibility to deal with fractional seasonality and reduces the parameters of the model when the frequencies of seasonalities are high.
-We fit a TBATS model using the \texttt{tbats()} function from the \texttt{forecast} package in R \citep{forecastpackage2020}.
+\citet{de2011forecasting} proposed a model to deal with time series exhibiting
+multiple complex seasonalities called ``TBATS''. It includes a Box-Cox
+Transformation, ARMA model for residuals and a trigonometric expression
+of seasonal terms. The latter gives the model more flexibility to deal
+with fractional seasonality and reduces the parameters of the model when
+the frequencies of seasonalities are high. We fit a TBATS model using
+the \texttt{tbats()} function from the \texttt{forecast} package in R
+\citep{forecastpackage2020}.
 
@@
 
-Quantile regression allows the production of density forecasts without assuming a fixed distributions shape controlled by a small number of parameters, such as the Poisson distribution with parameter \(\lambda\).
-By producing forecasts of multiple quantiles the full predictive distribution can be constructed.
-Quantile regression is particularly useful where data do not follow a simple distribution, of the distribution shape changes over time of with some covariate, which may be the case with ED arrivals.
+Quantile regression allows the production of density forecasts without
+assuming a fixed distributions shape controlled by a small number of
+parameters, such as the Poisson distribution with parameter \(\lambda\).
+By producing forecasts of multiple quantiles the full predictive
+distribution can be constructed. Quantile regression is particularly
+useful where data do not follow a simple distribution, the
+distribution shape changes over time with some covariate, which may
+be the case with ED arrivals.
 
-Gradient Boosting Machines (GBM) are a tree-based machine learning model for regression and classification.
-Here we produce probabilistic forecasts of ED arrivals via multiple quantile regression using GBMs as implemented in the R package \texttt{gbm} \citep{Greenwell2020}.
-GBMs are a best-in-class algorithm for similar regression problems characterised by modest volumes of training data and possible interactions between input features, and are therefore an appealing choice for ED arrival forecasting.
-Here GBMs are fit with the following features: hour of the day, day of the week, school holiday, temperature and day of the year.
-Hyperparameters are chosen via grid search on the training data, and were chosen as follows: 500 trees, an interaction depth of 2, and a shrinkage 0.1.
-An advantage of GBMs is their ability to learn interactions in comparison to the additive models presented later which require the user to specify possible interactions between inputs.
-Here GBM is included as a reference for the performance of `out-of-the-box' machine learning models applied with minimal effort.
-We also explored the possibility of performing linear regression with additive models, but these did not perform as well as GBM so are omited for brevity.
+Gradient Boosting Machines (GBM) are a tree-based machine learning model
+for regression and classification. Here we produce probabilistic
+forecasts of ED arrivals via multiple quantile regression using GBMs as
+implemented in the R package \texttt{gbm} \citep{Greenwell2020}. GBMs are a
+best-in-class algorithm for similar regression problems characterised by
+modest volumes of training data and possible interactions between input
+features, and are therefore an appealing choice for ED arrival
+forecasting. Here GBMs are fit with the following features: hour of the
+day, day of the week, school holiday, temperature and day of the year.
+Hyperparameters are chosen via grid search on the training data, and
+were chosen as follows: 500 trees, an interaction depth of 2, and a
+shrinkage 0.1. An advantage of GBMs is their ability to learn
+interactions in comparison to the additive models presented later which
+require the user to specify possible interactions between inputs. Here
+GBM is included as a reference for the performance of `out-of-the-box'
+machine learning models applied with minimal effort. We also explored
+the possibility of performing linear regression with additive models,
+but these did not perform as well as GBM so are omited for brevity.
 
@@
 
-\citet{SvetunkovAdam2021} proposed a framework for dynamic models called the Augmented Dynamic Adaptive Model (ADAM).
-This framework encompasses ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting multiple frequencies, non-normal distributions and intermittent demand \citep{Svetunkov2019a}.
-Based on this framework, we use the ETS(M,N,M) model with frequencies 24 (hour of the day) and 168 (hour of the week), adding dummy variables for the week of the year, holidays and lagged holidays.
-This way, we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters.
-Given that the data exhibits randomly occurring zeroes, we use the direct probability model of \citet{Svetunkov2019a} to treat those values.
-Finally, given the skewness of the empirical distribution observed in the preliminary analysis, we use the Gamma distribution for the error term.
-This model can be formulated as a set of the following equations:
-\begin{equation}
+\citet{SvetunkovAdam2021} proposed a framework for dynamic models called the
+Augmented Dynamic Adaptive Model (ADAM). This framework encompasses
+ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting
+multiple frequencies, non-normal distributions and intermittent demand
+\citep{Svetunkov2019a}. Based on this framework, we use the ETS(M,N,M) model
+with frequencies 24 (hour of the day) and 168 (hour of the week), adding
+dummy variables for the week of the year, holidays and lagged holidays.
+This way, we update the hour of day and day of week seasonal indices,
+keeping the week of year one fixed, thus reducing the number of
+estimated parameters. Given that the data exhibits randomly occurring
+zeroes, we use the direct probability model of \citet{Svetunkov2019a} to treat
+those values. Finally, given the skewness of the empirical distribution
+observed in the preliminary analysis, we use the Gamma distribution for
+the error term. This model can be formulated as a set of the following
+equations: \begin{equation}
     \begin{aligned}
@@
     \label{eq:ADAMModel}
-\end{equation}
-where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and \(\alpha_a\) are the smoothing parameters, defining how adaptive the components of the model should be, \(l_t\) is the level component for the demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components, \(\boldsymbol{\beta}\) is the vector of parameters for the explanatory variables, \(o_t\) is the binary variable, which is equal to one, when demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component for the occurrence part of the model, and \(\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)\), where \(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
-Finally, \(a_t\) is an unobservable series, underlying the occurrence part of the model and \((1 + \epsilon_{a,t})\) is an unobservable error term for \(a_t\).
-\citet{Svetunkov2019a} discuss how to estimate such a model.
-We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic nature of the model (level and seasonal components).
-Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it.
-If integer-valued quantiles are needed, then rounding up can be done for them (see Appendix \ref{quantilesceiling} for the explanation).
-This model is implemented in the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in our experiment as ``ADAM-iETSX''.
+\end{equation} where \(\alpha\), \(\beta\), \(\gamma_1\), \(\gamma_2\) and
+\(\alpha_a\) are the smoothing parameters, defining how adaptive the
+components of the model should be, \(l_t\) is the level component for the
+demand sizes, \(s_{1,t}\) and \(s_{2,t}\) are the seasonal components,
+\(\boldsymbol{\beta}\) is the vector of parameters for the explanatory
+variables, \(o_t\) is the binary variable, which is equal to one, when
+demand occurs and to zero otherwise, \(l_{a,t-1}\) is the level component
+for the occurrence part of the model, and
+\(\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)\), where
+\(s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2\) is the scale of the distribution.
+Finally, \(a_t\) is an unobservable series, underlying the occurrence part
+of the model and \((1 + \epsilon_{a,t})\) is an unobservable error term
+for \(a_t\). \citet{Svetunkov2019a} discuss how to estimate such a model. We
+expect this model to perform on par with the Poisson regression,
+potentially outperforming it in some instances, due to the dynamic
+nature of the model (level and seasonal components). Although the data
+is integer-valued, we expect that Gamma distribution will be a good
+approximation for it. If integer-valued quantiles are needed, then
+rounding up can be done for them (see Appendix \ref{quantilesceiling}
+for the explanation). This model is implemented in the \texttt{adam()} function
+from the \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in
+our experiment as ``ADAM-iETSX''.
 
@@
 
-Suppose we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above.
-In that case, the forecasting task becomes one of predicting the future values of that distribution's parameters.
-We then can use Generalised Additive Models for Location, Scale and Shape (GAMLSS).
-These are the distributional regression models where the parameters are modelled as additive functions of explanatory variables.
-This provides a powerful and flexible framework for probabilistic forecasting, provided that suitable distribution and additive model structures can be found.
-In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.
+Suppose we assume that our predictive distribution follows a given
+parametric distribution, as in Poisson regression discussed above. In
+that case, the forecasting task becomes one of predicting the future
+values of that distribution's parameters. We then can use Generalised
+Additive Models for Location, Scale and Shape (GAMLSS). These are the
+distributional regression models where the parameters are modelled as
+additive function of explanatory variables. This provides a powerful
+and flexible framework for probabilistic forecasting, provided that
+suitable distribution and additive model structures can be found. In
+practice, this means employing expert judgement and experimenting with
+various distributions and evaluating their suitability using available
+training data.
 
-Let \(F_t(y_t)\) be a predictive cumulative probability distribution of \(y_t\).
-In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters.
-In a GAMLSS framework of \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled as
+Let \(F_t(y_t)\) be a predictive cumulative probability distribution of
+\(y_t\). In a distributional regression context, \(F_t(y_t)\) is modelled
+via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is
+an \(m\)-dimensional vector of parameters. In a GAMLSS framework of
+\citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled as
 \begin{equation}
@@
     \label{eq:basicGAM}
-\end{equation}
-where \(g_j\) is a monotonic link function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\) \textbf{is \ldots{}}.
-If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third elements of \(\bm x_t\).
-Each \(f_{j,i}\) is a smooth function, constructed as
-\begin{equation}
+\end{equation} where \(g_j\) is a monotonic link function,
+\(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix
+\(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients,
+\(\bm x_t\) is a \(d\)-dimensional vector of covariates and
+\(S_{j,i} \subset \{1, \dots, d\}\) \textbf{is \ldots{}}. If \(S_{j,i} = \{1, 3\}\),
+then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional
+vector formed by the first and third elements of \(\bm x_t\). Each
+\(f_{j,i}\) is a smooth function, constructed as \begin{equation}
     f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
     \label{eq:smmothfunction}
-\end{equation}
-where \(b^{ji}_k\) are spline basis functions of dimension \(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients.
-The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used.
-See \citet{Wood2017} for a detailed introduction to \emph{GAM/GAMLSS} models, smoothing splines bases and penalties.
+\end{equation} where \(b^{ji}_k\) are spline basis functions of dimension
+\(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients.
+The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the
+definition of smoothness being dependent on the type of effect and
+penalty being used. See \citet{Wood2017} for a detailed introduction to
+\emph{GAM/GAMLSS} models, smoothing splines bases and penalties.
 
-As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form
+As our data are counts, the natural starting point is the Poisson
+distribution with an additive model for \(\log \lambda_t\) of the form
 \begin{equation}
@@
 
-The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time \(t\), respectively, and \(C_t\) is the temperature at time \(t\).
-This model is called Poisson-1 in discussions below.
+The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day
+(1--24), day of the week (1--7), and day of the year (1-366) at time
+\(t\), respectively, and \(C_t\) is the temperature at time \(t\). This model
+is called Poisson-1 in discussions below.
 
-However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application.
-In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables.
-Therefore, we consider more flexible, two-parameter distributions to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0.
-The resulting density forecasts are given by
-\begin{equation}
+However, experiments on the training data reveal that calibration of
+forecasts based on the Poisson distribution is poor, suggesting that the
+shape of the distribution is unsuitable for the present application. In
+particular, we observe that forecast uncertainty appears to vary
+depending on the time of day and possibly other explanatory variables.
+Therefore, we consider more flexible, two-parameter distributions to
+specify additive models for both location and scale parameters,
+specifically the truncated Normal distribution, with truncation at 0.
+The resulting density forecasts are given by \begin{equation}
 F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
 \label{eq:truncatedn}
-\end{equation}
-with additive models
+\end{equation} with additive models
 \begin{align*}
@@
     & log(\sigma_t) = \sum_{i=1}^{10} D^{+}_i(t) f(H(t))
-\end{align*}
-for the mean and variance parameters. This model is referred to as NOtr-1 below.
+\end{align*} for the mean and variance parameters. This model is
+referred to as NOtr-1 below.
 
-Furthermore, we consider an extension to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\).
-These models are labelled Poisson-2 and NOtr-2.
-We also performed experiments with the truncated \(t\) distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.
+Furthermore, we consider an extension to the additive models for
+\(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays
+into \(D\). These models are labelled Poisson-2 and NOtr-2. We also
+performed experiments with the truncated \(t\) distribution (Ttr-2) and
+negative binomial distribution (NBI-2), but these did not result in
+forecasts as well calibrated as the truncated normal.
 
@@
 
-In order to assess the performance of models, we evaluate predictive quantiles at probability levels 0.05 to 0.95 in steps of 0.05 and conditional expectations for 0 to 48 hours ahead produced by each model.
-We forecast up to 48 hours because this is the operational horizon in the ED, for which it is possible to make short-term changes in the shifts for nurses and doctors.
-The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000}, resulting in 727 origins.
-Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and expectation.
-The latter is measured via Root Mean Squared Error (RMSE):
-\begin{equation}
+In order to assess the performance of models, we evaluate predictive
+quantiles at probability levels 0.05 to 0.95 in steps of 0.05 and
+conditional expectations for 0 to 48 hours ahead produced by each model.
+We forecast up to 48 hours because this is the operational horizon in
+the ED, for which it is possible to make short-term changes in the
+shifts for nurses and doctors. The forecasts are produced every 12 hours
+for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000},
+resulting in 727 origins. Based on these values, several error measures
+are calculated to evaluate the performance of models in terms of
+specific quantiles and expectation. The latter is measured via Root Mean
+Squared Error (RMSE): \begin{equation}
   \mathrm{RMSE} = \sqrt{\frac{1}{h} \sum_{j=1}^h e_{t+j}^2} ,
   \label{eq:RMSE}
-\end{equation}
-where \(h\) is the forecast horizon and \(e_{t+j}\) is the point forecast error \(j\) steps ahead.
+\end{equation} where \(h\) is the forecast horizon and \(e_{t+j}\) is the
+point forecast error \(j\) steps ahead.
 
-The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated \citep{gneiting2007probabilistic}.
-A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision-makers provided the forecast is calibrated.
-Calibration, also called reliability, is the property that forecast probabilities match the observed frequency of realisations.
-If a forecast is calibrated, then, for example, \(20\%\) of observations should fall below the \(\alpha=0.2\) predictive quantile (with some tolerance based on the finite sample size).
-This property is necessary for forecast probabilities to be used in quantitative decision-making.
-Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, \(\alpha\), against observed frequency mean (\(\mathbf{1}(y_{t}\leq q_{\alpha,t})\)).
-We use several scores to assess the quantile performance of models.
+The objective of density forecasts is to be as sharp as possible while
+remaining reliable/calibrated \citep{gneiting2007probabilistic}. A forecast
+is said to be sharp if the predictive distribution has a relatively
+small spread, indicating low uncertainty, which is valuable to
+decision-makers provided the forecast is calibrated. Calibration, also
+called reliability, is the property that forecast probabilities match
+the observed frequency of realisations. If a forecast is calibrated,
+then, for example, \(20\%\) of observations should fall below the
+\(\alpha=0.2\) predictive quantile (with some tolerance based on the
+finite sample size). This property is necessary for forecast
+probabilities to be used in quantitative decision-making. Calibration is
+typically evaluated visually using reliability diagrams, which plot the
+nominal coverage, \(\alpha\), against observed frequency mean
+(\(\mathbf{1}(y_{t}\leq q_{\alpha,t})\)). We use several scores to assess
+the quantile performance of models.
 
-First, to measure quantile performance, we need to calculate the pinball score, which is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score \citep{hyndman2021forecasting}.
-It rewards sharpness and penalises miscalibration, so it measures overall performance.
-However, calibration should still be verified separately.
-Furthermore, The Pinball Score for an individual quantile matches the loss function minimised in a quantile regression model.
-The Pinball Score is given by
-\begin{equation}
+First, to measure quantile performance, we need to calculate the pinball
+score, which is a strictly proper score used to evaluate quantile
+forecasts and is the discrete form of the Continuous Rank Probability
+Score \citep{hyndman2021forecasting}. It rewards sharpness and penalises
+miscalibration, so it measures overall performance. However, calibration
+should still be verified separately. Furthermore, The Pinball Score for
+an individual quantile matches the loss function minimised in a quantile
+regression model. The Pinball Score is given by \begin{equation}
     \text{Pinball} = 
@@
  \label{eq:pinball}
-\end{equation}
-where \(\mathcal{A} = \{0.05,0.1,...,0.95\}\) is the set of quantiles being estimated.
+\end{equation} where \(\mathcal{A} = \{0.05,0.1,...,0.95\}\) is the set of
+quantiles being estimated.
 
-To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
-\begin{equation}
+To compare model performance, and the significance of any apparent
+difference in performance, we will use skill scores, which can be
+calculated for any metric via: \begin{equation}
   \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} \label{eq:skillscore}
-\end{equation}
-where \(M\) is the metric's value for the method being considered, \(M_\mathrm{ref}\) is the metric's value for a reference method.
-The skill score shows by how many per cent the reference approach is worse than the one under consideration.
-We will use bootstrap re-sampling of skill scores to determine if the differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}.
-Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model and employ a block-bootstrap with blocks of length 24h to account for temporal correlations of the underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
+\end{equation} where \(M\) is the metric's value for the method being
+considered, \(M_\mathrm{ref}\) is the metric's value for a reference
+method. The skill score shows by how many per cent the reference
+approach is worse than the one under consideration. We will use
+bootstrap re-sampling of skill scores to determine if the differences in
+forecast performance (i.e.~positive or negative skill) are significantly
+different from zero \citep{Efron1981Bootstrap}. Here we use the best
+performing simple benchmark, Naive (explained in Subsection
+\ref{naive}), as the reference model and employ a block-bootstrap
+with blocks of length 24h to account for temporal correlations of the
+underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
 
-Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function.
-All functions were re-estimated on each iteration.
-ADAM and Poisson regression estimated the parameters, taking the ones obtained in the initial model application to the data in the first origin as the pre-initials.
-This allowed to speed up the computation for these two models.
-The initial estimation of ADAM took approximately one hour and 25 minutes.
-Each step in the experiment took the time shown in Section \ref{result}.
+Finally, we have calculated the computational time for one iteration on
+the first rolling origin to compare the speed of each function. All
+functions were re-estimated on each iteration. ADAM and Poisson
+regression estimated the parameters, taking the ones obtained in the
+initial model application to the data in the first origin as the
+pre-initials. This allowed to speed up the computation for these two
+models. The initial estimation of ADAM took approximately one hour and
+25 minutes. Each step in the experiment took the time shown in Section
+\ref{result}.
 
@@
 
-The data is portioned into training (from 2014-04-01 to 2018-02-28) and test (from 2018-03-01 to 2019-02-28) sets, with all model development and hyper-parameter tuning performed using training data only.
-The rolling origin advances in 12-hour steps and the forecast horizon is set for 48 steps ahead.
+The data is portioned into training (from 2014-04-01 to 2018-02-28) and
+test (from 2018-03-01 to 2019-02-28) sets, with all model development
+and hyper-parameter tuning performed using training data only. The
+rolling origin advances in 12-hour steps and the forecast horizon is set
+for 48 steps ahead.
 
-Figure \ref{fig:Pinball} presents pinball score aggregated across forecasting horizons for each quantile.
-It shows that the difference in performance among the models mainly comes from the middle of the distribution and somewhat from the upper tail.
-There is very little difference in performance for the lower tail (except for TBATS, which has a consistently higher pinball value than the other models).
-This is interesting and reassuring that the better models are better at probabilities that matter more to decision-makers.
+Figure \ref{fig:Pinball} presents Pinball score aggregated across
+forecasting horizons for each quantile. It shows that the difference in
+performance among the models mainly comes from the middle of the
+distribution and somewhat from the upper tail. There is very little
+difference in performance for the lower tail (except for TBATS, which
+has a consistently higher Pinball value than the other models). This is
+interesting and reassuring that the better models are better at
+probabilities that matter more to decision-makers.
 
-Probabilistic forecasts are evaluated following the principle of \emph{sharpness subject to calibration}, meaning that the sharper forecast is prefered provided that it is calibrated.
-Mis-calibrated forecasts are unsuitable for use in decision-making, so they should be excluded.
-Calibration is evaluated visually in Figure \ref{fig:quantile-bias}, which highlights a systematic negative bias across all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels.
-Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.
-This could result in poor staffing decisions for a user as historic data fail to characterise present conditions accurately.
+Probabilistic forecasts are evaluated following the principle of
+\emph{sharpness subject to calibration}, meaning that the sharper forecast is
+prefered provided that it is calibrated. Mis-calibrated forecasts are
+unsuitable for use in decision-making, so they should be excluded.
+Calibration is evaluated visually in Figure \ref{fig:quantile-bias},
+which highlights a systematic negative bias across all probability
+levels in many models, with only the truncated normal and \(t\) family
+GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
+calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
+exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals.
+This could result in poor staffing decisions.
+This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}
 
@@
 
-Figure \ref{fig:lead-time-rmse} reports the RMSE for each forecast horizon.
-It illustrates the times of day that are harder to predict -- morning pick-up and afternoon peak.
-We can see that some models perform much better than others on specific forecast horizons.
-For example, Benchmark does an excellent job in predicting 1 - 5 steps ahead ED arrivals and, in general, doing well in forecasting arrivals in the night.
-In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, Prophet, TBATS and GBM do for some lead times.
+Figure \ref{fig:lead-time-rmse} reports the RMSE for each forecast
+horizon. It illustrates the times of day that are harder to predict --
+morning pick-up and afternoon peak. We can see that some models perform
+much better than others on specific forecast horizons. For example,
+Benchmark does an excellent job in predicting 1 - 5 steps ahead ED
+arrivals and, in general, doing well in forecasting arrivals in the
+night. In fact, the Benchmark model performs consistently well in terms
+of the conditional mean, not making as huge mistakes as, for example,
+Prophet, TBATS and GBM do for some lead times.
 
@@
 
-In above figures (Figures \ref{fig:Pinball}- \ref{fig:lead-time-rmse}), we only present the performance of the top seven methods.
-However, we have evaluated the performance of 14 methods from the test period.
-Evaluation metrics and computational time (in minutes) for all methods are presented in Table \ref{tab:tab-results}.
-They are ordered by Quantile Bias.
-The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.
+In above figures (Figures \ref{fig:Pinball}-
+\ref{fig:lead-time-rmse}), we only present the performance of the top
+seven methods. However, we have evaluated the performance of 14 methods
+from the test period. Evaluation metrics and computational time (in
+minutes) for all methods are presented in Table \ref{tab:tab-results}.
+They are ordered by Quantile Bias. The five models identified above have
+a Quantile Bias of 0.014 or less, which is substantially lower than the
+next group of forecasts with Quantile Biases of 0.037 and above, ETS
+being the only exception with a value of 0.019.
 
-One more thing to notice is that the ADAM-iETSX model with rounded up quantiles did not perform better than the simpler one with continuous ones (Table \ref{tab:tab-results}).
-This implies that the rounding is not necessary in general, but if integer values are needed (for example, to decide how many nurses to have), then using the continuous model and then rounding up the quantiles could be considered a reasonable strategy.
+One more thing to notice is that the ADAM-iETSX model with rounded up
+quantiles did not perform better than the simpler one with continuous
+ones (Table \ref{tab:tab-results}). This implies that the rounding is
+not necessary in general, but if integer values are needed (for example,
+to decide how many nurses to have), then using the continuous model and
+then rounding up the quantiles could be considered a reasonable
+strategy.
 
@@
 
-Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours, presented in Table \ref{tab:tab-results}.
-The X-axis shows the speed of each method presented as slow, moderate or fast.
-We observe that there is no clear association between speed and accuracy improvement of the models used in this study.
+Finally, Figure \ref{fig:time-accuracy} reports the forecast
+performance of each approach versus the computational time required to
+generate the forecast for a given forecasting horizon of 48 hours,
+presented in Table \ref{tab:tab-results}. The X-axis shows the speed of
+each method presented as slow, moderate or fast. We observe that there
+is no clear association between speed and accuracy improvement of the
+models used in this study.
 
-We found that most of the methods considered in this paper fall into the fast category with a different range of performance, depending on the type of error measure.
-Among the fastest methods, ETS is very competitive when assessed using Quantile Bias and RMSE.
-Moreover, Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by RMSE and Pinball.
+We found that most of the methods considered in this paper fall into the
+fast category with a different range of performance, depending on the
+type of error measure. Among the fastest methods, ETS is very
+competitive when assessed using Quantile Bias and RMSE. Moreover,
+Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by
+RMSE and Pinball.
 
-Figure \ref{fig:time-accuracy} shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures.
-The figure also indicates that while the \(t\) family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures.
-Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
+Figure \ref{fig:time-accuracy} shows that NOtr-2 is the fastest method
+that provides consistent accuracy assessed via all three accuracy
+measures. The figure also indicates that while the \(t\) family GAMLSS
+model (Ttr-2) is the slowest method, it has a very good performance
+across the three presented accuracy measures. Therefore, if the speed is
+not a major constraint when generating the forecasts, one may employ
+this approach to generate forecasts.
+
+\textcolor{blue}{The benefits of the proposed model and its managerial aspects include
+the use of probabilistic forecast to inform decision making and provide
+insights on the running time of each model to generate forecasts. Probabilistic forecasts contain all potential future outcomes and help
+planners to achieve a more efficient decision making by not only allocate a probability to the most likely outcome but also to unlikely
+outcomes such as extremely high or low arrivals in the Emergency Departments.
+The running time could be also an important aspects for managers depending on the frequency of generating forecasts.
+If the frequency of forecast generation is high, one may employ models such as
+Poisson regression (Poisson 1). If the frequency is lower and running time is not a big concern, a model like GAMLSS (Ttr-2) should be used.}
+
+\textcolor{blue}{Probabilistic forecasts can be considered as a useful tool for planners to manage the risk of under and over resource allocation, which consequently may reduce both costs and risks of service delivery in Emergency Departments.}
 
@@
 
-Short-term forecasting of arrivals at emergency departments is an important element of hospital staff and resource management.
-Furthermore, due to the asymmetric impact of an excess resource shortage, especially in emergency departments, quantifying forecast uncertainty is also of value as it enables planners to manage associated risks.
-In this paper, we have developed methods for producing a probabilistic forecast of hourly arrivals up to 48 hours ahead, comparing different state-of-the-art approaches.
+Short-term forecasting of arrivals at emergency departments is an
+important element of hospital staff and resource management.
+Furthermore, due to the asymmetric impact of an excess resource
+shortage, especially in emergency departments, quantifying forecast
+uncertainty is also of value as it enables planners to manage associated
+risks. In this paper, we have developed methods for producing a
+probabilistic forecast of hourly arrivals up to 48 hours ahead,
+comparing different state-of-the-art approaches.
 
-Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression.
-The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution.
-The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature.
-Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score.
-This suggests that the latter may be preferred if the whole distribution is used in decision-making.
+Two approaches produced highly accurate, calibrated probabilistic
+forecasts: a time series model and a model based on distributional
+regression. The first is ADAM-iESTX, an extension of exponential
+smoothing incorporating multiple seasonalities, explanatory variables
+and assuming a Gamma predictive distribution. The second, labeled
+NOtr-2, regressed the two parameters of a truncated (at zero) normal
+distribution on the date and time features and temperature. Both
+approaches produced calibrated probabilistic forecasts, but the point
+prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while
+NOtr-2 produced forecasts with a lower Pinball score. This suggests that
+the latter may be preferred if the whole distribution is used in
+decision-making.
 
-Having compared the performance of a wide range of methods, we make the following observations: the choice of distribution assumed for probabilistic forecasts and choice of model features are as if not more important than the type of model employed; methods based on quantile regression, which do not assume a parametric distribution for forecasts, do not perform as well as those based on parametric distributions; and the best performing models handled the non-negative and skewed nature of the data automatically without the need for post-processing.
-These observations reflect the characteristics of the data, which is representative of ED arrivals, but determining the extent to which they generalise is beyond the scope of this paper.
-Furthermore, methods based on continuous-valued distributions are not adversely affected by the fact that the data are integer-valued.
-Rounding up predictive quantiles to the next integer does not make predictions worse.
+Having compared the performance of a wide range of methods, we make the
+following observations: the choice of distribution assumed for
+probabilistic forecasts and choice of model features are as if not more
+important than the type of model employed; methods based on quantile
+regression, which do not assume a parametric distribution for forecasts,
+do not perform as well as those based on parametric distributions; and
+the best performing models handled the non-negative and skewed nature of
+the data automatically without the need for post-processing. These
+observations reflect the characteristics of the data, which is
+representative of ED arrivals, but determining the extent to which they
+generalise is beyond the scope of this paper. Furthermore, methods based
+on continuous-valued distributions are not adversely affected by the
+fact that the data are integer-valued. Rounding up predictive quantiles
+to the next integer does not make predictions worse.
 
-Finally, we have found that out-of-the-box models, which require minimal tuning or manual development, do not perform as well as well-considered statistical approaches.
-The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks.
-Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well.
-ETS produces reasonably well-calibrated forecasts, in contrast to the poorly calibrated benchmarks and has highly accurate point forecasts.
-However, its probabilistic forecasts were considerably worse than NOtr-2 in terms of pinball score.
+Finally, we have found that out-of-the-box models, which require minimal
+tuning or manual development, do not perform as well as well-considered
+statistical approaches. The popular TBATS, Prophet and Gradient Boosting
+Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and
+even the benchmarks. Of the models requiring a modest amount of user
+input and expertise, exponential smoothing (ETS) was found to perform
+well. ETS produces reasonably well-calibrated forecasts, in contrast to
+the poorly calibrated benchmarks and has highly accurate point
+forecasts. However, its probabilistic forecasts were considerably worse
+than NOtr-2 in terms of Pinball score.
 
-Probabilistic forecasting opens the door to more sophisticated resource management in healthcare settings by providing decision-makers with uncertainty information and enabling quantitative risk management.
-Linking forecasts of arrivals with upstream (ambulance call-outs) and downstream (length of stay, medical outcomes) analytics offers an opportunity to improve forecasting skills and may also be necessary to maximise benefits through more holistic decision-making.
+\textcolor{blue}{
+The dataset used in this study does not include the COVID-19 pandemic.
+During COVID-19, the dynamic of ED has changed.
+That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period.
+However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}
 
-Further research is required to investigate the practical benefits of probabilistic forecasts in healthcare and how they can inform planning and decision making.
-This may require employing discrete simulation or application to the newsvendor problems.
-While this study has focused on hourly short-term forecasting, producing longer-term daily forecast (e.g.~180-270 days ahead) is often required by planners to support winter planning in ED and Ambulance services which requires more investigation.
-Moreover, more research is needed in the forecasting of other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions.
-This may require considering the dynamics among various services, including General Practitioners, Emergency Departments, Ambulance and Fire \& Rescue services.
+Probabilistic forecasting opens the door to more sophisticated resource
+management in healthcare settings by providing decision-makers with
+uncertainty information and enabling quantitative risk management.
+Linking forecasts of arrivals with upstream (ambulance call-outs) and
+downstream (length of stay, medical outcomes) analytics offers an
+opportunity to improve forecasting skills and may also be necessary to
+maximise benefits through more holistic decision-making.
+
+Further research is required to investigate the practical benefits of
+probabilistic forecasts in healthcare and how they can inform planning
+and decision making. This may require employing discrete event simulation or
+application to the newsvendor problems. While this study has focused on
+hourly short-term forecasting, producing longer-term daily forecast
+(e.g.~180-270 days ahead) is often required by planners to support
+winter planning in ED and Ambulance services which requires more
+investigation. Moreover, more research is needed in the forecasting of
+other important variables such as length of stay, bed occupancy and
+waiting time, in addition to patient arrivals and admissions. This may
+require considering the dynamics among various services, including
+General Practitioners, Emergency Departments, Ambulance and Fire \&
+Rescue services.
 
@@
 
-Before proceeding with the proof, we need to define the quantiles of the continuous and rounded up random variables:
+Before proceeding with the proof, we need to define the quantiles of the
+continuous and rounded up random variables:
 \begin{equation} \label{eq:quantCeil1}
     P \left(y_t < k \right) = 1 - \alpha ,
-\end{equation}
-and
-\begin{equation} \label{eq:quantCeil2}
+\end{equation} and \begin{equation} \label{eq:quantCeil2}
     P \left(\lceil y_t \rceil \leq n \right) \geq 1 - \alpha ,
-\end{equation}
-where \(n\) is the quantile of the distribution of rounded up values (the smallest integer number that satisfies the inequality \eqref{eq:quantCeil2}) and \(k\) is the quantile of the continuous distribution of the variable.
+\end{equation} where \(n\) is the quantile of the distribution of rounded
+up values (the smallest integer number that satisfies the inequality
+\eqref{eq:quantCeil2}) and \(k\) is the quantile of the continuous
+distribution of the variable.
 
-In order to prove that \(n = \lceil k \rceil\), we need to use the following basic property:
-\begin{equation} \label{eq:quantCeil3}
+In order to prove that \(n = \lceil k \rceil\), we need to use the
+following basic property: \begin{equation} \label{eq:quantCeil3}
     \lceil y_t \rceil \leq n \iff  y_t \leq n,
-\end{equation}
-which means that the rounded up value will always be less than or equal to \(n\) if and only if the original value is less than or equal to \(n\).
-Taking into account \eqref{eq:quantCeil3}, the probability \eqref{eq:quantCeil2} can be rewritten as:
+\end{equation} which means that the rounded up value will always be less
+than or equal to \(n\) if and only if the original value is less than or
+equal to \(n\). Taking into account \eqref{eq:quantCeil3}, the probability
+\eqref{eq:quantCeil2} can be rewritten as:
 \begin{equation} \label{eq:quantCeil4}
@@
 
-Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2}, \eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the following can be summarised:
-\begin{equation} \label{eq:quantCeil6}
+Taking the inequalities \eqref{eq:quantCeil1}, \eqref{eq:quantCeil2},
+\eqref{eq:quantCeil4} and \eqref{eq:quantCeil5} into account, the
+following can be summarised: \begin{equation} \label{eq:quantCeil6}
     P \left(y_t \leq n-1 \right) < P \left(y_t < k \right) \leq P \left(y_t \leq n \right) ,
-\end{equation}
-which is possible only when \(k \in (n-1, n]\), which means that \(\lceil k \rceil = n\).
-So the rounded up quantile of continuous random variable \(y_t\) will always be equal to the quantile of the descritised value of \(y_t\):
+\end{equation} which is possible only when \(k \in (n-1, n]\), which means
+that \(\lceil k \rceil = n\). So the rounded up quantile of continuous
+random variable \(y_t\) will always be equal to the quantile of the
+descritised value of \(y_t\):
 \begin{equation} \label{eq:ceilingAndQuantiles1}
@@
 
-It is also worth noting that the same results can be obtained with the floor function instead of ceiling, following the same logic. So the following equation will hold for all \(y_t\) as well:
+It is also worth noting that the same results can be obtained with the
+floor function instead of ceiling, following the same logic. So the
+following equation will hold for all \(y_t\) as well:
 \begin{equation} \label{eq:floorAndQuantiles1}

### 2022-12-28 r_script/admision_viz.R

@@ -9,7 +9,7 @@ ride_data <- h2_hourly1 %>%
 
 ride_data_ampm <- ride_data %>% 
   mutate(hour_ampm = case_when(
-    hour == 0 ~ "Minnight",
+    hour == 0 ~ "Midnight",
     hour == 1 ~ "1 am",
     hour == 2 ~ "2 am",
     hour == 3 ~ "3 am",
@@ -37,7 +37,7 @@ ride_data_ampm <- ride_data %>%
   mutate(hour_ampm=factor(hour_ampm))
   
 
-l <- c("Minnight", "1 am",  "2 am",     "3 am",    "4 am",    "5 am",    "6 am",   
+l <- c("Midnight", "1 am",  "2 am",     "3 am",    "4 am",    "5 am",    "6 am",   
 "7 am",     "8 am",     "9 am",     "10 am",     "11 am",     "Midday",    
 "1 pm",     "2 pm",     "3 pm",     "4 pm",     "5 pm",     "6 pm",    
 "7 pm" ,    "8 pm"  ,   "9 pm" ,    "10 pm" ,    "11 pm")
@@ -66,7 +66,7 @@ p <- ggplot(ride_data_ampm , aes(y=hour_ampm,
                       scale = 2) +
   scale_x_continuous(breaks = seq(0,50,5), 
                      expand = c(0, 0))+
-  theme_ipsum_rc(grid=FALSE)+
+  theme_few()+
   scale_fill_manual(values = my_colorblind)+
   theme(legend.position = "bottom",
         axis.line = element_line(colour = "black", 
@@ -90,7 +90,7 @@ daily_density <- ride_data %>%
 #after_stat(density),
 ggplot(ride_data, aes(n_attendance, colour = dow)) +
   geom_freqpoly(binwidth = 1)+
-  theme_ipsum_rc(grid=FALSE)+
+  theme_few()+
   labs(colour=NULL, 
        y = "Count\n", 
        x = "\n ED arrivals")
@@ -105,12 +105,13 @@ ride_data_tsbl <- ride_data %>%
   summarise(n_attendance=sum(n_attendance)) %>%  as_tsibble()
 
 ride_data_tsbl %>% 
-  gg_season(n_attendance, period = "week")+
+  gg_subseries(n_attendance, period = "week")+
   geom_point(size=1)+
-  theme_bw()+
+  theme_few()+
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
   labs(
[…]

### 2022-12-28 r_script/result_viz.R
adding jitter to figure 8

@@ -269,7 +269,7 @@ time_long <- time_long %>% mutate(Time = Time+rnorm(1,0,1))
 ggplot(time_long,aes(x=Time,
                      y=Accuracy,
                      shape=Method))+
-  geom_point(position = "jitter", size=1)+
+  geom_jitter(width = 100,size=1)+
   scale_shape_manual(values = seq(0,13,1))+
   scale_x_continuous(breaks = c(50,500,850) ,
                      labels = c("Fast","Moderate","Slow"))+

### 2022-12-28 paper/paper.Rmd
added missing references

@@
 return visits to services, increased health care costs, inaccuracy in
-electronic medical records and other [@rostami2020anticipating].
+electronic medical records and other [@ROSTAMITABAR20221197].
 
@@
 Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
-Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient [@wright2006reexamining].
+Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently [@ramos2013probabilistic;@ROSTAMITABAR20221197].
 Furthermore, if the impact of under- and overstaffing can be quantified,
@@
 
-The variables _1--3_ allow modelling the seasonal patterns on the
+The variables _1--3_ allow to model the seasonal patterns on the
 appropriate level of detail throughout the year, while _4_ covers the
@@
 interactions in comparison to the additive models presented later which
-require the user to specify possible interactions between inputs. Here
-GBM is included as a reference for the performance of \`out-of-the-box'
-machine learning models applied with minimal effort. We also explored
+require the user to specify possible interactions between inputs. Here, GBM is included as a reference for the performance of _out-of-the-box_
+machine learning models applied with minimal effort [@ridgeway2007generalized]. We also explored
 the possibility of performing linear regression with additive models,
@@
 considered, $M_\mathrm{ref}$ is the metric's value for a reference
-method. The skill score shows by how many per cent the reference
+method. The skill score shows by how many percent the reference
 approach is worse than the one under consideration. We will use
@@
 \textcolor{blue}{
-The dataset used in this study does not include the COVID-19 pandemic.
+The dataset used in this study does not include the period of the COVID-19 pandemic.
 During COVID-19, the dynamic of ED has changed.

### 2022-12-28 paper/paper.tex
added missing references

@@
 return visits to services, increased health care costs, inaccuracy in
-electronic medical records and other \citep{rostami2020anticipating}.
+electronic medical records and other \citep{ROSTAMITABAR20221197}.
 
@@
 Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient.
-Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}.
+Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently \citep{ramos2013probabilistic, ROSTAMITABAR20221197}.
 Furthermore, if the impact of under- and overstaffing can be quantified,
@@
 interactions in comparison to the additive models presented later which
-require the user to specify possible interactions between inputs. Here
-GBM is included as a reference for the performance of `out-of-the-box'
-machine learning models applied with minimal effort. We also explored
+require the user to specify possible interactions between inputs. Here, GBM is included as a reference for the performance of \emph{out-of-the-box}
+machine learning models applied with minimal effort \citep{ridgeway2007generalized}. We also explored
 the possibility of performing linear regression with additive models,
@@
 considered, \(M_\mathrm{ref}\) is the metric's value for a reference
-method. The skill score shows by how many per cent the reference
+method. The skill score shows by how many percent the reference
 approach is worse than the one under consideration. We will use

### 2022-12-29 paper/paper.Rmd
incorporating changes proposed by Ivan in the response letter

@@
 These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
-forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}.
+forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e.
+having more staff than needed is costly, but
+having less staff than required may lead to higher death rate}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient [@wright2006reexamining].
@@
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-\textcolor{blue}{Another drawback of existing studies is that the
-datasets used are relatively small (e.g.
-time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation.
+\textcolor{blue}{nother drawback of existing studies is that the datasets used are relatively small (e.g.
+time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation.
 Such results might not be generalisable.
 Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals.
-Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
-Finally, all previous publications
-referenced in this paper are not fully reproducible as underlying data and code are not available.}
+Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches.
+Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}
 
@@
 There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
-Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention.
-We refer interested readers to  some extensive reviews of the relevant literature} [@mingliterature2022;@gul2020review;@ibrahim2016modeling].
-\textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare.
-However, given
-the focus of this study, we only discuss hourly ED forecasting.}
+Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention.
+We refer interested readers to some extensive reviews of the relevant literature by @mingliterature2022, @gul2020review and @ibrahim2016modeling.
+\textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
+However, given the focus of this study, we only discuss hourly ED forecasting.}
 
@@
 
-Figure \@ref(fig:seasonplot-dofw) \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals.
-Each individual plot illustrates how arrivals change over time for a each day of week (i.e.
-seven days) from Monday to Saturday.
-The blue line shows the average arrival for the given day.
-It is clear that ED arrivals on Monday are higher followed by Saturday.
-This indicates that there are
-significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}
+Figure \@ref(fig:seasonplot-dofw) \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
+the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
+of week from Monday to Saturday. The blue line shows the average arrival for the given day.
+It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
+Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
+compared to the rest of the week. This might be due to the closure of General Practitioners
+outpatient clinics over the weekends.}
 
@@
 package [@fable2020]. Note that the input data is assigned with an
-hourly and daily seasonality.
-\textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our benchmarks.}
+hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
+benchmarks.}
 
@@
 GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
-calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
-exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals.
-This could result in poor staffing decisions.
-This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}
+calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks exhibit nega-
+tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
+This could result in poor staffing decisions. This is because the empirical distribution of
+whole data fails to characterise how arrivals in ED may change over time.}
 
@@
 
-\textcolor{blue}{The benefits of the proposed model and its managerial aspects include
-the use of probabilistic forecast to inform decision making and provide
-insights on the running time of each model to generate forecasts. Probabilistic forecasts contain all potential future outcomes and help
-planners to achieve a more efficient decision making by not only allocate a probability to the most likely outcome but also to unlikely
-outcomes such as extremely high or low arrivals in the Emergency Departments.
-The running time could be also an important aspects for managers depending on the frequency of generating forecasts.
-If the frequency of forecast generation is high, one may employ models such as
-Poisson regression (Poisson 1). If the frequency is lower and running time is not a big concern, a model like GAMLSS (Ttr-2) should be used.}
-
-\textcolor{blue}{Probabilistic forecasts can be considered as a useful tool for planners to manage the risk of under and over resource allocation, which consequently may reduce both costs and risks of service delivery in Emergency Departments.}
+\textcolor{blue}{he main benefit of the applied models is the use of probabilistic forecasts to inform decision
+making. Probabilistic forecasts contain all potential future outcomes and help planners to
+achieve more efficient decisions by not only allocating a probability to the most likely outcome
+but also considering less likely outcomes, including extremely high or extremely low arrivals
+in the Emergency Departments. Based on these forecasts, hospitals can decide how many
+nurses to have for each shift to make the work of the ED more efficient, e.g. to meet the
+service level targets set by the National Health Service (NHS) in the United Kingdom.
+Practically speaking, the running time can be an important aspect for managers depending
+on the frequency of generating forecasts. If it is high, one may employ models that do not
+require a lot of computational time, such as Poisson regression (Poisson 1), sacrificing the
+accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
+like GAMLSS (Ttr-2) should be used}
+
+\textcolor{blue}{Probabilistic forecasts of daily ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
+Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.}
 
@@
 The dataset used in this study does not include the period of the COVID-19 pandemic.
-During COVID-19, the dynamic of ED has changed.
-That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period.
-However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}
+During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
+any forecasting model used for ED arrivals forecasting during that period would need to be
+modified to reflect those changes for that specific period. One of the simplest modifications
+would be to include a set of dummy variables, capturing different stages of the pandemic.
+However, this is outside of the scope of this paper and can be considered as a direction for
+future research.}
 

### 2022-12-29 paper/paper.tex
incorporating changes proposed by Ivan in the response letter

@@
 These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
-forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals.
-Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}.
+forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals.
+Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e.
+having more staff than needed is costly, but
+having less staff than required may lead to higher death rate}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}.
@@
 
-\textcolor{blue}{
+\textbackslash textcolor\{blue\}\{
 There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
-Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention.
-We refer interested readers to  some extensive reviews of the relevant literature} \citep{mingliterature2022, gul2020review, ibrahim2016modeling}.
-\textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare.
-However, given
-the focus of this study, we only discuss hourly ED forecasting.}
+Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention.
+We refer interested readers to some extensive reviews of the relevant literature by \citet{mingliterature2022}, \citet{gul2020review} and \citet{ibrahim2016modeling}.
+\textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
+However, given the focus of this study, we only discuss hourly ED forecasting.}
 
@@
 
-Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals.
-Each individual plot illustrates how arrivals change over time for a each day of week (i.e.
-seven days) from Monday to Saturday.
-The blue line shows the average arrival for the given day.
-It is clear that ED arrivals on Monday are higher followed by Saturday.
-This indicates that there are
-significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}
+Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
+the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
+of week from Monday to Saturday. The blue line shows the average arrival for the given day.
+It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
+Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
+compared to the rest of the week. This might be due to the closure of General Practitioners
+outpatient clinics over the weekends.}
 
@@
 
-The variables \emph{1--3} allow modelling the seasonal patterns on the
+The variables \emph{1--3} allow to model the seasonal patterns on the
 appropriate level of detail throughout the year, while \emph{4} covers the
@@
 package \citep{fable2020}. Note that the input data is assigned with an
-hourly and daily seasonality.
-\textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our benchmarks.}
+hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
+benchmarks.}
 
@@
 \textcolor{blue}{
-The dataset used in this study does not include the COVID-19 pandemic.
-During COVID-19, the dynamic of ED has changed.
-That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period.
-However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}
+The dataset used in this study does not include the period of the COVID-19 pandemic.
+During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
+any forecasting model used for ED arrivals forecasting during that period would need to be
+modified to reflect those changes for that specific period. One of the simplest modifications
+would be to include a set of dummy variables, capturing different stages of the pandemic.
+However, this is outside of the scope of this paper and can be considered as a direction for
+future research.}
 

### 2023-01-11 paper/paper.Rmd
submitted revised article

@@
 title:  Probabilistic forecasting of hourly Emergency Department arrivals
-# author:
-#   - name: Bahmna Rostami Tabar
-#     email: rostami-tabarb@cardiff.ac.uk
-#     affiliation: Cardiff Business School
-#     footnote: 12
-#   - name: Jethro Browell
-#     email: jethro.browell@glasgow.ac.uk
-#     affiliation: University of Glasgow
-#     footnote: 1
-#   - name: Ivan Svetunkov
-#     email: i.svetunkov@lancaster.ac.uk
-#     affiliation: Lancaster University
-#     footnote: 1
-# address:
-#   - code: Cardiff Business School
-#     address: Cardiff Business School, Cardiff University, UK
-#   - code: University of Glasgow
-#     address: School of Mathematics and Statistics, University of Glasgow, UK
-#   - code: Lancaster University
-#     address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
-# footnote:
-#   - code: 12
-#     text: "Corresponding Author"
-#   - code: 1
-#     text: "Equal contribution"
+author:
+  - name: Bahman Rostami-Tabar
+    email: rostami-tabarb@cardiff.ac.uk
+    affiliation: CBS
+    correspondingauthor: true
+  - name: Jethro Browell
+    email: jethro.browell@glasgow.ac.uk
+    affiliation: UG
+  - name: Ivan Svetunkov
+    email: i.svetunkov@lancaster.ac.uk
+    affiliation: LU
+address:
+  - code: CBS
+    address: Cardiff Business School, Cardiff University, UK.
+  - code: UG
+    address: School of Mathematics and Statistics, University of Glasgow, UK.
+  - code: LU
+    address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK.
+
 abstract: |
@@
 Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
-keywords:  "Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing"
-
-journal: "Annals of Emergency Medicine"
+keywords:
+- Emergency Department
+- Poisson Regression
+- Probabilistic Forecasting
+- Generalised Additive Models
+- Intermittent Exponential Smoothing
+
+journal: "Health Systems"
 date: "`r Sys.Date()`"
@@
 having more staff than needed is costly, but
-having less staff than required may lead to higher death rate}.
+having less staff than required may lead to worse outcomes for patients}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient [@wright2006reexamining].
@@
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-\textcolor{blue}{nother drawback of existing studies is that the datasets used are relatively small (e.g.
+\textcolor{blue}{Another drawback of existing studies is that the datasets used are relatively small (e.g.
 time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation.
@@
 # Research background: hourly ED forecasting {#lit}
-\textcolor{blue}{
-There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
+\textcolor{blue}{There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
 Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention.
-We refer interested readers to some extensive reviews of the relevant literature by @mingliterature2022, @gul2020review and @ibrahim2016modeling.
+We refer interested readers to some extensive reviews of the relevant literature by} @mingliterature2022, @gul2020review and @ibrahim2016modeling.
 \textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
@@
 
-\textcolor{blue}{he main benefit of the applied models is the use of probabilistic forecasts to inform decision
+\textcolor{blue}{The main benefit of the applied models is the use of probabilistic forecasts to inform decision
 making. Probabilistic forecasts contain all potential future outcomes and help planners to
-achieve more efficient decisions by not only allocating a probability to the most likely outcome
-but also considering less likely outcomes, including extremely high or extremely low arrivals
-in the Emergency Departments. Based on these forecasts, hospitals can decide how many
-nurses to have for each shift to make the work of the ED more efficient, e.g. to meet the
-service level targets set by the National Health Service (NHS) in the United Kingdom.
-Practically speaking, the running time can be an important aspect for managers depending
+achieve more efficient decisions by not only predicting the most likely outcome but also
+quantifying the probability of all possible outcomes including extremely high or extremely
+low arrivals in the Emergency Departments. This information enables decision makers to
+manage risk associated with low-probability-high-impact events. Based on these forecasts,
+hospitals can decide how many nurses to have for each shift to make the work of the ED more
+efficient, e.g. to meet the service level targets set by the National Health Service (NHS) in
+the United Kingdom.
+
+\textcolor{blue}{Practically speaking, the running time can be an important aspect for managers depending
 on the frequency of generating forecasts. If it is high, one may employ models that do not
@@
 accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
-like GAMLSS (Ttr-2) should be used}
-
-\textcolor{blue}{Probabilistic forecasts of daily ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
+like GAMLSS (Ttr-2) should be used.}
+
+\textcolor{blue}{Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
 Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.}
@@
 
+# Reproducibility {.unnumbered}
+
+R code to produce all results in this paper is available at
+
 # References {#references .unnumbered}

### 2023-01-11 paper/paper.tex
submitted revised article

@@
 
-  \journal{Annals of Emergency Medicine} % Sets Journal name
+  \journal{Health Systems} % Sets Journal name
 
@@
   \title{Probabilistic forecasting of hourly Emergency Department arrivals}
+    \author[CBS]{Bahman Rostami-Tabar%
+  \corref{cor1}%
+  }
+   \ead{rostami-tabarb@cardiff.ac.uk} 
+    \author[UG]{Jethro Browell%
+  %
+  }
+   \ead{jethro.browell@glasgow.ac.uk} 
+    \author[LU]{Ivan Svetunkov%
+  %
+  }
+   \ead{i.svetunkov@lancaster.ac.uk} 
+      \affiliation[CBS]{Cardiff Business School, Cardiff University, UK.}
+    \affiliation[UG]{School of Mathematics and Statistics, University of Glasgow, UK.}
+    \affiliation[LU]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK.}
     \cortext[cor1]{Corresponding author}
@@
     \begin{keyword}
-    Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing \sep 
-    Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing
+    Emergency Department \sep Poisson Regression \sep Probabilistic Forecasting \sep Generalised Additive Models \sep 
+    Intermittent Exponential Smoothing
   \end{keyword}
@@
 having more staff than needed is costly, but
-having less staff than required may lead to higher death rate}.
+having less staff than required may lead to worse outcomes for patients}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}.
@@
 
-\textbackslash textcolor\{blue\}\{
-There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
+\textcolor{blue}{There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
 Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention.
-We refer interested readers to some extensive reviews of the relevant literature by \citet{mingliterature2022}, \citet{gul2020review} and \citet{ibrahim2016modeling}.
+We refer interested readers to some extensive reviews of the relevant literature by} \citet{mingliterature2022}, \citet{gul2020review} and \citet{ibrahim2016modeling}.
 \textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
@@
 
-\textcolor{blue}{he main benefit of the applied models is the use of probabilistic forecasts to inform decision
+\textbackslash textcolor\{blue\}\{The main benefit of the applied models is the use of probabilistic forecasts to inform decision
 making. Probabilistic forecasts contain all potential future outcomes and help planners to
-achieve more efficient decisions by not only allocating a probability to the most likely outcome
-but also considering less likely outcomes, including extremely high or extremely low arrivals
-in the Emergency Departments. Based on these forecasts, hospitals can decide how many
-nurses to have for each shift to make the work of the ED more efficient, e.g. to meet the
-service level targets set by the National Health Service (NHS) in the United Kingdom.
-Practically speaking, the running time can be an important aspect for managers depending
+achieve more efficient decisions by not only predicting the most likely outcome but also
+quantifying the probability of all possible outcomes including extremely high or extremely
+low arrivals in the Emergency Departments. This information enables decision makers to
+manage risk associated with low-probability-high-impact events. Based on these forecasts,
+hospitals can decide how many nurses to have for each shift to make the work of the ED more
+efficient, e.g.~to meet the service level targets set by the National Health Service (NHS) in
+the United Kingdom.
+
+\textcolor{blue}{Practically speaking, the running time can be an important aspect for managers depending
 on the frequency of generating forecasts. If it is high, one may employ models that do not
@@
 
+\hypertarget{reproducibility}{%
+\section*{Reproducibility}\label{reproducibility}}
+\addcontentsline{toc}{section}{Reproducibility}
+
+R code to produce all results in this paper is available at
+
 \renewcommand\refname{References}

### 2023-04-13 journal_submission/revised/latex_with_authors/paper_with_authors.tex
prepare a version to share in the journal

--- 2023-01-11 paper/paper.tex
@@
 \usepackage[top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
-\usepackage[]{natbib}
-\bibliographystyle{plainnat}
 
@@
 \makesavenoteenv{longtable}
+
+% Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 problem that causes challenges for the patient flow
-\citep{derlet2002overcrowding}. Also, it is related to increasing length of
-stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected
+(Derlet, 2002). Also, it is related to increasing length of
+stay (Muhammet and Guneri, 2015), low patient satisfaction, unexpected
 return visits to services, increased health care costs, inaccuracy in
-electronic medical records and other \citep{ROSTAMITABAR20221197}.
+electronic medical records (Rostami-Tabar and Ziel, 2022).
 
@@
 events. These studies use multiple approaches, including Exponential
-Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
-(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
-Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
-model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
-Neural Networks \citep{hyndman2021forecasting}.
-\textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study.
-These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
-forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals.
+Smoothing (Svetunkov, 2021a), Autoregressive Integrated Moving Average
+(ARIMA) (Hyndman and Athanasopoulos, 2021), Autoregressive Conditional
+Heteroskedasticity (ARCH) (Bollerslev et al., 1994), Vector Autoregressive
+model (Lütkepohl, 2013), TBATS (De Livera et al., 2011) and Artificial
+Neural Networks (Hyndman and Athanasopoulos, 2021).We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study.
+These gaps and their importance are discussed below.
+Most of these studies are limited to only predicting future arrivals as a point
+forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
+There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
 Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e.
 having more staff than needed is costly, but
-having less staff than required may lead to worse outcomes for patients}.
-This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}.
-Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently \citep{ramos2013probabilistic, ROSTAMITABAR20221197}.
+having less staff than required may lead to worse outcomes for patients.
+This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient (Wright et al., 2006).
+Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently (Ramos et al., 2013; Rostami-Tabar and Ziel, 2022).
 Furthermore, if the impact of under- and overstaffing can be quantified,
@@
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-\textcolor{blue}{Another drawback of existing studies is that the datasets used are relatively small (e.g.
+Another drawback of existing studies is that the datasets used are relatively small (e.g.
 time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation.
@@
 Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches.
-Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}
+Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.
 
@@
   We develop an advanced dynamic model to forecast ED arrivals based
-  on iETS \citep{Svetunkov2019a} and ETSX models with a modification for
+  on iETS (Svetunkov and Boylan, 2019) and ETSX models with a modification for
   multiple frequencies, which produced highly-accurate point
@@
   be generalised to forecast hourly requirements for longer horizons
-  and in other services \citep{al2021empirical}, such as inpatient and
+  and in other services (Al-Azzani et al., 2021), such as inpatient and
   outpatient care services, the number of attended incidents in
@@
 
-\textcolor{blue}{There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
+There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
 Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention.
-We refer interested readers to some extensive reviews of the relevant literature by} \citet{mingliterature2022}, \citet{gul2020review} and \citet{ibrahim2016modeling}.
-\textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
-However, given the focus of this study, we only discuss hourly ED forecasting.}
-
-\citet{hertzum2017forecasting} used linear regression, ARIMA, and Naïve to
+We refer interested readers to some extensive reviews of the relevant literature by Shi et al.
+(2022), Gul and Celik (2020) and Ibrahim et al.
+(2016).
+The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
+However, given the focus of this study, we only discuss hourly ED forecasting.
+
+Hertzum (2017) used linear regression, ARIMA, and Naïve to
 investigate whether accurate hourly accident and emergency department
 patient arrivals and occupancy forecasts can be generated using calendar
-variables. \citet{hertzum2017forecasting} study found that patient arrival
+variables. Hertzum (2017) study found that patient arrival
 variation is larger across the hours of the day than across the days of
@@
 
-\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neural
+Choudhury and Urena (2020) used ARIMA, Holt-Winters, TBATS, and neural
 network methods to forecast hourly accident and emergency department
@@
 claimed that ARIMA provided high and acceptable hourly ED forecasting
-accuracy, even outperforming TBATS. \citet{Cheng2021} developed an ARIMA model
+accuracy, even outperforming TBATS. Cheng et al. (2021a) developed an ARIMA model
 for ED occupancy with a seasonal component and exogenous variables,
@@
 
-\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in
+Morzuch and Allen (2006) used the Unobserved Components Model (UCM), in
 which each component of the time series is separately modelled as
@@
 
-\citet{mccarthy2008challenge} employed a Poisson log-linear regression model,
+McCarthy et al. (2008) employed a Poisson log-linear regression model,
 including independent variables such as temporal factors (e.g.,
@@
 
-\citet{schweigler2009forecasting} investigated whether time series methods
+Schweigler et al. (2009) investigated whether time series methods
 could accurately generate short-term forecasts of ED bed occupancy. A
@@
 
-\citet{kim2014predicting} compared different univariate and multivariate time
+Kim et al. (2014) compared different univariate and multivariate time
 series forecasting techniques to predict patient volume for a Hospital
@@
 
-\citet{asheim2019real} developed a Poisson time-series regression model with
+Asheim et al. (2019) developed a Poisson time-series regression model with
 continuous day-of-week and week-of-year effects to implement a real-time
 system that could forecast ED arrivals on 1, 2, 3 hours ahead. Measuring
-the accuracy using the MAPE, \citet{asheim2019real} noticed that significant
+the accuracy using the MAPE, Asheim et al. (2019) noticed that significant
 improvement happened when the time of notification was incorporated into
@@
 
-\citet{cheng2021forecasting} used one year of ED visits time series to evaluate
+Cheng et al. (2021b) used one year of ED visits time series to evaluate
 the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED
@@
 accuracy of forecasting models on hourly accident and emergency
-department data is low \citep{boyle2012predicting, hertzum2017forecasting},
+department data is low (Boyle et al., 2012; Hertzum, 2017),
 while others mentioned that the accuracy of ED hourly forecast is at an
-acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.
+acceptable level (Choudhury and Urena, 2020; McCarthy et al., 2008; Schweigler et al., 2009).
 
@@
 It is clear that the number of arrivals has a sub-daily structure,
-similar to the one summarised by \citet{hertzum2017forecasting}. The ED
+similar to the one summarised by Hertzum (2017). The ED
 arrivals decrease between midnight and early morning and then increase
@@
 
-Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
+Figure \ref{fig:seasonplot-dofw} illustrates the daily subseries plot, with the x-axis representing the date and y-axis
 the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
@@
 compared to the rest of the week. This might be due to the closure of General Practitioners
-outpatient clinics over the weekends.}
+outpatient clinics over the weekends.
 
@@
 Storm Emma with heavy snowfall that resulted in travel disruption, mass
-power outages, and schools closed in the UK \citep{stormemma2018}.
+power outages, and schools closed in the UK (Wales, n.d.).
 
@@
 empirical distribution of the hourly arrival time series is used to
-forecast the future arrival distribution \citep{la2021new}. We consider the
+forecast the future arrival distribution (La Salle et al., 2021). We consider the
 empirical distribution of all available historic data (Benchmark-1) and
@@
 should be substituted by some other model. One of the models that is
-frequently used in practice is the Poisson regression \citep[see, for example,][]{mccarthy2008challenge}, which can be summarised as \begin{equation}
+frequently used in practice is the Poisson regression (see, for example, McCarthy et al., 2008), which can be summarised as \begin{equation}
   \begin{aligned}
@@
 and day of the week are fixed and do not change over time. We use the
-\texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for
-R \citep{RTeam2021} for the experiments and denote this model as ``Poisson
+\texttt{alm()} function from the \texttt{greybox} package (Svetunkov, 2021b) for
+R (Team, 2021) for the experiments and denote this model as ``Poisson
 Regression''.
@@
 
-\citet{Hyndman2008b} developed a state space approach for exponential smoothing
+Hyndman et al. (2008) developed a state space approach for exponential smoothing
 models. The model can have a set of elements, including different types
@@
 frequency of 24 (hour of the day) as a benchmark. This was done using
-the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth}
+the \texttt{adam()} function from the \texttt{smooth} package (Svetunkov, 2021c)
 for R and denoted as \emph{ETS}. This model does not capture the day of week
@@
 Prophet is a forecasting procedure created by Facebook
-\citep{taylor2018forecasting} that accounts for multiple seasonality,
+(Taylor and Letham, 2018) that accounts for multiple seasonality,
 piecewise trend and holiday effects. Prophet is robust to missing data
@@
 flexible than other methods. The model itself relies on the Multiple
-Source of Error state space model, initially proposed by \citet{Kalman1960}.
+Source of Error state space model, initially proposed by Kalman (1960).
 The model is incorporated using the corresponding implementation of the
 Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
-package \citep{fable2020}. Note that the input data is assigned with an
-hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
-benchmarks.}
+package (O'Hara-Wild et al., 2020). Note that the input data is assigned with an
+hourly and daily seasonality. This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
+benchmarks.
 
@@
 
-\citet{de2011forecasting} proposed a model to deal with time series exhibiting
+De Livera et al. (2011) proposed a model to deal with time series exhibiting
 multiple complex seasonalities called ``TBATS''. It includes a Box-Cox
@@
 the \texttt{tbats()} function from the \texttt{forecast} package in R
-\citep{forecastpackage2020}.
+(Hyndman et al., 2020).
 
@@
 forecasts of ED arrivals via multiple quantile regression using GBMs as
-implemented in the R package \texttt{gbm} \citep{Greenwell2020}. GBMs are a
+implemented in the R package \texttt{gbm} (Greenwell et al., 2020). GBMs are a
 best-in-class algorithm for similar regression problems characterised by
@@
 require the user to specify possible interactions between inputs. Here, GBM is included as a reference for the performance of \emph{out-of-the-box}
-machine learning models applied with minimal effort \citep{ridgeway2007generalized}. We also explored
+machine learning models applied with minimal effort (Ridgeway, 2007). We also explored
 the possibility of performing linear regression with additive models,
@@
 
-\citet{SvetunkovAdam2021} proposed a framework for dynamic models called the
+Svetunkov (2022) proposed a framework for dynamic models called the
 Augmented Dynamic Adaptive Model (ADAM). This framework encompasses
-ARIMA \citep{Box1976}, ETS \citep{Hyndman2008b} and regression, supporting
+ARIMA (Box and Jenkins, 1976), ETS (Hyndman et al., 2008) and regression, supporting
 multiple frequencies, non-normal distributions and intermittent demand
-\citep{Svetunkov2019a}. Based on this framework, we use the ETS(M,N,M) model
+(Svetunkov and Boylan, 2019). Based on this framework, we use the ETS(M,N,M) model
 with frequencies 24 (hour of the day) and 168 (hour of the week), adding
@@
 estimated parameters. Given that the data exhibits randomly occurring
-zeroes, we use the direct probability model of \citet{Svetunkov2019a} to treat
+zeroes, we use the direct probability model of Svetunkov and Boylan (2019) to treat
 those values. Finally, given the skewness of the empirical distribution
@@
 of the model and \((1 + \epsilon_{a,t})\) is an unobservable error term
-for \(a_t\). \citet{Svetunkov2019a} discuss how to estimate such a model. We
+for \(a_t\). Svetunkov and Boylan (2019) discuss how to estimate such a model. We
 expect this model to perform on par with the Poisson regression,
@@
 for the explanation). This model is implemented in the \texttt{adam()} function
-from the \texttt{smooth} package \citep{Svetunkov2021Smooth} for R and is denoted in
+from the \texttt{smooth} package (Svetunkov, 2021c) for R and is denoted in
 our experiment as ``ADAM-iETSX''.
@@
 an \(m\)-dimensional vector of parameters. In a GAMLSS framework of
-\citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled as
+Rigby and Stasinopoulos (2005) the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled as
 \begin{equation}
@@
 definition of smoothness being dependent on the type of effect and
-penalty being used. See \citet{Wood2017} for a detailed introduction to
+penalty being used. See Wood (2017) for a detailed introduction to
 \emph{GAM/GAMLSS} models, smoothing splines bases and penalties.
@@
 shifts for nurses and doctors. The forecasts are produced every 12 hours
-for the holdout of 365 days in a rolling origin fashion \citep{Tashman2000},
+for the holdout of 365 days in a rolling origin fashion (Tashman, 2000),
 resulting in 727 origins. Based on these values, several error measures
@@
 The objective of density forecasts is to be as sharp as possible while
-remaining reliable/calibrated \citep{gneiting2007probabilistic}. A forecast
+remaining reliable/calibrated (Gneiting et al., 2007). A forecast
 is said to be sharp if the predictive distribution has a relatively
@@
 forecasts and is the discrete form of the Continuous Rank Probability
-Score \citep{hyndman2021forecasting}. It rewards sharpness and penalises
+Score (Hyndman and Athanasopoulos, 2021). It rewards sharpness and penalises
 miscalibration, so it measures overall performance. However, calibration
@@
 forecast performance (i.e.~positive or negative skill) are significantly
-different from zero \citep{Efron1981Bootstrap}. Here we use the best
+different from zero (Efron, 1981). Here we use the best
 performing simple benchmark, Naive (explained in Subsection
@@
 with blocks of length 24h to account for temporal correlations of the
-underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.
+underlying data (Bergmeir et al., 2016; Hongyi Li and Maddala, 1996).
 
@@
 GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
-calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks exhibit nega-
+calibration across most probability levels. Notably, both benchmarks exhibit nega-
 tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
 This could result in poor staffing decisions. This is because the empirical distribution of
-whole data fails to characterise how arrivals in ED may change over time.}
+whole data fails to characterise how arrivals in ED may change over time.
 
@@
 
-\textbackslash textcolor\{blue\}\{The main benefit of the applied models is the use of probabilistic forecasts to inform decision
-making. Probabilistic forecasts contain all potential future outcomes and help planners to
+\textbackslash The main benefit of the applied models is the use of probabilistic forecasts to inform decision making.
+Probabilistic forecasts contain all potential future outcomes and help planners to
 achieve more efficient decisions by not only predicting the most likely outcome but also
@@
 
-\textcolor{blue}{Practically speaking, the running time can be an important aspect for managers depending
+Practically speaking, the running time can be an important aspect for managers depending
 on the frequency of generating forecasts. If it is high, one may employ models that do not
@@
 accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
-like GAMLSS (Ttr-2) should be used.}
-
-\textcolor{blue}{Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
-Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.}
+like GAMLSS (Ttr-2) should be used.
+
+Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
+Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.
 
@@
 
-\textcolor{blue}{
 The dataset used in this study does not include the period of the COVID-19 pandemic.
@@
 However, this is outside of the scope of this paper and can be considered as a direction for
-future research.}
+future research.
 
@@
 
-\renewcommand\refname{References}
-\bibliography{mybibfile.bib}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-al2021empirical}{}}%
+Al-Azzani, M.A., Davari, S., England, T.J., 2021. An empirical investigation of forecasting methods for ambulance calls-a case study. Health Systems 10, 268--285.
+
+\leavevmode\vadjust pre{\hypertarget{ref-asheim2019real}{}}%
+Asheim, A., Bjørnsen, L.P.B.-W., Næss-Pleym, L.E., Uleberg, O., Dale, J., Nilsen, S.M., 2019.
+Real-time forecasting of emergency department arrivals using prehospital data.
+BMC emergency medicine 19, 42.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bergmeir2016303}{}}%
+Bergmeir, C., Hyndman, R.J., Benítez, J.M., 2016. Bagging exponential smoothing methods using STL decomposition and box-cox transformation. International Journal of Forecasting.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bollerslev1994arch}{}}%
+Bollerslev, T., Engle, R.F., Nelson, D.B., 1994. ARCH models. Handbook of econometrics 4, 2959--3038.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Box1976}{}}%
+Box, G., Jenkins, G., 1976. {Time series analysis: forecasting and control}. Holden-day, Oakland, California.
+
+\leavevmode\vadjust pre{\hypertarget{ref-boyle2012predicting}{}}%
+Boyle, J., Jessup, M., Crilly, J., Green, D., Lind, J., Wallis, M., Miller, P., Fitzgerald, G., 2012. Predicting emergency department admissions. Emergency Medicine Journal 29, 358--365.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cheng2021forecasting}{}}%
+Cheng, Q., Argon, N.T., Evans, C.S., Liu, Y., Platts-Mills, Timothy F., Ziya, S., 2021b.
+Forecasting emergency department hourly occupancy using time series analysis.
+The American Journal of Emergency Medicine 48, 177--182.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Cheng2021}{}}%
+Cheng, Q., Argon, N.T., Evans, C.S., Liu, Y., Platts-Mills, Timothy F., Ziya, S., 2021a.
+Forecasting emergency department hourly occupancy using time series analysis.
+The American Journal of Emergency Medicine 48, 177--182.
+doi:\url{https://doi.org/10.1016/j.ajem.2021.04.075}
+
+\leavevmode\vadjust pre{\hypertarget{ref-choudhury2020forecasting}{}}%
+Choudhury, A., Urena, E., 2020. Forecasting hourly emergency department arrival using time series analysis. British Journal of Healthcare Management 26, 34--43.
+
+\leavevmode\vadjust pre{\hypertarget{ref-de2011forecasting}{}}%
+De Livera, A.M., Hyndman, R.J., Snyder, R.D., 2011. Forecasting time series with complex seasonal patterns using exponential smoothing. Journal of the American Statistical Association 106, 1513--1527.
+
+\leavevmode\vadjust pre{\hypertarget{ref-derlet2002overcrowding}{}}%
+Derlet, R.W., 2002. Overcrowding in emergency departments: Increased demand and decreased capacity. Annals of emergency medicine 39, 430--432.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Efron1981Bootstrap}{}}%
+Efron, B., 1981. Nonparametric estimates of standard error: The jackknife, the bootstrap and other methods. Biometrika 68, 589--599.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gneiting2007probabilistic}{}}%
+Gneiting, T., Balabdaoui, F., Raftery, A.E., 2007. Probabilistic forecasts, calibration and sharpness. Journal of the Royal Statistical Society: Series B (Statistical Methodology) 69, 243--268.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Greenwell2020}{}}%
+Greenwell, B., Boehmke, B., Cunningham, J., Developers, G., 2020. \href{https://CRAN.R-project.org/package=gbm}{Gbm: Generalized boosted regression models}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gul2020review}{}}%
+Gul, M., Celik, E., 2020. An exhaustive review and analysis on applications of statistical forecasting in hospital emergency departments. Health Systems 9, 263--284.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hertzum2017forecasting}{}}%
+Hertzum, M., 2017. Forecasting hourly patient visits in the emergency department to counteract crowding. The Ergonomics Open Journal 10.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hongyi1996bootstrapping}{}}%
+Hongyi Li, G., Maddala, 1996. Bootstrapping time series models. Econometric reviews 15, 115--158.
+
+\leavevmode\vadjust pre{\hypertarget{ref-forecastpackage2020}{}}%
+Hyndman, R., Athanasopoulos, G., Bergmeir, C., Caceres, G., Chhay, L., O'Hara-Wild, M., Petropoulos, F., Razbash, S., Wang, E., Yasmeen, F., 2020.
+\href{http://pkg.robjhyndman.com/forecast}{{forecast}: Forecasting functions for time series and linear models}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hyndman2021forecasting}{}}%
+Hyndman, R.J., Athanasopoulos, G., 2021. Forecasting: Principles and practice. OTexts.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hyndman2008b}{}}%
+Hyndman, R.J., Koehler, A.B., Ord, J.K., Snyder, R.D., 2008. {Forecasting with Exponential Smoothing}. Springer Berlin Heidelberg.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ibrahim2016modeling}{}}%
+Ibrahim, R., Ye, H., L'Ecuyer, P., Shen, H., 2016. Modeling and forecasting call center arrivals: A literature survey and a case study. International Journal of Forecasting 32, 865--874.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Kalman1960}{}}%
+Kalman, R.E., 1960. {A New Approach to Linear Filtering and Prediction Problems}. Journal of Basic Engineering 82, 35. doi:\href{https://doi.org/10.1115/1.3662552}{10.1115/1.3662552}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kim2014predicting}{}}%
+Kim, K., Lee, C., O'Leary, K., Rosenauer, S., Mehrotra, S., 2014.
+Predicting patient volumes in hospital medicine: A comparative study of different time series forecasting methods.
+Northwestern University, Illinois, USA, Scientific Report.
+
+\leavevmode\vadjust pre{\hypertarget{ref-la2021new}{}}%
+La Salle, J.L.G., David, M., Lauret, P., 2021. A new climatology reference model to benchmark probabilistic solar forecasts. Solar Energy 223, 398--414.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lutkepohl2013vector}{}}%
+Lütkepohl, H., 2013. Vector autoregressive models, in: Handbook of Research Methods and Applications in Empirical Macroeconomics. Edward Elgar Publishing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mccarthy2008challenge}{}}%
+McCarthy, M.L., Zeger, S.L., Ding, R., Aronsky, D., Hoot, N.R., Kelen, G.D., 2008. The challenge of predicting demand for emergency department services. Academic Emergency Medicine 15, 337--346.
+
+\leavevmode\vadjust pre{\hypertarget{ref-morzuch2006forecasting}{}}%
+Morzuch, B.J., Allen, P.G., 2006. Forecasting hospital emergency department arrivals. 26th Annual Symposium on Forecasting, Santander, Spain.
+
+\leavevmode\vadjust pre{\hypertarget{ref-muhammet2015forecasting}{}}%
+Muhammet, G., Guneri, A.F., 2015.
+Forecasting patient length of stay in an emergency department by artificial neural networks.
+Journal of aeronautics and space technologies (Havacilik ve uzay teknolojileri dergisi) 8, 1--6.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fable2020}{}}%
+O'Hara-Wild, M., Hyndman, R., Wang, E., Caceres, G., 2020. \href{https://fable.tidyverts.org/}{{fable}: Forecasting models for tidy time serie}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ramos2013probabilistic}{}}%
+Ramos, M.H., Van Andel, S.J., Pappenberger, F., 2013. Do probabilistic forecasts lead to better decisions? Hydrology and Earth System Sciences 17, 2219--2232.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ridgeway2007generalized}{}}%
+Ridgeway, G., 2007. Generalized boosted models: A guide to the gbm package. Update 1, 2007.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Rigby2005}{}}%
+Rigby, R.A., Stasinopoulos, D.M., 2005. {Generalized additive models for location, scale and shape}. Journal of the Royal Statistical Society: Series C (Applied Statistics) 54, 554--507.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ROSTAMITABAR20221197}{}}%
+Rostami-Tabar, B., Ziel, F., 2022.
+Anticipating special events in emergency department forecasting.
+International Journal of Forecasting 38, 1197--1213.
+doi:\url{https://doi.org/10.1016/j.ijforecast.2020.01.001}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schweigler2009forecasting}{}}%
+Schweigler, L.M., Desmond, J.S., McCarthy, M.L., Bukowski, K.J., Ionides, E.L., Younger, J.G., 2009. Forecasting models of emergency department crowding. Academic Emergency Medicine 16, 301--308.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mingliterature2022}{}}%
+Shi, M., Rostami-Tabar, B., Gartner, D., 2022. Forecasting for unschduled care services: A literature review (Working Paper), Under Review in European Journal of Operational Research.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Svetunkov2021Greybox}{}}%
+Svetunkov, I., 2021b. \href{https://github.com/config-i1/greybox}{Greybox: Toolbox for model building and forecasting}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Svetunkov2021Smooth}{}}%
+Svetunkov, I., 2021c. \href{https://github.com/config-i1/smooth}{Smooth: Forecasting using state space models}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SvetunkovAdam}{}}%
+Svetunkov, I., 2021a. \href{https://openforecast.org/adam/}{Forecasting and analytics with ADAM}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SvetunkovAdam2021}{}}%
+Svetunkov, I., 2022. \href{https://openforecast.org/adam/}{Forecasting and analytics with ADAM}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Svetunkov2019a}{}}%
+Svetunkov, I., Boylan, J.E., 2019. {Multiplicative state-space models for intermittent time series}. doi:\href{https://doi.org/10.13140/RG.2.2.35897.06242}{10.13140/RG.2.2.35897.06242}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Tashman2000}{}}%
+Tashman, L.J., 2000.
+{Out-of-sample tests of forecasting accuracy: An analysis and review}.
+International Journal of Forecasting 16, 437--450.
+doi:\href{https://doi.org/10.1016/S0169-2070(00)00065-0}{10.1016/S0169-2070(00)00065-0}
+
+\leavevmode\vadjust pre{\hypertarget{ref-taylor2018forecasting}{}}%
+Taylor, S.J., Letham, B., 2018. Forecasting at scale. The American Statistician 72, 37--45.
+
+\leavevmode\vadjust pre{\hypertarget{ref-RTeam2021}{}}%
+Team, R.C., 2021. \href{https://www.R-project.org/}{R: A language and environment for statistical computing}. R Foundation for Statistical Computing, Vienna, Austria.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stormemma2018}{}}%
+Wales, B., n.d. Wales returning to normal as snow thaws and temperatures rise.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Wood2017}{}}%
+Wood, S.N., 2017. Generalized additive models: An introduction with r. chapman; hall/CRC.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wright2006reexamining}{}}%
+Wright, P.D., Bretthauer, K.M., Côté, M.J., 2006. Reexamining the nurse scheduling problem: Staffing ratios and nursing shortages. Decision Sciences 37, 39--70.
+
+\end{CSLReferences}
 

### 2023-04-13 journal_submission/revised/latex_without_author/paper_without_authors.tex
prepare a version to share in the journal

--- 2023-04-13 journal_submission/revised/latex_with_authors/paper_with_authors.tex
@@
   \title{Probabilistic forecasting of hourly Emergency Department arrivals}
-    \author[CBS]{Bahman Rostami-Tabar%
-  \corref{cor1}%
-  }
-   \ead{rostami-tabarb@cardiff.ac.uk} 
-    \author[UG]{Jethro Browell%
-  %
-  }
-   \ead{jethro.browell@glasgow.ac.uk} 
-    \author[LU]{Ivan Svetunkov%
-  %
-  }
-   \ead{i.svetunkov@lancaster.ac.uk} 
-      \affiliation[CBS]{Cardiff Business School, Cardiff University, UK.}
-    \affiliation[UG]{School of Mathematics and Statistics, University of Glasgow, UK.}
-    \affiliation[LU]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK.}
       \cortext[cor1]{Corresponding author}
@@
 model (Lütkepohl, 2013), TBATS (De Livera et al., 2011) and Artificial
-Neural Networks (Hyndman and Athanasopoulos, 2021).We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study.
-These gaps and their importance are discussed below.
-Most of these studies are limited to only predicting future arrivals as a point
-forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.
-There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals.
+Neural Networks (Hyndman and Athanasopoulos, 2021).
+\textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study.
+These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
+forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals.
 Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e.
 having more staff than needed is costly, but
-having less staff than required may lead to worse outcomes for patients.
+having less staff than required may lead to worse outcomes for patients}.
 This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient (Wright et al., 2006).
@@
 Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task.
-Another drawback of existing studies is that the datasets used are relatively small (e.g.
+\textcolor{blue}{Another drawback of existing studies is that the datasets used are relatively small (e.g.
 time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation.
@@
 Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches.
-Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.
+Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}
 
@@
 
-There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
+\textcolor{blue}{There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare.
 Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention.
-We refer interested readers to some extensive reviews of the relevant literature by Shi et al.
+We refer interested readers to some extensive reviews of the relevant literature by} Shi et al.
 (2022), Gul and Celik (2020) and Ibrahim et al.
 (2016).
-The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
-However, given the focus of this study, we only discuss hourly ED forecasting.
+\textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare.
+However, given the focus of this study, we only discuss hourly ED forecasting.}
 
@@
 
-Figure \ref{fig:seasonplot-dofw} illustrates the daily subseries plot, with the x-axis representing the date and y-axis
+Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
 the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
@@
 compared to the rest of the week. This might be due to the closure of General Practitioners
-outpatient clinics over the weekends.
+outpatient clinics over the weekends.}
 
@@
 package (O'Hara-Wild et al., 2020). Note that the input data is assigned with an
-hourly and daily seasonality. This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
-benchmarks.
+hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
+benchmarks.}
 
@@
 GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
-calibration across most probability levels. Notably, both benchmarks exhibit nega-
+calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks exhibit nega-
 tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
 This could result in poor staffing decisions. This is because the empirical distribution of
-whole data fails to characterise how arrivals in ED may change over time.
+whole data fails to characterise how arrivals in ED may change over time.}
 
@@
 
-\textbackslash The main benefit of the applied models is the use of probabilistic forecasts to inform decision making.
-Probabilistic forecasts contain all potential future outcomes and help planners to
+\textbackslash textcolor\{blue\}\{The main benefit of the applied models is the use of probabilistic forecasts to inform decision
+making. Probabilistic forecasts contain all potential future outcomes and help planners to
 achieve more efficient decisions by not only predicting the most likely outcome but also
@@
 
-Practically speaking, the running time can be an important aspect for managers depending
+\textcolor{blue}{Practically speaking, the running time can be an important aspect for managers depending
 on the frequency of generating forecasts. If it is high, one may employ models that do not
@@
 accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
-like GAMLSS (Ttr-2) should be used.
-
-Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
-Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.
+like GAMLSS (Ttr-2) should be used.}
+
+\textcolor{blue}{Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning.
+Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.}
 
@@
 
+\textcolor{blue}{
 The dataset used in this study does not include the period of the COVID-19 pandemic.
@@
 However, this is outside of the scope of this paper and can be considered as a direction for
-future research.
+future research.}
 

### 2023-04-13 manuscript/paper.Rmd
prepare a version to share in the journal

--- 2023-01-11 paper/paper.Rmd
@@
     base_format: rticles::elsevier_article
-    citation_package: natbib
+    citation_package: default
 bibliography: mybibfile.bib
-#biblio-style: authoryear
-#csl: elsevier-without-titles.csl
+biblio-style: apa
+csl: elsevier-harvard.csl
 #biblio-style: abbrevnat
@@
 return visits to services, increased health care costs, inaccuracy in
-electronic medical records and other [@ROSTAMITABAR20221197].
+electronic medical records [@ROSTAMITABAR20221197].
 
@@
 
-```{r seasonplot-dofw, fig.cap= "Subseries plot: day of weekly arrivals", fig.align='center',out.width="70%"}
-```
+Clear weekday effects
+are notable with larger volumes of incidents observed on
+Fridays and Saturdays. All such observations will become
+of key importance when designing schedules for ambulance
+crews.
 
@@
 
-Finally, Figure \@ref(fig:date-plot) presents the time plot of daily
-arrivals. Each point represents one day, and points are shape-coded by
-day-of-week to show the weekly cycle. The figure shows more arrivals on
-Mondays than on other weekdays, agreeing with the previous findings.
-Moreover, we can observe a long-term trend line and the significant
-effect of some holidays. We can see that arrivals near Christmas and New
-Year's day are significantly lower than other days of the year.
-Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact
-of special events on arrivals. For instance, we see that arrivals are
-significantly low for 01-03 of March 2018. These days correspond to the
-Storm Emma with heavy snowfall that resulted in travel disruption, mass
-power outages, and schools closed in the UK [@stormemma2018].
+In summary, we observe an increase in demand levels for
+EMS between 2005 and 2009, and note that demand is
+more volatile on weekends and in months associated with
+school holidays, such as July and December.
 

### 2023-04-13 manuscript/paper.tex
prepare a version to share in the journal

--- 2023-04-13 journal_submission/revised/latex_without_author/paper_without_authors.tex
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/seasonplot-dofw-1}
-
-}
-
-\caption{Subseries plot: day of weekly arrivals}\label{fig:seasonplot-dofw}
-\end{figure}
+Clear weekday effects
+are notable with larger volumes of incidents observed on
+Fridays and Saturdays. All such observations will become
+of key importance when designing schedules for ambulance
+crews.
 
@@
 
-Finally, Figure \ref{fig:date-plot} presents the time plot of daily
-arrivals. Each point represents one day, and points are shape-coded by
-day-of-week to show the weekly cycle. The figure shows more arrivals on
-Mondays than on other weekdays, agreeing with the previous findings.
-Moreover, we can observe a long-term trend line and the significant
-effect of some holidays. We can see that arrivals near Christmas and New
-Year's day are significantly lower than other days of the year.
-Moreover, Figure \ref{fig:date-plot} allows us to identify the impact
-of special events on arrivals. For instance, we see that arrivals are
-significantly low for 01-03 of March 2018. These days correspond to the
-Storm Emma with heavy snowfall that resulted in travel disruption, mass
-power outages, and schools closed in the UK (Wales, n.d.).
+In summary, we observe an increase in demand levels for
+EMS between 2005 and 2009, and note that demand is
+more volatile on weekends and in months associated with
+school holidays, such as July and December.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-stormemma2018}{}}%
-Wales, B., n.d. Wales returning to normal as snow thaws and temperatures rise.
-
 \leavevmode\vadjust pre{\hypertarget{ref-Wood2017}{}}%

### 2023-04-13 adam-model/2021-03-25-experiment.R

@@ -1,259 +1,0 @@
-# ADAM experiment
-remotes::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
-remotes::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
-
-require(data.table)
-require(readxl)
-
-library(zoo)
-library(greybox)
-library(smooth)
-library(foreach)
-library(doMC)
-library(ProbCast)
-
-## Load and Prep Data ####
-# load("../data/hw_hourly.rds")
-h2 <- fread("data/h2_hourly_gb.csv")
-h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-
-add_calendar_variables(h2,datetimecol = "targetTime_UK")
-
-# Load Holidays
-hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
-hols[,Date:=as.Date(Date)]
-h2[,Date:=as.Date(targetTime_UK)]
-h2 <- merge(h2,hols,by="Date",all.x=T)
-rm(hols); h2[,Date:=NULL]
-h2[,school_holiday:=as.factor(school_holiday)]
-h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
-h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
-h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
-
-# Prepare explanatory variables
-xreg <- as.character(h2$holiday_festive_day)
-# Do leads and lags of the variables
-xreg <- xregExpander(xreg, lags=c(-2:1)*24, gaps="NAs")
-xreg <- as.data.table(xreg)
-xreg$x[is.na(xreg$x)] <- "none"
-xreg$xLag24[is.na(xreg$xLag24)] <- "none"
-xreg$xLag48[is.na(xreg$xLag48)] <- "none"
-xreg$xLead24[is.na(xreg$xLead24)] <- "none"
-xreg$x <- factor(xreg$x)
-xreg$xLag24 <- factor(xreg$xLag24)
-xreg$xLag48 <- factor(xreg$xLag48)
-xreg$xLead24 <- factor(xreg$xLead24)
-
-# Create zoo objects
-# The first observations is from 31st March
[…]

### 2023-04-13 adam-model/adam-times.R

@@ -1,226 +1,0 @@
-# ADAM experiment
-remotes::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
-remotes::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
-
-require(data.table)
-require(readxl)
-
-library(zoo)
-library(greybox)
-library(smooth)
-library(foreach)
-library(doMC)
-library(ProbCast)
-
-## Load and Prep Data ####
-# load("../data/hw_hourly.rds")
-h2 <- fread("data/h2_hourly_gb.csv")
-h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-
-add_calendar_variables(h2,datetimecol = "targetTime_UK")
-
-# Load Holidays
-hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
-hols[,Date:=as.Date(Date)]
-h2[,Date:=as.Date(targetTime_UK)]
-h2 <- merge(h2,hols,by="Date",all.x=T)
-rm(hols); h2[,Date:=NULL]
-h2[,school_holiday:=as.factor(school_holiday)]
-h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
-h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
-h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
-
-# Prepare explanatory variables
-xreg <- as.character(h2$holiday_festive_day)
-# Do leads and lags of the variables
-xreg <- xregExpander(xreg, lags=c(-2:1)*24, gaps="NAs")
-xreg <- as.data.table(xreg)
-xreg$x[is.na(xreg$x)] <- "none"
-xreg$xLag24[is.na(xreg$xLag24)] <- "none"
-xreg$xLag48[is.na(xreg$xLag48)] <- "none"
-xreg$xLead24[is.na(xreg$xLead24)] <- "none"
-xreg$x <- factor(xreg$x)
-xreg$xLag24 <- factor(xreg$xLag24)
-xreg$xLag48 <- factor(xreg$xLag48)
-xreg$xLead24 <- factor(xreg$xLead24)
-
-# Create zoo objects
-# The first observations is from 31st March
[…]

### 2023-04-13 adam-model/daylightSavingExploration.R

@@ -1,17 +1,0 @@
-install.packages(c("lubridate","tsutils"))
-
-library(lubridate)
-library(data.table)
-
-h2 <- fread("data/h2_hourly.csv")
-
-# Get the 2018 data
-test <- h2[year(h2$arrival_1h)==2018,]
-# Get to March
-testMarch <- test[month(test$arrival_1h)==3,]
-# Season plot of 24 - 27 of March (25th March, 1am)
-tsutils::seasplot(ts(testMarch$n_attendance[1:(24*4)+552],frequency=24))
-# Get to October
-testOctober <- test[month(test$arrival_1h)==10,]
-# Season plot of 26 - 29 of October (28th October)
-tsutils::seasplot(ts(testOctober$n_attendance[1:(24*4)+600],frequency=24))

### 2023-04-13 adam-model/dstplots.R

@@ -1,46 +1,0 @@
-devtools::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE, ref="Profile-with-DST")
-library(smooth)
-
-#### Examples with shifts via DST ####
-
-x <- zoo(h2$n_attendance, order.by=h2$arrival_1h)
-testDST1 <- adam(x, "ANA", lags=c(1,24,24*7), silent=F, initial="back", persistence=rep(0,3))
-testDST2 <- adam(as.vector(x), "ANA", lags=c(1,24,24*7), silent=F, initial="back", persistence=rep(0,3))
-
-mean((actuals(testDST1)-fitted(testDST1))^2)
-mean((actuals(testDST2)-fitted(testDST2))^2)
-
-pdf("/home/config/Dropbox/R/dstPlots.pdf",width=10,height=7)
-# End of DST in 2014
-plot(actuals(testDST1)[1:(24*7)+4994-24*3],type="b")
-lines(fitted(testDST1)[1:(24*7)+4994-24*3],col="red")
-points(fitted(testDST1)[1:(24*7)+4994-24*3],col="red")
-abline(v=time(actuals(testDST1))[4994],col="red",lwd=2)
-
-plot(actuals(testDST2)[1:(24*7)+4994-24*3],type="b")
-lines(fitted(testDST2)[1:(24*7)+4994-24*3],col="blue")
-points(fitted(testDST2)[1:(24*7)+4994-24*3],col="blue")
-abline(v=24*3,col="red",lwd=2)
-
-
-# Start of DST in 2015
-plot(actuals(testDST1)[1:(24*7)+8690-24*3],type="b")
-lines(fitted(testDST1)[1:(24*7)+8690-24*3],col="red")
-points(fitted(testDST1)[1:(24*7)+8690-24*3],col="red")
-abline(v=time(actuals(testDST1))[8690],col="red",lwd=2)
-
-plot(actuals(testDST2)[1:(24*7)+8690-24*3],type="b")
-lines(fitted(testDST2)[1:(24*7)+8690-24*3],col="blue")
-points(fitted(testDST2)[1:(24*7)+8690-24*3],col="blue")
-abline(v=24*3,col="red",lwd=2)
-
-
-# Start of DST in 2017
-plot(actuals(testDST1)[1:(24*7)+26162-24*3],type="l")
-lines(fitted(testDST1)[1:(24*7)+26162-24*3],col="red")
[…]

### 2023-04-13 adam-model/experiment.R

@@ -1,236 +1,0 @@
-# ADAM experiment
-devtools::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
-devtools::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
-
-require(data.table)
-require(readxl)
-
-library(zoo)
-library(greybox)
-library(smooth)
-library(forecast)
-library(foreach)
-library(doMC)
-library(ProbCast)
-
-## Load and Prep Data ####
-# load("../data/hw_hourly.rds")
-h2 <- fread("data/h2_hourly_gb.csv")
-h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-
-add_calendar_variables(h2,datetimecol = "targetTime_UK")
-
-# Load Holidays
-hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
-hols[,Date:=as.Date(Date)]
-h2[,Date:=as.Date(targetTime_UK)]
-h2 <- merge(h2,hols,by="Date",all.x=T)
-rm(hols); h2[,Date:=NULL]
-h2[,school_holiday:=as.factor(school_holiday)]
-h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
-h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
-h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
-
-# Prepare explanatory variables
-xreg <- as.character(h2$holiday_festive_day)
-# Do leads and lags of the variables
-xreg <- xregExpander(xreg, lags=c(-2:1)*24, gaps="NAs")
-xreg <- as.data.table(xreg)
-xreg$x[is.na(xreg$x)] <- "none"
-xreg$xLag24[is.na(xreg$xLag24)] <- "none"
-xreg$xLag48[is.na(xreg$xLag48)] <- "none"
-xreg$xLead24[is.na(xreg$xLead24)] <- "none"
-xreg$x <- factor(xreg$x)
-xreg$xLag24 <- factor(xreg$xLag24)
-xreg$xLag48 <- factor(xreg$xLag48)
-xreg$xLead24 <- factor(xreg$xLead24)
-
-# Create zoo objects
[…]

### 2023-04-13 adam-model/exploration.R

@@ -1,57 +1,0 @@
-devtools::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
-devtools::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
-devtools::install_github("config-i1/gnorm", upgrade="never", dependencies=FALSE)
-
-require(data.table)
-require(readxl)
-library(greybox)
-library(smooth)
-library(zoo)
-library(statmod)
-library(gnorm)
-
-## Load and Prep Data ####
-# load("../data/hw_hourly.rds")
-h2 <- fread("data/h2_hourly.csv")
-h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-
-add_calendar_variables(h2,datetimecol = "targetTime_UK")
-
-# Load Holidays
-hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
-hols[,Date:=as.Date(Date)]
-h2[,Date:=as.Date(targetTime_UK)]
-h2 <- merge(h2,hols,by="Date",all.x=T)
-rm(hols); h2[,Date:=NULL]
-h2[,school_holiday:=as.factor(school_holiday)]
-h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
-h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
-h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
-
-# Prepare explanatory variables
-xreg <- as.character(h2$holiday_festive_day)
-xreg[is.na(h2$holiday_festive_day)] <- "none"
-xreg <- factor(xreg)
-
-# Create zoo objects
-x <- zoo(h2$n_attendance, order.by=h2$targetTime)
-xreg <- zoo(xreg, order.by=h2$targetTime)
-xregData <- data.frame(x=x,xreg=xreg)
-
-# Fit the occurrence model
-oesModel <- oes(x, "MNN", h=24*7, holdout=TRUE, occurrence="direct")
-# Fit the demand sizes model
-adamModel <- adam(x, "MNM", lags=c(24,24*7), h=24*7, holdout=TRUE, initial="b", occurrence=oesModel)
[…]

### 2023-04-13 gam-mgcv/exploration.R

@@ -1,121 +1,0 @@
-
-## Forecast every 12h (00 and 12 local time)
-## Lead times from 0 to 48h
-## Quantiles from 0.05 to 0.95, Pinball loss
-## Test Data: 1/3/2018 to 28/2/2019
-
-require(rstudioapi)
-require(data.table)
-require(mgcv)
-require(mgcViz)
-require(ggplot2)
-# library(devtools)
-# install_github("jbrowell/ProbCast")
-require(ProbCast)
-require(readxl)
-
-setwd(dirname(getActiveDocumentContext()$path))
-
-## Load and Prep Data ####
-# load("../data/hw_hourly.rds")
-# h2 <- fread("../data/h2_hourly.csv")
-# h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-# h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-h2 <- fread("../data/h2_hourly_gb.csv")
-h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-
-add_calendar_variables(h2,datetimecol = "targetTime_UK")
-
-# Load Holidays
-hols <- as.data.table(read_xlsx("../data/holiday_rugby_all.xlsx"))
-hols[,Date:=as.Date(Date)]
-h2[,Date:=as.Date(targetTime_UK)]
-h2 <- merge(h2,hols,by="Date",all.x=T)
-rm(hols); h2[,Date:=NULL]
-h2[,school_holiday:=as.factor(school_holiday)]
-h2[,holiday_festive_day:=as.factor(festive_day)]
-h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
-h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
-
-#load weather data
-weather_data <- data.table()
-for(f in c(list.files(path="../data/",pattern = "weather",full.names = T))){
-  load(f)
-  weather_data <- rbind(weather_data,features)
-}
-h2 <- merge(weather_data,h2,by = "targetTime")
-rm(f,features,weather_data)
-h2[,wind10m:=sqrt(`10U`^2+`10V`^2)]
-
-
[…]

### 2023-04-13 gam-mgcv/study_v1.R

@@ -1,1085 +1,0 @@
-
-## Forecast every 12h (00 and 12 local time) <<< Circulate template!!!
-## Lead times from 0 to 48h
-## Quantiles from 0.05 to 0.95, Pinball loss
-## Test Data: 1/3/2018 to 28/2/2019
-
-
-## TO DO ####
-#
-# Models for sigma in NOtr
-# Alternative dist in gamlss
-# qgam()
-# Extract expectation for RMSE. Column name "expectation".
-
-## Begin... ####
-require(rstudioapi)
-require(data.table)
-require(mgcv)
-require(mgcViz)
-require(ggplot2)
-# library(devtools); Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
-# install_github("jbrowell/ProbCast")
-require(ProbCast)
-require(readxl)
-
-
-## Utilities for fitting truncated normal
-require(gamlss.tr)
-require(gamlss.add)
-gen.trun(par = c(0),family = NO,type="left")
-gen.trun(par = c(0),family = TF2,type="left")
-# Function to calculate mean of truncated normal and t, and estimate from quantiles
-mean_t_norm <- function(mu,sigma,a,b){
-  
-  mu + sigma*(dnorm((a-mu)/sigma) - dnorm((b-mu)/sigma))/
-    (pnorm(b)-pnorm(a))
-  
-}
-
-mean_t_t <- function(mu,sigma,nu,a,b){
-  # from https://doi.org/10.1016/j.jspi.2011.06.006
-  
-  if(any(nu<=1,na.rm = T)){stop("All nu must be greater than 1 in mean_t_t()")}
-  
-  ## For standard ~ mu=1; sigma=0
-  a_x <- a*sigma+mu
-  b_x <- b*sigma+mu
-  tau_0 <- 1
-  alp_0 <- pt(b*tau_0^0.5,df = nu) - pt(a*tau_0^0.5,df = nu)
-  k <- gamma((nu+1)/2)/(alp_0*gamma(nu/2)*(nu*pi)^0.5)
-  X_bar <- (k*nu/(nu-1))*( (1+a_x^2/nu)^((1-nu)/2) - (1+b_x^2/nu)^((1-nu)/2) )
-  
-  ## Re-scale:
-  mu + sigma*X_bar
-  
-}
-
-mean_from_qs <- function(mqr){
-  
-  q_cols <- grep("q",colnames(mqr))
-  qs <- as.numeric(gsub("q","",colnames(mqr)[q_cols]))/100
[…]

### 2023-04-13 gam-mgcv/study_v1_timing.R

@@ -1,836 +1,0 @@
-
-## Forecast every 12h (00 and 12 local time) <<< Circulate template!!!
-## Lead times from 0 to 48h
-## Quantiles from 0.05 to 0.95, Pinball loss
-## Test Data: 1/3/2018 to 28/2/2019
-
-
-## TO DO ####
-#
-# Models for sigma in NOtr
-# Alternative dist in gamlss
-# qgam()
-# Extract expectation for RMSE. Column name "expectation".
-
-## Begin... ####
-require(rstudioapi)
-require(data.table)
-require(mgcv)
-require(mgcViz)
-require(ggplot2)
-# library(devtools); Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
-remotes::install_github("jbrowell/ProbCast")
-require(ProbCast)
-require(readxl)
-
-
-## Utilities for fitting truncated normal
-require(gamlss.tr)
-require(gamlss.add)
-gen.trun(par = c(0),family = NO,type="left")
-gen.trun(par = c(0),family = TF2,type="left")
-# Function to calculate mean of truncated normal and t, and estimate from quantiles
-mean_t_norm <- function(mu,sigma,a,b){
-  
-  mu + sigma*(dnorm((a-mu)/sigma) - dnorm((b-mu)/sigma))/
-    (pnorm(b)-pnorm(a))
-  
-}
-
-mean_t_t <- function(mu,sigma,nu,a,b){
-  # from https://doi.org/10.1016/j.jspi.2011.06.006
-  
-  if(any(nu<=1,na.rm = T)){stop("All nu must be greater than 1 in mean_t_t()")}
-  
-  ## For standard ~ mu=1; sigma=0
-  a_x <- a*sigma+mu
-  b_x <- b*sigma+mu
-  tau_0 <- 1
-  alp_0 <- pt(b*tau_0^0.5,df = nu) - pt(a*tau_0^0.5,df = nu)
-  k <- gamma((nu+1)/2)/(alp_0*gamma(nu/2)*(nu*pi)^0.5)
-  X_bar <- (k*nu/(nu-1))*( (1+a_x^2/nu)^((1-nu)/2) - (1+b_x^2/nu)^((1-nu)/2) )
-  
-  ## Re-scale:
-  mu + sigma*X_bar
-  
-}
-
-mean_from_qs <- function(mqr){
-  
-  q_cols <- grep("q",colnames(mqr))
[…]

### 2023-04-13 r_script/adam-model/2021-03-25-experiment.R

@@ -1,0 +1,259 @@
+# ADAM experiment
+remotes::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
+remotes::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
+
+require(data.table)
+require(readxl)
+
+library(zoo)
+library(greybox)
+library(smooth)
+library(foreach)
+library(doMC)
+library(ProbCast)
+
+## Load and Prep Data ####
+# load("../data/hw_hourly.rds")
+h2 <- fread("data/h2_hourly_gb.csv")
+h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
+h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
+
+add_calendar_variables(h2,datetimecol = "targetTime_UK")
+
+# Load Holidays
+hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
+hols[,Date:=as.Date(Date)]
+h2[,Date:=as.Date(targetTime_UK)]
+h2 <- merge(h2,hols,by="Date",all.x=T)
+rm(hols); h2[,Date:=NULL]
+h2[,school_holiday:=as.factor(school_holiday)]
+h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
+h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
+h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
+
+# Prepare explanatory variables
+xreg <- as.character(h2$holiday_festive_day)
+# Do leads and lags of the variables
+xreg <- xregExpander(xreg, lags=c(-2:1)*24, gaps="NAs")
+xreg <- as.data.table(xreg)
+xreg$x[is.na(xreg$x)] <- "none"
+xreg$xLag24[is.na(xreg$xLag24)] <- "none"
+xreg$xLag48[is.na(xreg$xLag48)] <- "none"
+xreg$xLead24[is.na(xreg$xLead24)] <- "none"
+xreg$x <- factor(xreg$x)
+xreg$xLag24 <- factor(xreg$xLag24)
+xreg$xLag48 <- factor(xreg$xLag48)
+xreg$xLead24 <- factor(xreg$xLead24)
+
+# Create zoo objects
+# The first observations is from 31st March
[…]

### 2023-04-13 r_script/adam-model/experiment.R

@@ -1,0 +1,236 @@
+# ADAM experiment
+devtools::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
+devtools::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
+
+require(data.table)
+require(readxl)
+
+library(zoo)
+library(greybox)
+library(smooth)
+library(forecast)
+library(foreach)
+library(doMC)
+library(ProbCast)
+
+## Load and Prep Data ####
+# load("../data/hw_hourly.rds")
+h2 <- fread("data/h2_hourly_gb.csv")
+h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
+h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
+
+add_calendar_variables(h2,datetimecol = "targetTime_UK")
+
+# Load Holidays
+hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
+hols[,Date:=as.Date(Date)]
+h2[,Date:=as.Date(targetTime_UK)]
+h2 <- merge(h2,hols,by="Date",all.x=T)
+rm(hols); h2[,Date:=NULL]
+h2[,school_holiday:=as.factor(school_holiday)]
+h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
+h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
+h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
+
+# Prepare explanatory variables
+xreg <- as.character(h2$holiday_festive_day)
+# Do leads and lags of the variables
+xreg <- xregExpander(xreg, lags=c(-2:1)*24, gaps="NAs")
+xreg <- as.data.table(xreg)
+xreg$x[is.na(xreg$x)] <- "none"
+xreg$xLag24[is.na(xreg$xLag24)] <- "none"
+xreg$xLag48[is.na(xreg$xLag48)] <- "none"
+xreg$xLead24[is.na(xreg$xLead24)] <- "none"
+xreg$x <- factor(xreg$x)
+xreg$xLag24 <- factor(xreg$xLag24)
+xreg$xLag48 <- factor(xreg$xLag48)
+xreg$xLead24 <- factor(xreg$xLead24)
+
+# Create zoo objects
[…]

### 2023-04-13 r_script/adam-model/exploration.R

@@ -1,0 +1,57 @@
+devtools::install_github("config-i1/greybox", upgrade="never", dependencies=FALSE)
+devtools::install_github("config-i1/smooth", upgrade="never", dependencies=FALSE)
+devtools::install_github("config-i1/gnorm", upgrade="never", dependencies=FALSE)
+
+require(data.table)
+require(readxl)
+library(greybox)
+library(smooth)
+library(zoo)
+library(statmod)
+library(gnorm)
+
+## Load and Prep Data ####
+# load("../data/hw_hourly.rds")
+h2 <- fread("data/h2_hourly.csv")
+h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
+h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
+
+add_calendar_variables(h2,datetimecol = "targetTime_UK")
+
+# Load Holidays
+hols <- as.data.table(read_xlsx("data/holiday_rugby.xlsx"))
+hols[,Date:=as.Date(Date)]
+h2[,Date:=as.Date(targetTime_UK)]
+h2 <- merge(h2,hols,by="Date",all.x=T)
+rm(hols); h2[,Date:=NULL]
+h2[,school_holiday:=as.factor(school_holiday)]
+h2[,holiday_festive_day:=as.factor(holiday_festive_day)]
+h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
+h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
+
+# Prepare explanatory variables
+xreg <- as.character(h2$holiday_festive_day)
+xreg[is.na(h2$holiday_festive_day)] <- "none"
+xreg <- factor(xreg)
+
+# Create zoo objects
+x <- zoo(h2$n_attendance, order.by=h2$targetTime)
+xreg <- zoo(xreg, order.by=h2$targetTime)
+xregData <- data.frame(x=x,xreg=xreg)
+
+# Fit the occurrence model
+oesModel <- oes(x, "MNN", h=24*7, holdout=TRUE, occurrence="direct")
+# Fit the demand sizes model
+adamModel <- adam(x, "MNM", lags=c(24,24*7), h=24*7, holdout=TRUE, initial="b", occurrence=oesModel)
[…]

### 2023-04-13 r_script/gam-mgcv/exploration.R

@@ -1,0 +1,121 @@
+
+## Forecast every 12h (00 and 12 local time)
+## Lead times from 0 to 48h
+## Quantiles from 0.05 to 0.95, Pinball loss
+## Test Data: 1/3/2018 to 28/2/2019
+
+require(rstudioapi)
+require(data.table)
+require(mgcv)
+require(mgcViz)
+require(ggplot2)
+# library(devtools)
+# install_github("jbrowell/ProbCast")
+require(ProbCast)
+require(readxl)
+
+setwd(dirname(getActiveDocumentContext()$path))
+
+## Load and Prep Data ####
+# load("../data/hw_hourly.rds")
+# h2 <- fread("../data/h2_hourly.csv")
+# h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
+# h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
+h2 <- fread("../data/h2_hourly_gb.csv")
+h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
+h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
+
+add_calendar_variables(h2,datetimecol = "targetTime_UK")
+
+# Load Holidays
+hols <- as.data.table(read_xlsx("../data/holiday_rugby_all.xlsx"))
+hols[,Date:=as.Date(Date)]
+h2[,Date:=as.Date(targetTime_UK)]
+h2 <- merge(h2,hols,by="Date",all.x=T)
+rm(hols); h2[,Date:=NULL]
+h2[,school_holiday:=as.factor(school_holiday)]
+h2[,holiday_festive_day:=as.factor(festive_day)]
+h2[,is_rug_in_Cardiff:=is_rug_in_Cardiff==1]
+h2[,is_rug_out_Cardiff:=is_rug_out_Cardiff==1]
+
+#load weather data
+weather_data <- data.table()
+for(f in c(list.files(path="../data/",pattern = "weather",full.names = T))){
+  load(f)
+  weather_data <- rbind(weather_data,features)
+}
+h2 <- merge(weather_data,h2,by = "targetTime")
+rm(f,features,weather_data)
+h2[,wind10m:=sqrt(`10U`^2+`10V`^2)]
+
+
[…]

### 2023-04-13 r_script/gam-mgcv/study_v1.R

@@ -1,0 +1,1085 @@
+
+## Forecast every 12h (00 and 12 local time) <<< Circulate template!!!
+## Lead times from 0 to 48h
+## Quantiles from 0.05 to 0.95, Pinball loss
+## Test Data: 1/3/2018 to 28/2/2019
+
+
+## TO DO ####
+#
+# Models for sigma in NOtr
+# Alternative dist in gamlss
+# qgam()
+# Extract expectation for RMSE. Column name "expectation".
+
+## Begin... ####
+require(rstudioapi)
+require(data.table)
+require(mgcv)
+require(mgcViz)
+require(ggplot2)
+# library(devtools); Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
+# install_github("jbrowell/ProbCast")
+require(ProbCast)
+require(readxl)
+
+
+## Utilities for fitting truncated normal
+require(gamlss.tr)
+require(gamlss.add)
+gen.trun(par = c(0),family = NO,type="left")
+gen.trun(par = c(0),family = TF2,type="left")
+# Function to calculate mean of truncated normal and t, and estimate from quantiles
+mean_t_norm <- function(mu,sigma,a,b){
+  
+  mu + sigma*(dnorm((a-mu)/sigma) - dnorm((b-mu)/sigma))/
+    (pnorm(b)-pnorm(a))
+  
+}
+
+mean_t_t <- function(mu,sigma,nu,a,b){
+  # from https://doi.org/10.1016/j.jspi.2011.06.006
+  
+  if(any(nu<=1,na.rm = T)){stop("All nu must be greater than 1 in mean_t_t()")}
+  
+  ## For standard ~ mu=1; sigma=0
+  a_x <- a*sigma+mu
+  b_x <- b*sigma+mu
+  tau_0 <- 1
+  alp_0 <- pt(b*tau_0^0.5,df = nu) - pt(a*tau_0^0.5,df = nu)
+  k <- gamma((nu+1)/2)/(alp_0*gamma(nu/2)*(nu*pi)^0.5)
+  X_bar <- (k*nu/(nu-1))*( (1+a_x^2/nu)^((1-nu)/2) - (1+b_x^2/nu)^((1-nu)/2) )
+  
+  ## Re-scale:
+  mu + sigma*X_bar
+  
+}
+
+mean_from_qs <- function(mqr){
+  
+  q_cols <- grep("q",colnames(mqr))
+  qs <- as.numeric(gsub("q","",colnames(mqr)[q_cols]))/100
[…]

### 2023-04-13 r_script/gam-mgcv/study_v1_timing.R

@@ -1,0 +1,836 @@
+
+## Forecast every 12h (00 and 12 local time) <<< Circulate template!!!
+## Lead times from 0 to 48h
+## Quantiles from 0.05 to 0.95, Pinball loss
+## Test Data: 1/3/2018 to 28/2/2019
+
+
+## TO DO ####
+#
+# Models for sigma in NOtr
+# Alternative dist in gamlss
+# qgam()
+# Extract expectation for RMSE. Column name "expectation".
+
+## Begin... ####
+require(rstudioapi)
+require(data.table)
+require(mgcv)
+require(mgcViz)
+require(ggplot2)
+# library(devtools); Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
+remotes::install_github("jbrowell/ProbCast")
+require(ProbCast)
+require(readxl)
+
+
+## Utilities for fitting truncated normal
+require(gamlss.tr)
+require(gamlss.add)
+gen.trun(par = c(0),family = NO,type="left")
+gen.trun(par = c(0),family = TF2,type="left")
+# Function to calculate mean of truncated normal and t, and estimate from quantiles
+mean_t_norm <- function(mu,sigma,a,b){
+  
+  mu + sigma*(dnorm((a-mu)/sigma) - dnorm((b-mu)/sigma))/
+    (pnorm(b)-pnorm(a))
+  
+}
+
+mean_t_t <- function(mu,sigma,nu,a,b){
+  # from https://doi.org/10.1016/j.jspi.2011.06.006
+  
+  if(any(nu<=1,na.rm = T)){stop("All nu must be greater than 1 in mean_t_t()")}
+  
+  ## For standard ~ mu=1; sigma=0
+  a_x <- a*sigma+mu
+  b_x <- b*sigma+mu
+  tau_0 <- 1
+  alp_0 <- pt(b*tau_0^0.5,df = nu) - pt(a*tau_0^0.5,df = nu)
+  k <- gamma((nu+1)/2)/(alp_0*gamma(nu/2)*(nu*pi)^0.5)
+  X_bar <- (k*nu/(nu-1))*( (1+a_x^2/nu)^((1-nu)/2) - (1+b_x^2/nu)^((1-nu)/2) )
+  
+  ## Re-scale:
+  mu + sigma*X_bar
+  
+}
+
+mean_from_qs <- function(mqr){
+  
+  q_cols <- grep("q",colnames(mqr))
[…]

### 2023-04-13 r_script/result_viz.R

@@ -226,7 +226,7 @@ ggplot(data=merge(PB[kfold=="Test" & Horizon=="All",.(Pinball=mean(Loss)),by=c("
         legend.title = element_text(face = "bold"))
 
 ## ---- time
-results_table <- read_rds("results_table.rds")
+results_table <- read_rds("../results/results_table.rds")
 time <- results_table %>% as_tibble() %>% select(Method,Time)
 time_selected <- time %>% 
   mutate(Method=factor(Method)) %>% 
@@ -244,7 +244,7 @@ ggplot(time_selected,
   labs(x="Method", y = "Running time in second")
 
 ## ---- time-accuracy
-results_table <- read_rds("results_table.rds")
+results_table <- read_rds("../results/results_table.rds")
 results_table <- results_table %>% mutate(Time=if_else(is.na(Time),1.279187,Time))
 time <- results_table %>% as_tibble()
 # time_selected <- time %>% 
@@ -286,4 +286,3 @@ ggplot(time_long,aes(x=Time,
 #   facet_wrap(vars(Measure), ncol = 1, scales = "free")+
 #   theme_few()+
 #   labs(x="Time (in seconds)")
-

### 2023-04-13 results/Evaluatoin_Script.R

@@ -1,293 +1,0 @@
-rm(list=ls())
-## Forecast every 12h (00 and 12 local time) <<< Circulate template!!!
-## Lead times from 0 to 48h
-## Quantiles from 0.05 to 0.95, Pinball loss
-## Test Data: 1/3/2018 to 28/2/2019
-
-## To do:
-# add checks that exactly same data are being evaluated
-# evaluate training data as well as test
-# test ceiling/floor options for non-integer forecasts ! << Easy now...
-# report RMSE
-
-## For consistent plots in paper:
-# use ggthemes: theme_few()
-
-
-## Link to use case:
-# "Process 95% of patients in 4h"
-
-require(rstudioapi)
-require(data.table)
-require(ggplot2)
-# library(devtools); Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
-# install_github("jbrowell/ProbCast")
-require(ProbCast)
-setwd(dirname(getActiveDocumentContext()$path))
-
-
-
-## Load A&E Arrival Data ####
-h2 <- fread("../data/h2_hourly_gb.csv")
-h2[,targetTime:=as.POSIXct(arrival_1h,tz="UTC",format="%Y-%m-%dT%H:%M:%SZ")]
-h2[,targetTime_UK:=targetTime]; attributes(h2$targetTime_UK)$tzone <- "Europe/London"
-
-
-## Initialis Containers ####
-test_start <- "2018-03-01"
-last_issue <- "2019-02-26 00:00:00" #### <<<<< Impliment!!! Make test for data matching
-PB <- data.table()
-REL <- data.table()
-RMSE <- data.table()
-
-## Eval function
-big_eval_function <- function(forecast_DT,h2_actuals,method_name){
-  
-  forecast_DT <- forecast_DT[targetTime_UK>=test_start & issueTime<=last_issue,]
-  
-  ## Get Actuals
-  setkey(forecast_DT,issueTime,targetTime_UK)
-  actuals <- merge(forecast_DT[,.(issueTime,targetTime_UK)],h2_actuals[,.(targetTime_UK,n_attendance)],
-                   by="targetTime_UK",all.x=T,no.dups = F)
-  setkey(actuals,issueTime,targetTime_UK)
-  
-  ## All
-  temp <- data.table(pinball(forecast_DT[,grep(names(forecast_DT),pattern = "q"),with=F],actuals[,n_attendance],plot.it = F))
[…]
